---
title: "PENIP Queen Conch Habitat Classification"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

input_file <- knitr::current_input()
if (!is.null(input_file)) {
  knitr::opts_knit$set(root.dir = normalizePath(file.path(dirname(input_file), "..")))
}
```

# Setup

```{r libraries}
library(dplyr)
library(gstat)
library(ggplot2)
library(purrr)
library(readxl)
library(rvest)
library(sf)
library(tibble)
library(tidyr)
library(xml2)
```

```{r paths}
conch_xlsx <- "data/conch/PENIP_QueenConchSurveys.xlsx"
conch_sheet <- "PENIP_Conch_10mRadial"
drift_sheet <- "PENIP_DriftTransects"
drift_position_sheet <- "PENIP_ConchDriftPosition"
radial_counts_csv <- "data/conch/PENIP_circular_counts.csv"

output_map <- "outputs/conch_point_habitat_map.png"
output_radial_abundance_points <- "outputs/conch_radial_abundance_points.png"
output_radial_abundance_heatmap <- "outputs/conch_radial_abundance_heatmap.png"
output_radial_mating_map <- "outputs/conch_radial_mating_map.png"
output_adult_mating_halo_map <- "outputs/conch_adult_mating_halo_map.png"
output_adult_mating_distance_plot <- "outputs/conch_adult_mating_distance_plot.png"
output_adult_mating_distance_map <- "outputs/conch_adult_mating_distance_map.png"
output_adult_mating_radius_plot <- "outputs/conch_adult_mating_radius_plot.png"
output_adult_mating_radius_map <- "outputs/conch_adult_mating_radius_map.png"
output_conch_blob_prediction_map <- "outputs/conch_blob_adult_juvenile_prediction_map.png"
output_drift_background_density_plot <- "outputs/conch_drift_background_density_by_habitat.png"
output_drift_map <- "outputs/conch_drift_transect_map.png"
output_drift_habitat_histogram <- "outputs/conch_drift_habitat_area_histogram.png"
output_drift_position_counts <- "outputs/conch_drift_position_habitat_counts.png"
output_drift_conch_map <- "outputs/conch_drift_transects_observations_map.png"
output_density_model_plot <- "outputs/conch_density_model_estimates.png"
output_zinb_density_model_plot <- "outputs/conch_zinb_density_model_estimates.png"

type_cols <- c(
  "Aggregated Patch Reef" = "#8dd3c7",
  "Artificial" = "#ffffb3",
  "Inner Reef" = "#bebada",
  "Middle Reef" = "#fb8072",
  "Nearshore Ridge Complex" = "#80b1d3",
  "Outer Reef" = "#fdb462",
  "Ridge" = "#b3de69",
  "Sand" = "#f2e8c9",
  "Scattered Rock in Unconsolidated Sediment" = "#bc80bd"
)
```

# Habitat Polygons

This recreates `polygons_final` from `analysis/PENIP_analysis.Rmd`.

```{r habitat-polygons}
# Load final habitat polygons created in analysis/PENIP_analysis.Rmd
polygons_final <- readRDS("data/processed/polygons_final.rds")

old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

polygons_final <- polygons_final %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)
```

# Radial survey Point Habitat Classification

```{r conch-points}
conch_points <- readxl::read_xlsx(conch_xlsx, sheet = conch_sheet)

conch_points_sf <- conch_points %>%
  mutate(
    .record = row_number(),
    .row_excel = row_number() + 1L
  ) %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE)

polygons_for_join <- polygons_final %>%
  st_transform(st_crs(conch_points_sf)) %>%
  select(Type)

point_matches <- st_join(
  conch_points_sf,
  polygons_for_join,
  join = st_within,
  left = TRUE
) %>%
  st_drop_geometry() %>%
  select(
    .record,
    .row_excel,
    Waypoint,
    `Site Code`,
    Lat,
    Lon,
    Habitat_Type = Type
  )

ambiguous_points <- point_matches %>%
  count(.record, .row_excel, Waypoint, `Site Code`, name = "n_matches") %>%
  filter(n_matches > 1)

point_assignments <- point_matches %>%
  group_by(.record) %>%
  slice(1) %>%
  ungroup()

point_assignments %>%
  count(Habitat_Type, sort = TRUE)

ambiguous_points
```

## Radial Survey Data

This creates the full `PENIP_Conch_10mRadial` data frame with the habitat classifications populated in `Habitat (class at point)`.

```{r radial-survey-data}
conch_radial <- conch_points %>%
  mutate(`Habitat (class at point)` = point_assignments$Habitat_Type)

stopifnot(
  nrow(conch_radial) == nrow(conch_points),
  sum(is.na(conch_radial[["Habitat (class at point)"]])) == 0
)

table(conch_radial[["Habitat (class at point)"]], useNA = "ifany")
```

## Radial Survey Abundance Points

```{r radial-survey-abundance-points, fig.width=8, fig.height=6}
radial_counts <- read.csv(
  radial_counts_csv,
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
) %>%
  filter(!is.na(`Site Code`), `Site Code` != "") %>%
  mutate(
    juvenile_count = as.numeric(`# Juv`),
    adult_count = as.numeric(`# Adult`),
    mating_count = as.numeric(ReproductiveActivity)
  )

radial_survey_area_m2 <- pi * 10^2

radial_abundance <- conch_radial %>%
  select(`Site Code`, Lat, Lon, `Habitat (class at point)`) %>%
  left_join(
    radial_counts %>%
      select(`Site Code`, juvenile_count, adult_count, mating_count, `Habitat Description`),
    by = "Site Code"
  ) %>%
  mutate(
    juvenile_density_100m2 = juvenile_count / radial_survey_area_m2 * 100,
    adult_density_100m2 = adult_count / radial_survey_area_m2 * 100
  )

stopifnot(
  nrow(radial_abundance) == nrow(conch_radial),
  sum(is.na(radial_abundance$juvenile_count)) == 0,
  sum(is.na(radial_abundance$adult_count)) == 0
)

radial_abundance_long <- radial_abundance %>%
  select(
    `Site Code`,
    Lat,
    Lon,
    Adult = adult_count,
    Juvenile = juvenile_count
  ) %>%
  pivot_longer(
    cols = c(Adult, Juvenile),
    names_to = "age_class",
    values_to = "n_animals"
  )

radial_map_xlim <- range(radial_abundance$Lon, na.rm = TRUE) + c(-0.001, 0.001)
radial_map_ylim <- range(radial_abundance$Lat, na.rm = TRUE) + c(-0.001, 0.001)

radial_abundance_points_plot <- ggplot() +
  geom_sf(
    data = polygons_final,
    aes(fill = Type),
    color = "grey35",
    linewidth = 0.05,
    alpha = 0.55
  ) +
  geom_point(
    data = radial_abundance_long,
    aes(x = Lon, y = Lat, size = n_animals),
    shape = 21,
    fill = "#FDE725",
    color = "black",
    alpha = 0.85,
    stroke = 0.3
  ) +
  facet_wrap(~ age_class) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  scale_size_area(max_size = 8, breaks = c(0, 5, 10, 20), name = "Observed conch") +
  coord_sf(
    xlim = radial_map_xlim,
    ylim = radial_map_ylim,
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude", fill = "Habitat type") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

radial_abundance_points_plot

ggsave(
  filename = output_radial_abundance_points,
  plot = radial_abundance_points_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r radial-survey-abundance-heatmap, fig.width=8, fig.height=6}
radial_abundance_count_long <- radial_abundance %>%
  select(
    `Site Code`,
    Lat,
    Lon,
    Adult = adult_count,
    Juvenile = juvenile_count
  ) %>%
  pivot_longer(
    cols = c(Adult, Juvenile),
    names_to = "age_class",
    values_to = "n_animals"
  )

radial_abundance_count_sf <- radial_abundance_count_long %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32617)

radial_heatmap_grid <- radial_abundance_count_sf %>%
  st_union() %>%
  st_convex_hull() %>%
  st_buffer(150) %>%
  st_make_grid(cellsize = 20, what = "centers") %>%
  st_sf(geometry = .)

make_count_idw_surface <- function(age_class_name) {
  observed <- radial_abundance_count_sf %>%
    filter(age_class == age_class_name)

  gstat::idw(
    n_animals ~ 1,
    locations = observed,
    newdata = radial_heatmap_grid,
    idp = 2
  ) %>%
    mutate(age_class = age_class_name)
}

radial_abundance_count_surface <- bind_rows(
  lapply(c("Adult", "Juvenile"), make_count_idw_surface)
) %>%
  mutate(n_animals = pmax(var1.pred, 0))

radial_abundance_count_surface_df <- radial_abundance_count_surface %>%
  mutate(
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

radial_abundance_count_points_df <- radial_abundance_count_sf %>%
  mutate(
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

radial_heatmap_limits <- st_bbox(
  radial_abundance_count_sf %>%
    st_union() %>%
    st_convex_hull() %>%
    st_buffer(150)
)

radial_count_breaks <- c(0, 1, 2, 5, 10, 20, 40)

radial_abundance_heatmap <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, 32617),
    fill = "grey85",
    color = "grey70",
    linewidth = 0.05,
    alpha = 0.35
  ) +
  geom_tile(
    data = radial_abundance_count_surface_df,
    aes(x = x, y = y, fill = n_animals),
    width = 20,
    height = 20,
    alpha = 0.9
  ) +
  geom_point(
    data = radial_abundance_count_points_df,
    aes(x = x, y = y),
    shape = 21,
    fill = "white",
    color = "black",
    size = 1.2,
    stroke = 0.25
  ) +
  facet_wrap(~ age_class) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, max(radial_abundance_count_long$n_animals, na.rm = TRUE)),
    breaks = radial_count_breaks,
    trans = "sqrt",
    name = "Observed conch"
  ) +
  coord_sf(
    crs = st_crs(32617),
    xlim = c(radial_heatmap_limits["xmin"], radial_heatmap_limits["xmax"]),
    ylim = c(radial_heatmap_limits["ymin"], radial_heatmap_limits["ymax"]),
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

radial_abundance_heatmap

ggsave(
  filename = output_radial_abundance_heatmap,
  plot = radial_abundance_heatmap,
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r radial-survey-mating-map, fig.width=8, fig.height=6}
radial_mating_plot_data <- radial_abundance %>%
  filter(mating_count > 0)

radial_mating_map <- ggplot() +
  geom_sf(
    data = polygons_final,
    aes(fill = Type),
    color = "grey35",
    linewidth = 0.05,
    alpha = 0.55
  ) +
  geom_point(
    data = radial_abundance,
    aes(x = Lon, y = Lat),
    shape = 21,
    fill = "white",
    color = "grey30",
    size = 1.4,
    stroke = 0.25,
    alpha = 0.7
  ) +
  geom_point(
    data = radial_mating_plot_data,
    aes(x = Lon, y = Lat, size = mating_count),
    shape = 21,
    fill = "#D7301F",
    color = "black",
    alpha = 0.9,
    stroke = 0.3
  ) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  scale_size_area(max_size = 9, breaks = c(1, 2, 3, 5), name = "Mating observations") +
  coord_sf(
    xlim = radial_map_xlim,
    ylim = radial_map_ylim,
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude", fill = "Habitat type") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

radial_mating_map

ggsave(
  filename = output_radial_mating_map,
  plot = radial_mating_map,
  width = 8,
  height = 6,
  dpi = 300
)

radial_abundance %>%
  summarize(
    mating_sites = sum(mating_count > 0),
    total_mating_observations = sum(mating_count)
  )
```

```{r adult-mating-halo-model, fig.width=8, fig.height=6}
radial_abundance_sf_utm <- radial_abundance %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32617)

mating_sites_sf_utm <- radial_abundance_sf_utm %>%
  filter(mating_count > 0)

radial_to_mating_dist <- st_distance(radial_abundance_sf_utm, mating_sites_sf_utm) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(radial_abundance_sf_utm), ncol = nrow(mating_sites_sf_utm))

mating_weights <- mating_sites_sf_utm$mating_count

calculate_mating_halo <- function(dist_matrix, weights, scale_m) {
  as.numeric(exp(-dist_matrix / scale_m) %*% weights)
}

halo_scales_m <- seq(25, 500, by = 25)

adult_mating_halo_model_selection <- tibble(scale_m = halo_scales_m) %>%
  mutate(
    model = map(scale_m, function(scale_m) {
      model_data <- radial_abundance %>%
        mutate(
          survey_area_m2 = radial_survey_area_m2,
          mating_halo = calculate_mating_halo(
            radial_to_mating_dist,
            mating_weights,
            scale_m
          )
        )

      MASS::glm.nb(
        adult_count ~ mating_halo + offset(log(survey_area_m2)),
        data = model_data
      )
    }),
    AIC = map_dbl(model, AIC)
  ) %>%
  arrange(AIC)

best_halo_scale_m <- adult_mating_halo_model_selection$scale_m[[1]]

adult_mating_halo_data <- radial_abundance %>%
  mutate(
    survey_area_m2 = radial_survey_area_m2,
    distance_nearest_mating_m = as.numeric(apply(radial_to_mating_dist, 1, min)),
    mating_halo = calculate_mating_halo(
      radial_to_mating_dist,
      mating_weights,
      best_halo_scale_m
    )
  )

adult_mating_halo_points <- radial_abundance_sf_utm %>%
  mutate(
    adult_count = radial_abundance$adult_count,
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

adult_distance_model <- MASS::glm.nb(
  adult_count ~ distance_nearest_mating_m + offset(log(survey_area_m2)),
  data = adult_mating_halo_data
)

adult_mating_halo_model <- MASS::glm.nb(
  adult_count ~ mating_halo + offset(log(survey_area_m2)),
  data = adult_mating_halo_data
)

adult_mating_model_comparison <- tibble(
  model = c("nearest mating distance", "additive mating halo"),
  AIC = c(AIC(adult_distance_model), AIC(adult_mating_halo_model))
) %>%
  mutate(delta_AIC = AIC - min(AIC))

halo_prediction_grid <- radial_abundance_sf_utm %>%
  st_union() %>%
  st_convex_hull() %>%
  st_buffer(150) %>%
  st_make_grid(cellsize = 20, what = "centers") %>%
  st_sf(geometry = .)

grid_to_mating_dist <- st_distance(halo_prediction_grid, mating_sites_sf_utm) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(halo_prediction_grid), ncol = nrow(mating_sites_sf_utm))

halo_prediction_grid <- halo_prediction_grid %>%
  mutate(
    mating_halo = calculate_mating_halo(
      grid_to_mating_dist,
      mating_weights,
      best_halo_scale_m
    ),
    survey_area_m2 = 1,
    adult_density_m2 = predict(
      adult_mating_halo_model,
      newdata = tibble(mating_halo = mating_halo, survey_area_m2 = survey_area_m2),
      type = "response"
    ),
    adult_density_100m2 = adult_density_m2 * 100,
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  )

halo_map_limits <- st_bbox(
  radial_abundance_sf_utm %>%
    st_union() %>%
    st_convex_hull() %>%
    st_buffer(150)
)

adult_mating_halo_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, 32617),
    fill = "grey85",
    color = "grey70",
    linewidth = 0.05,
    alpha = 0.35
  ) +
  geom_tile(
    data = st_drop_geometry(halo_prediction_grid),
    aes(x = x, y = y, fill = adult_density_100m2),
    width = 20,
    height = 20,
    alpha = 0.9
  ) +
  geom_point(
    data = adult_mating_halo_points,
    aes(x = x, y = y, size = adult_count),
    shape = 21,
    fill = "white",
    color = "black",
    stroke = 0.25,
    alpha = 0.8
  ) +
  geom_sf(
    data = mating_sites_sf_utm,
    shape = 24,
    fill = "#D7301F",
    color = "black",
    size = 3,
    stroke = 0.35
  ) +
  scale_fill_viridis_c(
    option = "magma",
    name = expression("Predicted adults per 100 " * m^2)
  ) +
  scale_size_area(max_size = 7, name = "Observed adults") +
  coord_sf(
    crs = st_crs(32617),
    xlim = c(halo_map_limits["xmin"], halo_map_limits["xmax"]),
    ylim = c(halo_map_limits["ymin"], halo_map_limits["ymax"]),
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

adult_mating_halo_map

ggsave(
  filename = output_adult_mating_halo_map,
  plot = adult_mating_halo_map,
  width = 8,
  height = 6,
  dpi = 300
)

adult_mating_halo_model_selection %>%
  select(scale_m, AIC) %>%
  slice_head(n = 5)

adult_mating_model_comparison
summary(adult_mating_halo_model)
```

```{r adult-mating-distance-model-plots, fig.width=8, fig.height=6}
adult_distance_prediction <- tibble(
  distance_nearest_mating_m = seq(
    0,
    max(adult_mating_halo_data$distance_nearest_mating_m, na.rm = TRUE),
    length.out = 200
  ),
  survey_area_m2 = radial_survey_area_m2
)

adult_distance_prediction_link <- predict(
  adult_distance_model,
  newdata = adult_distance_prediction,
  type = "link",
  se.fit = TRUE
)

adult_distance_prediction <- adult_distance_prediction %>%
  mutate(
    adult_count_fit = exp(adult_distance_prediction_link$fit),
    adult_count_lwr = exp(adult_distance_prediction_link$fit - 1.96 * adult_distance_prediction_link$se.fit),
    adult_count_upr = exp(adult_distance_prediction_link$fit + 1.96 * adult_distance_prediction_link$se.fit),
    adult_density_100m2_fit = adult_count_fit / radial_survey_area_m2 * 100,
    adult_density_100m2_lwr = adult_count_lwr / radial_survey_area_m2 * 100,
    adult_density_100m2_upr = adult_count_upr / radial_survey_area_m2 * 100
  )

adult_mating_distance_plot <- ggplot() +
  geom_point(
    data = adult_mating_halo_data,
    aes(
      x = distance_nearest_mating_m,
      y = adult_count / radial_survey_area_m2 * 100
    ),
    shape = 21,
    fill = "white",
    color = "black",
    alpha = 0.8
  ) +
  geom_ribbon(
    data = adult_distance_prediction,
    aes(
      x = distance_nearest_mating_m,
      ymin = adult_density_100m2_lwr,
      ymax = adult_density_100m2_upr
    ),
    fill = "#2C7FB8",
    alpha = 0.2
  ) +
  geom_line(
    data = adult_distance_prediction,
    aes(x = distance_nearest_mating_m, y = adult_density_100m2_fit),
    color = "#2C7FB8",
    linewidth = 1
  ) +
  labs(
    x = "Distance to nearest mating observation (m)",
    y = expression("Adult conch density (adults per 100 " * m^2 * ")")
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

adult_mating_distance_plot

ggsave(
  filename = output_adult_mating_distance_plot,
  plot = adult_mating_distance_plot,
  width = 8,
  height = 6,
  dpi = 300
)

grid_to_nearest_mating_dist <- apply(grid_to_mating_dist, 1, min)

distance_prediction_grid <- halo_prediction_grid %>%
  mutate(
    distance_nearest_mating_m = grid_to_nearest_mating_dist,
    survey_area_m2 = 1,
    adult_density_m2 = predict(
      adult_distance_model,
      newdata = tibble(
        distance_nearest_mating_m = distance_nearest_mating_m,
        survey_area_m2 = survey_area_m2
      ),
      type = "response"
    ),
    adult_density_100m2 = adult_density_m2 * 100,
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  )

adult_mating_distance_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, 32617),
    fill = "grey85",
    color = "grey70",
    linewidth = 0.05,
    alpha = 0.35
  ) +
  geom_tile(
    data = st_drop_geometry(distance_prediction_grid),
    aes(x = x, y = y, fill = adult_density_100m2),
    width = 20,
    height = 20,
    alpha = 0.9
  ) +
  geom_point(
    data = adult_mating_halo_points,
    aes(x = x, y = y, size = adult_count),
    shape = 21,
    fill = "white",
    color = "black",
    stroke = 0.25,
    alpha = 0.8
  ) +
  geom_sf(
    data = mating_sites_sf_utm,
    shape = 24,
    fill = "#D7301F",
    color = "black",
    size = 3,
    stroke = 0.35
  ) +
  scale_fill_viridis_c(
    option = "magma",
    name = expression("Predicted adults per 100 " * m^2)
  ) +
  scale_size_area(max_size = 7, name = "Observed adults") +
  coord_sf(
    crs = st_crs(32617),
    xlim = c(halo_map_limits["xmin"], halo_map_limits["xmax"]),
    ylim = c(halo_map_limits["ymin"], halo_map_limits["ymax"]),
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

adult_mating_distance_map

ggsave(
  filename = output_adult_mating_distance_map,
  plot = adult_mating_distance_map,
  width = 8,
  height = 6,
  dpi = 300
)

adult_mating_model_comparison
summary(adult_distance_model)
```

```{r adult-mating-radius-density-model, fig.width=8, fig.height=6}
calculate_mating_within_radius <- function(dist_matrix, weights, radius_m) {
  as.numeric((dist_matrix <= radius_m) %*% weights)
}

mating_radii_m <- seq(25, 500, by = 25)

adult_mating_radius_model_selection <- tibble(radius_m = mating_radii_m) %>%
  mutate(
    model = map(radius_m, function(radius_m) {
      model_data <- adult_mating_halo_data %>%
        mutate(
          mating_within_radius = calculate_mating_within_radius(
            radial_to_mating_dist,
            mating_weights,
            radius_m
          )
        )

      MASS::glm.nb(
        adult_count ~ mating_within_radius + offset(log(survey_area_m2)),
        data = model_data
      )
    }),
    AIC = map_dbl(model, AIC)
  ) %>%
  arrange(AIC)

best_mating_radius_m <- adult_mating_radius_model_selection$radius_m[[1]]

adult_mating_radius_data <- adult_mating_halo_data %>%
  mutate(
    mating_within_radius = calculate_mating_within_radius(
      radial_to_mating_dist,
      mating_weights,
      best_mating_radius_m
    )
  )

adult_mating_radius_model <- MASS::glm.nb(
  adult_count ~ mating_within_radius + offset(log(survey_area_m2)),
  data = adult_mating_radius_data
)

adult_mating_model_comparison <- bind_rows(
  adult_mating_model_comparison,
  tibble(
    model = paste0("mating observations within ", best_mating_radius_m, " m"),
    AIC = AIC(adult_mating_radius_model)
  )
) %>%
  mutate(delta_AIC = AIC - min(AIC)) %>%
  arrange(AIC)

adult_radius_prediction <- tibble(
  mating_within_radius = seq(
    0,
    max(adult_mating_radius_data$mating_within_radius, na.rm = TRUE),
    by = 1
  ),
  survey_area_m2 = radial_survey_area_m2
)

adult_radius_prediction_link <- predict(
  adult_mating_radius_model,
  newdata = adult_radius_prediction,
  type = "link",
  se.fit = TRUE
)

adult_radius_prediction <- adult_radius_prediction %>%
  mutate(
    adult_count_fit = exp(adult_radius_prediction_link$fit),
    adult_count_lwr = exp(adult_radius_prediction_link$fit - 1.96 * adult_radius_prediction_link$se.fit),
    adult_count_upr = exp(adult_radius_prediction_link$fit + 1.96 * adult_radius_prediction_link$se.fit),
    adult_density_100m2_fit = adult_count_fit / radial_survey_area_m2 * 100,
    adult_density_100m2_lwr = adult_count_lwr / radial_survey_area_m2 * 100,
    adult_density_100m2_upr = adult_count_upr / radial_survey_area_m2 * 100
  )

adult_mating_radius_plot <- ggplot() +
  geom_point(
    data = adult_mating_radius_data,
    aes(
      x = mating_within_radius,
      y = adult_count / radial_survey_area_m2 * 100
    ),
    shape = 21,
    fill = "white",
    color = "black",
    alpha = 0.8
  ) +
  geom_ribbon(
    data = adult_radius_prediction,
    aes(
      x = mating_within_radius,
      ymin = adult_density_100m2_lwr,
      ymax = adult_density_100m2_upr
    ),
    fill = "#7B3294",
    alpha = 0.2
  ) +
  geom_line(
    data = adult_radius_prediction,
    aes(x = mating_within_radius, y = adult_density_100m2_fit),
    color = "#7B3294",
    linewidth = 1
  ) +
  labs(
    x = paste0("Mating observations within ", best_mating_radius_m, " m"),
    y = expression("Adult conch density (adults per 100 " * m^2 * ")")
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

adult_mating_radius_plot

ggsave(
  filename = output_adult_mating_radius_plot,
  plot = adult_mating_radius_plot,
  width = 8,
  height = 6,
  dpi = 300
)

radius_prediction_grid <- halo_prediction_grid %>%
  mutate(
    mating_within_radius = calculate_mating_within_radius(
      grid_to_mating_dist,
      mating_weights,
      best_mating_radius_m
    ),
    survey_area_m2 = 1,
    adult_density_m2 = predict(
      adult_mating_radius_model,
      newdata = tibble(
        mating_within_radius = mating_within_radius,
        survey_area_m2 = survey_area_m2
      ),
      type = "response"
    ),
    adult_density_100m2 = adult_density_m2 * 100,
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  )

adult_radius_map_max <- max(radius_prediction_grid$adult_density_100m2, na.rm = TRUE)

adult_mating_radius_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, 32617),
    fill = "grey85",
    color = "grey70",
    linewidth = 0.05,
    alpha = 0.35
  ) +
  geom_tile(
    data = st_drop_geometry(radius_prediction_grid),
    aes(x = x, y = y, fill = adult_density_100m2),
    width = 20,
    height = 20,
    alpha = 0.9
  ) +
  geom_point(
    data = adult_mating_halo_points,
    aes(x = x, y = y, size = adult_count),
    shape = 21,
    fill = "white",
    color = "black",
    stroke = 0.25,
    alpha = 0.8
  ) +
  geom_sf(
    data = mating_sites_sf_utm,
    shape = 24,
    fill = "#D7301F",
    color = "black",
    size = 3,
    stroke = 0.35
  ) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, adult_radius_map_max),
    trans = "sqrt",
    name = expression("Predicted adults per 100 " * m^2)
  ) +
  scale_size_area(max_size = 7, name = "Observed adults") +
  coord_sf(
    crs = st_crs(32617),
    xlim = c(halo_map_limits["xmin"], halo_map_limits["xmax"]),
    ylim = c(halo_map_limits["ymin"], halo_map_limits["ymax"]),
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

adult_mating_radius_map

ggsave(
  filename = output_adult_mating_radius_map,
  plot = adult_mating_radius_map,
  width = 8,
  height = 6,
  dpi = 300
)

adult_mating_radius_model_selection %>%
  select(radius_m, AIC) %>%
  slice_head(n = 5)

adult_mating_model_comparison
summary(adult_mating_radius_model)
```

```{r blob-population-and-background-density, fig.width=8, fig.height=6}
blob_buffer_m <- 150
prediction_cellsize_m <- 20

radial_blob_candidates <- radial_abundance_sf_utm %>%
  st_buffer(blob_buffer_m) %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  st_cast("POLYGON") %>%
  mutate(
    blob_id = row_number(),
    area_m2 = as.numeric(st_area(geometry)),
    n_radial_sites = lengths(st_intersects(., radial_abundance_sf_utm))
  ) %>%
  arrange(desc(n_radial_sites), desc(area_m2))

radial_blob <- radial_blob_candidates %>%
  slice(1)

blob_grid <- st_make_grid(
  radial_blob,
  cellsize = prediction_cellsize_m,
  what = "polygons",
  square = TRUE
) %>%
  st_sf(geometry = .) %>%
  st_intersection(radial_blob) %>%
  mutate(cell_area_m2 = as.numeric(st_area(geometry))) %>%
  filter(cell_area_m2 > 0)

blob_grid_points <- st_point_on_surface(blob_grid)

blob_to_mating_dist <- st_distance(blob_grid_points, mating_sites_sf_utm) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(blob_grid_points), ncol = nrow(mating_sites_sf_utm))

blob_adult_prediction <- blob_grid %>%
  mutate(
    mating_within_radius = calculate_mating_within_radius(
      blob_to_mating_dist,
      mating_weights,
      best_mating_radius_m
    ),
    survey_area_m2 = 1,
    adult_density_m2 = predict(
      adult_mating_radius_model,
      newdata = tibble(
        mating_within_radius = mating_within_radius,
        survey_area_m2 = survey_area_m2
      ),
      type = "response"
    ),
    adult_total = adult_density_m2 * cell_area_m2
  )

blob_total_adults <- sum(blob_adult_prediction$adult_total, na.rm = TRUE)

blob_grid_points_df <- blob_grid_points %>%
  mutate(
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

blob_to_radial_dist <- st_distance(blob_grid_points, radial_abundance_count_sf) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(blob_grid_points), ncol = nrow(radial_abundance_count_sf))

calculate_idw_prediction <- function(dist_matrix, values, power = 2) {
  apply(dist_matrix, 1, function(d) {
    if (any(d == 0)) {
      mean(values[d == 0], na.rm = TRUE)
    } else {
      weights <- 1 / d^power
      sum(weights * values, na.rm = TRUE) / sum(weights, na.rm = TRUE)
    }
  })
}

juvenile_radial_values <- radial_abundance_count_sf %>%
  filter(age_class == "Juvenile") %>%
  pull(n_animals)

blob_to_juvenile_radial_dist <- st_distance(
  blob_grid_points,
  radial_abundance_count_sf %>% filter(age_class == "Juvenile")
) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(blob_grid_points), ncol = length(juvenile_radial_values))

blob_juvenile_prediction <- blob_grid %>%
  mutate(
    juvenile_count_idw = calculate_idw_prediction(
      blob_to_juvenile_radial_dist,
      juvenile_radial_values,
      power = 2
    ),
    juvenile_density_m2 = juvenile_count_idw / radial_survey_area_m2,
    juvenile_total = juvenile_density_m2 * cell_area_m2
  )

blob_total_juveniles <- sum(blob_juvenile_prediction$juvenile_total, na.rm = TRUE)

blob_prediction_long <- bind_rows(
  blob_adult_prediction %>%
    mutate(
      age_class = "Adult",
      density_100m2 = adult_density_m2 * 100,
      total_animals = adult_total
    ) %>%
    select(age_class, density_100m2, total_animals, geometry),
  blob_juvenile_prediction %>%
    mutate(
      age_class = "Juvenile",
      density_100m2 = juvenile_density_m2 * 100,
      total_animals = juvenile_total
    ) %>%
    select(age_class, density_100m2, total_animals, geometry)
)

blob_prediction_points <- blob_prediction_long %>%
  st_point_on_surface() %>%
  mutate(
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

blob_density_max <- max(blob_prediction_points$density_100m2, na.rm = TRUE)

conch_blob_prediction_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, 32617),
    fill = "grey85",
    color = "grey70",
    linewidth = 0.05,
    alpha = 0.35
  ) +
  geom_tile(
    data = blob_prediction_points,
    aes(x = x, y = y, fill = density_100m2),
    width = prediction_cellsize_m,
    height = prediction_cellsize_m,
    alpha = 0.9
  ) +
  geom_sf(
    data = radial_blob,
    fill = NA,
    color = "black",
    linewidth = 0.6
  ) +
  geom_point(
    data = adult_mating_halo_points,
    aes(x = x, y = y),
    shape = 21,
    fill = "white",
    color = "black",
    size = 1,
    stroke = 0.2,
    alpha = 0.7
  ) +
  facet_wrap(~ age_class) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, blob_density_max),
    trans = "sqrt",
    name = expression("Conch per 100 " * m^2)
  ) +
  coord_sf(
    crs = st_crs(32617),
    xlim = st_bbox(radial_blob)[c("xmin", "xmax")],
    ylim = st_bbox(radial_blob)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

conch_blob_prediction_map

ggsave(
  filename = output_conch_blob_prediction_map,
  plot = conch_blob_prediction_map,
  width = 8,
  height = 6,
  dpi = 300
)

if (!exists("drift_rectangles")) {
  drift_surveys <- readxl::read_xlsx(conch_xlsx, sheet = drift_sheet) %>%
    mutate(
      start_lat = as.numeric(`StartLat (dec deg)`),
      start_lon = as.numeric(`StartLong (dec deg)`),
      end_lat = as.numeric(`EndLat (dec deg)`),
      end_lon = as.numeric(`EndLon (dec deg)`)
    ) %>%
    filter(
      !is.na(start_lat),
      !is.na(start_lon),
      !is.na(end_lat),
      !is.na(end_lon)
    ) %>%
    mutate(.record = row_number())

  drift_start_sf <- drift_surveys %>%
    st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
    st_transform(32617)

  drift_end_sf <- drift_surveys %>%
    st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
    st_transform(32617)

  start_xy <- st_coordinates(drift_start_sf)
  end_xy <- st_coordinates(drift_end_sf)

  make_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
    direction <- end_point - start_point
    transect_length <- sqrt(sum(direction^2))

    if (transect_length == 0) {
      return(st_polygon())
    }

    unit_direction <- direction / transect_length
    perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

    corners <- rbind(
      start_point + perpendicular,
      end_point + perpendicular,
      end_point - perpendicular,
      start_point - perpendicular,
      start_point + perpendicular
    )

    st_polygon(list(corners))
  }

  drift_rectangles <- st_sf(
    drift_surveys %>% select(-start_lat, -start_lon, -end_lat, -end_lon),
    geometry = st_sfc(
      map2(
        seq_len(nrow(start_xy)),
        seq_len(nrow(end_xy)),
        ~ make_transect_rectangle(start_xy[.x, ], end_xy[.y, ], half_width_m = 2)
      ),
      crs = 32617
    )
  ) %>%
    filter(!st_is_empty(.)) %>%
    st_transform(4326)
}

if (!exists("drift_habitat_area")) {
  drift_habitat_intersections <- st_intersection(
    drift_rectangles %>% st_transform(32617) %>% st_make_valid(),
    polygons_final %>% st_transform(32617) %>% st_make_valid() %>% select(Type)
  ) %>%
    mutate(area_m2 = as.numeric(st_area(geometry)))

  drift_habitat_area <- drift_habitat_intersections %>%
    st_drop_geometry() %>%
    group_by(.record, Transect, Line, Type) %>%
    summarize(area_m2 = sum(area_m2), .groups = "drop") %>%
    arrange(.record, desc(area_m2), Type)
}

if (!exists("conch_drift_positions_out")) {
  conch_drift_positions <- readxl::read_xlsx(conch_xlsx, sheet = drift_position_sheet) %>%
    mutate(
      .record = row_number(),
      lat = as.numeric(`Lat (dec deg)`),
      lon = as.numeric(`Long (dec deg)`),
      animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`),
      animal_stage = case_when(
        grepl("^A", animal_id) ~ "Adult",
        grepl("^J", animal_id) ~ "Juvenile",
        TRUE ~ NA_character_
      ),
      transect = sub("([AJ][0-9]+)$", "", `WPT (B/E)`)
    ) %>%
    filter(!is.na(lat), !is.na(lon))

  conch_drift_positions_sf <- conch_drift_positions %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  conch_drift_position_matches <- st_join(
    conch_drift_positions_sf,
    polygons_for_join,
    join = st_within,
    left = TRUE
  ) %>%
    st_drop_geometry() %>%
    select(.record, Habitat_Type = Type)

  conch_drift_positions_out <- conch_drift_positions %>%
    left_join(
      conch_drift_position_matches %>%
        group_by(.record) %>%
        slice(1) %>%
        ungroup(),
      by = ".record"
    ) %>%
    mutate(`Habitat (Class at point)` = Habitat_Type) %>%
    select(-Habitat_Type)
}

drift_rectangles_utm <- drift_rectangles %>%
  st_transform(32617) %>%
  st_make_valid()

drift_background_records <- drift_rectangles_utm %>%
  mutate(
    transect_group = sub(" cont\\.$", "", Transect),
    intersects_blob = lengths(st_intersects(., radial_blob)) > 0
  ) %>%
  st_drop_geometry() %>%
  select(.record, Transect, transect_group, Line, intersects_blob)

background_drift_records <- drift_background_records %>%
  group_by(transect_group) %>%
  summarize(intersects_blob = any(intersects_blob), .groups = "drop") %>%
  filter(!intersects_blob)

background_drift_area <- drift_habitat_area %>%
  mutate(transect_group = sub(" cont\\.$", "", Transect)) %>%
  semi_join(background_drift_records, by = "transect_group")

background_drift_counts <- conch_drift_positions_out %>%
  semi_join(background_drift_records, by = c("transect" = "transect_group")) %>%
  count(
    transect,
    `Habitat (Class at point)`,
    animal_stage,
    name = "n_animals"
  ) %>%
  rename(
    Transect = transect,
    Type = `Habitat (Class at point)`
  )

background_drift_density <- background_drift_area %>%
  expand_grid(animal_stage = c("Adult", "Juvenile")) %>%
  left_join(
    background_drift_counts,
    by = c("Transect", "Type", "animal_stage")
  ) %>%
  mutate(n_animals = replace_na(n_animals, 0L)) %>%
  group_by(Type, animal_stage) %>%
  summarize(
    n_animals = sum(n_animals),
    area_m2 = sum(area_m2),
    density_m2 = n_animals / area_m2,
    density_100m2 = density_m2 * 100,
    .groups = "drop"
  ) %>%
  arrange(Type, animal_stage)

background_drift_density_plot <- ggplot(
  background_drift_density,
  aes(x = Type, y = density_100m2, fill = animal_stage)
) +
  geom_col(position = position_dodge(width = 0.75), color = "grey30", linewidth = 0.2) +
  coord_flip() +
  labs(
    x = "Habitat type",
    y = expression("Background density (conch per 100 " * m^2 * ")"),
    fill = "Conch stage"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

background_drift_density_plot

ggsave(
  filename = output_drift_background_density_plot,
  plot = background_drift_density_plot,
  width = 8,
  height = 6,
  dpi = 300
)

tibble(
  blob_buffer_m = blob_buffer_m,
  blob_area_m2 = sum(blob_adult_prediction$cell_area_m2),
  blob_area_ha = sum(blob_adult_prediction$cell_area_m2) / 10000,
  n_radial_sites_in_blob = radial_blob$n_radial_sites,
  best_mating_radius_m = best_mating_radius_m,
  estimated_adults_in_blob = blob_total_adults,
  estimated_juveniles_in_blob = blob_total_juveniles,
  background_drift_transects = n_distinct(background_drift_records$transect_group),
  excluded_drift_transects = drift_background_records %>%
    group_by(transect_group) %>%
    summarize(intersects_blob = any(intersects_blob), .groups = "drop") %>%
    summarize(n = sum(intersects_blob)) %>%
    pull(n)
)

background_drift_density
```

## Map

```{r habitat-map, fig.width=8, fig.height=8}
conch_map <- ggplot() +
  geom_sf(
    data = polygons_final,
    aes(fill = Type),
    color = "grey35",
    linewidth = 0.05,
    alpha = 0.65
  ) +
  geom_sf(
    data = conch_points_sf,
    shape = 21,
    fill = "black",
    color = "white",
    stroke = 0.25,
    size = 2
  ) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  coord_sf(
    xlim = c(-80.1116, -80.077),
    ylim = c(26.058, 26.112),
    expand = FALSE
  ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "Habitat type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

conch_map +
  coord_sf(xlim = c(-80.1116, -80.1),
    ylim = c(26.08, 26.095))

ggsave(
  filename = output_map,
  plot = conch_map,
  width = 8,
  height = 8,
  dpi = 300
)
```

# Drift Survey Transects

```{r drift-survey-transects, fig.width=8, fig.height=8}
drift_surveys <- readxl::read_xlsx(conch_xlsx, sheet = drift_sheet) %>%
  mutate(
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(
    !is.na(start_lat),
    !is.na(start_lon),
    !is.na(end_lat),
    !is.na(end_lon)
  ) %>%
  mutate(.record = row_number())

drift_start_sf <- drift_surveys %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32617)

drift_end_sf <- drift_surveys %>%
  st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(32617)

start_xy <- st_coordinates(drift_start_sf)
end_xy <- st_coordinates(drift_end_sf)

make_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(st_polygon())
  }

  unit_direction <- direction / transect_length
  perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

  corners <- rbind(
    start_point + perpendicular,
    end_point + perpendicular,
    end_point - perpendicular,
    start_point - perpendicular,
    start_point + perpendicular
  )

  st_polygon(list(corners))
}

drift_rectangles <- st_sf(
  drift_surveys %>% select(-start_lat, -start_lon, -end_lat, -end_lon),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ make_transect_rectangle(start_xy[.x, ], end_xy[.y, ], half_width_m = 2)
    ),
    crs = 32617
  )
) %>%
  filter(!st_is_empty(.)) %>%
  st_transform(4326)

mid_xy <- (start_xy + end_xy) / 2

drift_label_points <- drift_surveys %>%
  mutate(
    label_x = mid_xy[, "X"],
    label_y = mid_xy[, "Y"]
  ) %>%
  st_as_sf(coords = c("label_x", "label_y"), crs = 32617) %>%
  st_transform(4326)

drift_label_df <- drift_label_points %>%
  mutate(
    label_lon = st_coordinates(.)[, "X"],
    label_lat = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

drift_map <- ggplot() +
  geom_sf(
    data = polygons_final,
    aes(fill = Type),
    color = "grey35",
    linewidth = 0.05,
    alpha = 0.65
  ) +
  geom_sf(
    data = drift_rectangles,
    fill = NA,
    color = "black",
    linewidth = 0.45
  ) +
  geom_text(
    data = drift_label_df,
    aes(x = label_lon, y = label_lat, label = Line),
    size = 2.5,
    color = "black"
  ) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  coord_sf(
    xlim = c(-80.1116, -80.077),
    ylim = c(26.058, 26.112),
    expand = FALSE
  ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "Habitat type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

drift_map

ggsave(
  filename = output_drift_map,
  plot = drift_map,
  width = 8,
  height = 8,
  dpi = 300
)
```

## Drift Survey Habitat Areas

```{r drift-survey-habitat-areas, fig.width=8, fig.height=5}
drift_rectangles_utm <- drift_rectangles %>%
  st_transform(32617) %>%
  st_make_valid()

polygons_final_utm <- polygons_final %>%
  st_transform(32617) %>%
  st_make_valid()

drift_habitat_intersections <- st_intersection(
  drift_rectangles_utm,
  polygons_final_utm %>% select(Type)
) %>%
  mutate(area_m2 = as.numeric(st_area(geometry)))

drift_habitat_area <- drift_habitat_intersections %>%
  st_drop_geometry() %>%
  group_by(.record, Transect, Line, Type) %>%
  summarize(area_m2 = sum(area_m2), .groups = "drop") %>%
  arrange(.record, desc(area_m2), Type)

drift_habitat_totals <- drift_habitat_area %>%
  group_by(Type) %>%
  summarize(area_m2 = sum(area_m2), .groups = "drop") %>%
  arrange(desc(area_m2))

drift_habitat_histogram <- ggplot(
  drift_habitat_totals,
  aes(x = reorder(Type, area_m2), y = area_m2, fill = Type)
) +
  geom_col(color = "grey30", linewidth = 0.2) +
  coord_flip() +
  scale_fill_manual(values = type_cols, na.value = "grey80", guide = "none") +
  labs(
    x = "Habitat type",
    y = expression("Area within drift rectangles (" * m^2 * ")")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

drift_habitat_histogram

ggsave(
  filename = output_drift_habitat_histogram,
  plot = drift_habitat_histogram,
  width = 8,
  height = 5,
  dpi = 300
)

drift_habitat_area
```

## Conch Drift Positions

```{r conch-drift-positions, fig.width=9, fig.height=5}
conch_drift_positions <- readxl::read_xlsx(conch_xlsx, sheet = drift_position_sheet) %>%
  mutate(
    .record = row_number(),
    lat = as.numeric(`Lat (dec deg)`),
    lon = as.numeric(`Long (dec deg)`),
    animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`),
    animal_stage = case_when(
      grepl("^A", animal_id) ~ "Adult",
      grepl("^J", animal_id) ~ "Juvenile",
      TRUE ~ NA_character_
    ),
    transect = sub("([AJ][0-9]+)$", "", `WPT (B/E)`)
  ) %>%
  filter(!is.na(lat), !is.na(lon))

conch_drift_positions_sf <- conch_drift_positions %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

conch_drift_position_matches <- st_join(
  conch_drift_positions_sf,
  polygons_for_join,
  join = st_within,
  left = TRUE
) %>%
  st_drop_geometry() %>%
  select(
    .record,
    `WPT (B/E)`,
    transect,
    animal_id,
    animal_stage,
    lat,
    lon,
    Habitat_Type = Type
  )

ambiguous_drift_positions <- conch_drift_position_matches %>%
  count(.record, `WPT (B/E)`, transect, animal_id, name = "n_matches") %>%
  filter(n_matches > 1)

conch_drift_position_assignments <- conch_drift_position_matches %>%
  group_by(.record) %>%
  slice(1) %>%
  ungroup()

conch_drift_positions_out <- conch_drift_positions %>%
  left_join(
    conch_drift_position_assignments %>% select(.record, Habitat_Type),
    by = ".record"
  ) %>%
  mutate(`Habitat (Class at point)` = Habitat_Type) %>%
  select(-Habitat_Type)

transect_levels <- drift_surveys$Transect %>%
  sub(" cont\\.$", "", .) %>%
  unique()

conch_drift_position_counts <- conch_drift_positions_out %>%
  mutate(
    transect = factor(transect, levels = transect_levels),
    `Habitat (Class at point)` = if_else(
      is.na(`Habitat (Class at point)`),
      "Unmatched",
      `Habitat (Class at point)`
    )
  ) %>%
  count(transect, `Habitat (Class at point)`, name = "n_animals")

conch_drift_position_counts_plot <- ggplot(
  conch_drift_position_counts,
  aes(x = transect, y = n_animals, fill = `Habitat (Class at point)`)
) +
  geom_col(color = "grey30", linewidth = 0.2) +
  scale_fill_manual(values = c(type_cols, "Unmatched" = "grey70"), na.value = "grey70") +
  labs(
    x = "Transect",
    y = "Observed conch",
    fill = "Habitat type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position = "right"
  )

conch_drift_position_counts_plot

ggsave(
  filename = output_drift_position_counts,
  plot = conch_drift_position_counts_plot,
  width = 9,
  height = 5,
  dpi = 300
)

conch_drift_position_counts
ambiguous_drift_positions
```

```{r conch-drift-observation-map, fig.width=8, fig.height=8}
drift_conch_map <- ggplot() +
  geom_sf(
    data = polygons_final,
    aes(fill = Type),
    color = "grey35",
    linewidth = 0.05,
    alpha = 0.65
  ) +
  geom_sf(
    data = drift_rectangles,
    fill = NA,
    color = "black",
    linewidth = 0.45
  ) +
  geom_sf(
    data = conch_drift_positions_sf,
    aes(color = animal_stage),
    size = 2,
    alpha = 0.9
  ) +
  geom_text(
    data = drift_label_df,
    aes(x = label_lon, y = label_lat, label = Line),
    size = 2.5,
    color = "black"
  ) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  scale_color_manual(
    values = c("Adult" = "#D95F02", "Juvenile" = "#1B9E77"),
    na.value = "grey20"
  ) +
  coord_sf(
    xlim = c(-80.1116, -80.077),
    ylim = c(26.058, 26.112),
    expand = FALSE
  ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "Habitat type",
    color = "Conch stage"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    legend.position = "right"
  )

drift_conch_map

ggsave(
  filename = output_drift_conch_map,
  plot = drift_conch_map,
  width = 8,
  height = 8,
  dpi = 300
)
```

# Conch Density Model

```{r conch-density-model, fig.width=8, fig.height=5}
conch_counts_by_habitat <- conch_drift_positions_out %>%
  count(
    transect,
    `Habitat (Class at point)`,
    animal_stage,
    name = "n_animals"
  ) %>%
  rename(
    Transect = transect,
    Type = `Habitat (Class at point)`
  )

age_classes <- c("Adult", "Juvenile")

conch_density_model_data <- drift_habitat_area %>%
  expand_grid(animal_stage = age_classes) %>%
  left_join(
    conch_counts_by_habitat,
    by = c("Transect", "Type", "animal_stage")
  ) %>%
  mutate(
    n_animals = replace_na(n_animals, 0L),
    Type = factor(Type, levels = names(type_cols)),
    animal_stage = factor(animal_stage, levels = age_classes)
  ) %>%
  filter(area_m2 > 0)

conch_density_model <- MASS::glm.nb(
  n_animals ~ Type + animal_stage + offset(log(area_m2)),
  data = conch_density_model_data
)

density_prediction_grid <- conch_density_model_data %>%
  st_drop_geometry() %>%
  distinct(Type, animal_stage) %>%
  arrange(Type, animal_stage) %>%
  mutate(area_m2 = 1)

density_prediction <- predict(
  conch_density_model,
  newdata = density_prediction_grid,
  type = "link",
  se.fit = TRUE
)

observed_density_summary <- conch_density_model_data %>%
  group_by(Type, animal_stage) %>%
  summarize(
    observed_count = sum(n_animals),
    sampled_area_m2 = sum(area_m2),
    observed_density_per_m2 = observed_count / sampled_area_m2,
    .groups = "drop"
  )

conch_density_estimates <- density_prediction_grid %>%
  mutate(
    density_per_m2 = exp(density_prediction$fit),
    density_lwr_per_m2 = exp(density_prediction$fit - 1.96 * density_prediction$se.fit),
    density_upr_per_m2 = exp(density_prediction$fit + 1.96 * density_prediction$se.fit),
    density_per_100m2 = density_per_m2 * 100,
    density_lwr_per_100m2 = density_lwr_per_m2 * 100,
    density_upr_per_100m2 = density_upr_per_m2 * 100
  ) %>%
  left_join(observed_density_summary, by = c("Type", "animal_stage")) %>%
  mutate(
    estimable = observed_count > 0,
    density_per_m2 = if_else(estimable, density_per_m2, 0),
    density_lwr_per_m2 = if_else(estimable, density_lwr_per_m2, NA_real_),
    density_upr_per_m2 = if_else(estimable, density_upr_per_m2, NA_real_),
    density_per_100m2 = if_else(estimable, density_per_100m2, 0),
    density_lwr_per_100m2 = if_else(estimable, density_lwr_per_100m2, NA_real_),
    density_upr_per_100m2 = if_else(estimable, density_upr_per_100m2, NA_real_)
  ) %>%
  arrange(Type, animal_stage)

conch_density_plot <- ggplot(
  conch_density_estimates,
  aes(x = Type, y = density_per_100m2, fill = animal_stage)
) +
  geom_col(position = position_dodge(width = 0.75), color = "grey30", linewidth = 0.2) +
  geom_errorbar(
    aes(ymin = density_lwr_per_100m2, ymax = density_upr_per_100m2),
    position = position_dodge(width = 0.75),
    width = 0.2,
    na.rm = TRUE
  ) +
  coord_flip() +
  labs(
    x = "Habitat type",
    y = expression("Estimated conch density (animals per 100 " * m^2 * ")"),
    fill = "Conch stage"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

conch_density_plot

ggsave(
  filename = output_density_model_plot,
  plot = conch_density_plot,
  width = 8,
  height = 5,
  dpi = 300
)

summary(conch_density_model)
conch_density_estimates
```

## Zero-Inflated Density Model

```{r conch-zero-inflated-density-model, fig.width=8, fig.height=5}
conch_zinb_density_model <- pscl::zeroinfl(
  n_animals ~ Type + animal_stage + offset(log(area_m2)) | 1,
  data = conch_density_model_data,
  dist = "negbin",
  EM = TRUE
)

nb_expected_zero_prob <- dnbinom(
  0,
  mu = fitted(conch_density_model),
  size = conch_density_model$theta
)

zinb_zero_prob <- predict(
  conch_zinb_density_model,
  type = "prob"
)[, "0"]

zero_inflation_assessment <- tibble(
  model = c("negative binomial", "zero-inflated negative binomial"),
  AIC = c(AIC(conch_density_model), AIC(conch_zinb_density_model)),
  observed_zero_proportion = mean(conch_density_model_data$n_animals == 0),
  expected_zero_proportion = c(
    mean(nb_expected_zero_prob),
    mean(zinb_zero_prob)
  )
) %>%
  mutate(delta_AIC = AIC - min(AIC))

zero_inflation_interpretation <- if (
  zero_inflation_assessment$delta_AIC[zero_inflation_assessment$model == "zero-inflated negative binomial"] < -2
) {
  "The zero-inflated model is clearly favored by AIC."
} else if (
  abs(diff(zero_inflation_assessment$expected_zero_proportion)) < 0.05 &&
    zero_inflation_assessment$delta_AIC[zero_inflation_assessment$model == "zero-inflated negative binomial"] > 0
) {
  "The negative-binomial model already matches the observed zero proportion well, and the zero-inflated model is not favored by AIC."
} else {
  "The zero-inflated model does not provide a clear improvement over the negative-binomial model."
}

zinb_density_prediction_grid <- density_prediction_grid

zinb_density_prediction <- predict(
  conch_zinb_density_model,
  newdata = zinb_density_prediction_grid,
  type = "response"
)

simulate_zinb_response <- function(model, newdata, n_sims = 5000, seed = 1) {
  set.seed(seed)

  coef_hat <- coef(model)
  coef_vcov <- as.matrix(vcov(model))

  finite_diag <- diag(coef_vcov)[is.finite(diag(coef_vcov)) & diag(coef_vcov) > 0]
  fallback_var <- if (length(finite_diag) > 0) max(finite_diag) else 1
  bad_diag <- !is.finite(diag(coef_vcov)) | diag(coef_vcov) <= 0

  coef_vcov[!is.finite(coef_vcov)] <- 0
  diag(coef_vcov)[bad_diag] <- fallback_var
  coef_vcov <- coef_vcov + diag(1e-8, nrow(coef_vcov))

  eig <- eigen((coef_vcov + t(coef_vcov)) / 2, symmetric = TRUE)
  eig$values <- pmax(eig$values, 0)
  coef_draws <- matrix(rnorm(n_sims * length(coef_hat)), nrow = n_sims) %*%
    (eig$vectors %*% diag(sqrt(eig$values), nrow = length(eig$values))) +
    matrix(coef_hat, nrow = n_sims, ncol = length(coef_hat), byrow = TRUE)
  colnames(coef_draws) <- names(coef_hat)

  count_coef_names <- grep("^count_", names(coef_hat), value = TRUE)
  zero_coef_names <- grep("^zero_", names(coef_hat), value = TRUE)

  x_count <- model.matrix(
    ~ Type + animal_stage,
    data = newdata
  )
  x_count <- x_count[, sub("^count_", "", count_coef_names), drop = FALSE]

  x_zero <- matrix(1, nrow = nrow(newdata), ncol = length(zero_coef_names))
  colnames(x_zero) <- sub("^zero_", "", zero_coef_names)

  count_eta <- x_count %*% t(coef_draws[, count_coef_names, drop = FALSE]) +
    log(newdata$area_m2)
  zero_eta <- x_zero %*% t(coef_draws[, zero_coef_names, drop = FALSE])

  count_mu <- exp(pmin(count_eta, 30))
  zero_prob <- plogis(zero_eta)

  response_draws <- (1 - zero_prob) * count_mu

  tibble(
    density_lwr_per_m2 = apply(response_draws, 1, quantile, probs = 0.025, na.rm = TRUE),
    density_upr_per_m2 = apply(response_draws, 1, quantile, probs = 0.975, na.rm = TRUE)
  )
}

zinb_density_intervals <- simulate_zinb_response(
  conch_zinb_density_model,
  zinb_density_prediction_grid
)

conch_zinb_density_estimates <- zinb_density_prediction_grid %>%
  mutate(
    density_per_m2 = as.numeric(zinb_density_prediction),
    density_lwr_per_m2 = zinb_density_intervals$density_lwr_per_m2,
    density_upr_per_m2 = zinb_density_intervals$density_upr_per_m2,
    density_per_100m2 = density_per_m2 * 100,
    density_lwr_per_100m2 = density_lwr_per_m2 * 100,
    density_upr_per_100m2 = density_upr_per_m2 * 100
  ) %>%
  left_join(observed_density_summary, by = c("Type", "animal_stage")) %>%
  mutate(estimable = observed_count > 0) %>%
  arrange(Type, animal_stage)

conch_zinb_density_plot <- ggplot(
  conch_zinb_density_estimates,
  aes(x = Type, y = density_per_100m2, fill = animal_stage)
) +
  geom_col(position = position_dodge(width = 0.75), color = "grey30", linewidth = 0.2) +
  geom_errorbar(
    aes(ymin = density_lwr_per_100m2, ymax = density_upr_per_100m2),
    position = position_dodge(width = 0.75),
    width = 0.2
  ) +
  coord_flip() +
  labs(
    x = "Habitat type",
    y = expression("ZINB estimated conch density (animals per 100 " * m^2 * ")"),
    fill = "Conch stage"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

conch_zinb_density_plot

ggsave(
  filename = output_zinb_density_model_plot,
  plot = conch_zinb_density_plot,
  width = 8,
  height = 5,
  dpi = 300
)

zero_inflation_assessment
zero_inflation_interpretation
summary(conch_zinb_density_model)
conch_zinb_density_estimates
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
