---
title: "PENIP Conch IDW Abundance Estimates"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

find_project_root <- function(start = getwd()) {
  start <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- c(start, normalizePath(file.path(start, ".."), winslash = "/", mustWork = TRUE))

  while (length(candidates) > 0) {
    candidate <- candidates[[1]]
    if (file.exists(file.path(candidate, "data/processed/polygons_final.rds"))) {
      return(candidate)
    }

    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not find the PENIP project root from ", start, call. = FALSE)
    }

    candidates <- c(candidates[-1], parent)
  }
}

knitr::opts_knit$set(root.dir = find_project_root())
```

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(ggnewscale)
library(purrr)
library(readxl)
library(sf)
library(stringr)
library(tibble)
library(tidyr)
library(xml2)
```

```{r paths-and-constants}
conch_xlsx <- "data/conch/PENIP_QueenConchSurveys.xlsx"
radial_counts_csv <- "data/conch/PENIP_circular_counts.csv"
berry_broad_xlsx <- "data/conch/Berry_Conch_per_site.xlsx"
roving_xlsx <- "data/conch/Roving Diver Survey Data Sept 2025.xlsx"
conch_quest_txt <- "data/conch/ConchQuest2026_CombinedSurveys.txt"
conch_quest_excluded_sites <- c("333", "373")
conch_quest_latitude_clusters <- 3
polygons_rds <- "data/processed/polygons_final.rds"
impact_kml <- "data/spatial/Impact_zones_Scenario 2.kml"
nmfs_kml <- "data/spatial/Scenario2_NMFSConsiderationAreas.kml"

radial_sheet <- "PENIP_Conch_10mRadial"
drift_sheet <- "PENIP_DriftTransects"
drift_position_sheet <- "PENIP_ConchDriftPosition"

utm_crs <- 32617
radial_survey_radius_m <- 10
radial_survey_area_m2 <- pi * radial_survey_radius_m^2
drift_half_width_m <- 2
drift_segment_length_m <- 20
berry_broad_site_area_m2 <- 2 * 30 * 4
idw_cellsize_m <- 25
idw_power <- 2
idw_search_radius_m <- Inf
kriging_nmax <- 20
kriging_maxdist_m <- Inf
kriging_nmin <- 1
kriging_sensitivity_nmax <- c(8, 12, 20)
kriging_sensitivity_maxdist_m <- c(250, 330, 500, 1000)
kriging_sensitivity_nmin <- c(3, 5)
kriging_selected_nmax <- 12
kriging_selected_maxdist_m <- 330
kriging_selected_nmin <- 3
kriging_cv_cores <- 4
breeding_activity_radius_m <- 330
conch_quest_search_radius_m <- 330
penip_cv_search_radii_m <- c(100, 330, 500, 1000)
penip_adult_cv_search_radius_m <- 330
penip_juvenile_cv_search_radius_m <- Inf

output_data_collection_map <- "outputs/conch_data_collection_overview_map.png"
output_idw_map <- "outputs/conch_idw_density_map.png"
output_kriging_map <- "outputs/conch_kriging_density_map.png"
output_kriging_loocv_plot <- "outputs/conch_kriging_loocv.png"
output_kriging_sensitivity_plot <- "outputs/conch_kriging_sensitivity.png"
output_kriging_selected_map <- "outputs/conch_kriging_selected_constrained_map.png"
output_kriging_adult_parameter_maps <- "outputs/conch_kriging_adult_parameter_maps.png"
output_kriging_juvenile_parameter_maps <- "outputs/conch_kriging_juvenile_parameter_maps.png"
output_adult_breeding_radius_idw_map <- "outputs/conch_adult_breeding_radius_idw_map.png"
output_idw_parameter_sensitivity_plot <- "outputs/conch_idw_parameter_sensitivity.png"
output_idw_selected_parameter_maps <- "outputs/conch_idw_selected_parameter_maps.png"
output_conch_quest_coordinate_map <- "outputs/conch_quest_distinct_coordinates_map.png"
output_conch_quest_pairwise_decay_plot <- "outputs/conch_quest_pairwise_decay.png"
output_conch_quest_idw_power_cv_plot <- "outputs/conch_quest_idw_power_cv.png"
output_penip_idw_power_cv_plot <- "outputs/penip_idw_power_cv.png"
output_penip_no_drift_idw_power_cv_plot <- "outputs/penip_no_drift_idw_power_cv.png"
output_penip_age_specific_radius_idw_power_cv_plot <- "outputs/penip_age_specific_radius_idw_power_cv.png"

survey_effort_xlim <- c(-80.1116, -80.078)
survey_effort_ylim <- c(26.058, 26.112)
```

# Input Data

```{r import-data}
polygons_final <- readRDS(polygons_rds) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)

type_cols <- RColorBrewer::brewer.pal(nlevels(factor(polygons_final$Type)) + 1, "Set3")[-1]
names(type_cols) <- levels(factor(polygons_final$Type))

old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

polygons_utm <- polygons_final %>%
  st_transform(utm_crs) %>%
  st_make_valid()

polygons_conch_habitat_utm <- polygons_utm %>%
  mutate(
    conch_habitat = case_when(
      Type %in% c("Artificial", "Nearshore Ridge Complex") ~ "Nearshore Ridge Complex",
      Type == "Inner Reef" ~ "Inner Reef",
      TRUE ~ NA_character_
    )
  )

target_habitats_utm <- polygons_conch_habitat_utm %>%
  filter(!is.na(conch_habitat)) %>%
  group_by(conch_habitat) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

extract_placemark_info <- function(kml_file, source_label) {
  ns <- c(kml = "http://www.opengis.net/kml/2.2")
  kml <- read_xml(kml_file)
  placemarks <- xml_find_all(kml, ".//kml:Placemark", ns)

  map(placemarks, function(pm) {
    name <- xml_text(xml_find_first(pm, ".//kml:name", ns))
    folder_node <- xml_parent(xml_parent(pm))
    folder_name <- xml_text(xml_find_first(folder_node, ".//kml:name", ns))
    if (is.na(folder_name) || folder_name == "") folder_name <- "Top Level"

    coord_nodes <- xml_find_all(pm, ".//kml:coordinates", ns)
    rings <- map(coord_nodes, function(cn) {
      coords <- strsplit(trimws(xml_text(cn)), "\\s+")[[1]]
      coords_mat <- do.call(rbind, lapply(coords, function(x) {
        as.numeric(strsplit(x, ",")[[1]][1:2])
      }))
      if (nrow(coords_mat) >= 3) coords_mat else NULL
    }) %>%
      compact()

    if (length(rings) == 0) return(NULL)

    tibble(
      folder = folder_name,
      name = name,
      geometry = st_sfc(st_multipolygon(lapply(rings, function(r) list(r))), crs = 4326),
      source = source_label
    )
  }) %>%
    compact() %>%
    bind_rows() %>%
    st_as_sf()
}

impact_zones_sf <- extract_placemark_info(impact_kml, "Impact Zone")
nmfs_sf <- extract_placemark_info(nmfs_kml, "NMFS Consideration Area")

impact_zones_utm <- bind_rows(impact_zones_sf, nmfs_sf) %>%
  mutate(
    name = if_else(
      is.na(name) | name == "",
      str_remove(str_remove(folder, "^Scenario_2_"), "\\.shp$"),
      name
    ),
    scenario = if_else(
      name %in% c("North", "South", "East", "Nearshore"),
      "NMFS Consideration",
      "Scenario 2"
    )
  ) %>%
  st_make_valid() %>%
  st_transform(utm_crs) %>%
  group_by(name, scenario, source) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

channel_poly_utm <- impact_zones_utm %>%
  filter(str_detect(tolower(as.character(name)), "channel"))

channel_bbox <- st_bbox(channel_poly_utm)
habitat_bbox <- st_bbox(polygons_utm)

impact_monitoring_box <- st_as_sfc(st_bbox(
  c(
    xmin = habitat_bbox[["xmin"]],
    xmax = habitat_bbox[["xmax"]],
    ymin = channel_bbox[["ymin"]] - 1200,
    ymax = channel_bbox[["ymax"]] + 1200
  ),
  crs = st_crs(polygons_utm)
))

impact_monitoring_habitat_utm <- st_intersection(
  impact_monitoring_box,
  st_union(polygons_utm)
) %>%
  st_union() %>%
  st_make_valid()

existing_impact_zone_hull_utm <- impact_zones_utm %>%
  st_union() %>%
  st_buffer(20) %>%
  st_union() %>%
  st_buffer(-20) %>%
  st_make_valid()

impact_monitoring_area_utm <- st_sf(
  name = "Impact Monitoring Area",
  scenario = "Impact Monitoring Area",
  source = "Generated",
  geometry = st_sfc(
    st_difference(impact_monitoring_habitat_utm, existing_impact_zone_hull_utm),
    crs = st_crs(polygons_utm)
  )
) %>%
  st_make_valid()

impact_zones_utm <- bind_rows(
  impact_zones_utm,
  impact_monitoring_area_utm
)

impact_zone_area_utm <- impact_zones_utm %>%
  st_union() %>%
  st_make_valid()

make_exterior_lines <- function(x) {
  polygon_parts <- x %>%
    st_collection_extract("POLYGON") %>%
    st_cast("POLYGON", warn = FALSE)

  st_sf(
    name = polygon_parts$name,
    scenario = polygon_parts$scenario,
    source = polygon_parts$source,
    geometry = st_sfc(
      map(st_geometry(polygon_parts), ~ st_linestring(.x[[1]])),
      crs = st_crs(polygon_parts)
    )
  )
}

impact_zone_plot_outlines_utm <- bind_rows(
  impact_zones_utm %>%
    filter(name != "Impact Monitoring Area") %>%
    st_boundary() %>%
    st_as_sf(),
  impact_monitoring_area_utm %>%
    make_exterior_lines()
) %>%
  st_make_valid()

radial_sites <- readxl::read_xlsx(conch_xlsx, sheet = radial_sheet)

radial_counts <- read.csv(
  radial_counts_csv,
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
) %>%
  filter(!is.na(`Site Code`), `Site Code` != "") %>%
  mutate(
    juvenile_count = as.numeric(`# Juv`),
    adult_count = as.numeric(`# Adult`),
    breeding_activity_count = as.numeric(ReproductiveActivity)
  )

radial_abundance <- radial_sites %>%
  select(`Site Code`, Lat, Lon, `Depth (ft)`) %>%
  left_join(
    radial_counts %>%
      select(`Site Code`, juvenile_count, adult_count, breeding_activity_count),
    by = "Site Code"
  )

stopifnot(
  nrow(radial_abundance) == nrow(radial_sites),
  sum(is.na(radial_abundance$adult_count)) == 0,
  sum(is.na(radial_abundance$juvenile_count)) == 0
)

radial_observations_sf_utm <- radial_abundance %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  mutate(
    observation_source = "PENIP radial survey",
    survey_area_m2 = radial_survey_area_m2
  ) %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

radial_breeding_activity_sites_sf_utm <- radial_abundance %>%
  filter(breeding_activity_count > 0) %>%
  mutate(
    breeding_activity_source = "PENIP radial survey",
    breeding_activity_type = "Breeding activity"
  ) %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  select(breeding_activity_source, breeding_activity_type, geometry)

roving_breeding_activity_sites_sf_utm <- readxl::read_xlsx(roving_xlsx, sheet = 1) %>%
  mutate(
    latitude = as.numeric(`Lat (dec deg)`),
    longitude = as.numeric(`Long (dec deg)`),
    comments_clean = str_squish(str_to_lower(replace_na(Comments, ""))),
    eggs_observed = str_detect(comments_clean, "egg"),
    mating_observed = str_detect(comments_clean, "mating|mate"),
    breeding_activity_type = case_when(
      eggs_observed & mating_observed ~ "Eggs and mating",
      eggs_observed ~ "Eggs",
      mating_observed ~ "Mating",
      TRUE ~ NA_character_
    ),
    breeding_activity_source = "Roving diver survey"
  ) %>%
  filter(!is.na(latitude), !is.na(longitude), eggs_observed | mating_observed) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  select(breeding_activity_source, breeding_activity_type, geometry)

breeding_activity_sites_sf_utm <- bind_rows(
  radial_breeding_activity_sites_sf_utm,
  roving_breeding_activity_sites_sf_utm
)

breeding_activity_summary <- breeding_activity_sites_sf_utm %>%
  st_drop_geometry() %>%
  count(breeding_activity_source, breeding_activity_type, name = "observations")

radial_summary <- radial_observations_sf_utm %>%
  st_drop_geometry() %>%
  summarize(
    radial_surveys = n(),
    radial_area_m2 = sum(survey_area_m2),
    adult_count = sum(adult_count),
    juvenile_count = sum(juvenile_count)
  )

knitr::kable(radial_summary, digits = 2)
knitr::kable(breeding_activity_summary)
```

# Drift Transects

Drift transects are converted to 4 m-wide rectangles and then divided into `r drift_segment_length_m` m segments. Observed conch from the drift survey point sheet are assigned to the nearest segment within the same transect.

```{r drift-segments}
make_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(st_polygon())
  }

  unit_direction <- direction / transect_length
  perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

  corners <- rbind(
    start_point + perpendicular,
    end_point + perpendicular,
    end_point - perpendicular,
    start_point - perpendicular,
    start_point + perpendicular
  )

  st_polygon(list(corners))
}

make_segment_rectangles <- function(start_point, end_point, segment_length_m, half_width_m) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(tibble())
  }

  unit_direction <- direction / transect_length
  segment_starts <- seq(0, transect_length, by = segment_length_m)
  segment_ends <- pmin(segment_starts + segment_length_m, transect_length)
  keep_segments <- segment_ends > segment_starts

  tibble(
    segment_index = seq_along(segment_starts)[keep_segments],
    segment_start_m = segment_starts[keep_segments],
    segment_end_m = segment_ends[keep_segments],
    geometry = map2(
      segment_start_m,
      segment_end_m,
      ~ make_transect_rectangle(
        start_point + unit_direction * .x,
        start_point + unit_direction * .y,
        half_width_m = half_width_m
      )
    )
  )
}

drift_surveys <- readxl::read_xlsx(conch_xlsx, sheet = drift_sheet) %>%
  mutate(
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon), !is.na(end_lat), !is.na(end_lon)) %>%
  mutate(
    .record = row_number(),
    transect_group = sub(" cont\\.$", "", Transect)
  )

drift_start_sf <- drift_surveys %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

drift_end_sf <- drift_surveys %>%
  st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

start_xy <- st_coordinates(drift_start_sf)
end_xy <- st_coordinates(drift_end_sf)

drift_rectangles <- st_sf(
  drift_surveys %>% select(-start_lat, -start_lon, -end_lat, -end_lon),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ make_transect_rectangle(start_xy[.x, ], end_xy[.y, ], half_width_m = drift_half_width_m)
    ),
    crs = utm_crs
  )
) %>%
  filter(!st_is_empty(.)) %>%
  mutate(transect_area_m2 = as.numeric(st_area(geometry)))

drift_transect_lines <- st_sf(
  drift_surveys %>% select(.record, Transect, transect_group, Line),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ st_linestring(rbind(start_xy[.x, ], end_xy[.y, ]))
    ),
    crs = utm_crs
  )
)

drift_segment_table <- map_dfr(seq_len(nrow(drift_surveys)), function(i) {
  make_segment_rectangles(
    start_xy[i, ],
    end_xy[i, ],
    segment_length_m = drift_segment_length_m,
    half_width_m = drift_half_width_m
  ) %>%
    mutate(
      .record = drift_surveys$.record[[i]],
      Transect = drift_surveys$Transect[[i]],
      transect_group = drift_surveys$transect_group[[i]],
      Line = drift_surveys$Line[[i]],
      segment_id = paste(.record, segment_index, sep = "_")
    )
})

drift_segments <- st_as_sf(
  drift_segment_table,
  sf_column_name = "geometry",
  crs = utm_crs
) %>%
  select(segment_id, .record, Transect, transect_group, Line, segment_index, segment_start_m, segment_end_m) %>%
  mutate(segment_area_m2 = as.numeric(st_area(geometry)))

drift_segment_summary <- tibble(
  drift_transects = nrow(drift_rectangles),
  drift_segment_count = nrow(drift_segments),
  drift_area_m2 = sum(drift_rectangles$transect_area_m2),
  drift_segment_area_m2 = sum(drift_segments$segment_area_m2)
)

knitr::kable(drift_segment_summary, digits = 2)
```

```{r drift-position-counts}
conch_drift_positions <- readxl::read_xlsx(conch_xlsx, sheet = drift_position_sheet) %>%
  mutate(
    .record = row_number(),
    lat = as.numeric(`Lat (dec deg)`),
    lon = as.numeric(`Long (dec deg)`),
    animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`),
    animal_stage = case_when(
      grepl("^A", animal_id) ~ "Adult",
      grepl("^J", animal_id) ~ "Juvenile",
      TRUE ~ NA_character_
    ),
    transect_group = sub("([AJ][0-9]+)$", "", `WPT (B/E)`)
  ) %>%
  filter(!is.na(lat), !is.na(lon), !is.na(animal_stage))

conch_drift_positions_sf_utm <- conch_drift_positions %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

assign_to_nearest_segment <- function(points_sf, segments_sf) {
  map_dfr(seq_len(nrow(points_sf)), function(i) {
    point_i <- points_sf[i, ]
    point_dat <- point_i %>% st_drop_geometry()
    candidate_segments <- segments_sf %>%
      filter(transect_group == point_dat$transect_group[[1]])

    if (nrow(candidate_segments) == 0) {
      return(tibble(
        .record = point_dat$.record[[1]],
        animal_stage = point_dat$animal_stage[[1]],
        transect_group = point_dat$transect_group[[1]],
        segment_id = NA_character_,
        segment_index = NA_integer_,
        segment_area_m2 = NA_real_,
        distance_to_segment_m = NA_real_
      ))
    }

    distances_m <- as.numeric(st_distance(point_i, candidate_segments))
    nearest_i <- which.min(distances_m)
    nearest_segment <- candidate_segments[nearest_i, ] %>% st_drop_geometry()

    tibble(
      .record = point_dat$.record[[1]],
      animal_stage = point_dat$animal_stage[[1]],
      transect_group = point_dat$transect_group[[1]],
      segment_id = nearest_segment$segment_id[[1]],
      segment_index = nearest_segment$segment_index[[1]],
      segment_area_m2 = nearest_segment$segment_area_m2[[1]],
      distance_to_segment_m = distances_m[[nearest_i]]
    )
  })
}

drift_observation_segment_assignments <- assign_to_nearest_segment(
  conch_drift_positions_sf_utm,
  drift_segments
)

drift_segment_counts <- drift_observation_segment_assignments %>%
  filter(!is.na(segment_id)) %>%
  count(segment_id, animal_stage, name = "n_animals") %>%
  pivot_wider(
    names_from = animal_stage,
    values_from = n_animals,
    values_fill = 0
  )

drift_segment_observations_sf_utm <- drift_segments %>%
  st_point_on_surface() %>%
  left_join(drift_segment_counts, by = "segment_id") %>%
  mutate(
    Adult = replace_na(Adult, 0),
    Juvenile = replace_na(Juvenile, 0),
    observation_source = "PENIP drift segment",
    survey_area_m2 = segment_area_m2,
    adult_count = Adult,
    juvenile_count = Juvenile
  ) %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

drift_observation_assignment_summary <- tibble(
  observed_conch_total = nrow(conch_drift_positions),
  assigned_to_segment = sum(!is.na(drift_observation_segment_assignments$segment_id)),
  unassigned_to_segment = observed_conch_total - assigned_to_segment,
  median_distance_to_assigned_segment_m = median(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  ),
  max_distance_to_assigned_segment_m = max(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  )
)

knitr::kable(drift_observation_assignment_summary, digits = 2)
```

# Broad-Scale Survey Counts

The Berry broad-scale survey counts are classified as adult = `Adult + Old` and juvenile/non-adult = `Juv + Sub + Young`. Each broad-scale site is treated as 240 m2 of surveyed area, from two 30 m transects surveyed 2 m on each side of the tape.

```{r berry-broad-data}
berry_broad_raw <- readxl::read_xlsx(berry_broad_xlsx, sheet = 1)

berry_broad_sites <- berry_broad_raw %>%
  mutate(
    .record = row_number(),
    adult_count = as.numeric(Adult) + as.numeric(Old),
    juvenile_count = as.numeric(Juv) + as.numeric(Sub) + as.numeric(Young),
    total_count = adult_count + juvenile_count,
    survey_area_m2 = berry_broad_site_area_m2
  ) %>%
  select(.record, Site, LatDD, LonDD, Adult, Juv, Old, Sub, Young, adult_count, juvenile_count, total_count, survey_area_m2)

berry_observations_sf_utm <- berry_broad_sites %>%
  st_as_sf(coords = c("LonDD", "LatDD"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  mutate(observation_source = "Berry broad-scale site") %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

berry_broad_summary <- berry_broad_sites %>%
  summarize(
    berry_sites = n(),
    berry_area_m2 = sum(survey_area_m2, na.rm = TRUE),
    berry_adults = sum(adult_count, na.rm = TRUE),
    berry_juveniles = sum(juvenile_count, na.rm = TRUE),
    berry_total = sum(total_count, na.rm = TRUE)
  )

knitr::kable(berry_broad_summary, digits = 2)
```

# Data Collection Overview

This figure shows the spatial distribution of the three conch survey inputs using the same extent, habitat colors, and survey-point style as Figure 1B in the coral report.

```{r data-collection-overview-map, fig.width=7, fig.height=3.5}
survey_map_theme <- theme_minimal(base_size = 8) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.size = unit(0.3, "cm"),
    legend.text = element_text(size = 6),
    legend.title = element_blank(),
    strip.text = element_text(size = 6),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(2, 0, 0, 0),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

survey_effort_levels <- c(
  "Berry broad-scale survey",
  "PENIP radial survey",
  "PENIP drift survey"
)

survey_effort_polygons <- crossing(survey_effort = survey_effort_levels) %>%
  pull(survey_effort) %>%
  map_dfr(function(panel_name) {
    polygons_final %>%
      mutate(survey_effort = panel_name)
  }) %>%
  st_as_sf() %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels))

berry_effort_points <- berry_observations_sf_utm %>%
  st_transform(4326) %>%
  mutate(survey_effort = "Berry broad-scale survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

radial_effort_points <- radial_observations_sf_utm %>%
  st_transform(4326) %>%
  mutate(survey_effort = "PENIP radial survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

drift_effort_lines <- drift_transect_lines %>%
  st_transform(4326) %>%
  mutate(survey_effort = "PENIP drift survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

data_collection_map <- ggplot() +
  geom_sf(
    data = survey_effort_polygons,
    aes(fill = Type),
    color = NA,
    linewidth = 0.1,
    alpha = 0.5
  ) +
  geom_sf(
    data = drift_effort_lines,
    color = "black",
    linewidth = 0.25,
    alpha = 0.7
  ) +
  geom_sf(
    data = berry_effort_points,
    pch = 4,
    alpha = 0.6,
    size = 0.65,
    stroke = 0.3
  ) +
  geom_sf(
    data = radial_effort_points,
    pch = 4,
    alpha = 0.6,
    size = 0.65,
    stroke = 0.3
  ) +
  scale_fill_manual(values = type_cols, na.value = "gray80") +
  facet_wrap(
    ~ survey_effort,
    ncol = 3,
    labeller = as_labeller(c(
      "Berry broad-scale survey" = "Berry Broad-scale Survey",
      "PENIP radial survey" = "PENIP Radial Surveys",
      "PENIP drift survey" = "PENIP Drift Surveys"
    ))
  ) +
  coord_sf(
    xlim = survey_effort_xlim,
    ylim = survey_effort_ylim,
    expand = FALSE
  ) +
  guides(fill = guide_legend(nrow = 2, override.aes = list(alpha = 0.5))) +
  survey_map_theme

data_collection_map

ggsave(output_data_collection_map, data_collection_map, width = 180, height = 90, units = "mm", dpi = 300, bg = "white")
```

# IDW Estimates

The final analysis combines radial surveys, segmented drift transects, and broad-scale survey sites as observation units. Adult and juvenile densities are interpolated separately using inverse distance weighting. Prediction is limited to Inner Reef and Nearshore Ridge Complex habitat within the Figure 1B map extent; Artificial habitat is included with Nearshore Ridge Complex.

```{r idw-functions}
calculate_idw_prediction <- function(
  distance_matrix_m,
  observed_values,
  power = idw_power,
  search_radius_m = idw_search_radius_m
) {
  apply(distance_matrix_m, 1, function(distances_m) {
    valid_observations <- !is.na(observed_values) & !is.na(distances_m)

    if (is.finite(search_radius_m)) {
      valid_observations <- valid_observations & distances_m <= search_radius_m
    }

    if (!any(valid_observations)) {
      return(NA_real_)
    }

    zero_distance <- which(distances_m == 0 & valid_observations)

    if (length(zero_distance) > 0) {
      return(mean(observed_values[zero_distance], na.rm = TRUE))
    }

    weights <- 1 / distances_m[valid_observations]^power
    sum(weights * observed_values[valid_observations], na.rm = TRUE) / sum(weights, na.rm = TRUE)
  })
}
```

# Conch Quest Spatial Decay Check

The Conch Quest 2026 dataset is used here as an independent check on spatial decay. Positive longitude values are assumed to be missing a negative sign and are corrected before distances are calculated. Waypoints 333 and 373 are excluded pending coordinate confirmation because they form an isolated northern pair far from the main aggregation. The remaining sites are split into three latitude clusters, and only the middle-latitude cluster is used for the decay analysis below. The pairwise analysis estimates how adult and juvenile abundance differences increase with distance within 330 m, while the leave-one-out IDW analysis estimates which IDW power best predicts held-out Conch Quest survey sites using the same 330 m search radius.

```{r conch-quest-pairwise-decay}
conch_quest_raw <- read.delim(conch_quest_txt, check.names = FALSE)

conch_quest_sites_all <- conch_quest_raw %>%
  mutate(
    site_id = as.character(Waypoint),
    latitude = as.numeric(Lat),
    longitude_raw = as.numeric(Lon),
    longitude = if_else(longitude_raw > 0, -longitude_raw, longitude_raw),
    longitude_corrected = longitude_raw != longitude,
    adult_abundance_ha = as.numeric(AdultAbundance),
    juvenile_abundance_ha = as.numeric(JuvenileAbundance)
  ) %>%
  filter(
    !is.na(site_id),
    !site_id %in% conch_quest_excluded_sites,
    !is.na(latitude),
    !is.na(longitude),
    !is.na(adult_abundance_ha),
    !is.na(juvenile_abundance_ha)
  )

set.seed(1)
conch_quest_latitude_kmeans <- kmeans(
  conch_quest_sites_all$latitude,
  centers = conch_quest_latitude_clusters,
  nstart = 50
)

conch_quest_sites_clustered <- conch_quest_sites_all %>%
  mutate(latitude_cluster = conch_quest_latitude_kmeans$cluster)

conch_quest_latitude_cluster_summary <- conch_quest_sites_clustered %>%
  group_by(latitude_cluster) %>%
  summarize(
    sites = n(),
    minimum_latitude = min(latitude, na.rm = TRUE),
    median_latitude = median(latitude, na.rm = TRUE),
    maximum_latitude = max(latitude, na.rm = TRUE),
    minimum_longitude = min(longitude, na.rm = TRUE),
    maximum_longitude = max(longitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(median_latitude) %>%
  mutate(latitude_cluster_position = row_number())

conch_quest_middle_latitude_cluster <- conch_quest_latitude_cluster_summary %>%
  filter(latitude_cluster_position == ceiling(conch_quest_latitude_clusters / 2)) %>%
  pull(latitude_cluster)

conch_quest_sites <- conch_quest_sites_clustered %>%
  filter(latitude_cluster == conch_quest_middle_latitude_cluster)

knitr::kable(
  conch_quest_latitude_cluster_summary,
  digits = 5,
  caption = "Latitude clusters for Conch Quest 2026 sites after longitude correction and site exclusions. The middle-latitude cluster is used in the decay analysis."
)

conch_quest_sites_utm <- conch_quest_sites %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

conch_quest_dist_m <- st_distance(conch_quest_sites_utm) %>%
  as.matrix()

conch_quest_pairwise <- combn(seq_len(nrow(conch_quest_sites)), 2, simplify = FALSE) %>%
  map_dfr(function(idx) {
    i <- idx[[1]]
    j <- idx[[2]]

    tibble(
      site_1 = conch_quest_sites$site_id[[i]],
      site_2 = conch_quest_sites$site_id[[j]],
      distance_m = as.numeric(conch_quest_dist_m[i, j]),
      adult_abundance_difference_ha = abs(conch_quest_sites$adult_abundance_ha[[i]] - conch_quest_sites$adult_abundance_ha[[j]]),
      juvenile_abundance_difference_ha = abs(conch_quest_sites$juvenile_abundance_ha[[i]] - conch_quest_sites$juvenile_abundance_ha[[j]]),
      adult_semivariance = 0.5 * adult_abundance_difference_ha^2,
      juvenile_semivariance = 0.5 * juvenile_abundance_difference_ha^2
    )
  }) %>%
  filter(distance_m <= conch_quest_search_radius_m)

conch_quest_pairwise_binned <- conch_quest_pairwise %>%
  mutate(distance_bin_m = cut(
    distance_m,
    breaks = pretty(distance_m, n = 12),
    include.lowest = TRUE
  )) %>%
  group_by(distance_bin_m) %>%
  summarize(
    distance_m = median(distance_m, na.rm = TRUE),
    pairwise_comparisons = n(),
    adult_mean_abs_difference_ha = mean(adult_abundance_difference_ha, na.rm = TRUE),
    juvenile_mean_abs_difference_ha = mean(juvenile_abundance_difference_ha, na.rm = TRUE),
    adult_semivariance = mean(adult_semivariance, na.rm = TRUE),
    juvenile_semivariance = mean(juvenile_semivariance, na.rm = TRUE),
    .groups = "drop"
  )

fit_exponential_variogram <- function(distance_m, semivariance) {
  keep <- is.finite(distance_m) & is.finite(semivariance)
  distance_m <- distance_m[keep]
  semivariance <- semivariance[keep]

  start <- c(
    nugget = max(min(semivariance, na.rm = TRUE), 1e-6),
    partial_sill = max(max(semivariance, na.rm = TRUE) - min(semivariance, na.rm = TRUE), 1e-6),
    range_m = max(median(distance_m, na.rm = TRUE), 1)
  )
  upper <- c(
    nugget = max(max(semivariance, na.rm = TRUE) * 10, 1),
    partial_sill = max(max(semivariance, na.rm = TRUE) * 10, 1),
    range_m = max(max(distance_m, na.rm = TRUE) * 100, 10)
  )

  tryCatch(
    optim(
      par = log(start),
      fn = function(log_par) {
        pars <- exp(log_par)
        fitted_semivariance <- pars[["nugget"]] +
          pars[["partial_sill"]] * (1 - exp(-distance_m / pars[["range_m"]]))
        sum((semivariance - fitted_semivariance)^2, na.rm = TRUE)
      },
      method = "L-BFGS-B",
      lower = log(c(nugget = 1e-6, partial_sill = 1e-6, range_m = 1)),
      upper = log(upper)
    ),
    error = function(e) NULL
  )
}

adult_variogram_fit <- fit_exponential_variogram(
  conch_quest_pairwise_binned$distance_m,
  conch_quest_pairwise_binned$adult_semivariance
)

juvenile_variogram_fit <- fit_exponential_variogram(
  conch_quest_pairwise_binned$distance_m,
  conch_quest_pairwise_binned$juvenile_semivariance
)

extract_variogram_summary <- function(fit, age_class) {
  if (is.null(fit)) {
    return(tibble(
      age_class = age_class,
      nugget = NA_real_,
      partial_sill = NA_real_,
      range_m = NA_real_,
      practical_range_m = NA_real_
    ))
  }

  pars <- exp(fit$par)
  tibble(
    age_class = age_class,
    nugget = unname(pars[["nugget"]]),
    partial_sill = unname(pars[["partial_sill"]]),
    range_m = unname(pars[["range_m"]]),
    practical_range_m = 3 * range_m
  )
}

conch_quest_variogram_summary <- bind_rows(
  extract_variogram_summary(adult_variogram_fit, "Adult"),
  extract_variogram_summary(juvenile_variogram_fit, "Juvenile")
)

knitr::kable(
  conch_quest_variogram_summary,
  digits = 2,
  caption = "Exponential variogram-style fit to binned pairwise abundance differences from the Conch Quest 2026 surveys. Practical range is approximately three times the fitted exponential range."
)
```

```{r conch-quest-coordinate-map, fig.width=7.5, fig.height=6}
conch_quest_distinct_coordinates <- conch_quest_sites %>%
  distinct(latitude, longitude, .keep_all = TRUE)

conch_quest_coordinate_summary <- tibble(
  rows = nrow(conch_quest_sites),
  distinct_lat_lon = nrow(conch_quest_distinct_coordinates),
  excluded_sites = paste(conch_quest_excluded_sites, collapse = ", "),
  minimum_latitude = min(conch_quest_distinct_coordinates$latitude, na.rm = TRUE),
  maximum_latitude = max(conch_quest_distinct_coordinates$latitude, na.rm = TRUE),
  minimum_longitude = min(conch_quest_distinct_coordinates$longitude, na.rm = TRUE),
  maximum_longitude = max(conch_quest_distinct_coordinates$longitude, na.rm = TRUE),
  corrected_positive_longitudes = sum(conch_quest_distinct_coordinates$longitude_corrected, na.rm = TRUE),
  search_radius_m = conch_quest_search_radius_m
)

knitr::kable(
  conch_quest_coordinate_summary,
  digits = 5,
  caption = "Coordinate summary for distinct Conch Quest 2026 latitude/longitude combinations."
)

conch_quest_coordinate_map <- ggplot(conch_quest_distinct_coordinates) +
  geom_point(
    aes(x = longitude, y = latitude),
    size = 2,
    alpha = 0.75,
    color = "#2B6CB0"
  ) +
  geom_text(
    aes(x = longitude, y = latitude, label = site_id),
    size = 2.2,
    check_overlap = TRUE,
    nudge_y = 0.00008
  ) +
  coord_equal() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Conch Quest 2026 distinct survey coordinates",
    subtitle = paste0(nrow(conch_quest_distinct_coordinates), " distinct lat/lon combinations; middle latitude cluster only")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

conch_quest_coordinate_map

ggsave(
  output_conch_quest_coordinate_map,
  conch_quest_coordinate_map,
  width = 7.5,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

```{r conch-quest-pairwise-decay-plot, fig.width=8, fig.height=4.5}
conch_quest_pairwise_plot_data <- conch_quest_pairwise %>%
  select(distance_m, adult_abundance_difference_ha, juvenile_abundance_difference_ha) %>%
  pivot_longer(
    cols = c(adult_abundance_difference_ha, juvenile_abundance_difference_ha),
    names_to = "age_class",
    values_to = "abundance_difference_ha"
  ) %>%
  mutate(
    age_class = recode(
      age_class,
      adult_abundance_difference_ha = "Adult",
      juvenile_abundance_difference_ha = "Juvenile"
    )
  )

conch_quest_pairwise_decay_plot <- ggplot(
  conch_quest_pairwise_plot_data,
  aes(x = distance_m, y = abundance_difference_ha, color = age_class)
) +
  geom_point(size = 1.1, alpha = 0.25) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "Age class") +
  labs(
    x = "Distance between survey sites (m)",
    y = "Absolute abundance difference (conch per ha)",
    title = "Conch Quest pairwise abundance differences",
    subtitle = "Middle latitude cluster only; all pairwise site comparisons within 330 m"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

conch_quest_pairwise_decay_plot

ggsave(
  output_conch_quest_pairwise_decay_plot,
  conch_quest_pairwise_decay_plot,
  width = 8,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

```{r conch-quest-idw-power-cross-validation}
calculate_loocv_idw <- function(distance_matrix_m, observed_values, power, search_radius_m = Inf) {
  loocv_distance_matrix_m <- distance_matrix_m
  diag(loocv_distance_matrix_m) <- Inf

  apply(loocv_distance_matrix_m, 1, function(distances_m) {
    valid_observations <- !is.na(observed_values) & is.finite(distances_m)

    if (is.finite(search_radius_m)) {
      valid_observations <- valid_observations & distances_m <= search_radius_m
    }

    if (!any(valid_observations)) {
      return(NA_real_)
    }

    zero_distance <- which(distances_m == 0 & valid_observations)

    if (length(zero_distance) > 0) {
      return(mean(observed_values[zero_distance], na.rm = TRUE))
    }

    weights <- 1 / distances_m[valid_observations]^power
    sum(weights * observed_values[valid_observations], na.rm = TRUE) / sum(weights, na.rm = TRUE)
  })
}

conch_quest_idw_power_grid <- tidyr::expand_grid(
  power = seq(1, 6, by = 0.25),
  search_radius_m = conch_quest_search_radius_m
) %>%
  mutate(
    search_radius_label = paste0(search_radius_m, " m"),
    search_radius_label = factor(search_radius_label, levels = paste0(conch_quest_search_radius_m, " m"))
  )

summarize_conch_quest_loocv <- function(power, search_radius_m, search_radius_label) {
  adult_predicted <- calculate_loocv_idw(
    conch_quest_dist_m,
    conch_quest_sites$adult_abundance_ha,
    power = power,
    search_radius_m = search_radius_m
  )

  juvenile_predicted <- calculate_loocv_idw(
    conch_quest_dist_m,
    conch_quest_sites$juvenile_abundance_ha,
    power = power,
    search_radius_m = search_radius_m
  )

  tibble(
    power = power,
    search_radius_m = search_radius_m,
    search_radius_label = search_radius_label,
    adult_rmse = sqrt(mean((adult_predicted - conch_quest_sites$adult_abundance_ha)^2, na.rm = TRUE)),
    adult_mae = mean(abs(adult_predicted - conch_quest_sites$adult_abundance_ha), na.rm = TRUE),
    adult_predicted_sites = sum(!is.na(adult_predicted)),
    juvenile_rmse = sqrt(mean((juvenile_predicted - conch_quest_sites$juvenile_abundance_ha)^2, na.rm = TRUE)),
    juvenile_mae = mean(abs(juvenile_predicted - conch_quest_sites$juvenile_abundance_ha), na.rm = TRUE),
    juvenile_predicted_sites = sum(!is.na(juvenile_predicted))
  )
}

conch_quest_idw_power_cv <- pmap_dfr(
  conch_quest_idw_power_grid,
  summarize_conch_quest_loocv
)

conch_quest_idw_best_power <- bind_rows(
  conch_quest_idw_power_cv %>%
    slice_min(adult_rmse, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Adult",
      best_power_by_rmse = power,
      search_radius = search_radius_label,
      rmse = adult_rmse,
      mae = adult_mae,
      predicted_sites = adult_predicted_sites
    ),
  conch_quest_idw_power_cv %>%
    slice_min(juvenile_rmse, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Juvenile",
      best_power_by_rmse = power,
      search_radius = search_radius_label,
      rmse = juvenile_rmse,
      mae = juvenile_mae,
      predicted_sites = juvenile_predicted_sites
    )
)

knitr::kable(
  conch_quest_idw_best_power,
  digits = 2,
  caption = "Leave-one-out IDW cross-validation on Conch Quest 2026 sites. Lower RMSE indicates better predictive performance for held-out survey sites."
)
```

```{r conch-quest-idw-power-cross-validation-plot, fig.width=8, fig.height=4.5}
conch_quest_idw_power_cv_long <- conch_quest_idw_power_cv %>%
  select(power, search_radius_label, adult_rmse, juvenile_rmse) %>%
  pivot_longer(
    cols = c(adult_rmse, juvenile_rmse),
    names_to = "age_class",
    values_to = "rmse"
  ) %>%
  mutate(
    age_class = recode(
      age_class,
      adult_rmse = "Adult",
      juvenile_rmse = "Juvenile"
    )
  )

conch_quest_idw_power_cv_plot <- ggplot(
  conch_quest_idw_power_cv_long,
  aes(x = power, y = rmse, color = search_radius_label, group = search_radius_label)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~ age_class, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "Search radius") +
  labs(
    x = "IDW power",
    y = "Leave-one-out RMSE (conch per ha)",
    title = "Conch Quest IDW power cross-validation",
    subtitle = "Middle latitude cluster only; search radius fixed at 330 m"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

conch_quest_idw_power_cv_plot

ggsave(
  output_conch_quest_idw_power_cv_plot,
  conch_quest_idw_power_cv_plot,
  width = 8,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

```{r idw-observations}
idw_observations_sf_utm <- bind_rows(
  radial_observations_sf_utm,
  drift_segment_observations_sf_utm,
  berry_observations_sf_utm
) %>%
  mutate(
    adult_density_m2 = adult_count / survey_area_m2,
    juvenile_density_m2 = juvenile_count / survey_area_m2
  )

observation_summary <- idw_observations_sf_utm %>%
  st_drop_geometry() %>%
  group_by(observation_source) %>%
  summarize(
    observations = n(),
    area_m2 = sum(survey_area_m2, na.rm = TRUE),
    adult_count = sum(adult_count, na.rm = TRUE),
    juvenile_count = sum(juvenile_count, na.rm = TRUE),
    adult_density_per_ha = adult_count / area_m2 * 10000,
    juvenile_density_per_ha = juvenile_count / area_m2 * 10000,
    .groups = "drop"
  )

knitr::kable(observation_summary, digits = 2)
```

```{r penip-idw-power-cross-validation}
penip_idw_cv_points_utm <- idw_observations_sf_utm %>%
  st_point_on_surface()

penip_idw_cv_distance_m <- st_distance(penip_idw_cv_points_utm) %>%
  as.matrix()

penip_idw_power_grid <- tidyr::expand_grid(
  power = seq(1, 6, by = 0.25),
  search_radius_m = penip_cv_search_radii_m
) %>%
  mutate(
    search_radius_label = paste0(search_radius_m, " m"),
    search_radius_label = factor(search_radius_label, levels = paste0(penip_cv_search_radii_m, " m"))
  )

summarize_penip_loocv <- function(power, search_radius_m, search_radius_label) {
  adult_predicted <- calculate_loocv_idw(
    penip_idw_cv_distance_m,
    idw_observations_sf_utm$adult_density_m2,
    power = power,
    search_radius_m = search_radius_m
  )

  juvenile_predicted <- calculate_loocv_idw(
    penip_idw_cv_distance_m,
    idw_observations_sf_utm$juvenile_density_m2,
    power = power,
    search_radius_m = search_radius_m
  )

  tibble(
    power = power,
    search_radius_m = search_radius_m,
    search_radius_label = search_radius_label,
    adult_rmse_per_ha = sqrt(mean((adult_predicted - idw_observations_sf_utm$adult_density_m2)^2, na.rm = TRUE)) * 10000,
    adult_mae_per_ha = mean(abs(adult_predicted - idw_observations_sf_utm$adult_density_m2), na.rm = TRUE) * 10000,
    adult_predicted_sites = sum(!is.na(adult_predicted)),
    juvenile_rmse_per_ha = sqrt(mean((juvenile_predicted - idw_observations_sf_utm$juvenile_density_m2)^2, na.rm = TRUE)) * 10000,
    juvenile_mae_per_ha = mean(abs(juvenile_predicted - idw_observations_sf_utm$juvenile_density_m2), na.rm = TRUE) * 10000,
    juvenile_predicted_sites = sum(!is.na(juvenile_predicted))
  )
}

penip_idw_power_cv <- pmap_dfr(
  penip_idw_power_grid,
  summarize_penip_loocv
)

penip_idw_best_power <- bind_rows(
  penip_idw_power_cv %>%
    slice_min(adult_rmse_per_ha, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Adult",
      best_power_by_rmse = power,
      search_radius = search_radius_label,
      rmse_per_ha = adult_rmse_per_ha,
      mae_per_ha = adult_mae_per_ha,
      predicted_sites = adult_predicted_sites
    ),
  penip_idw_power_cv %>%
    slice_min(juvenile_rmse_per_ha, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Juvenile",
      best_power_by_rmse = power,
      search_radius = search_radius_label,
      rmse_per_ha = juvenile_rmse_per_ha,
      mae_per_ha = juvenile_mae_per_ha,
      predicted_sites = juvenile_predicted_sites
    )
)

knitr::kable(
  penip_idw_best_power,
  digits = 2,
  caption = "Leave-one-out IDW cross-validation using the Port Everglades observation units in the main abundance analysis. Error is calculated against observed density and reported as conch per hectare."
)
```

```{r penip-idw-power-cross-validation-plot, fig.width=8, fig.height=4.5}
penip_idw_power_cv_long <- penip_idw_power_cv %>%
  select(power, search_radius_label, adult_rmse_per_ha, juvenile_rmse_per_ha) %>%
  pivot_longer(
    cols = c(adult_rmse_per_ha, juvenile_rmse_per_ha),
    names_to = "age_class",
    values_to = "rmse_per_ha"
  ) %>%
  mutate(
    age_class = recode(
      age_class,
      adult_rmse_per_ha = "Adult",
      juvenile_rmse_per_ha = "Juvenile"
    )
  )

penip_idw_power_cv_plot <- ggplot(
  penip_idw_power_cv_long,
  aes(x = power, y = rmse_per_ha, color = search_radius_label, group = search_radius_label)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~ age_class, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "Search radius") +
  labs(
    x = "IDW power",
    y = "Leave-one-out RMSE (conch per ha)",
    title = "Port Everglades IDW power cross-validation",
    subtitle = "Radial surveys, 20 m drift segments, and broad-scale sites"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

penip_idw_power_cv_plot

ggsave(
  output_penip_idw_power_cv_plot,
  penip_idw_power_cv_plot,
  width = 8,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

```{r penip-no-drift-idw-power-cross-validation}
penip_no_drift_idw_observations_sf_utm <- idw_observations_sf_utm %>%
  filter(observation_source != "PENIP drift segment")

penip_no_drift_idw_cv_points_utm <- penip_no_drift_idw_observations_sf_utm %>%
  st_point_on_surface()

penip_no_drift_idw_cv_distance_m <- st_distance(penip_no_drift_idw_cv_points_utm) %>%
  as.matrix()

summarize_penip_no_drift_loocv <- function(power, search_radius_m, search_radius_label) {
  adult_predicted <- calculate_loocv_idw(
    penip_no_drift_idw_cv_distance_m,
    penip_no_drift_idw_observations_sf_utm$adult_density_m2,
    power = power,
    search_radius_m = search_radius_m
  )

  juvenile_predicted <- calculate_loocv_idw(
    penip_no_drift_idw_cv_distance_m,
    penip_no_drift_idw_observations_sf_utm$juvenile_density_m2,
    power = power,
    search_radius_m = search_radius_m
  )

  tibble(
    power = power,
    search_radius_m = search_radius_m,
    search_radius_label = search_radius_label,
    adult_rmse_per_ha = sqrt(mean((adult_predicted - penip_no_drift_idw_observations_sf_utm$adult_density_m2)^2, na.rm = TRUE)) * 10000,
    adult_mae_per_ha = mean(abs(adult_predicted - penip_no_drift_idw_observations_sf_utm$adult_density_m2), na.rm = TRUE) * 10000,
    adult_predicted_sites = sum(!is.na(adult_predicted)),
    juvenile_rmse_per_ha = sqrt(mean((juvenile_predicted - penip_no_drift_idw_observations_sf_utm$juvenile_density_m2)^2, na.rm = TRUE)) * 10000,
    juvenile_mae_per_ha = mean(abs(juvenile_predicted - penip_no_drift_idw_observations_sf_utm$juvenile_density_m2), na.rm = TRUE) * 10000,
    juvenile_predicted_sites = sum(!is.na(juvenile_predicted))
  )
}

penip_no_drift_idw_power_cv <- pmap_dfr(
  penip_idw_power_grid,
  summarize_penip_no_drift_loocv
)

penip_no_drift_idw_best_power <- bind_rows(
  penip_no_drift_idw_power_cv %>%
    slice_min(adult_rmse_per_ha, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Adult",
      best_power_by_rmse = power,
      search_radius = search_radius_label,
      rmse_per_ha = adult_rmse_per_ha,
      mae_per_ha = adult_mae_per_ha,
      predicted_sites = adult_predicted_sites
    ),
  penip_no_drift_idw_power_cv %>%
    slice_min(juvenile_rmse_per_ha, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Juvenile",
      best_power_by_rmse = power,
      search_radius = search_radius_label,
      rmse_per_ha = juvenile_rmse_per_ha,
      mae_per_ha = juvenile_mae_per_ha,
      predicted_sites = juvenile_predicted_sites
    )
)

knitr::kable(
  penip_no_drift_idw_best_power,
  digits = 2,
  caption = "Leave-one-out IDW cross-validation using Port Everglades radial and broad-scale observation units only, excluding drift transect segments. Error is calculated against observed density and reported as conch per hectare."
)
```

```{r penip-no-drift-idw-power-cross-validation-plot, fig.width=8, fig.height=4.5}
penip_no_drift_idw_power_cv_long <- penip_no_drift_idw_power_cv %>%
  select(power, search_radius_label, adult_rmse_per_ha, juvenile_rmse_per_ha) %>%
  pivot_longer(
    cols = c(adult_rmse_per_ha, juvenile_rmse_per_ha),
    names_to = "age_class",
    values_to = "rmse_per_ha"
  ) %>%
  mutate(
    age_class = recode(
      age_class,
      adult_rmse_per_ha = "Adult",
      juvenile_rmse_per_ha = "Juvenile"
    )
  )

penip_no_drift_idw_power_cv_plot <- ggplot(
  penip_no_drift_idw_power_cv_long,
  aes(x = power, y = rmse_per_ha, color = search_radius_label, group = search_radius_label)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~ age_class, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "Search radius") +
  labs(
    x = "IDW power",
    y = "Leave-one-out RMSE (conch per ha)",
    title = "Port Everglades IDW power cross-validation without drift segments",
    subtitle = "Radial surveys and broad-scale sites only"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

penip_no_drift_idw_power_cv_plot

ggsave(
  output_penip_no_drift_idw_power_cv_plot,
  penip_no_drift_idw_power_cv_plot,
  width = 8,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

```{r penip-age-specific-radius-idw-power-cross-validation}
penip_age_specific_radius_grid <- tidyr::expand_grid(
  power = seq(1, 6, by = 0.25)
)

format_search_radius_label <- function(search_radius_m) {
  if_else(is.infinite(search_radius_m), "No limit", paste0(search_radius_m, " m"))
}

summarize_penip_age_specific_radius_loocv <- function(power) {
  adult_predicted <- calculate_loocv_idw(
    penip_idw_cv_distance_m,
    idw_observations_sf_utm$adult_density_m2,
    power = power,
    search_radius_m = penip_adult_cv_search_radius_m
  )

  juvenile_predicted <- calculate_loocv_idw(
    penip_idw_cv_distance_m,
    idw_observations_sf_utm$juvenile_density_m2,
    power = power,
    search_radius_m = penip_juvenile_cv_search_radius_m
  )

  tibble(
    power = power,
    adult_search_radius = format_search_radius_label(penip_adult_cv_search_radius_m),
    juvenile_search_radius = format_search_radius_label(penip_juvenile_cv_search_radius_m),
    adult_rmse_per_ha = sqrt(mean((adult_predicted - idw_observations_sf_utm$adult_density_m2)^2, na.rm = TRUE)) * 10000,
    adult_mae_per_ha = mean(abs(adult_predicted - idw_observations_sf_utm$adult_density_m2), na.rm = TRUE) * 10000,
    adult_predicted_sites = sum(!is.na(adult_predicted)),
    juvenile_rmse_per_ha = sqrt(mean((juvenile_predicted - idw_observations_sf_utm$juvenile_density_m2)^2, na.rm = TRUE)) * 10000,
    juvenile_mae_per_ha = mean(abs(juvenile_predicted - idw_observations_sf_utm$juvenile_density_m2), na.rm = TRUE) * 10000,
    juvenile_predicted_sites = sum(!is.na(juvenile_predicted))
  )
}

penip_age_specific_radius_idw_power_cv <- map_dfr(
  penip_age_specific_radius_grid$power,
  summarize_penip_age_specific_radius_loocv
)

penip_age_specific_radius_idw_best_power <- bind_rows(
  penip_age_specific_radius_idw_power_cv %>%
    slice_min(adult_rmse_per_ha, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Adult",
      best_power_by_rmse = power,
      search_radius = adult_search_radius,
      rmse_per_ha = adult_rmse_per_ha,
      mae_per_ha = adult_mae_per_ha,
      predicted_sites = adult_predicted_sites
    ),
  penip_age_specific_radius_idw_power_cv %>%
    slice_min(juvenile_rmse_per_ha, n = 1, with_ties = FALSE) %>%
    transmute(
      age_class = "Juvenile",
      best_power_by_rmse = power,
      search_radius = juvenile_search_radius,
      rmse_per_ha = juvenile_rmse_per_ha,
      mae_per_ha = juvenile_mae_per_ha,
      predicted_sites = juvenile_predicted_sites
    )
)

knitr::kable(
  penip_age_specific_radius_idw_best_power,
  digits = 2,
  caption = "Leave-one-out IDW cross-validation using all Port Everglades observation units, including drift segments, with adult search radius fixed at 330 m and juvenile search radius set to no limit."
)
```

```{r penip-age-specific-radius-idw-power-cross-validation-plot, fig.width=8, fig.height=4.5}
penip_age_specific_radius_idw_power_cv_long <- bind_rows(
  penip_age_specific_radius_idw_power_cv %>%
    transmute(
      power,
      age_class = "Adult",
      search_radius = adult_search_radius,
      rmse_per_ha = adult_rmse_per_ha
    ),
  penip_age_specific_radius_idw_power_cv %>%
    transmute(
      power,
      age_class = "Juvenile",
      search_radius = juvenile_search_radius,
      rmse_per_ha = juvenile_rmse_per_ha
    )
)

penip_age_specific_radius_idw_power_cv_plot <- ggplot(
  penip_age_specific_radius_idw_power_cv_long,
  aes(x = power, y = rmse_per_ha, color = age_class, group = age_class)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~ age_class, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "Age class") +
  labs(
    x = "IDW power",
    y = "Leave-one-out RMSE (conch per ha)",
    title = "Port Everglades IDW power cross-validation with age-specific search radii",
    subtitle = "All observation units included; adult radius 330 m, juvenile radius no limit"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

penip_age_specific_radius_idw_power_cv_plot

ggsave(
  output_penip_age_specific_radius_idw_power_cv_plot,
  penip_age_specific_radius_idw_power_cv_plot,
  width = 8,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

```{r idw-prediction}
survey_effort_extent_utm <- st_as_sfc(st_bbox(
  c(
    xmin = survey_effort_xlim[[1]],
    xmax = survey_effort_xlim[[2]],
    ymin = survey_effort_ylim[[1]],
    ymax = survey_effort_ylim[[2]]
  ),
  crs = st_crs(4326)
)) %>%
  st_transform(utm_crs)

idw_area <- st_intersection(
  target_habitats_utm %>% st_union() %>% st_make_valid(),
  survey_effort_extent_utm
) %>%
  st_make_valid()

idw_bbox <- st_bbox(survey_effort_extent_utm)

conch_data_bbox <- st_bbox(idw_observations_sf_utm)
idw_plot_bbox <- idw_bbox
idw_plot_bbox[c("ymin", "ymax")] <- conch_data_bbox[c("ymin", "ymax")] + c(-150, 150)

idw_grid <- st_make_grid(
  idw_area,
  cellsize = idw_cellsize_m,
  what = "polygons",
  square = TRUE
) %>%
  st_sf(geometry = .) %>%
  st_intersection(st_sf(geometry = idw_area)) %>%
  mutate(cell_area_m2 = as.numeric(st_area(geometry))) %>%
  filter(cell_area_m2 > 0)

idw_grid_points <- st_point_on_surface(idw_grid)

idw_distance_matrix_m <- st_distance(
  idw_grid_points,
  idw_observations_sf_utm
) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(idw_grid_points), ncol = nrow(idw_observations_sf_utm))

idw_prediction <- idw_grid %>%
  mutate(
    adult_density_m2 = calculate_idw_prediction(
      idw_distance_matrix_m,
      idw_observations_sf_utm$adult_density_m2,
      power = idw_power,
      search_radius_m = idw_search_radius_m
    ),
    juvenile_density_m2 = calculate_idw_prediction(
      idw_distance_matrix_m,
      idw_observations_sf_utm$juvenile_density_m2,
      power = idw_power,
      search_radius_m = idw_search_radius_m
    ),
    adult_total = adult_density_m2 * cell_area_m2,
    juvenile_total = juvenile_density_m2 * cell_area_m2
  )

idw_population_summary <- tibble(
  modeled_area_m2 = sum(idw_prediction$cell_area_m2),
  modeled_area_ha = modeled_area_m2 / 10000,
  estimated_adults = sum(idw_prediction$adult_total, na.rm = TRUE),
  estimated_juveniles = sum(idw_prediction$juvenile_total, na.rm = TRUE),
  adult_density_per_ha = estimated_adults / modeled_area_ha,
  juvenile_density_per_ha = estimated_juveniles / modeled_area_ha,
  observation_units = nrow(idw_observations_sf_utm)
)

knitr::kable(idw_population_summary, digits = 2)
```

# IDW Parameter Sensitivity

This sensitivity analysis varies IDW power and search radius. All other IDW choices are held at their current baseline values.

```{r idw-fixed-settings}
idw_fixed_settings <- tibble(
  setting = c(
    "Baseline power",
    "Baseline search radius",
    "Prediction cell size",
    "Prediction mask",
    "Observation units",
    "Density response",
    "Density transform",
    "Observation weighting",
    "Maximum neighbors",
    "Minimum neighbors",
    "Distance metric",
    "Drift segment length",
    "Drift survey width",
    "Radial survey area",
    "Berry broad-scale site area"
  ),
  value = c(
    as.character(idw_power),
    if_else(is.infinite(idw_search_radius_m), "No limit / all observations", paste0(idw_search_radius_m, " m")),
    paste0(idw_cellsize_m, " m"),
    "Inner Reef and Nearshore Ridge Complex within the survey map extent; Artificial included with Nearshore Ridge Complex",
    "PENIP radial surveys, 20 m PENIP drift segments, and Berry broad-scale sites",
    "Observed count divided by survey area, interpolated separately for adults and juveniles",
    "None; raw density is interpolated",
    "None beyond converting each survey unit to density; each observation unit contributes equally to IDW",
    "No limit",
    "One observation; cells with no observations inside a finite search radius are left unpredicted",
    paste0("Straight-line Euclidean distance in UTM EPSG:", utm_crs),
    paste0(drift_segment_length_m, " m"),
    paste0(drift_half_width_m * 2, " m"),
    paste0(round(radial_survey_area_m2, 2), " m2"),
    paste0(berry_broad_site_area_m2, " m2")
  )
)

knitr::kable(idw_fixed_settings)
```

```{r idw-power-radius-grid}
idw_sensitivity_grid <- tidyr::expand_grid(
  power = c(2, 3, 4),
  search_radius_m = c(100, 250, 1000)
) %>%
  mutate(
    search_radius_label = paste0(search_radius_m, " m"),
    search_radius_label = factor(search_radius_label, levels = c("100 m", "250 m", "1000 m")),
    parameter_label = paste0(
      "Power ", power,
      ", ", search_radius_label
    )
  )

predict_idw_sensitivity_set <- function(
  power,
  search_radius_m,
  search_radius_label,
  parameter_label
) {
  adult_density_m2 <- calculate_idw_prediction(
    idw_distance_matrix_m,
    idw_observations_sf_utm$adult_density_m2,
    power = power,
    search_radius_m = search_radius_m
  )

  juvenile_density_m2 <- calculate_idw_prediction(
    idw_distance_matrix_m,
    idw_observations_sf_utm$juvenile_density_m2,
    power = power,
    search_radius_m = search_radius_m
  )

  idw_grid %>%
    mutate(
      power = power,
      search_radius_m = search_radius_m,
      search_radius_label = search_radius_label,
      parameter_label = parameter_label,
      adult_density_m2 = adult_density_m2,
      juvenile_density_m2 = juvenile_density_m2,
      adult_total = adult_density_m2 * cell_area_m2,
      juvenile_total = juvenile_density_m2 * cell_area_m2
    )
}

idw_sensitivity_predictions <- pmap_dfr(
  idw_sensitivity_grid,
  predict_idw_sensitivity_set
)

impact_zone_sensitivity_order <- c(
  "Channel",
  "Side Slopes",
  "Depth_10cm",
  "Depth_5cm",
  "Depth_1cm",
  "Depth_0_5cm",
  "Depth_0_1cm",
  "Nearshore",
  "North",
  "East",
  "South",
  "Impact Monitoring Area"
)

impact_zone_sensitivity_labels <- c(
  "Channel" = "Channel",
  "Side Slopes" = "Side Slopes",
  "Depth_10cm" = "Sediment deposition >10 cm",
  "Depth_5cm" = "Sediment deposition 5-10 cm",
  "Depth_1cm" = "Sediment deposition 1-5 cm",
  "Depth_0_5cm" = "Sediment deposition 0.5-1 cm",
  "Depth_0_1cm" = "Sediment deposition 0.1-0.5 cm",
  "Nearshore" = "Nearshore",
  "North" = "North",
  "East" = "East",
  "South" = "South",
  "Impact Monitoring Area" = "Rest of impact monitoring area"
)

idw_impact_zone_cell_area <- st_intersection(
  idw_grid %>%
    mutate(cell_id = row_number()) %>%
    select(cell_id),
  impact_zones_utm %>% select(name)
) %>%
  mutate(zone_cell_area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  filter(name %in% impact_zone_sensitivity_order)

summarize_idw_parameter_zones <- function(parameter_prediction) {
  idw_impact_zone_cell_area %>%
    left_join(parameter_prediction, by = "cell_id") %>%
    group_by(
      power,
      search_radius_m,
      search_radius_label,
      parameter_label,
      name
    ) %>%
    summarize(
      impact_area_ha = sum(zone_cell_area_m2, na.rm = TRUE) / 10000,
      adult_predicted_area_ha = sum(zone_cell_area_m2[!is.na(adult_density_m2)], na.rm = TRUE) / 10000,
      juvenile_predicted_area_ha = sum(zone_cell_area_m2[!is.na(juvenile_density_m2)], na.rm = TRUE) / 10000,
      estimated_adults = sum(adult_density_m2 * zone_cell_area_m2, na.rm = TRUE),
      estimated_juveniles = sum(juvenile_density_m2 * zone_cell_area_m2, na.rm = TRUE),
      estimated_total = estimated_adults + estimated_juveniles,
      .groups = "drop"
    ) %>%
    tidyr::complete(
      nesting(
        power,
        search_radius_m,
        search_radius_label,
        parameter_label
      ),
      name = impact_zone_sensitivity_order
    ) %>%
    mutate(
      impact_area_ha = replace_na(impact_area_ha, 0),
      adult_predicted_area_ha = replace_na(adult_predicted_area_ha, 0),
      juvenile_predicted_area_ha = replace_na(juvenile_predicted_area_ha, 0),
      estimated_adults = replace_na(estimated_adults, 0),
      estimated_juveniles = replace_na(estimated_juveniles, 0),
      estimated_total = replace_na(estimated_total, 0),
      impact_area = impact_zone_sensitivity_labels[name],
      name = factor(name, levels = impact_zone_sensitivity_order)
    ) %>%
    arrange(power, search_radius_m, name) %>%
    select(
      power,
      search_radius_m,
      search_radius_label,
      parameter_label,
      name,
      impact_area,
      impact_area_ha,
      adult_predicted_area_ha,
      juvenile_predicted_area_ha,
      estimated_adults,
      estimated_juveniles,
      estimated_total
    )
}

idw_sensitivity_cell_predictions <- idw_sensitivity_predictions %>%
  st_drop_geometry() %>%
  group_by(power, search_radius_m, search_radius_label, parameter_label) %>%
  mutate(cell_id = row_number()) %>%
  ungroup()

idw_parameter_sensitivity_zones <- summarize_idw_parameter_zones(
  idw_sensitivity_cell_predictions %>%
    select(
      power,
      search_radius_m,
      search_radius_label,
      parameter_label,
      cell_id,
      adult_density_m2,
      juvenile_density_m2
    )
) %>%
  mutate(
    search_radius_label = factor(search_radius_label, levels = c("100 m", "250 m", "1000 m")),
    baseline = power == 2 & search_radius_m == 1000
  )

idw_parameter_sensitivity <- idw_parameter_sensitivity_zones %>%
  group_by(power, search_radius_m, search_radius_label, parameter_label, baseline) %>%
  summarize(
    impact_area_ha = sum(impact_area_ha, na.rm = TRUE),
    adult_predicted_area_ha = sum(adult_predicted_area_ha, na.rm = TRUE),
    juvenile_predicted_area_ha = sum(juvenile_predicted_area_ha, na.rm = TRUE),
    adult_unpredicted_area_ha = impact_area_ha - adult_predicted_area_ha,
    juvenile_unpredicted_area_ha = impact_area_ha - juvenile_predicted_area_ha,
    estimated_adults = sum(estimated_adults, na.rm = TRUE),
    estimated_juveniles = sum(estimated_juveniles, na.rm = TRUE),
    estimated_total = sum(estimated_total, na.rm = TRUE),
    adult_density_per_ha = estimated_adults / adult_predicted_area_ha,
    juvenile_density_per_ha = estimated_juveniles / juvenile_predicted_area_ha,
    .groups = "drop"
  )

idw_parameter_sensitivity_display <- idw_parameter_sensitivity %>%
  transmute(
    Power = power,
    `Search radius` = search_radius_label,
    `Adult conch in impact/monitoring areas` = round(estimated_adults),
    `Juvenile conch in impact/monitoring areas` = round(estimated_juveniles),
    `Total conch in impact/monitoring areas` = round(estimated_total),
    `Adult density per predicted ha` = round(adult_density_per_ha, 2),
    `Juvenile density per predicted ha` = round(juvenile_density_per_ha, 2),
    `Predicted impact/monitoring area ha` = round(adult_predicted_area_ha, 2),
    `Unpredicted impact/monitoring area ha` = round(adult_unpredicted_area_ha, 2),
    Baseline = if_else(baseline, "Current setting", "")
  )

idw_baseline_power_search_radius_zone_display <- idw_parameter_sensitivity_zones %>%
  filter(power == idw_power) %>%
  transmute(
    `Impact area` = impact_area,
    `Search radius` = search_radius_label,
    `Adult conch` = round(estimated_adults),
    `Juvenile conch` = round(estimated_juveniles),
    `Total conch` = round(estimated_total)
  ) %>%
  pivot_wider(
    names_from = `Search radius`,
    values_from = c(`Adult conch`, `Juvenile conch`, `Total conch`)
  )

knitr::kable(
  idw_parameter_sensitivity_display,
  caption = "Sensitivity of total conch estimates summed across the same impact and monitoring areas used in the main abundance table."
)

knitr::kable(
  idw_baseline_power_search_radius_zone_display,
  caption = paste0("Impact-area estimates across search radii at the baseline IDW power of ", idw_power, ".")
)
```

```{r idw-parameter-sensitivity-plot, fig.width=8, fig.height=5}
idw_parameter_sensitivity_long <- idw_parameter_sensitivity %>%
  select(power, search_radius_label, estimated_adults, estimated_juveniles, baseline) %>%
  pivot_longer(
    cols = c(estimated_adults, estimated_juveniles),
    names_to = "age_class",
    values_to = "estimated_conch"
  ) %>%
  mutate(
    age_class = recode(
      age_class,
      estimated_adults = "Adult",
      estimated_juveniles = "Juvenile"
    )
  )

idw_parameter_sensitivity_plot <- ggplot(
  idw_parameter_sensitivity_long,
  aes(x = power, y = estimated_conch, color = search_radius_label, group = search_radius_label)
) +
  geom_line(linewidth = 0.7) +
  geom_point(aes(shape = baseline), size = 2.1) +
  facet_wrap(~ age_class, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "Search radius") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8), guide = "none") +
  labs(
    x = "IDW power",
    y = "Estimated conch in impact/monitoring areas",
    title = "IDW sensitivity for impact and monitoring area totals"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

idw_parameter_sensitivity_plot

ggsave(
  output_idw_parameter_sensitivity_plot,
  idw_parameter_sensitivity_plot,
  width = 8,
  height = 5,
  dpi = 300,
  bg = "white"
)
```
```{r idw-habitat-estimates}
idw_prediction_cells <- idw_prediction %>%
  mutate(cell_id = row_number())

habitat_prediction_area <- st_intersection(
  idw_prediction_cells,
  target_habitats_utm
) %>%
  mutate(intersection_area_m2 = as.numeric(st_area(geometry)))

habitat_idw_summary <- habitat_prediction_area %>%
  st_drop_geometry() %>%
  group_by(conch_habitat) %>%
  summarize(
    area_m2 = sum(intersection_area_m2, na.rm = TRUE),
    area_ha = area_m2 / 10000,
    estimated_adults = sum(adult_density_m2 * intersection_area_m2, na.rm = TRUE),
    estimated_juveniles = sum(juvenile_density_m2 * intersection_area_m2, na.rm = TRUE),
    adult_density_per_100m2 = estimated_adults / area_m2 * 100,
    juvenile_density_per_100m2 = estimated_juveniles / area_m2 * 100,
    adult_density_per_ha = estimated_adults / area_ha,
    juvenile_density_per_ha = estimated_juveniles / area_ha,
    .groups = "drop"
  ) %>%
  arrange(conch_habitat)

habitat_zone_prediction_area <- st_intersection(
  habitat_prediction_area,
  impact_zones_utm %>% select(name, scenario)
) %>%
  mutate(zone_area_m2 = as.numeric(st_area(geometry)))

habitat_zone_idw_summary <- habitat_zone_prediction_area %>%
  st_drop_geometry() %>%
  group_by(scenario, name, conch_habitat) %>%
  summarize(
    area_m2 = sum(zone_area_m2, na.rm = TRUE),
    area_ha = area_m2 / 10000,
    estimated_adults = sum(adult_density_m2 * zone_area_m2, na.rm = TRUE),
    estimated_juveniles = sum(juvenile_density_m2 * zone_area_m2, na.rm = TRUE),
    adult_density_per_100m2 = estimated_adults / area_m2 * 100,
    juvenile_density_per_100m2 = estimated_juveniles / area_m2 * 100,
    adult_density_per_ha = estimated_adults / area_ha,
    juvenile_density_per_ha = estimated_juveniles / area_ha,
    .groups = "drop"
  ) %>%
  arrange(scenario, name, conch_habitat)

knitr::kable(habitat_idw_summary, digits = 2)
knitr::kable(habitat_zone_idw_summary, digits = 2)
```

```{r idw-impact-zone-total-table}
impact_zone_order <- c(
  "Channel",
  "Side Slopes",
  "Depth_10cm",
  "Depth_5cm",
  "Depth_1cm",
  "Depth_0_5cm",
  "Depth_0_1cm",
  "Nearshore",
  "North",
  "East",
  "South",
  "Impact Monitoring Area"
)

impact_zone_labels <- c(
  "Channel" = "Channel",
  "Side Slopes" = "Side Slopes",
  "Depth_10cm" = "Sediment deposition >10 cm",
  "Depth_5cm" = "Sediment deposition 5-10 cm",
  "Depth_1cm" = "Sediment deposition 1-5 cm",
  "Depth_0_5cm" = "Sediment deposition 0.5-1 cm",
  "Depth_0_1cm" = "Sediment deposition 0.1-0.5 cm",
  "Nearshore" = "Nearshore",
  "North" = "North",
  "East" = "East",
  "South" = "South",
  "Impact Monitoring Area" = "Rest of impact monitoring area"
)

breeding_activity_buffer_utm <- breeding_activity_sites_sf_utm %>%
  st_buffer(breeding_activity_radius_m) %>%
  st_union() %>%
  st_make_valid()

adult_breeding_radius_prediction <- st_intersection(
  idw_prediction %>%
    mutate(
      density_per_ha = adult_density_m2 * 10000,
      density_class = case_when(
        density_per_ha <= 0 ~ "0",
        density_per_ha <= 10 ~ "1-10",
        density_per_ha <= 50 ~ "10-50",
        density_per_ha <= 100 ~ "50-100",
        TRUE ~ ">100"
      ),
      density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
    ),
  st_sf(geometry = breeding_activity_buffer_utm)
) %>%
  mutate(
    clipped_area_m2 = as.numeric(st_area(geometry)),
    adult_total_clipped = adult_density_m2 * clipped_area_m2
  )

adult_breeding_activity_zone_totals <- st_intersection(
  adult_breeding_radius_prediction %>% select(adult_density_m2),
  impact_zones_utm %>% select(name)
) %>%
  mutate(zone_area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  group_by(name) %>%
  summarize(
    estimated_adults_near_breeding = sum(adult_density_m2 * zone_area_m2, na.rm = TRUE),
    .groups = "drop"
  )

conch_impact_zone_totals <- habitat_zone_idw_summary %>%
  group_by(name) %>%
  summarize(
    estimated_adults = sum(estimated_adults, na.rm = TRUE),
    estimated_juveniles = sum(estimated_juveniles, na.rm = TRUE),
    estimated_total = estimated_adults + estimated_juveniles,
    .groups = "drop"
  ) %>%
  left_join(adult_breeding_activity_zone_totals, by = "name") %>%
  right_join(tibble(name = impact_zone_order), by = "name") %>%
  mutate(
    estimated_adults = replace_na(estimated_adults, 0),
    estimated_adults_near_breeding = replace_na(estimated_adults_near_breeding, 0),
    estimated_juveniles = replace_na(estimated_juveniles, 0),
    estimated_total = replace_na(estimated_total, 0),
    impact_area = impact_zone_labels[name],
    name = factor(name, levels = impact_zone_order)
  ) %>%
  arrange(name) %>%
  mutate(
    cumulative_adults = cumsum(estimated_adults),
    cumulative_juveniles = cumsum(estimated_juveniles),
    cumulative_total = cumsum(estimated_total)
  ) %>%
  select(
    impact_area,
    estimated_adults,
    estimated_adults_near_breeding,
    estimated_juveniles,
    estimated_total,
    cumulative_adults,
    cumulative_juveniles,
    cumulative_total
  )

conch_impact_zone_totals_with_total <- bind_rows(
  conch_impact_zone_totals,
  conch_impact_zone_totals %>%
    summarize(
      impact_area = "Total",
      estimated_adults = sum(estimated_adults, na.rm = TRUE),
      estimated_adults_near_breeding = sum(estimated_adults_near_breeding, na.rm = TRUE),
      estimated_juveniles = sum(estimated_juveniles, na.rm = TRUE),
      estimated_total = sum(estimated_total, na.rm = TRUE),
      cumulative_adults = sum(estimated_adults, na.rm = TRUE),
      cumulative_juveniles = sum(estimated_juveniles, na.rm = TRUE),
      cumulative_total = sum(estimated_total, na.rm = TRUE)
    )
)

conch_impact_zone_totals_display <- conch_impact_zone_totals_with_total %>%
  mutate(across(
    where(is.numeric),
    ~ format(round(.), big.mark = ",", scientific = FALSE, trim = TRUE)
  )) %>%
  rename(
    `Impact area` = impact_area,
    `Adult conch` = estimated_adults,
    `Adult conch within 330 m of breeding activity` = estimated_adults_near_breeding,
    `Juvenile conch` = estimated_juveniles,
    `Total conch` = estimated_total,
    `Cumulative adult conch` = cumulative_adults,
    `Cumulative juvenile conch` = cumulative_juveniles,
    `Cumulative total conch` = cumulative_total
  )

knitr::kable(
  conch_impact_zone_totals_display,
  caption = "Estimated total adult and juvenile conch in each impact zone. Estimates are based on IDW-predicted habitat-specific densities within Inner Reef and Nearshore Ridge Complex habitat, with Artificial habitat included in Nearshore Ridge Complex. Adults near breeding activity are the adult IDW estimates restricted to areas within 330 m of sites where breeding activity was recorded. Cumulative totals sum across impact zones moving away from the channel."
)
```

```{r idw-map, fig.width=8, fig.height=7}
idw_prediction_long <- bind_rows(
  idw_prediction %>%
    mutate(
      age_class = "Adult",
      density_per_ha = adult_density_m2 * 10000,
      total_animals = adult_total
    ) %>%
    select(age_class, density_per_ha, total_animals, geometry),
  idw_prediction %>%
    mutate(
      age_class = "Juvenile",
      density_per_ha = juvenile_density_m2 * 10000,
      total_animals = juvenile_total
    ) %>%
    select(age_class, density_per_ha, total_animals, geometry)
) %>%
  mutate(
    density_class = case_when(
      is.na(density_per_ha) ~ NA_character_,
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
  )

target_habitat_outline <- st_intersection(
  target_habitats_utm,
  st_sf(geometry = survey_effort_extent_utm)
) %>%
  st_make_valid()

make_inner_boundary_lines <- function(x, offset_m = 15) {
  map_dfr(seq_len(nrow(x)), function(i) {
    feature_i <- x[i, ]
    inner_i <- st_buffer(feature_i, -offset_m)

    if (all(st_is_empty(inner_i))) {
      inner_i <- feature_i
    }

    inner_i %>%
      st_boundary() %>%
      st_collection_extract("LINESTRING")
  }) %>%
    st_as_sf() %>%
    st_make_valid()
}

target_habitat_inner_lines <- make_inner_boundary_lines(target_habitat_outline, offset_m = 15)

idw_map <- ggplot() +
  geom_sf(
    data = idw_prediction_long,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  facet_wrap(~ age_class) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    drop = FALSE,
    name = "Conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(face = "bold", size = 11),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

idw_map

ggsave(output_idw_map, idw_map, width = 9, height = 4.5, dpi = 300, bg = "white")
```

# Kriging Alternative

This section provides an exploratory alternative to IDW using local ordinary kriging. The same observation units and prediction grid are used as in the IDW analysis. Densities are modeled as conch per hectare, predictions use the nearest 20 observation units, and negative kriging predictions are clipped to zero before abundance totals are calculated.

```{r kriging-functions}
prepare_kriging_observations <- function(observations_sf, density_column) {
  observation_points <- observations_sf %>%
    st_point_on_surface() %>%
    mutate(
      density_per_ha = .data[[density_column]] * 10000
    ) %>%
    filter(is.finite(density_per_ha))

  observation_xy <- st_coordinates(observation_points)

  observation_points %>%
    st_drop_geometry() %>%
    mutate(
      kriging_x = observation_xy[, "X"],
      kriging_y = observation_xy[, "Y"]
    ) %>%
    group_by(kriging_x, kriging_y) %>%
    summarize(
      density_per_ha = sum(density_per_ha * survey_area_m2, na.rm = TRUE) / sum(survey_area_m2, na.rm = TRUE),
      survey_area_m2 = sum(survey_area_m2, na.rm = TRUE),
      observation_units = n(),
      .groups = "drop"
    ) %>%
    st_as_sf(coords = c("kriging_x", "kriging_y"), crs = st_crs(observations_sf))
}

fit_best_variogram <- function(observations_sp) {
  empirical_variogram <- gstat::variogram(
    density_per_ha ~ 1,
    observations_sp
  )

  initial_partial_sill <- max(stats::var(observations_sp$density_per_ha, na.rm = TRUE) * 0.75, 0.001)
  initial_nugget <- max(stats::var(observations_sp$density_per_ha, na.rm = TRUE) * 0.25, 0.001)
  initial_range <- max(max(empirical_variogram$dist, na.rm = TRUE) / 3, 1)

  candidate_models <- c("Exp", "Sph", "Gau")

  fitted_models <- map(candidate_models, function(model_name) {
    candidate <- gstat::vgm(
      psill = initial_partial_sill,
      model = model_name,
      range = initial_range,
      nugget = initial_nugget
    )

    tryCatch(
      gstat::fit.variogram(empirical_variogram, candidate),
      error = function(e) NULL,
      warning = function(w) suppressWarnings(gstat::fit.variogram(empirical_variogram, candidate))
    )
  })

  keep_models <- !map_lgl(fitted_models, is.null)

  if (!any(keep_models)) {
    stop("No valid variogram model could be fit for kriging.")
  }

  fitted_models <- fitted_models[keep_models]
  model_scores <- map_dbl(fitted_models, ~ attr(.x, "SSErr"))
  fitted_models[[which.min(model_scores)]]
}

predict_ordinary_kriging <- function(
  observations_sf,
  prediction_grid_sf,
  density_column,
  age_class,
  nmax = kriging_nmax,
  maxdist = kriging_maxdist_m,
  nmin = kriging_nmin
) {
  observations_for_kriging <- prepare_kriging_observations(observations_sf, density_column)
  observations_sp <- as(observations_for_kriging, "Spatial")

  prediction_points <- prediction_grid_sf %>%
    mutate(cell_id = row_number()) %>%
    st_point_on_surface()

  prediction_points_sp <- as(prediction_points, "Spatial")
  best_variogram <- fit_best_variogram(observations_sp)

  kriging_prediction <- gstat::krige(
    density_per_ha ~ 1,
    observations_sp,
    prediction_points_sp,
    model = best_variogram,
    nmax = nmax,
    maxdist = maxdist,
    nmin = nmin,
    debug.level = 0
  )

  prediction_grid_sf %>%
    mutate(
      cell_id = row_number(),
      age_class = age_class,
      density_per_ha_raw = kriging_prediction$var1.pred,
      kriging_variance_per_ha = kriging_prediction$var1.var,
      density_per_ha = pmax(density_per_ha_raw, 0),
      density_m2 = density_per_ha / 10000,
      estimated_total = density_m2 * cell_area_m2,
      variogram_model = as.character(best_variogram$model[[2]]),
      variogram_nugget = best_variogram$psill[[1]],
      variogram_partial_sill = best_variogram$psill[[2]],
      variogram_range_m = best_variogram$range[[2]],
      kriging_nmax = nmax,
      kriging_maxdist_m = maxdist,
      kriging_nmin = nmin
    )
}
```

# Kriging Cross-Validation

Leave-one-out cross-validation is used here to evaluate the same local kriging parameters considered in the sensitivity analysis. Adult and juvenile densities are evaluated separately, so the preferred kriging settings can differ between age classes. Exact duplicate observation coordinates are collapsed before cross-validation using the same area-weighted density calculation used for prediction.

```{r kriging-loocv, cache=TRUE}
kriging_cv_grid <- tidyr::expand_grid(
  kriging_nmax = kriging_sensitivity_nmax,
  kriging_maxdist_m = kriging_sensitivity_maxdist_m,
  kriging_nmin = kriging_sensitivity_nmin
) %>%
  mutate(
    kriging_maxdist_label = paste0(kriging_maxdist_m, " m"),
    kriging_parameter_label = paste0(
      "nmax ", kriging_nmax,
      ", maxdist ", kriging_maxdist_label,
      ", nmin ", kriging_nmin
    )
  )

calculate_kriging_loocv <- function(observations_sf, density_column, age_class, parameter_grid) {
  observations_for_kriging <- prepare_kriging_observations(observations_sf, density_column)
  observations_sp <- as(observations_for_kriging, "Spatial")
  best_variogram <- fit_best_variogram(observations_sp)
  null_rmse <- sqrt(mean(
    (observations_for_kriging$density_per_ha - mean(observations_for_kriging$density_per_ha, na.rm = TRUE))^2,
    na.rm = TRUE
  ))

  run_parameter_set <- function(parameter_row) {
    kriging_nmax <- parameter_row$kriging_nmax
    kriging_maxdist_m <- parameter_row$kriging_maxdist_m
    kriging_nmin <- parameter_row$kriging_nmin
    kriging_maxdist_label <- parameter_row$kriging_maxdist_label
    kriging_parameter_label <- parameter_row$kriging_parameter_label

    cv_prediction <- tryCatch(
      gstat::krige.cv(
        density_per_ha ~ 1,
        observations_sp,
        model = best_variogram,
        nfold = nrow(observations_for_kriging),
        nmax = kriging_nmax,
        maxdist = kriging_maxdist_m,
        nmin = kriging_nmin,
        debug.level = 0
      ),
      error = function(e) NULL
    )

    if (is.null(cv_prediction)) {
      return(tibble(
        age_class = age_class,
        kriging_nmax = kriging_nmax,
        kriging_maxdist_m = kriging_maxdist_m,
        kriging_maxdist_label = kriging_maxdist_label,
        kriging_nmin = kriging_nmin,
        kriging_parameter_label = kriging_parameter_label,
        observation_locations = nrow(observations_for_kriging),
        predicted_locations = 0,
        prediction_coverage_percent = 0,
        rmse_per_ha = NA_real_,
        mae_per_ha = NA_real_,
        mean_error_per_ha = NA_real_,
        rms_standardized_error = NA_real_,
        rmse_skill_vs_mean = NA_real_,
        variogram_model = as.character(best_variogram$model[[2]]),
        variogram_nugget = best_variogram$psill[[1]],
        variogram_partial_sill = best_variogram$psill[[2]],
        variogram_range_m = best_variogram$range[[2]]
      ))
    }

    cv_output <- as.data.frame(cv_prediction)

    tibble(
      age_class = age_class,
      kriging_nmax = kriging_nmax,
      kriging_maxdist_m = kriging_maxdist_m,
      kriging_maxdist_label = kriging_maxdist_label,
      kriging_nmin = kriging_nmin,
      kriging_parameter_label = kriging_parameter_label,
      observation_locations = nrow(cv_output),
      predicted_locations = sum(is.finite(cv_output$var1.pred)),
      prediction_coverage_percent = predicted_locations / observation_locations * 100,
      rmse_per_ha = sqrt(mean(cv_output$residual^2, na.rm = TRUE)),
      mae_per_ha = mean(abs(cv_output$residual), na.rm = TRUE),
      mean_error_per_ha = mean(cv_output$residual, na.rm = TRUE),
      rms_standardized_error = sqrt(mean(cv_output$zscore^2, na.rm = TRUE)),
      rmse_skill_vs_mean = 1 - rmse_per_ha / null_rmse,
      variogram_model = as.character(best_variogram$model[[2]]),
      variogram_nugget = best_variogram$psill[[1]],
      variogram_partial_sill = best_variogram$psill[[2]],
      variogram_range_m = best_variogram$range[[2]]
    )
  }

  parameter_rows <- split(parameter_grid, seq_len(nrow(parameter_grid)))

  if (.Platform$OS.type == "windows" || kriging_cv_cores <= 1) {
    map_dfr(parameter_rows, run_parameter_set)
  } else {
    parallel::mclapply(
      parameter_rows,
      run_parameter_set,
      mc.cores = min(kriging_cv_cores, length(parameter_rows))
    ) %>%
      bind_rows()
  }
}

kriging_loocv_results <- bind_rows(
  calculate_kriging_loocv(
    idw_observations_sf_utm,
    "adult_density_m2",
    "Adult",
    kriging_cv_grid
  ),
  calculate_kriging_loocv(
    idw_observations_sf_utm,
    "juvenile_density_m2",
    "Juvenile",
    kriging_cv_grid
  )
) %>%
  arrange(age_class, rmse_per_ha)

kriging_loocv_best <- kriging_loocv_results %>%
  filter(is.finite(rmse_per_ha)) %>%
  group_by(age_class) %>%
  slice_min(rmse_per_ha, n = 1, with_ties = FALSE) %>%
  ungroup()

knitr::kable(
  kriging_loocv_best %>%
    transmute(
      `Age class` = age_class,
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_label,
      nmin = kriging_nmin,
      `Observation locations` = observation_locations,
      `Predicted locations` = predicted_locations,
      `Coverage %` = round(prediction_coverage_percent, 1),
      `RMSE conch/ha` = round(rmse_per_ha, 2),
      `MAE conch/ha` = round(mae_per_ha, 2),
      `Mean error conch/ha` = round(mean_error_per_ha, 2),
      `RMSE skill vs mean` = round(rmse_skill_vs_mean, 3)
    ),
  caption = "Best local ordinary kriging settings by leave-one-out RMSE. Adult and juvenile density models are evaluated separately."
)

knitr::kable(
  kriging_loocv_results %>%
    transmute(
      `Age class` = age_class,
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_label,
      nmin = kriging_nmin,
      `Coverage %` = round(prediction_coverage_percent, 1),
      `RMSE conch/ha` = round(rmse_per_ha, 2),
      `MAE conch/ha` = round(mae_per_ha, 2),
      `Mean error conch/ha` = round(mean_error_per_ha, 2),
      `RMS standardized error` = round(rms_standardized_error, 2),
      `RMSE skill vs mean` = round(rmse_skill_vs_mean, 3)
    ),
  caption = "Leave-one-out kriging cross-validation for all local-neighborhood parameter sets. Lower RMSE and MAE indicate better held-out prediction; coverage shows the percentage of observation locations that could be predicted under each neighborhood constraint."
)
```

```{r kriging-loocv-plot, fig.width=9, fig.height=6}
kriging_loocv_plot <- ggplot(
  kriging_loocv_results,
  aes(
    x = kriging_maxdist_m,
    y = rmse_per_ha,
    color = factor(kriging_nmax),
    group = kriging_nmax
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(aes(shape = prediction_coverage_percent < 100), size = 2) +
  facet_grid(age_class ~ kriging_nmin, labeller = label_both) +
  scale_x_continuous(breaks = kriging_sensitivity_maxdist_m) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "nmax") +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 1),
    labels = c(`FALSE` = "100%", `TRUE` = "<100%"),
    name = "CV coverage"
  ) +
  labs(
    x = "Maximum search distance for kriging (m)",
    y = "Leave-one-out RMSE (conch per ha)",
    title = "Kriging parameter cross-validation"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

kriging_loocv_plot

ggsave(
  output_kriging_loocv_plot,
  kriging_loocv_plot,
  width = 9,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

# Drift Segment Length Check for Kriging

The best kriging settings from the leave-one-out cross-validation are used here to test whether increasing drift transect segment length from 20 m to 60 m changes adult or juvenile estimates. Only the drift observation units are rebuilt; radial surveys, broad-scale surveys, habitat mask, impact zones, and kriging settings are held constant.

```{r kriging-drift-segment-length-check}
kriging_drift_segment_lengths_m <- c(20, 60)

make_drift_segment_observations <- function(segment_length_m) {
  drift_segment_table_i <- map_dfr(seq_len(nrow(drift_surveys)), function(i) {
    make_segment_rectangles(
      start_xy[i, ],
      end_xy[i, ],
      segment_length_m = segment_length_m,
      half_width_m = drift_half_width_m
    ) %>%
      mutate(
        .record = drift_surveys$.record[[i]],
        Transect = drift_surveys$Transect[[i]],
        transect_group = drift_surveys$transect_group[[i]],
        Line = drift_surveys$Line[[i]],
        segment_id = paste(segment_length_m, .record, segment_index, sep = "_")
      )
  })

  drift_segments_i <- st_as_sf(
    drift_segment_table_i,
    sf_column_name = "geometry",
    crs = utm_crs
  ) %>%
    select(segment_id, .record, Transect, transect_group, Line, segment_index, segment_start_m, segment_end_m) %>%
    mutate(segment_area_m2 = as.numeric(st_area(geometry)))

  drift_observation_segment_assignments_i <- assign_to_nearest_segment(
    conch_drift_positions_sf_utm,
    drift_segments_i
  )

  drift_segment_counts_i <- drift_observation_segment_assignments_i %>%
    filter(!is.na(segment_id)) %>%
    count(segment_id, animal_stage, name = "n_animals") %>%
    pivot_wider(
      names_from = animal_stage,
      values_from = n_animals,
      values_fill = 0
    )

  drift_segments_i %>%
    st_point_on_surface() %>%
    left_join(drift_segment_counts_i, by = "segment_id") %>%
    mutate(
      Adult = replace_na(Adult, 0),
      Juvenile = replace_na(Juvenile, 0),
      observation_source = paste0("PENIP drift segment ", segment_length_m, " m"),
      survey_area_m2 = segment_area_m2,
      adult_count = Adult,
      juvenile_count = Juvenile,
      drift_segment_length_m = segment_length_m
    ) %>%
    select(observation_source, survey_area_m2, adult_count, juvenile_count, drift_segment_length_m, geometry)
}

predict_best_kriging_for_segment_length <- function(segment_length_m) {
  observations_i <- bind_rows(
    radial_observations_sf_utm %>% mutate(drift_segment_length_m = segment_length_m),
    make_drift_segment_observations(segment_length_m),
    berry_observations_sf_utm %>% mutate(drift_segment_length_m = segment_length_m)
  ) %>%
    mutate(
      adult_density_m2 = adult_count / survey_area_m2,
      juvenile_density_m2 = juvenile_count / survey_area_m2
    )

  adult_params <- kriging_loocv_best %>% filter(age_class == "Adult")
  juvenile_params <- kriging_loocv_best %>% filter(age_class == "Juvenile")

  bind_rows(
    predict_ordinary_kriging(
      observations_i,
      idw_grid,
      "adult_density_m2",
      "Adult",
      nmax = adult_params$kriging_nmax[[1]],
      maxdist = adult_params$kriging_maxdist_m[[1]],
      nmin = adult_params$kriging_nmin[[1]]
    ),
    predict_ordinary_kriging(
      observations_i,
      idw_grid,
      "juvenile_density_m2",
      "Juvenile",
      nmax = juvenile_params$kriging_nmax[[1]],
      maxdist = juvenile_params$kriging_maxdist_m[[1]],
      nmin = juvenile_params$kriging_nmin[[1]]
    )
  ) %>%
    mutate(drift_segment_length_m = segment_length_m)
}

kriging_drift_length_predictions <- map_dfr(
  kriging_drift_segment_lengths_m,
  predict_best_kriging_for_segment_length
)

kriging_drift_length_modeled_totals <- kriging_drift_length_predictions %>%
  st_drop_geometry() %>%
  group_by(drift_segment_length_m, age_class) %>%
  summarize(
    predicted_area_ha = sum(cell_area_m2[!is.na(density_per_ha)], na.rm = TRUE) / 10000,
    estimated_conch_modeled_habitat = sum(estimated_total, na.rm = TRUE),
    density_per_predicted_ha = estimated_conch_modeled_habitat / predicted_area_ha,
    .groups = "drop"
  )

kriging_drift_length_impact_totals <- kriging_drift_length_predictions %>%
  st_drop_geometry() %>%
  select(drift_segment_length_m, age_class, cell_id, density_m2) %>%
  left_join(idw_impact_zone_cell_area, by = "cell_id") %>%
  group_by(drift_segment_length_m, age_class) %>%
  summarize(
    estimated_conch_impact_areas = sum(density_m2 * zone_cell_area_m2, na.rm = TRUE),
    .groups = "drop"
  )

kriging_drift_length_comparison <- kriging_drift_length_modeled_totals %>%
  left_join(
    kriging_drift_length_impact_totals,
    by = c("drift_segment_length_m", "age_class")
  ) %>%
  group_by(age_class) %>%
  mutate(
    impact_difference_from_20m = estimated_conch_impact_areas - estimated_conch_impact_areas[drift_segment_length_m == 20][[1]],
    impact_percent_difference_from_20m = impact_difference_from_20m / estimated_conch_impact_areas[drift_segment_length_m == 20][[1]] * 100,
    modeled_difference_from_20m = estimated_conch_modeled_habitat - estimated_conch_modeled_habitat[drift_segment_length_m == 20][[1]],
    modeled_percent_difference_from_20m = modeled_difference_from_20m / estimated_conch_modeled_habitat[drift_segment_length_m == 20][[1]] * 100
  ) %>%
  ungroup()

knitr::kable(
  kriging_drift_length_comparison %>%
    transmute(
      `Drift segment length` = paste0(drift_segment_length_m, " m"),
      `Age class` = age_class,
      `Predicted area ha` = round(predicted_area_ha, 2),
      `Modeled-habitat conch` = round(estimated_conch_modeled_habitat),
      `Modeled difference from 20 m` = round(modeled_difference_from_20m),
      `Modeled percent difference from 20 m` = round(modeled_percent_difference_from_20m, 1),
      `Impact/monitoring conch` = round(estimated_conch_impact_areas),
      `Impact difference from 20 m` = round(impact_difference_from_20m),
      `Impact percent difference from 20 m` = round(impact_percent_difference_from_20m, 1)
    ),
  caption = "Kriging sensitivity to drift transect segment length using the age-specific best LOOCV kriging settings."
)
```

```{r kriging-prediction}
adult_kriging_prediction <- predict_ordinary_kriging(
  idw_observations_sf_utm,
  idw_grid,
  "adult_density_m2",
  "Adult"
)

juvenile_kriging_prediction <- predict_ordinary_kriging(
  idw_observations_sf_utm,
  idw_grid,
  "juvenile_density_m2",
  "Juvenile"
)

kriging_prediction <- bind_rows(
  adult_kriging_prediction,
  juvenile_kriging_prediction
) %>%
  mutate(
    density_class = case_when(
      is.na(density_per_ha) ~ NA_character_,
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
  )

kriging_variogram_summary <- kriging_prediction %>%
  st_drop_geometry() %>%
  distinct(
    age_class,
    variogram_model,
    variogram_nugget,
    variogram_partial_sill,
    variogram_range_m,
    kriging_nmax,
    kriging_maxdist_m,
    kriging_nmin
  )

kriging_population_summary <- kriging_prediction %>%
  st_drop_geometry() %>%
  group_by(age_class) %>%
  summarize(
    modeled_area_m2 = sum(cell_area_m2, na.rm = TRUE),
    modeled_area_ha = modeled_area_m2 / 10000,
    predicted_area_m2 = sum(cell_area_m2[!is.na(density_per_ha)], na.rm = TRUE),
    predicted_area_ha = predicted_area_m2 / 10000,
    estimated_conch = sum(estimated_total, na.rm = TRUE),
    density_per_predicted_ha = estimated_conch / predicted_area_ha,
    .groups = "drop"
  )

knitr::kable(
  kriging_variogram_summary,
  digits = 3,
  caption = "Fitted local ordinary kriging variogram parameters for conch density in conch per hectare."
)

knitr::kable(
  kriging_population_summary,
  digits = 2,
  caption = "Kriging-estimated conch abundance over the same habitat-limited prediction grid used for IDW."
)
```

```{r kriging-impact-zone-table}
kriging_habitat_prediction_area <- st_intersection(
  kriging_prediction,
  target_habitats_utm
) %>%
  mutate(intersection_area_m2 = as.numeric(st_area(geometry)))

kriging_habitat_zone_prediction_area <- st_intersection(
  kriging_habitat_prediction_area,
  impact_zones_utm %>% select(name, scenario)
) %>%
  mutate(zone_area_m2 = as.numeric(st_area(geometry)))

kriging_impact_zone_totals <- kriging_habitat_zone_prediction_area %>%
  st_drop_geometry() %>%
  group_by(name, age_class) %>%
  summarize(
    estimated_conch = sum(density_m2 * zone_area_m2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = age_class,
    values_from = estimated_conch,
    values_fill = 0
  ) %>%
  right_join(tibble(name = impact_zone_order), by = "name") %>%
  mutate(
    Adult = replace_na(Adult, 0),
    Juvenile = replace_na(Juvenile, 0),
    Total = Adult + Juvenile,
    impact_area = impact_zone_labels[name],
    name = factor(name, levels = impact_zone_order)
  ) %>%
  arrange(name) %>%
  select(impact_area, Adult, Juvenile, Total)

kriging_impact_zone_totals_with_total <- bind_rows(
  kriging_impact_zone_totals,
  kriging_impact_zone_totals %>%
    summarize(
      impact_area = "Total",
      Adult = sum(Adult, na.rm = TRUE),
      Juvenile = sum(Juvenile, na.rm = TRUE),
      Total = sum(Total, na.rm = TRUE)
    )
)

kriging_impact_zone_totals_display <- kriging_impact_zone_totals_with_total %>%
  mutate(across(
    where(is.numeric),
    ~ format(round(.), big.mark = ",", scientific = FALSE, trim = TRUE)
  )) %>%
  rename(
    `Impact area` = impact_area,
    `Adult conch` = Adult,
    `Juvenile conch` = Juvenile,
    `Total conch` = Total
  )

knitr::kable(
  kriging_impact_zone_totals_display,
  caption = "Kriging-estimated total adult and juvenile conch in each impact zone. Estimates are based on local ordinary kriging over the same Inner Reef and Nearshore Ridge Complex habitat mask used for IDW, with Artificial habitat included in Nearshore Ridge Complex."
)
```

```{r kriging-map, fig.width=8, fig.height=4.5}
kriging_map <- ggplot() +
  geom_sf(
    data = kriging_prediction,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  facet_wrap(~ age_class) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    drop = FALSE,
    name = "Conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(face = "bold", size = 11),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

kriging_map

ggsave(output_kriging_map, kriging_map, width = 9, height = 4.5, dpi = 300, bg = "white")
```

# Kriging Sensitivity

This section explores the kriging levers most likely to control extrapolation artifacts: the maximum number of observations used (`nmax`), the maximum search distance (`maxdist`), and the minimum number of observations required (`nmin`). Cells without enough nearby observations are left unpredicted.

```{r kriging-sensitivity}
kriging_sensitivity_grid <- tidyr::expand_grid(
  kriging_nmax = kriging_sensitivity_nmax,
  kriging_maxdist_m = kriging_sensitivity_maxdist_m,
  kriging_nmin = kriging_sensitivity_nmin
) %>%
  mutate(
    kriging_maxdist_label = paste0(kriging_maxdist_m, " m"),
    kriging_parameter_label = paste0(
      "nmax ", kriging_nmax,
      ", maxdist ", kriging_maxdist_label,
      ", nmin ", kriging_nmin
    )
  )

predict_kriging_sensitivity_set <- function(kriging_nmax, kriging_maxdist_m, kriging_nmin, kriging_maxdist_label, kriging_parameter_label) {
  bind_rows(
    predict_ordinary_kriging(
      idw_observations_sf_utm,
      idw_grid,
      "adult_density_m2",
      "Adult",
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_m,
      nmin = kriging_nmin
    ),
    predict_ordinary_kriging(
      idw_observations_sf_utm,
      idw_grid,
      "juvenile_density_m2",
      "Juvenile",
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_m,
      nmin = kriging_nmin
    )
  ) %>%
    mutate(
      kriging_maxdist_label = kriging_maxdist_label,
      kriging_parameter_label = kriging_parameter_label
    )
}

kriging_sensitivity_predictions <- pmap_dfr(
  kriging_sensitivity_grid,
  predict_kriging_sensitivity_set
)

kriging_sensitivity_summary <- kriging_sensitivity_predictions %>%
  st_drop_geometry() %>%
  group_by(age_class, kriging_nmax, kriging_maxdist_m, kriging_maxdist_label, kriging_nmin, kriging_parameter_label) %>%
  summarize(
    modeled_area_ha = sum(cell_area_m2, na.rm = TRUE) / 10000,
    predicted_area_ha = sum(cell_area_m2[!is.na(density_per_ha)], na.rm = TRUE) / 10000,
    predicted_area_percent = predicted_area_ha / modeled_area_ha * 100,
    estimated_conch = sum(estimated_total, na.rm = TRUE),
    density_per_predicted_ha = estimated_conch / predicted_area_ha,
    maximum_density_per_ha = max(density_per_ha, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(age_class, kriging_nmin, kriging_nmax, kriging_maxdist_m)

kriging_sensitivity_display <- kriging_sensitivity_summary %>%
  transmute(
    `Age class` = age_class,
    nmax = kriging_nmax,
    maxdist = kriging_maxdist_label,
    nmin = kriging_nmin,
    `Predicted area ha` = round(predicted_area_ha, 2),
    `Predicted area %` = round(predicted_area_percent, 1),
    `Estimated conch` = round(estimated_conch),
    `Density per predicted ha` = round(density_per_predicted_ha, 2),
    `Maximum density per ha` = round(maximum_density_per_ha, 2)
  )

knitr::kable(
  kriging_sensitivity_display,
  caption = "Kriging sensitivity across local-neighborhood parameters. Smaller maxdist and larger nmin constrain extrapolation by leaving unsupported cells unpredicted."
)
```

```{r kriging-sensitivity-plot, fig.width=9, fig.height=6}
kriging_sensitivity_plot <- ggplot(
  kriging_sensitivity_summary,
  aes(
    x = kriging_maxdist_m,
    y = estimated_conch,
    color = factor(kriging_nmax),
    group = kriging_nmax
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  facet_grid(age_class ~ kriging_nmin, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = kriging_sensitivity_maxdist_m) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "nmax") +
  labs(
    x = "Maximum search distance for kriging (m)",
    y = "Estimated conch over predicted habitat area",
    title = "Kriging sensitivity to local-neighborhood settings"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

kriging_sensitivity_plot

ggsave(
  output_kriging_sensitivity_plot,
  kriging_sensitivity_plot,
  width = 9,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

```{r kriging-selected-constrained-map, fig.width=8, fig.height=4.5}
kriging_selected_prediction <- kriging_sensitivity_predictions %>%
  filter(
    kriging_nmax == kriging_selected_nmax,
    kriging_maxdist_m == kriging_selected_maxdist_m,
    kriging_nmin == kriging_selected_nmin
  ) %>%
  mutate(
    density_class = case_when(
      is.na(density_per_ha) ~ NA_character_,
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
  )

kriging_selected_summary <- kriging_sensitivity_summary %>%
  filter(
    kriging_nmax == kriging_selected_nmax,
    kriging_maxdist_m == kriging_selected_maxdist_m,
    kriging_nmin == kriging_selected_nmin
  )

knitr::kable(
  kriging_selected_summary %>%
    transmute(
      `Age class` = age_class,
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_label,
      nmin = kriging_nmin,
      `Predicted area ha` = round(predicted_area_ha, 2),
      `Predicted area %` = round(predicted_area_percent, 1),
      `Estimated conch` = round(estimated_conch),
      `Density per predicted ha` = round(density_per_predicted_ha, 2)
    ),
  caption = "Selected constrained kriging setting used for the map below."
)

kriging_selected_map <- ggplot() +
  geom_sf(
    data = kriging_selected_prediction,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  facet_wrap(~ age_class) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    na.value = "transparent",
    drop = FALSE,
    name = "Conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  labs(
    title = paste0(
      "Constrained local kriging: nmax ",
      kriging_selected_nmax,
      ", maxdist ",
      kriging_selected_maxdist_m,
      " m, nmin ",
      kriging_selected_nmin
    )
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

kriging_selected_map

ggsave(
  output_kriging_selected_map,
  kriging_selected_map,
  width = 9,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

```{r kriging-parameter-maps, fig.width=16, fig.height=24}
kriging_breeding_impact_cell_area <- st_intersection(
  idw_grid %>%
    mutate(cell_id = row_number()) %>%
    select(cell_id),
  impact_zones_utm %>% select(name)
) %>%
  st_intersection(st_sf(geometry = breeding_activity_buffer_utm)) %>%
  mutate(zone_cell_area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  filter(name %in% impact_zone_order)

kriging_sensitivity_cell_predictions <- kriging_sensitivity_predictions %>%
  st_drop_geometry() %>%
  select(
    age_class,
    kriging_nmax,
    kriging_maxdist_m,
    kriging_maxdist_label,
    kriging_nmin,
    kriging_parameter_label,
    cell_id,
    density_m2
  )

kriging_sensitivity_impact_totals <- kriging_sensitivity_cell_predictions %>%
  left_join(idw_impact_zone_cell_area, by = "cell_id") %>%
  group_by(age_class, kriging_nmax, kriging_maxdist_m, kriging_maxdist_label, kriging_nmin, kriging_parameter_label) %>%
  summarize(
    estimated_conch_impact_areas = sum(density_m2 * zone_cell_area_m2, na.rm = TRUE),
    .groups = "drop"
  )

kriging_sensitivity_adult_breeding_totals <- kriging_sensitivity_cell_predictions %>%
  filter(age_class == "Adult") %>%
  left_join(kriging_breeding_impact_cell_area, by = "cell_id") %>%
  group_by(kriging_nmax, kriging_maxdist_m, kriging_maxdist_label, kriging_nmin, kriging_parameter_label) %>%
  summarize(
    estimated_adults_impact_areas_within_330m_breeding = sum(density_m2 * zone_cell_area_m2, na.rm = TRUE),
    .groups = "drop"
  )

kriging_sensitivity_annotation <- kriging_sensitivity_impact_totals %>%
  select(
    age_class,
    kriging_nmax,
    kriging_maxdist_m,
    kriging_maxdist_label,
    kriging_nmin,
    kriging_parameter_label,
    estimated_conch_impact_areas
  ) %>%
  pivot_wider(
    names_from = age_class,
    values_from = estimated_conch_impact_areas,
    values_fill = 0
  ) %>%
  left_join(
    kriging_sensitivity_adult_breeding_totals,
    by = c("kriging_nmax", "kriging_maxdist_m", "kriging_maxdist_label", "kriging_nmin", "kriging_parameter_label")
  ) %>%
  mutate(
    estimated_adults_impact_areas_within_330m_breeding = replace_na(estimated_adults_impact_areas_within_330m_breeding, 0),
    label = paste0(
      "Impact A: ", format(round(Adult), big.mark = ",", scientific = FALSE, trim = TRUE),
      "\nImpact J: ", format(round(Juvenile), big.mark = ",", scientific = FALSE, trim = TRUE),
      "\nImpact A within 330 m: ", format(
        round(estimated_adults_impact_areas_within_330m_breeding),
        big.mark = ",",
        scientific = FALSE,
        trim = TRUE
      )
    ),
    x = idw_plot_bbox[["xmin"]] + 60,
    y = idw_plot_bbox[["ymax"]] - 60
  ) %>%
  tidyr::crossing(age_class = factor(c("Adult", "Juvenile"), levels = c("Adult", "Juvenile")))

kriging_parameter_map_data <- kriging_sensitivity_predictions %>%
  mutate(
    age_class = factor(age_class, levels = c("Adult", "Juvenile")),
    kriging_parameter_label = factor(kriging_parameter_label, levels = kriging_sensitivity_grid$kriging_parameter_label),
    density_class = case_when(
      is.na(density_per_ha) ~ NA_character_,
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
  )

kriging_parameter_map_tiles <- kriging_parameter_map_data %>%
  mutate(tile_xy = st_coordinates(st_point_on_surface(geometry))) %>%
  st_drop_geometry() %>%
  mutate(
    x = tile_xy[, "X"],
    y = tile_xy[, "Y"]
  ) %>%
  select(-tile_xy)

make_kriging_parameter_map <- function(map_age_class) {
  ggplot() +
    geom_tile(
      data = kriging_parameter_map_tiles %>% filter(age_class == map_age_class),
      aes(x = x, y = y, fill = density_class),
      width = idw_cellsize_m,
      height = idw_cellsize_m,
      alpha = 0.95
    ) +
    geom_sf(
      data = target_habitat_inner_lines,
      aes(color = conch_habitat),
      linewidth = 0.42
    ) +
    geom_sf(
      data = impact_zone_plot_outlines_utm,
      fill = NA,
      color = "grey20",
      linewidth = 0.14
    ) +
    geom_sf(
      data = st_boundary(st_sf(geometry = breeding_activity_buffer_utm)),
      fill = NA,
      color = "#7B3294",
      linewidth = 0.18
    ) +
    geom_label(
      data = kriging_sensitivity_annotation %>% filter(age_class == map_age_class),
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 1,
      size = 1.65,
      label.size = 0.1,
      label.padding = unit(0.06, "lines"),
      fill = "white",
      alpha = 0.88,
      lineheight = 0.88
    ) +
    facet_wrap(~kriging_parameter_label, ncol = 4) +
    scale_fill_manual(
      values = c(
        "0" = "white",
        "1-10" = "#F0F0F0",
        "10-50" = "#BDBDBD",
        "50-100" = "#737373",
        ">100" = "#252525"
      ),
      na.value = "transparent",
      drop = FALSE,
      name = "Conch per ha"
    ) +
    scale_color_manual(
      values = c(
        "Inner Reef" = type_cols[["Inner Reef"]],
        "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
      ),
      name = "Habitat"
    ) +
    coord_sf(
      crs = st_crs(utm_crs),
      xlim = idw_plot_bbox[c("xmin", "xmax")],
      ylim = idw_plot_bbox[c("ymin", "ymax")],
      expand = FALSE,
      datum = NA
    ) +
    labs(title = paste(map_age_class, "kriging sensitivity maps")) +
    theme_void(base_size = 8) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      strip.text = element_text(face = "bold", size = 6),
      panel.spacing = unit(0.22, "lines"),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 7)
    )
}

kriging_adult_parameter_maps <- make_kriging_parameter_map("Adult")
kriging_juvenile_parameter_maps <- make_kriging_parameter_map("Juvenile")

kriging_adult_parameter_maps
kriging_juvenile_parameter_maps

ggsave(
  output_kriging_adult_parameter_maps,
  kriging_adult_parameter_maps,
  width = 14,
  height = 18,
  dpi = 200,
  bg = "white"
)

ggsave(
  output_kriging_juvenile_parameter_maps,
  kriging_juvenile_parameter_maps,
  width = 14,
  height = 18,
  dpi = 200,
  bg = "white"
)
```

```{r adult-breeding-radius-idw-map, fig.width=5.5, fig.height=7}
adult_breeding_radius_summary <- adult_breeding_radius_prediction %>%
  st_drop_geometry() %>%
  summarize(
    breeding_activity_sites = nrow(breeding_activity_sites_sf_utm),
    radius_m = breeding_activity_radius_m,
    area_m2 = sum(clipped_area_m2, na.rm = TRUE),
    area_ha = area_m2 / 10000,
    estimated_adults = sum(adult_total_clipped, na.rm = TRUE),
    adult_density_per_ha = estimated_adults / area_ha
  )

knitr::kable(adult_breeding_radius_summary, digits = 2)

adult_breeding_radius_map <- ggplot() +
  geom_sf(
    data = adult_breeding_radius_prediction,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  geom_sf(
    data = breeding_activity_sites_sf_utm,
    shape = 21,
    fill = "#7B3294",
    color = "white",
    size = 2.2,
    stroke = 0.35
  ) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    drop = FALSE,
    name = "Adult conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  labs(title = paste0("Adult conch density within ", breeding_activity_radius_m, " m of breeding activity")) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

adult_breeding_radius_map

ggsave(
  output_adult_breeding_radius_idw_map,
  adult_breeding_radius_map,
  width = 5.5,
  height = 7,
  dpi = 300,
  bg = "white"
)
```

# Selected IDW Parameter Sets

These maps and totals focus on nine parameter sets: IDW power 2, 3, and 4, each crossed with search radii of 100 m, 250 m, and 1000 m. Drift transects are kept at the baseline 20 m segment length.

```{r selected-idw-parameter-totals}
selected_idw_parameter_grid <- idw_sensitivity_grid
selected_idw_predictions <- idw_sensitivity_predictions

selected_breeding_impact_cell_area <- st_intersection(
  idw_grid %>%
    mutate(cell_id = row_number()) %>%
    select(cell_id),
  impact_zones_utm %>% select(name)
) %>%
  st_intersection(st_sf(geometry = breeding_activity_buffer_utm)) %>%
  mutate(zone_cell_area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  filter(name %in% impact_zone_order)

selected_idw_cell_predictions <- selected_idw_predictions %>%
  st_drop_geometry() %>%
  group_by(power, search_radius_m, search_radius_label, parameter_label) %>%
  mutate(cell_id = row_number()) %>%
  ungroup()

selected_idw_parameter_totals <- selected_idw_cell_predictions %>%
  select(power, search_radius_m, search_radius_label, parameter_label, cell_id, adult_density_m2, juvenile_density_m2) %>%
  left_join(idw_impact_zone_cell_area, by = "cell_id") %>%
  group_by(power, search_radius_m, search_radius_label, parameter_label) %>%
  summarize(
    estimated_adults_impact_areas = sum(adult_density_m2 * zone_cell_area_m2, na.rm = TRUE),
    estimated_juveniles_impact_areas = sum(juvenile_density_m2 * zone_cell_area_m2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    selected_idw_cell_predictions %>%
      select(power, search_radius_m, search_radius_label, parameter_label, cell_id, adult_density_m2, juvenile_density_m2) %>%
      left_join(selected_breeding_impact_cell_area, by = "cell_id") %>%
      group_by(power, search_radius_m, search_radius_label, parameter_label) %>%
      summarize(
        estimated_adults_impact_areas_within_330m_breeding = sum(adult_density_m2 * zone_cell_area_m2, na.rm = TRUE),
        estimated_juveniles_impact_areas_within_330m_breeding = sum(juvenile_density_m2 * zone_cell_area_m2, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("power", "search_radius_m", "search_radius_label", "parameter_label")
  ) %>%
  arrange(power, search_radius_m)

selected_idw_parameter_totals_display <- selected_idw_parameter_totals %>%
  transmute(
    `IDW power` = power,
    `Search radius` = search_radius_label,
    `Adult conch across impact/monitoring areas` = format(round(estimated_adults_impact_areas), big.mark = ",", scientific = FALSE, trim = TRUE),
    `Juvenile conch across impact/monitoring areas` = format(round(estimated_juveniles_impact_areas), big.mark = ",", scientific = FALSE, trim = TRUE),
    `Adult conch in impact/monitoring areas within 330 m of breeding` = format(
      round(estimated_adults_impact_areas_within_330m_breeding),
      big.mark = ",",
      scientific = FALSE,
      trim = TRUE
    ),
    `Juvenile conch in impact/monitoring areas within 330 m of breeding` = format(
      round(estimated_juveniles_impact_areas_within_330m_breeding),
      big.mark = ",",
      scientific = FALSE,
      trim = TRUE
    )
  )

knitr::kable(
  selected_idw_parameter_totals_display,
  caption = "Selected IDW parameter-set totals for adult conch across impact and monitoring areas, and the subset of those areas within 330 m of reproductive-behavior observations."
)
```

```{r selected-idw-parameter-maps, fig.width=11, fig.height=13}
selected_idw_prediction_long <- bind_rows(
  selected_idw_predictions %>%
    mutate(
      age_class = "Adult",
      density_per_ha = adult_density_m2 * 10000
    ) %>%
    select(power, search_radius_label, parameter_label, age_class, density_per_ha, geometry),
  selected_idw_predictions %>%
    mutate(
      age_class = "Juvenile",
      density_per_ha = juvenile_density_m2 * 10000
    ) %>%
    select(power, search_radius_label, parameter_label, age_class, density_per_ha, geometry)
) %>%
  mutate(
    parameter_label = factor(parameter_label, levels = selected_idw_parameter_grid$parameter_label),
    density_class = case_when(
      is.na(density_per_ha) ~ NA_character_,
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100")),
    age_class = factor(age_class, levels = c("Adult", "Juvenile"))
  )

selected_idw_parameter_map_labels <- selected_idw_parameter_totals %>%
  mutate(
    parameter_label = factor(parameter_label, levels = selected_idw_parameter_grid$parameter_label)
  ) %>%
  transmute(
    parameter_label,
    label = paste0(
      "Impact A: ", format(round(estimated_adults_impact_areas), big.mark = ",", scientific = FALSE, trim = TRUE),
      "\nImpact J: ", format(round(estimated_juveniles_impact_areas), big.mark = ",", scientific = FALSE, trim = TRUE),
      "\n330 m A: ", format(round(estimated_adults_impact_areas_within_330m_breeding), big.mark = ",", scientific = FALSE, trim = TRUE),
      "\n330 m J: ", format(round(estimated_juveniles_impact_areas_within_330m_breeding), big.mark = ",", scientific = FALSE, trim = TRUE)
    ),
    x = idw_plot_bbox[["xmin"]] + 60,
    y = idw_plot_bbox[["ymax"]] - 60
  ) %>%
  tidyr::crossing(age_class = factor(c("Adult", "Juvenile"), levels = c("Adult", "Juvenile")))

selected_idw_parameter_maps <- ggplot() +
  geom_sf(
    data = selected_idw_prediction_long,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.55
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.18
  ) +
  geom_sf(
    data = st_boundary(st_sf(geometry = breeding_activity_buffer_utm)),
    fill = NA,
    color = "#7B3294",
    linewidth = 0.3
  ) +
  geom_sf(
    data = breeding_activity_sites_sf_utm,
    shape = 21,
    fill = "#7B3294",
    color = "white",
    size = 0.8,
    stroke = 0.15
  ) +
  geom_label(
    data = selected_idw_parameter_map_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 1.6,
    label.size = 0.12,
    label.padding = unit(0.08, "lines"),
    fill = "white",
    alpha = 0.88,
    lineheight = 0.9
  ) +
  facet_wrap(age_class ~ parameter_label, ncol = 3) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    na.value = "transparent",
    na.translate = FALSE,
    drop = FALSE,
    name = "Conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  theme_void(base_size = 9) +
  theme(
    legend.position = "bottom",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(face = "bold", size = 7),
    panel.spacing = unit(0.3, "lines"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7)
  )

selected_idw_parameter_maps

ggsave(
  output_idw_selected_parameter_maps,
  selected_idw_parameter_maps,
  width = 11,
  height = 13,
  dpi = 300,
  bg = "white"
)
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
