---
title: "PENIP Conch Blob and Background Estimates"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

find_project_root <- function(start = getwd()) {
  start <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- c(start, normalizePath(file.path(start, ".."), winslash = "/", mustWork = TRUE))

  while (length(candidates) > 0) {
    candidate <- candidates[[1]]
    if (file.exists(file.path(candidate, "data/processed/polygons_final.rds"))) {
      return(candidate)
    }

    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not find the PENIP project root from ", start, call. = FALSE)
    }

    candidates <- c(candidates[-1], parent)
  }
}

knitr::opts_knit$set(root.dir = find_project_root())
```

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(purrr)
library(readxl)
library(sf)
library(tibble)
library(tidyr)
```

```{r paths-and-constants}
conch_xlsx <- "data/conch/PENIP_QueenConchSurveys.xlsx"
radial_counts_csv <- "data/conch/PENIP_circular_counts.csv"
polygons_rds <- "data/processed/polygons_final.rds"

radial_sheet <- "PENIP_Conch_10mRadial"
drift_sheet <- "PENIP_DriftTransects"
drift_position_sheet <- "PENIP_ConchDriftPosition"

utm_crs <- 32617
radial_survey_radius_m <- 10
radial_survey_area_m2 <- pi * radial_survey_radius_m^2
drift_half_width_m <- 2
drift_segment_length_m <- 20
blob_buffer_m <- 150
prediction_cellsize_m <- 20

output_blob_context_map <- "outputs/conch_blob_context_map.png"
output_adult_model_comparison_plot <- "outputs/conch_blob_adult_model_comparison.png"
output_blob_prediction_map <- "outputs/conch_blob_prediction_map.png"
output_drift_background_map <- "outputs/conch_background_drift_exclusion_map.png"
output_background_density_plot <- "outputs/conch_background_density_by_habitat.png"

type_cols <- c(
  "Aggregated Patch Reef" = "#8dd3c7",
  "Artificial" = "#ffffb3",
  "Inner Reef" = "#bebada",
  "Middle Reef" = "#fb8072",
  "Nearshore Ridge Complex" = "#80b1d3",
  "Outer Reef" = "#fdb462",
  "Ridge" = "#b3de69",
  "Sand" = "#f2e8c9",
  "Scattered Rock in Unconsolidated Sediment" = "#bc80bd"
)
```

# Input Data

```{r import-data}
polygons_final <- readRDS(polygons_rds) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)

old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

radial_sites <- readxl::read_xlsx(conch_xlsx, sheet = radial_sheet)

radial_counts <- read.csv(
  radial_counts_csv,
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
) %>%
  filter(!is.na(`Site Code`), `Site Code` != "") %>%
  mutate(
    juvenile_count = as.numeric(`# Juv`),
    adult_count = as.numeric(`# Adult`),
    mating_count = as.numeric(ReproductiveActivity)
  )

radial_abundance <- radial_sites %>%
  select(`Site Code`, Lat, Lon, `Depth (ft)`) %>%
  left_join(
    radial_counts %>%
      select(`Site Code`, juvenile_count, adult_count, mating_count, `Habitat Description`),
    by = "Site Code"
  )

stopifnot(
  nrow(radial_abundance) == nrow(radial_sites),
  sum(is.na(radial_abundance$adult_count)) == 0,
  sum(is.na(radial_abundance$juvenile_count)) == 0
)

radial_abundance_sf_utm <- radial_abundance %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

mating_sites_sf_utm <- radial_abundance_sf_utm %>%
  filter(mating_count > 0)

radial_summary <- radial_abundance %>%
  summarize(
    radial_surveys = n(),
    adult_count = sum(adult_count),
    juvenile_count = sum(juvenile_count),
    mating_sites = sum(mating_count > 0),
    mating_observations = sum(mating_count)
  )

knitr::kable(radial_summary, digits = 2)
```

# Blob Definition

The blob is defined as the largest connected polygon produced by buffering radial survey sites by `r blob_buffer_m` m and dissolving overlapping buffers. This constrains population prediction to the radial-survey-supported cluster rather than the broader rectangular map extent.

```{r blob-definition}
radial_blob_candidates <- radial_abundance_sf_utm %>%
  st_buffer(blob_buffer_m) %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  st_cast("POLYGON") %>%
  mutate(
    blob_id = row_number(),
    area_m2 = as.numeric(st_area(geometry)),
    n_radial_sites = lengths(st_intersects(., radial_abundance_sf_utm))
  ) %>%
  arrange(desc(n_radial_sites), desc(area_m2))

radial_blob <- radial_blob_candidates %>%
  slice(1)

blob_grid <- st_make_grid(
  radial_blob,
  cellsize = prediction_cellsize_m,
  what = "polygons",
  square = TRUE
) %>%
  st_sf(geometry = .) %>%
  st_intersection(radial_blob) %>%
  mutate(cell_area_m2 = as.numeric(st_area(geometry))) %>%
  filter(cell_area_m2 > 0)

blob_grid_points <- st_point_on_surface(blob_grid)

blob_summary <- tibble(
  blob_buffer_m = blob_buffer_m,
  blob_area_m2 = sum(blob_grid$cell_area_m2),
  blob_area_ha = blob_area_m2 / 10000,
  radial_sites_in_blob = radial_blob$n_radial_sites
)

knitr::kable(blob_summary, digits = 2)
```

```{r blob-context-map, fig.width=8, fig.height=6}
blob_context_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, utm_crs),
    aes(fill = Type),
    color = NA,
    alpha = 0.45
  ) +
  geom_sf(data = radial_blob, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(
    data = radial_abundance_sf_utm,
    shape = 21,
    fill = "white",
    color = "black",
    size = 1.4,
    stroke = 0.25
  ) +
  geom_sf(
    data = mating_sites_sf_utm,
    shape = 24,
    fill = "#D7301F",
    color = "black",
    size = 3,
    stroke = 0.35
  ) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = st_bbox(radial_blob)[c("xmin", "xmax")] + c(-100, 100),
    ylim = st_bbox(radial_blob)[c("ymin", "ymax")] + c(-100, 100),
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)", fill = "Habitat type") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

blob_context_map

ggsave(output_blob_context_map, blob_context_map, width = 8, height = 6, dpi = 300)
```

# Drift Transect Segments

Drift transects are divided into `r drift_segment_length_m` m by `r drift_half_width_m * 2` m segments. Segmenting lets the portions that fall inside the blob contribute to blob density estimates, while portions outside the blob remain available for background density estimates.

```{r drift-segments}
make_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(st_polygon())
  }

  unit_direction <- direction / transect_length
  perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

  corners <- rbind(
    start_point + perpendicular,
    end_point + perpendicular,
    end_point - perpendicular,
    start_point - perpendicular,
    start_point + perpendicular
  )

  st_polygon(list(corners))
}

make_segment_rectangles <- function(start_point, end_point, segment_length_m, half_width_m) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(tibble())
  }

  unit_direction <- direction / transect_length
  segment_starts <- seq(0, transect_length, by = segment_length_m)
  segment_ends <- pmin(segment_starts + segment_length_m, transect_length)
  keep_segments <- segment_ends > segment_starts

  tibble(
    segment_index = seq_along(segment_starts)[keep_segments],
    segment_start_m = segment_starts[keep_segments],
    segment_end_m = segment_ends[keep_segments],
    geometry = map2(
      segment_start_m,
      segment_end_m,
      ~ make_transect_rectangle(
        start_point + unit_direction * .x,
        start_point + unit_direction * .y,
        half_width_m = half_width_m
      )
    )
  )
}

drift_surveys <- readxl::read_xlsx(conch_xlsx, sheet = drift_sheet) %>%
  mutate(
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon), !is.na(end_lat), !is.na(end_lon)) %>%
  mutate(
    .record = row_number(),
    transect_group = sub(" cont\\.$", "", Transect)
  )

drift_start_sf <- drift_surveys %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

drift_end_sf <- drift_surveys %>%
  st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

start_xy <- st_coordinates(drift_start_sf)
end_xy <- st_coordinates(drift_end_sf)

drift_rectangles <- st_sf(
  drift_surveys %>% select(-start_lat, -start_lon, -end_lat, -end_lon),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ make_transect_rectangle(start_xy[.x, ], end_xy[.y, ], half_width_m = drift_half_width_m)
    ),
    crs = utm_crs
  )
) %>%
  filter(!st_is_empty(.)) %>%
  mutate(transect_area_m2 = as.numeric(st_area(geometry)))

drift_segment_table <- map_dfr(seq_len(nrow(drift_surveys)), function(i) {
  make_segment_rectangles(
    start_xy[i, ],
    end_xy[i, ],
    segment_length_m = drift_segment_length_m,
    half_width_m = drift_half_width_m
  ) %>%
    mutate(
      .record = drift_surveys$.record[[i]],
      Transect = drift_surveys$Transect[[i]],
      transect_group = drift_surveys$transect_group[[i]],
      Line = drift_surveys$Line[[i]],
      segment_id = paste(.record, segment_index, sep = "_")
    )
})

drift_segments <- st_as_sf(
  drift_segment_table,
  sf_column_name = "geometry",
  crs = utm_crs
) %>%
  select(segment_id, .record, Transect, transect_group, Line, segment_index, segment_start_m, segment_end_m) %>%
  mutate(segment_area_m2 = as.numeric(st_area(geometry)))

blob_boundary <- radial_blob %>%
  select(blob_id)

drift_segments_blob <- st_intersection(
  drift_segments %>% st_make_valid(),
  blob_boundary %>% st_make_valid()
) %>%
  mutate(
    portion = "Blob",
    portion_id = paste0(segment_id, "_blob"),
    segment_area_m2 = as.numeric(st_area(geometry))
  ) %>%
  filter(segment_area_m2 > 0)

drift_segments_background <- st_difference(
  drift_segments %>% st_make_valid(),
  st_union(blob_boundary)
) %>%
  st_collection_extract("POLYGON") %>%
  mutate(
    portion = "Background",
    portion_id = paste0(segment_id, "_background"),
    segment_area_m2 = as.numeric(st_area(geometry))
  ) %>%
  filter(segment_area_m2 > 0)

drift_segment_summary <- tibble(
  drift_transects = nrow(drift_rectangles),
  drift_segments = nrow(drift_segments),
  full_drift_area_m2 = sum(drift_rectangles$transect_area_m2),
  blob_segment_area_m2 = sum(drift_segments_blob$segment_area_m2),
  background_segment_area_m2 = sum(drift_segments_background$segment_area_m2)
)

knitr::kable(drift_segment_summary, digits = 2)
```

```{r drift-position-counts}
conch_drift_positions <- readxl::read_xlsx(conch_xlsx, sheet = drift_position_sheet) %>%
  mutate(
    .record = row_number(),
    lat = as.numeric(`Lat (dec deg)`),
    lon = as.numeric(`Long (dec deg)`),
    animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`),
    animal_stage = case_when(
      grepl("^A", animal_id) ~ "Adult",
      grepl("^J", animal_id) ~ "Juvenile",
      TRUE ~ NA_character_
    ),
    transect_group = sub("([AJ][0-9]+)$", "", `WPT (B/E)`)
  ) %>%
  filter(!is.na(lat), !is.na(lon), !is.na(animal_stage))

conch_drift_positions_sf_utm <- conch_drift_positions %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

polygons_utm <- polygons_final %>%
  st_transform(utm_crs) %>%
  st_make_valid()

conch_drift_position_habitats <- st_join(
  conch_drift_positions_sf_utm,
  polygons_utm %>% select(Type),
  join = st_within,
  left = TRUE
) %>%
  st_drop_geometry() %>%
  select(.record, Habitat_Type = Type) %>%
  group_by(.record) %>%
  slice(1) %>%
  ungroup()

conch_drift_positions_out <- conch_drift_positions %>%
  left_join(conch_drift_position_habitats, by = ".record") %>%
  mutate(`Habitat (Class at point)` = Habitat_Type) %>%
  select(-Habitat_Type)

drift_segment_portions <- bind_rows(
  drift_segments_blob %>% select(portion_id, segment_id, transect_group, segment_index, portion, segment_area_m2),
  drift_segments_background %>% select(portion_id, segment_id, transect_group, segment_index, portion, segment_area_m2)
)

assign_to_nearest_segment <- function(points_sf, segments_sf) {
  map_dfr(seq_len(nrow(points_sf)), function(i) {
    point_i <- points_sf[i, ]
    point_dat <- point_i %>% st_drop_geometry()
    candidate_segments <- segments_sf %>%
      filter(transect_group == point_dat$transect_group[[1]])

    if (nrow(candidate_segments) == 0) {
      return(tibble(
        .record = point_dat$.record[[1]],
        animal_stage = point_dat$animal_stage[[1]],
        transect_group = point_dat$transect_group[[1]],
        portion_id = NA_character_,
        segment_id = NA_character_,
        segment_index = NA_integer_,
        portion = NA_character_,
        segment_area_m2 = NA_real_,
        distance_to_segment_m = NA_real_
      ))
    }

    distances_m <- as.numeric(st_distance(point_i, candidate_segments))
    nearest_i <- which.min(distances_m)
    nearest_segment <- candidate_segments[nearest_i, ] %>% st_drop_geometry()

    tibble(
      .record = point_dat$.record[[1]],
      animal_stage = point_dat$animal_stage[[1]],
      transect_group = point_dat$transect_group[[1]],
      portion_id = nearest_segment$portion_id[[1]],
      segment_id = nearest_segment$segment_id[[1]],
      segment_index = nearest_segment$segment_index[[1]],
      portion = nearest_segment$portion[[1]],
      segment_area_m2 = nearest_segment$segment_area_m2[[1]],
      distance_to_segment_m = distances_m[[nearest_i]]
    )
  })
}

drift_observation_segment_assignments <- assign_to_nearest_segment(
  conch_drift_positions_sf_utm,
  drift_segment_portions
) %>%
  left_join(
    conch_drift_positions_out %>%
      select(.record, Type = `Habitat (Class at point)`),
    by = ".record"
  )

blob_drift_segment_counts <- drift_observation_segment_assignments %>%
  filter(portion == "Blob") %>%
  count(portion_id, segment_id, transect_group, animal_stage, name = "n_animals")

background_drift_segment_counts <- drift_observation_segment_assignments %>%
  filter(portion == "Background") %>%
  count(
    portion_id,
    segment_id,
    transect_group,
    Type,
    animal_stage,
    name = "n_animals"
  )

drift_observation_assignment_summary <- tibble(
  observed_conch_total = nrow(conch_drift_positions),
  assigned_to_segment = sum(!is.na(drift_observation_segment_assignments$segment_id)),
  assigned_to_blob_segment = sum(drift_observation_segment_assignments$portion == "Blob", na.rm = TRUE),
  assigned_to_background_segment = sum(drift_observation_segment_assignments$portion == "Background", na.rm = TRUE),
  unassigned_to_segment = observed_conch_total - assigned_to_segment,
  median_distance_to_assigned_segment_m = median(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  ),
  max_distance_to_assigned_segment_m = max(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  )
)

drift_count_centroids <- bind_rows(
  blob_drift_segment_counts %>%
    mutate(portion = "Blob"),
  background_drift_segment_counts %>%
    mutate(portion = "Background")
) %>%
  group_by(portion, portion_id, segment_id, animal_stage) %>%
  summarize(n_animals = sum(n_animals), .groups = "drop") %>%
  left_join(
    bind_rows(
      drift_segments_blob %>%
        st_drop_geometry() %>%
        select(portion_id, segment_id, portion, segment_area_m2),
      drift_segments_background %>%
        st_drop_geometry() %>%
        select(portion_id, segment_id, portion, segment_area_m2)
    ),
    by = c("portion", "portion_id", "segment_id")
  )

drift_count_centroids_sf <- bind_rows(
  drift_segments_blob %>% select(portion_id, segment_id, portion),
  drift_segments_background %>% select(portion_id, segment_id, portion)
) %>%
  inner_join(drift_count_centroids, by = c("portion", "portion_id", "segment_id")) %>%
  st_point_on_surface() %>%
  mutate(
    age_class = animal_stage,
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  )

knitr::kable(drift_observation_assignment_summary, digits = 2)
```

# Adult Blob Model

Adult abundance is estimated using the best of three mating-related negative-binomial count models. The model includes radial surveys and drift segment portions that intersect the blob, with each observation's surveyed area as an offset.

```{r adult-models}
blob_drift_observations_sf_utm <- drift_segments_blob %>%
  st_point_on_surface() %>%
  left_join(
    blob_drift_segment_counts %>%
      group_by(portion_id, animal_stage) %>%
      summarize(n_animals = sum(n_animals), .groups = "drop") %>%
      pivot_wider(
        names_from = animal_stage,
        values_from = n_animals,
        values_fill = 0
      ),
    by = "portion_id"
  ) %>%
  mutate(
    Adult = replace_na(Adult, 0),
    Juvenile = replace_na(Juvenile, 0),
    observation_type = "Drift segment",
    survey_area_m2 = segment_area_m2,
    adult_count = Adult,
    juvenile_count = Juvenile
  ) %>%
  select(observation_type, survey_area_m2, adult_count, juvenile_count, geometry)

radial_observations_sf_utm <- radial_abundance_sf_utm %>%
  mutate(
    observation_type = "Radial survey",
    survey_area_m2 = radial_survey_area_m2,
    adult_count = radial_abundance$adult_count,
    juvenile_count = radial_abundance$juvenile_count
  ) %>%
  select(observation_type, survey_area_m2, adult_count, juvenile_count, geometry)

blob_observations_sf_utm <- bind_rows(
  radial_observations_sf_utm,
  blob_drift_observations_sf_utm
)

distance_matrix_m <- st_distance(blob_observations_sf_utm, mating_sites_sf_utm) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(blob_observations_sf_utm), ncol = nrow(mating_sites_sf_utm))

mating_weights <- mating_sites_sf_utm$mating_count

calculate_mating_halo <- function(dist_matrix, weights, scale_m) {
  as.numeric(exp(-dist_matrix / scale_m) %*% weights)
}

calculate_mating_within_radius <- function(dist_matrix, weights, radius_m) {
  as.numeric((dist_matrix <= radius_m) %*% weights)
}

adult_model_data <- blob_observations_sf_utm %>%
  st_drop_geometry() %>%
  mutate(
    distance_nearest_mating_m = as.numeric(apply(distance_matrix_m, 1, min))
  )

adult_distance_model <- MASS::glm.nb(
  adult_count ~ distance_nearest_mating_m + offset(log(survey_area_m2)),
  data = adult_model_data,
  control = glm.control(maxit = 100)
)

halo_scales_m <- seq(25, 500, by = 25)

adult_halo_model_selection <- tibble(scale_m = halo_scales_m) %>%
  mutate(
    model = map(scale_m, function(scale_m) {
      dat <- adult_model_data %>%
        mutate(mating_halo = calculate_mating_halo(distance_matrix_m, mating_weights, scale_m))

      MASS::glm.nb(
        adult_count ~ mating_halo + offset(log(survey_area_m2)),
        data = dat,
        control = glm.control(maxit = 100)
      )
    }),
    AIC = map_dbl(model, AIC)
  ) %>%
  arrange(AIC)

best_halo_scale_m <- adult_halo_model_selection$scale_m[[1]]
adult_halo_model <- adult_halo_model_selection$model[[1]]

mating_radii_m <- seq(25, 500, by = 25)

adult_radius_model_selection <- tibble(radius_m = mating_radii_m) %>%
  mutate(
    model = map(radius_m, function(radius_m) {
      dat <- adult_model_data %>%
        mutate(
          mating_within_radius = calculate_mating_within_radius(
            distance_matrix_m,
            mating_weights,
            radius_m
          )
        )

      MASS::glm.nb(
        adult_count ~ mating_within_radius + offset(log(survey_area_m2)),
        data = dat,
        control = glm.control(maxit = 100)
      )
    }),
    AIC = map_dbl(model, AIC)
  ) %>%
  arrange(AIC)

best_mating_radius_m <- adult_radius_model_selection$radius_m[[1]]
adult_radius_model <- adult_radius_model_selection$model[[1]]

adult_model_comparison <- tibble(
  model = c(
    "nearest mating distance",
    paste0("additive mating halo, scale ", best_halo_scale_m, " m"),
    paste0("mating observations within ", best_mating_radius_m, " m")
  ),
  AIC = c(
    AIC(adult_distance_model),
    AIC(adult_halo_model),
    AIC(adult_radius_model)
  )
) %>%
  mutate(delta_AIC = AIC - min(AIC)) %>%
  arrange(AIC)

knitr::kable(adult_model_comparison, digits = 2)
```

```{r adult-model-plot, fig.width=8, fig.height=5}
adult_radius_data <- adult_model_data %>%
  mutate(
    mating_within_radius = calculate_mating_within_radius(
      distance_matrix_m,
      mating_weights,
      best_mating_radius_m
    )
  )

adult_radius_prediction <- tibble(
  mating_within_radius = seq(0, max(adult_radius_data$mating_within_radius), by = 1),
  survey_area_m2 = radial_survey_area_m2
)

adult_radius_prediction_link <- predict(
  adult_radius_model,
  newdata = adult_radius_prediction,
  type = "link",
  se.fit = TRUE
)

adult_radius_prediction <- adult_radius_prediction %>%
  mutate(
    density_100m2 = exp(adult_radius_prediction_link$fit) / radial_survey_area_m2 * 100,
    density_lwr_100m2 = exp(adult_radius_prediction_link$fit - 1.96 * adult_radius_prediction_link$se.fit) /
      radial_survey_area_m2 * 100,
    density_upr_100m2 = exp(adult_radius_prediction_link$fit + 1.96 * adult_radius_prediction_link$se.fit) /
      radial_survey_area_m2 * 100
  )

adult_model_comparison_plot <- ggplot() +
  geom_point(
    data = adult_radius_data,
    aes(x = mating_within_radius, y = adult_count / survey_area_m2 * 100, shape = observation_type),
    fill = "white",
    color = "black",
    alpha = 0.8
  ) +
  geom_ribbon(
    data = adult_radius_prediction,
    aes(x = mating_within_radius, ymin = density_lwr_100m2, ymax = density_upr_100m2),
    fill = "#7B3294",
    alpha = 0.2
  ) +
  geom_line(
    data = adult_radius_prediction,
    aes(x = mating_within_radius, y = density_100m2),
    color = "#7B3294",
    linewidth = 1
  ) +
  labs(
    x = paste0("Mating observations within ", best_mating_radius_m, " m"),
    y = expression("Adult density (adults per 100 " * m^2 * ")"),
    shape = "Observation"
  ) +
  theme_minimal(base_size = 11)

adult_model_comparison_plot

ggsave(output_adult_model_comparison_plot, adult_model_comparison_plot, width = 8, height = 5, dpi = 300)
```

# Blob Population Estimates

Adults are predicted with the selected mating-radius model. Juveniles are estimated in the same blob area by IDW interpolation of juvenile density from radial surveys and blob-intersecting drift segments, with no mating predictor.

```{r blob-prediction}
blob_to_mating_dist_m <- st_distance(blob_grid_points, mating_sites_sf_utm) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(blob_grid_points), ncol = nrow(mating_sites_sf_utm))

blob_adult_prediction <- blob_grid %>%
  mutate(
    mating_within_radius = calculate_mating_within_radius(
      blob_to_mating_dist_m,
      mating_weights,
      best_mating_radius_m
    ),
    survey_area_m2 = 1,
    adult_density_m2 = predict(
      adult_radius_model,
      newdata = tibble(mating_within_radius = mating_within_radius, survey_area_m2 = survey_area_m2),
      type = "response"
    ),
    adult_total = adult_density_m2 * cell_area_m2
  )

calculate_idw_prediction <- function(dist_matrix, values, power = 2) {
  apply(dist_matrix, 1, function(d) {
    if (any(d == 0)) {
      mean(values[d == 0], na.rm = TRUE)
    } else {
      weights <- 1 / d^power
      sum(weights * values, na.rm = TRUE) / sum(weights, na.rm = TRUE)
    }
  })
}

juvenile_observations_sf_utm <- blob_observations_sf_utm %>%
  mutate(juvenile_density_m2 = juvenile_count / survey_area_m2)

blob_to_juvenile_dist_m <- st_distance(blob_grid_points, juvenile_observations_sf_utm) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(blob_grid_points), ncol = nrow(juvenile_observations_sf_utm))

blob_juvenile_prediction <- blob_grid %>%
  mutate(
    juvenile_density_m2 = calculate_idw_prediction(
      blob_to_juvenile_dist_m,
      juvenile_observations_sf_utm$juvenile_density_m2,
      power = 2
    ),
    juvenile_total = juvenile_density_m2 * cell_area_m2
  )

blob_population_summary <- tibble(
  blob_area_ha = sum(blob_grid$cell_area_m2) / 10000,
  estimated_adults = sum(blob_adult_prediction$adult_total, na.rm = TRUE),
  estimated_juveniles = sum(blob_juvenile_prediction$juvenile_total, na.rm = TRUE),
  adult_density_per_ha = estimated_adults / blob_area_ha,
  juvenile_density_per_ha = estimated_juveniles / blob_area_ha
)

knitr::kable(blob_population_summary, digits = 2)
```

```{r blob-prediction-map, fig.width=8, fig.height=6}
blob_prediction_long <- bind_rows(
  blob_adult_prediction %>%
    mutate(
      age_class = "Adult",
      density_100m2 = adult_density_m2 * 100,
      total_animals = adult_total
    ) %>%
    select(age_class, density_100m2, total_animals, geometry),
  blob_juvenile_prediction %>%
    mutate(
      age_class = "Juvenile",
      density_100m2 = juvenile_density_m2 * 100,
      total_animals = juvenile_total
    ) %>%
    select(age_class, density_100m2, total_animals, geometry)
)

blob_prediction_points <- blob_prediction_long %>%
  st_point_on_surface() %>%
  mutate(
    x = st_coordinates(.)[, "X"],
    y = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()

blob_prediction_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, utm_crs),
    fill = "grey85",
    color = "grey70",
    linewidth = 0.05,
    alpha = 0.35
  ) +
  geom_tile(
    data = blob_prediction_points,
    aes(x = x, y = y, fill = density_100m2),
    width = prediction_cellsize_m,
    height = prediction_cellsize_m,
    alpha = 0.9
  ) +
  geom_sf(data = radial_blob, fill = NA, color = "black", linewidth = 0.6) +
  geom_sf(
    data = drift_segments_blob,
    fill = NA,
    color = "#2B8CBE",
    linewidth = 0.25,
    alpha = 0.6
  ) +
  geom_sf(
    data = radial_abundance_sf_utm,
    shape = 21,
    fill = "white",
    color = "black",
    size = 0.9,
    stroke = 0.2,
    alpha = 0.7
  ) +
  geom_point(
    data = drift_count_centroids_sf %>%
      filter(portion == "Blob") %>%
      st_drop_geometry(),
    aes(x = x, y = y, size = n_animals),
    shape = 21,
    fill = "#FED976",
    color = "black",
    stroke = 0.25,
    alpha = 0.9
  ) +
  facet_wrap(~ age_class) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, max(blob_prediction_points$density_100m2, na.rm = TRUE)),
    trans = "sqrt",
    name = expression("Conch per 100 " * m^2)
  ) +
  scale_size_continuous(name = "Observed drift conch in segment", range = c(1.5, 5)) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = st_bbox(radial_blob)[c("xmin", "xmax")],
    ylim = st_bbox(radial_blob)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

blob_prediction_map

ggsave(output_blob_prediction_map, blob_prediction_map, width = 8, height = 6, dpi = 300)
```

# Drift Transect Background Estimates

Background densities are estimated from the portions of drift transects that fall outside the blob. This keeps the blob/background split spatial rather than excluding an entire transect whenever any part touches the blob.

```{r background-density}
background_drift_habitat_area <- st_intersection(
  drift_segments_background %>% st_make_valid(),
  polygons_utm %>% select(Type)
) %>%
  mutate(area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  group_by(portion_id, segment_id, transect_group, Type) %>%
  summarize(area_m2 = sum(area_m2), .groups = "drop")

background_drift_density <- background_drift_habitat_area %>%
  expand_grid(animal_stage = c("Adult", "Juvenile")) %>%
  left_join(
    background_drift_segment_counts,
    by = c("portion_id", "segment_id", "transect_group", "Type", "animal_stage")
  ) %>%
  mutate(n_animals = replace_na(n_animals, 0L)) %>%
  group_by(Type, animal_stage) %>%
  summarize(
    n_animals = sum(n_animals),
    area_m2 = sum(area_m2),
    density_m2 = n_animals / area_m2,
    density_100m2 = density_m2 * 100,
    .groups = "drop"
  ) %>%
  arrange(Type, animal_stage)

background_summary <- tibble(
  drift_transects_total = nrow(drift_rectangles),
  drift_segments_total = nrow(drift_segments),
  blob_segment_portions = nrow(drift_segments_blob),
  background_segment_portions = nrow(drift_segments_background),
  blob_drift_area_m2 = sum(drift_segments_blob$segment_area_m2),
  background_drift_area_m2 = sum(drift_segments_background$segment_area_m2)
)

knitr::kable(background_summary, digits = 2)
knitr::kable(background_drift_density, digits = 4)
```

```{r drift-background-map, fig.width=8, fig.height=6}
drift_background_map <- ggplot() +
  geom_sf(
    data = st_transform(polygons_final, utm_crs),
    aes(fill = Type),
    color = NA,
    alpha = 0.45
  ) +
  geom_sf(data = radial_blob, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(
    data = drift_segments_background,
    aes(color = "Background portion"),
    fill = NA,
    linewidth = 0.35,
    alpha = 0.8
  ) +
  geom_sf(
    data = drift_segments_blob,
    aes(color = "Blob portion"),
    fill = NA,
    linewidth = 0.45,
    alpha = 0.9
  ) +
  geom_point(
    data = drift_count_centroids_sf %>% st_drop_geometry(),
    aes(x = x, y = y, size = n_animals, shape = animal_stage),
    fill = "#FED976",
    color = "black",
    stroke = 0.25,
    alpha = 0.9
  ) +
  scale_fill_manual(values = type_cols, na.value = "grey80") +
  scale_color_manual(values = c("Background portion" = "black", "Blob portion" = "#D7301F")) +
  scale_shape_manual(name = "Observed conch stage", values = c("Adult" = 21, "Juvenile" = 24)) +
  scale_size_continuous(name = "Observed conch in segment", range = c(1.5, 5)) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = st_bbox(drift_rectangles)[c("xmin", "xmax")] + c(-200, 200),
    ylim = st_bbox(drift_rectangles)[c("ymin", "ymax")] + c(-200, 200),
    expand = FALSE
  ) +
  labs(x = "Easting (m)", y = "Northing (m)", fill = "Habitat type", color = "Drift transects") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

drift_background_map

ggsave(output_drift_background_map, drift_background_map, width = 8, height = 6, dpi = 300)
```

```{r background-density-plot, fig.width=8, fig.height=5}
background_density_plot <- ggplot(
  background_drift_density,
  aes(x = Type, y = density_100m2, fill = animal_stage)
) +
  geom_col(position = position_dodge(width = 0.75), color = "grey30", linewidth = 0.2) +
  coord_flip() +
  labs(
    x = "Habitat type",
    y = expression("Background density (conch per 100 " * m^2 * ")"),
    fill = "Conch stage"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

background_density_plot

ggsave(output_background_density_plot, background_density_plot, width = 8, height = 5, dpi = 300)
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
