---
title: "PENIP Conch IDW Abundance Estimates"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

find_project_root <- function(start = getwd()) {
  start <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- c(start, normalizePath(file.path(start, ".."), winslash = "/", mustWork = TRUE))

  while (length(candidates) > 0) {
    candidate <- candidates[[1]]
    if (file.exists(file.path(candidate, "data/processed/polygons_final.rds"))) {
      return(candidate)
    }

    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not find the PENIP project root from ", start, call. = FALSE)
    }

    candidates <- c(candidates[-1], parent)
  }
}

knitr::opts_knit$set(root.dir = find_project_root())
```

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(ggnewscale)
library(purrr)
library(readxl)
library(sf)
library(stringr)
library(tibble)
library(tidyr)
library(xml2)
```

```{r paths-and-constants}
conch_xlsx <- "data/conch/PENIP_QueenConchSurveys.xlsx"
radial_counts_csv <- "data/conch/PENIP_circular_counts.csv"
berry_broad_xlsx <- "data/conch/Berry_Conch_per_site.xlsx"
polygons_rds <- "data/processed/polygons_final.rds"
impact_kml <- "data/spatial/Impact_zones_Scenario 2.kml"
nmfs_kml <- "data/spatial/Scenario2_NMFSConsiderationAreas.kml"

radial_sheet <- "PENIP_Conch_10mRadial"
drift_sheet <- "PENIP_DriftTransects"
drift_position_sheet <- "PENIP_ConchDriftPosition"

utm_crs <- 32617
radial_survey_radius_m <- 10
radial_survey_area_m2 <- pi * radial_survey_radius_m^2
drift_half_width_m <- 2
drift_segment_length_m <- 20
berry_broad_site_area_m2 <- 2 * 30 * 4
idw_cellsize_m <- 25
breeding_activity_radius_m <- 330

output_data_collection_map <- "outputs/conch_data_collection_overview_map.png"
output_idw_map <- "outputs/conch_idw_density_map.png"
output_adult_breeding_radius_idw_map <- "outputs/conch_adult_breeding_radius_idw_map.png"

survey_effort_xlim <- c(-80.1116, -80.078)
survey_effort_ylim <- c(26.058, 26.112)
```

# Input Data

```{r import-data}
polygons_final <- readRDS(polygons_rds) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)

type_cols <- RColorBrewer::brewer.pal(nlevels(factor(polygons_final$Type)) + 1, "Set3")[-1]
names(type_cols) <- levels(factor(polygons_final$Type))

old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

polygons_utm <- polygons_final %>%
  st_transform(utm_crs) %>%
  st_make_valid()

polygons_conch_habitat_utm <- polygons_utm %>%
  mutate(
    conch_habitat = case_when(
      Type %in% c("Artificial", "Nearshore Ridge Complex") ~ "Nearshore Ridge Complex",
      Type == "Inner Reef" ~ "Inner Reef",
      TRUE ~ NA_character_
    )
  )

target_habitats_utm <- polygons_conch_habitat_utm %>%
  filter(!is.na(conch_habitat)) %>%
  group_by(conch_habitat) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

extract_placemark_info <- function(kml_file, source_label) {
  ns <- c(kml = "http://www.opengis.net/kml/2.2")
  kml <- read_xml(kml_file)
  placemarks <- xml_find_all(kml, ".//kml:Placemark", ns)

  map(placemarks, function(pm) {
    name <- xml_text(xml_find_first(pm, ".//kml:name", ns))
    folder_node <- xml_parent(xml_parent(pm))
    folder_name <- xml_text(xml_find_first(folder_node, ".//kml:name", ns))
    if (is.na(folder_name) || folder_name == "") folder_name <- "Top Level"

    coord_nodes <- xml_find_all(pm, ".//kml:coordinates", ns)
    rings <- map(coord_nodes, function(cn) {
      coords <- strsplit(trimws(xml_text(cn)), "\\s+")[[1]]
      coords_mat <- do.call(rbind, lapply(coords, function(x) {
        as.numeric(strsplit(x, ",")[[1]][1:2])
      }))
      if (nrow(coords_mat) >= 3) coords_mat else NULL
    }) %>%
      compact()

    if (length(rings) == 0) return(NULL)

    tibble(
      folder = folder_name,
      name = name,
      geometry = st_sfc(st_multipolygon(lapply(rings, function(r) list(r))), crs = 4326),
      source = source_label
    )
  }) %>%
    compact() %>%
    bind_rows() %>%
    st_as_sf()
}

impact_zones_sf <- extract_placemark_info(impact_kml, "Impact Zone")
nmfs_sf <- extract_placemark_info(nmfs_kml, "NMFS Consideration Area")

impact_zones_utm <- bind_rows(impact_zones_sf, nmfs_sf) %>%
  mutate(
    name = if_else(
      is.na(name) | name == "",
      str_remove(str_remove(folder, "^Scenario_2_"), "\\.shp$"),
      name
    ),
    scenario = if_else(
      name %in% c("North", "South", "East", "Nearshore"),
      "NMFS Consideration",
      "Scenario 2"
    )
  ) %>%
  st_make_valid() %>%
  st_transform(utm_crs) %>%
  group_by(name, scenario, source) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

channel_poly_utm <- impact_zones_utm %>%
  filter(str_detect(tolower(as.character(name)), "channel"))

channel_bbox <- st_bbox(channel_poly_utm)
habitat_bbox <- st_bbox(polygons_utm)

impact_monitoring_box <- st_as_sfc(st_bbox(
  c(
    xmin = habitat_bbox[["xmin"]],
    xmax = habitat_bbox[["xmax"]],
    ymin = channel_bbox[["ymin"]] - 1200,
    ymax = channel_bbox[["ymax"]] + 1200
  ),
  crs = st_crs(polygons_utm)
))

impact_monitoring_habitat_utm <- st_intersection(
  impact_monitoring_box,
  st_union(polygons_utm)
) %>%
  st_union() %>%
  st_make_valid()

existing_impact_zone_hull_utm <- impact_zones_utm %>%
  st_union() %>%
  st_buffer(20) %>%
  st_union() %>%
  st_buffer(-20) %>%
  st_make_valid()

impact_monitoring_area_utm <- st_sf(
  name = "Impact Monitoring Area",
  scenario = "Impact Monitoring Area",
  source = "Generated",
  geometry = st_sfc(
    st_difference(impact_monitoring_habitat_utm, existing_impact_zone_hull_utm),
    crs = st_crs(polygons_utm)
  )
) %>%
  st_make_valid()

impact_zones_utm <- bind_rows(
  impact_zones_utm,
  impact_monitoring_area_utm
)

impact_zone_area_utm <- impact_zones_utm %>%
  st_union() %>%
  st_make_valid()

make_exterior_lines <- function(x) {
  polygon_parts <- x %>%
    st_collection_extract("POLYGON") %>%
    st_cast("POLYGON", warn = FALSE)

  st_sf(
    name = polygon_parts$name,
    scenario = polygon_parts$scenario,
    source = polygon_parts$source,
    geometry = st_sfc(
      map(st_geometry(polygon_parts), ~ st_linestring(.x[[1]])),
      crs = st_crs(polygon_parts)
    )
  )
}

impact_zone_plot_outlines_utm <- bind_rows(
  impact_zones_utm %>%
    filter(name != "Impact Monitoring Area") %>%
    st_boundary() %>%
    st_as_sf(),
  impact_monitoring_area_utm %>%
    make_exterior_lines()
) %>%
  st_make_valid()

radial_sites <- readxl::read_xlsx(conch_xlsx, sheet = radial_sheet)

radial_counts <- read.csv(
  radial_counts_csv,
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
) %>%
  filter(!is.na(`Site Code`), `Site Code` != "") %>%
  mutate(
    juvenile_count = as.numeric(`# Juv`),
    adult_count = as.numeric(`# Adult`),
    breeding_activity_count = as.numeric(ReproductiveActivity)
  )

radial_abundance <- radial_sites %>%
  select(`Site Code`, Lat, Lon, `Depth (ft)`) %>%
  left_join(
    radial_counts %>%
      select(`Site Code`, juvenile_count, adult_count, breeding_activity_count),
    by = "Site Code"
  )

stopifnot(
  nrow(radial_abundance) == nrow(radial_sites),
  sum(is.na(radial_abundance$adult_count)) == 0,
  sum(is.na(radial_abundance$juvenile_count)) == 0
)

radial_observations_sf_utm <- radial_abundance %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  mutate(
    observation_source = "PENIP radial survey",
    survey_area_m2 = radial_survey_area_m2
  ) %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

breeding_activity_sites_sf_utm <- radial_abundance %>%
  filter(breeding_activity_count > 0) %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

radial_summary <- radial_observations_sf_utm %>%
  st_drop_geometry() %>%
  summarize(
    radial_surveys = n(),
    radial_area_m2 = sum(survey_area_m2),
    adult_count = sum(adult_count),
    juvenile_count = sum(juvenile_count)
  )

knitr::kable(radial_summary, digits = 2)
```

# Drift Transects

Drift transects are converted to 4 m-wide rectangles and then divided into `r drift_segment_length_m` m segments. Observed conch from the drift survey point sheet are assigned to the nearest segment within the same transect.

```{r drift-segments}
make_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(st_polygon())
  }

  unit_direction <- direction / transect_length
  perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

  corners <- rbind(
    start_point + perpendicular,
    end_point + perpendicular,
    end_point - perpendicular,
    start_point - perpendicular,
    start_point + perpendicular
  )

  st_polygon(list(corners))
}

make_segment_rectangles <- function(start_point, end_point, segment_length_m, half_width_m) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(tibble())
  }

  unit_direction <- direction / transect_length
  segment_starts <- seq(0, transect_length, by = segment_length_m)
  segment_ends <- pmin(segment_starts + segment_length_m, transect_length)
  keep_segments <- segment_ends > segment_starts

  tibble(
    segment_index = seq_along(segment_starts)[keep_segments],
    segment_start_m = segment_starts[keep_segments],
    segment_end_m = segment_ends[keep_segments],
    geometry = map2(
      segment_start_m,
      segment_end_m,
      ~ make_transect_rectangle(
        start_point + unit_direction * .x,
        start_point + unit_direction * .y,
        half_width_m = half_width_m
      )
    )
  )
}

drift_surveys <- readxl::read_xlsx(conch_xlsx, sheet = drift_sheet) %>%
  mutate(
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon), !is.na(end_lat), !is.na(end_lon)) %>%
  mutate(
    .record = row_number(),
    transect_group = sub(" cont\\.$", "", Transect)
  )

drift_start_sf <- drift_surveys %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

drift_end_sf <- drift_surveys %>%
  st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

start_xy <- st_coordinates(drift_start_sf)
end_xy <- st_coordinates(drift_end_sf)

drift_rectangles <- st_sf(
  drift_surveys %>% select(-start_lat, -start_lon, -end_lat, -end_lon),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ make_transect_rectangle(start_xy[.x, ], end_xy[.y, ], half_width_m = drift_half_width_m)
    ),
    crs = utm_crs
  )
) %>%
  filter(!st_is_empty(.)) %>%
  mutate(transect_area_m2 = as.numeric(st_area(geometry)))

drift_transect_lines <- st_sf(
  drift_surveys %>% select(.record, Transect, transect_group, Line),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ st_linestring(rbind(start_xy[.x, ], end_xy[.y, ]))
    ),
    crs = utm_crs
  )
)

drift_segment_table <- map_dfr(seq_len(nrow(drift_surveys)), function(i) {
  make_segment_rectangles(
    start_xy[i, ],
    end_xy[i, ],
    segment_length_m = drift_segment_length_m,
    half_width_m = drift_half_width_m
  ) %>%
    mutate(
      .record = drift_surveys$.record[[i]],
      Transect = drift_surveys$Transect[[i]],
      transect_group = drift_surveys$transect_group[[i]],
      Line = drift_surveys$Line[[i]],
      segment_id = paste(.record, segment_index, sep = "_")
    )
})

drift_segments <- st_as_sf(
  drift_segment_table,
  sf_column_name = "geometry",
  crs = utm_crs
) %>%
  select(segment_id, .record, Transect, transect_group, Line, segment_index, segment_start_m, segment_end_m) %>%
  mutate(segment_area_m2 = as.numeric(st_area(geometry)))

drift_segment_summary <- tibble(
  drift_transects = nrow(drift_rectangles),
  drift_segment_count = nrow(drift_segments),
  drift_area_m2 = sum(drift_rectangles$transect_area_m2),
  drift_segment_area_m2 = sum(drift_segments$segment_area_m2)
)

knitr::kable(drift_segment_summary, digits = 2)
```

```{r drift-position-counts}
conch_drift_positions <- readxl::read_xlsx(conch_xlsx, sheet = drift_position_sheet) %>%
  mutate(
    .record = row_number(),
    lat = as.numeric(`Lat (dec deg)`),
    lon = as.numeric(`Long (dec deg)`),
    animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`),
    animal_stage = case_when(
      grepl("^A", animal_id) ~ "Adult",
      grepl("^J", animal_id) ~ "Juvenile",
      TRUE ~ NA_character_
    ),
    transect_group = sub("([AJ][0-9]+)$", "", `WPT (B/E)`)
  ) %>%
  filter(!is.na(lat), !is.na(lon), !is.na(animal_stage))

conch_drift_positions_sf_utm <- conch_drift_positions %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

assign_to_nearest_segment <- function(points_sf, segments_sf) {
  map_dfr(seq_len(nrow(points_sf)), function(i) {
    point_i <- points_sf[i, ]
    point_dat <- point_i %>% st_drop_geometry()
    candidate_segments <- segments_sf %>%
      filter(transect_group == point_dat$transect_group[[1]])

    if (nrow(candidate_segments) == 0) {
      return(tibble(
        .record = point_dat$.record[[1]],
        animal_stage = point_dat$animal_stage[[1]],
        transect_group = point_dat$transect_group[[1]],
        segment_id = NA_character_,
        segment_index = NA_integer_,
        segment_area_m2 = NA_real_,
        distance_to_segment_m = NA_real_
      ))
    }

    distances_m <- as.numeric(st_distance(point_i, candidate_segments))
    nearest_i <- which.min(distances_m)
    nearest_segment <- candidate_segments[nearest_i, ] %>% st_drop_geometry()

    tibble(
      .record = point_dat$.record[[1]],
      animal_stage = point_dat$animal_stage[[1]],
      transect_group = point_dat$transect_group[[1]],
      segment_id = nearest_segment$segment_id[[1]],
      segment_index = nearest_segment$segment_index[[1]],
      segment_area_m2 = nearest_segment$segment_area_m2[[1]],
      distance_to_segment_m = distances_m[[nearest_i]]
    )
  })
}

drift_observation_segment_assignments <- assign_to_nearest_segment(
  conch_drift_positions_sf_utm,
  drift_segments
)

drift_segment_counts <- drift_observation_segment_assignments %>%
  filter(!is.na(segment_id)) %>%
  count(segment_id, animal_stage, name = "n_animals") %>%
  pivot_wider(
    names_from = animal_stage,
    values_from = n_animals,
    values_fill = 0
  )

drift_segment_observations_sf_utm <- drift_segments %>%
  st_point_on_surface() %>%
  left_join(drift_segment_counts, by = "segment_id") %>%
  mutate(
    Adult = replace_na(Adult, 0),
    Juvenile = replace_na(Juvenile, 0),
    observation_source = "PENIP drift segment",
    survey_area_m2 = segment_area_m2,
    adult_count = Adult,
    juvenile_count = Juvenile
  ) %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

drift_observation_assignment_summary <- tibble(
  observed_conch_total = nrow(conch_drift_positions),
  assigned_to_segment = sum(!is.na(drift_observation_segment_assignments$segment_id)),
  unassigned_to_segment = observed_conch_total - assigned_to_segment,
  median_distance_to_assigned_segment_m = median(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  ),
  max_distance_to_assigned_segment_m = max(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  )
)

knitr::kable(drift_observation_assignment_summary, digits = 2)
```

# Broad-Scale Survey Counts

The Berry broad-scale survey counts are classified as adult = `Adult + Old` and juvenile/non-adult = `Juv + Sub + Young`. Each broad-scale site is treated as 240 m2 of surveyed area, from two 30 m transects surveyed 2 m on each side of the tape.

```{r berry-broad-data}
berry_broad_raw <- readxl::read_xlsx(berry_broad_xlsx, sheet = 1)

berry_broad_sites <- berry_broad_raw %>%
  mutate(
    .record = row_number(),
    adult_count = as.numeric(Adult) + as.numeric(Old),
    juvenile_count = as.numeric(Juv) + as.numeric(Sub) + as.numeric(Young),
    total_count = adult_count + juvenile_count,
    survey_area_m2 = berry_broad_site_area_m2
  ) %>%
  select(.record, Site, LatDD, LonDD, Adult, Juv, Old, Sub, Young, adult_count, juvenile_count, total_count, survey_area_m2)

berry_observations_sf_utm <- berry_broad_sites %>%
  st_as_sf(coords = c("LonDD", "LatDD"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  mutate(observation_source = "Berry broad-scale site") %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

berry_broad_summary <- berry_broad_sites %>%
  summarize(
    berry_sites = n(),
    berry_area_m2 = sum(survey_area_m2, na.rm = TRUE),
    berry_adults = sum(adult_count, na.rm = TRUE),
    berry_juveniles = sum(juvenile_count, na.rm = TRUE),
    berry_total = sum(total_count, na.rm = TRUE)
  )

knitr::kable(berry_broad_summary, digits = 2)
```

# Data Collection Overview

This figure shows the spatial distribution of the three conch survey inputs using the same extent, habitat colors, and survey-point style as Figure 1B in the coral report.

```{r data-collection-overview-map, fig.width=7, fig.height=3.5}
survey_map_theme <- theme_minimal(base_size = 8) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.size = unit(0.3, "cm"),
    legend.text = element_text(size = 6),
    legend.title = element_blank(),
    strip.text = element_text(size = 6),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(2, 0, 0, 0),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

survey_effort_levels <- c(
  "Berry broad-scale survey",
  "PENIP radial survey",
  "PENIP drift survey"
)

survey_effort_polygons <- crossing(survey_effort = survey_effort_levels) %>%
  pull(survey_effort) %>%
  map_dfr(function(panel_name) {
    polygons_final %>%
      mutate(survey_effort = panel_name)
  }) %>%
  st_as_sf() %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels))

berry_effort_points <- berry_observations_sf_utm %>%
  st_transform(4326) %>%
  mutate(survey_effort = "Berry broad-scale survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

radial_effort_points <- radial_observations_sf_utm %>%
  st_transform(4326) %>%
  mutate(survey_effort = "PENIP radial survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

drift_effort_lines <- drift_transect_lines %>%
  st_transform(4326) %>%
  mutate(survey_effort = "PENIP drift survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

data_collection_map <- ggplot() +
  geom_sf(
    data = survey_effort_polygons,
    aes(fill = Type),
    color = NA,
    linewidth = 0.1,
    alpha = 0.5
  ) +
  geom_sf(
    data = drift_effort_lines,
    color = "black",
    linewidth = 0.25,
    alpha = 0.7
  ) +
  geom_sf(
    data = berry_effort_points,
    pch = 4,
    alpha = 0.6,
    size = 0.65,
    stroke = 0.3
  ) +
  geom_sf(
    data = radial_effort_points,
    pch = 4,
    alpha = 0.6,
    size = 0.65,
    stroke = 0.3
  ) +
  scale_fill_manual(values = type_cols, na.value = "gray80") +
  facet_wrap(
    ~ survey_effort,
    ncol = 3,
    labeller = as_labeller(c(
      "Berry broad-scale survey" = "Berry Broad-scale Survey",
      "PENIP radial survey" = "PENIP Radial Surveys",
      "PENIP drift survey" = "PENIP Drift Surveys"
    ))
  ) +
  coord_sf(
    xlim = survey_effort_xlim,
    ylim = survey_effort_ylim,
    expand = FALSE
  ) +
  guides(fill = guide_legend(nrow = 2, override.aes = list(alpha = 0.5))) +
  survey_map_theme

data_collection_map

ggsave(output_data_collection_map, data_collection_map, width = 180, height = 90, units = "mm", dpi = 300, bg = "white")
```

# IDW Estimates

The final analysis combines radial surveys, segmented drift transects, and broad-scale survey sites as observation units. Adult and juvenile densities are interpolated separately using inverse distance weighting. Prediction is limited to Inner Reef and Nearshore Ridge Complex habitat within the Figure 1B map extent; Artificial habitat is included with Nearshore Ridge Complex.

```{r idw-functions}
calculate_idw_prediction <- function(distance_matrix_m, observed_values, power = 2) {
  apply(distance_matrix_m, 1, function(distances_m) {
    zero_distance <- which(distances_m == 0)

    if (length(zero_distance) > 0) {
      return(mean(observed_values[zero_distance], na.rm = TRUE))
    }

    weights <- 1 / distances_m^power
    sum(weights * observed_values, na.rm = TRUE) / sum(weights[!is.na(observed_values)], na.rm = TRUE)
  })
}
```

```{r idw-observations}
idw_observations_sf_utm <- bind_rows(
  radial_observations_sf_utm,
  drift_segment_observations_sf_utm,
  berry_observations_sf_utm
) %>%
  mutate(
    adult_density_m2 = adult_count / survey_area_m2,
    juvenile_density_m2 = juvenile_count / survey_area_m2
  )

observation_summary <- idw_observations_sf_utm %>%
  st_drop_geometry() %>%
  group_by(observation_source) %>%
  summarize(
    observations = n(),
    area_m2 = sum(survey_area_m2, na.rm = TRUE),
    adult_count = sum(adult_count, na.rm = TRUE),
    juvenile_count = sum(juvenile_count, na.rm = TRUE),
    adult_density_per_ha = adult_count / area_m2 * 10000,
    juvenile_density_per_ha = juvenile_count / area_m2 * 10000,
    .groups = "drop"
  )

knitr::kable(observation_summary, digits = 2)
```

```{r idw-prediction}
survey_effort_extent_utm <- st_as_sfc(st_bbox(
  c(
    xmin = survey_effort_xlim[[1]],
    xmax = survey_effort_xlim[[2]],
    ymin = survey_effort_ylim[[1]],
    ymax = survey_effort_ylim[[2]]
  ),
  crs = st_crs(4326)
)) %>%
  st_transform(utm_crs)

idw_area <- st_intersection(
  target_habitats_utm %>% st_union() %>% st_make_valid(),
  survey_effort_extent_utm
) %>%
  st_make_valid()

idw_bbox <- st_bbox(survey_effort_extent_utm)

conch_data_bbox <- st_bbox(idw_observations_sf_utm)
idw_plot_bbox <- idw_bbox
idw_plot_bbox[c("ymin", "ymax")] <- conch_data_bbox[c("ymin", "ymax")] + c(-150, 150)

idw_grid <- st_make_grid(
  idw_area,
  cellsize = idw_cellsize_m,
  what = "polygons",
  square = TRUE
) %>%
  st_sf(geometry = .) %>%
  st_intersection(st_sf(geometry = idw_area)) %>%
  mutate(cell_area_m2 = as.numeric(st_area(geometry))) %>%
  filter(cell_area_m2 > 0)

idw_grid_points <- st_point_on_surface(idw_grid)

idw_distance_matrix_m <- st_distance(
  idw_grid_points,
  idw_observations_sf_utm
) %>%
  as.matrix() %>%
  as.numeric() %>%
  matrix(nrow = nrow(idw_grid_points), ncol = nrow(idw_observations_sf_utm))

idw_prediction <- idw_grid %>%
  mutate(
    adult_density_m2 = calculate_idw_prediction(
      idw_distance_matrix_m,
      idw_observations_sf_utm$adult_density_m2,
      power = 2
    ),
    juvenile_density_m2 = calculate_idw_prediction(
      idw_distance_matrix_m,
      idw_observations_sf_utm$juvenile_density_m2,
      power = 2
    ),
    adult_total = adult_density_m2 * cell_area_m2,
    juvenile_total = juvenile_density_m2 * cell_area_m2
  )

idw_population_summary <- tibble(
  modeled_area_m2 = sum(idw_prediction$cell_area_m2),
  modeled_area_ha = modeled_area_m2 / 10000,
  estimated_adults = sum(idw_prediction$adult_total, na.rm = TRUE),
  estimated_juveniles = sum(idw_prediction$juvenile_total, na.rm = TRUE),
  adult_density_per_ha = estimated_adults / modeled_area_ha,
  juvenile_density_per_ha = estimated_juveniles / modeled_area_ha,
  observation_units = nrow(idw_observations_sf_utm)
)

knitr::kable(idw_population_summary, digits = 2)
```

```{r idw-habitat-estimates}
idw_prediction_cells <- idw_prediction %>%
  mutate(cell_id = row_number())

habitat_prediction_area <- st_intersection(
  idw_prediction_cells,
  target_habitats_utm
) %>%
  mutate(intersection_area_m2 = as.numeric(st_area(geometry)))

habitat_idw_summary <- habitat_prediction_area %>%
  st_drop_geometry() %>%
  group_by(conch_habitat) %>%
  summarize(
    area_m2 = sum(intersection_area_m2, na.rm = TRUE),
    area_ha = area_m2 / 10000,
    estimated_adults = sum(adult_density_m2 * intersection_area_m2, na.rm = TRUE),
    estimated_juveniles = sum(juvenile_density_m2 * intersection_area_m2, na.rm = TRUE),
    adult_density_per_100m2 = estimated_adults / area_m2 * 100,
    juvenile_density_per_100m2 = estimated_juveniles / area_m2 * 100,
    adult_density_per_ha = estimated_adults / area_ha,
    juvenile_density_per_ha = estimated_juveniles / area_ha,
    .groups = "drop"
  ) %>%
  arrange(conch_habitat)

habitat_zone_prediction_area <- st_intersection(
  habitat_prediction_area,
  impact_zones_utm %>% select(name, scenario)
) %>%
  mutate(zone_area_m2 = as.numeric(st_area(geometry)))

habitat_zone_idw_summary <- habitat_zone_prediction_area %>%
  st_drop_geometry() %>%
  group_by(scenario, name, conch_habitat) %>%
  summarize(
    area_m2 = sum(zone_area_m2, na.rm = TRUE),
    area_ha = area_m2 / 10000,
    estimated_adults = sum(adult_density_m2 * zone_area_m2, na.rm = TRUE),
    estimated_juveniles = sum(juvenile_density_m2 * zone_area_m2, na.rm = TRUE),
    adult_density_per_100m2 = estimated_adults / area_m2 * 100,
    juvenile_density_per_100m2 = estimated_juveniles / area_m2 * 100,
    adult_density_per_ha = estimated_adults / area_ha,
    juvenile_density_per_ha = estimated_juveniles / area_ha,
    .groups = "drop"
  ) %>%
  arrange(scenario, name, conch_habitat)

knitr::kable(habitat_idw_summary, digits = 2)
knitr::kable(habitat_zone_idw_summary, digits = 2)
```

```{r idw-impact-zone-total-table}
impact_zone_order <- c(
  "Channel",
  "Side Slopes",
  "Depth_10cm",
  "Depth_5cm",
  "Depth_1cm",
  "Depth_0_5cm",
  "Depth_0_1cm",
  "Nearshore",
  "North",
  "East",
  "South",
  "Impact Monitoring Area"
)

impact_zone_labels <- c(
  "Channel" = "Channel",
  "Side Slopes" = "Side Slopes",
  "Depth_10cm" = "Sediment deposition >10 cm",
  "Depth_5cm" = "Sediment deposition 5-10 cm",
  "Depth_1cm" = "Sediment deposition 1-5 cm",
  "Depth_0_5cm" = "Sediment deposition 0.5-1 cm",
  "Depth_0_1cm" = "Sediment deposition 0.1-0.5 cm",
  "Nearshore" = "Nearshore",
  "North" = "North",
  "East" = "East",
  "South" = "South",
  "Impact Monitoring Area" = "Rest of impact monitoring area"
)

breeding_activity_buffer_utm <- breeding_activity_sites_sf_utm %>%
  st_buffer(breeding_activity_radius_m) %>%
  st_union() %>%
  st_make_valid()

adult_breeding_radius_prediction <- st_intersection(
  idw_prediction %>%
    mutate(
      density_per_ha = adult_density_m2 * 10000,
      density_class = case_when(
        density_per_ha <= 0 ~ "0",
        density_per_ha <= 10 ~ "1-10",
        density_per_ha <= 50 ~ "10-50",
        density_per_ha <= 100 ~ "50-100",
        TRUE ~ ">100"
      ),
      density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
    ),
  st_sf(geometry = breeding_activity_buffer_utm)
) %>%
  mutate(
    clipped_area_m2 = as.numeric(st_area(geometry)),
    adult_total_clipped = adult_density_m2 * clipped_area_m2
  )

adult_breeding_activity_zone_totals <- st_intersection(
  adult_breeding_radius_prediction %>% select(adult_density_m2),
  impact_zones_utm %>% select(name)
) %>%
  mutate(zone_area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  group_by(name) %>%
  summarize(
    estimated_adults_near_breeding = sum(adult_density_m2 * zone_area_m2, na.rm = TRUE),
    .groups = "drop"
  )

conch_impact_zone_totals <- habitat_zone_idw_summary %>%
  group_by(name) %>%
  summarize(
    estimated_adults = sum(estimated_adults, na.rm = TRUE),
    estimated_juveniles = sum(estimated_juveniles, na.rm = TRUE),
    estimated_total = estimated_adults + estimated_juveniles,
    .groups = "drop"
  ) %>%
  left_join(adult_breeding_activity_zone_totals, by = "name") %>%
  right_join(tibble(name = impact_zone_order), by = "name") %>%
  mutate(
    estimated_adults = replace_na(estimated_adults, 0),
    estimated_adults_near_breeding = replace_na(estimated_adults_near_breeding, 0),
    estimated_juveniles = replace_na(estimated_juveniles, 0),
    estimated_total = replace_na(estimated_total, 0),
    impact_area = impact_zone_labels[name],
    name = factor(name, levels = impact_zone_order)
  ) %>%
  arrange(name) %>%
  mutate(
    cumulative_adults = cumsum(estimated_adults),
    cumulative_juveniles = cumsum(estimated_juveniles),
    cumulative_total = cumsum(estimated_total)
  ) %>%
  select(
    impact_area,
    estimated_adults,
    estimated_adults_near_breeding,
    estimated_juveniles,
    estimated_total,
    cumulative_adults,
    cumulative_juveniles,
    cumulative_total
  )

conch_impact_zone_totals_with_total <- bind_rows(
  conch_impact_zone_totals,
  conch_impact_zone_totals %>%
    summarize(
      impact_area = "Total",
      estimated_adults = sum(estimated_adults, na.rm = TRUE),
      estimated_adults_near_breeding = sum(estimated_adults_near_breeding, na.rm = TRUE),
      estimated_juveniles = sum(estimated_juveniles, na.rm = TRUE),
      estimated_total = sum(estimated_total, na.rm = TRUE),
      cumulative_adults = sum(estimated_adults, na.rm = TRUE),
      cumulative_juveniles = sum(estimated_juveniles, na.rm = TRUE),
      cumulative_total = sum(estimated_total, na.rm = TRUE)
    )
)

conch_impact_zone_totals_display <- conch_impact_zone_totals_with_total %>%
  mutate(across(
    where(is.numeric),
    ~ format(round(.), big.mark = ",", scientific = FALSE, trim = TRUE)
  )) %>%
  rename(
    `Impact area` = impact_area,
    `Adult conch` = estimated_adults,
    `Adult conch within 330 m of breeding activity` = estimated_adults_near_breeding,
    `Juvenile conch` = estimated_juveniles,
    `Total conch` = estimated_total,
    `Cumulative adult conch` = cumulative_adults,
    `Cumulative juvenile conch` = cumulative_juveniles,
    `Cumulative total conch` = cumulative_total
  )

knitr::kable(
  conch_impact_zone_totals_display,
  caption = "Estimated total adult and juvenile conch in each impact zone. Estimates are based on IDW-predicted habitat-specific densities within Inner Reef and Nearshore Ridge Complex habitat, with Artificial habitat included in Nearshore Ridge Complex. Adults near breeding activity are the adult IDW estimates restricted to areas within 330 m of sites where breeding activity was recorded. Cumulative totals sum across impact zones moving away from the channel."
)
```

```{r idw-map, fig.width=8, fig.height=7}
idw_prediction_long <- bind_rows(
  idw_prediction %>%
    mutate(
      age_class = "Adult",
      density_per_ha = adult_density_m2 * 10000,
      total_animals = adult_total
    ) %>%
    select(age_class, density_per_ha, total_animals, geometry),
  idw_prediction %>%
    mutate(
      age_class = "Juvenile",
      density_per_ha = juvenile_density_m2 * 10000,
      total_animals = juvenile_total
    ) %>%
    select(age_class, density_per_ha, total_animals, geometry)
) %>%
  mutate(
    density_class = case_when(
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100"))
  )

target_habitat_outline <- st_intersection(
  target_habitats_utm,
  st_sf(geometry = survey_effort_extent_utm)
) %>%
  st_make_valid()

make_inner_boundary_lines <- function(x, offset_m = 15) {
  map_dfr(seq_len(nrow(x)), function(i) {
    feature_i <- x[i, ]
    inner_i <- st_buffer(feature_i, -offset_m)

    if (all(st_is_empty(inner_i))) {
      inner_i <- feature_i
    }

    inner_i %>%
      st_boundary() %>%
      st_collection_extract("LINESTRING")
  }) %>%
    st_as_sf() %>%
    st_make_valid()
}

target_habitat_inner_lines <- make_inner_boundary_lines(target_habitat_outline, offset_m = 15)

idw_map <- ggplot() +
  geom_sf(
    data = idw_prediction_long,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  facet_wrap(~ age_class) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    drop = FALSE,
    name = "Conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(face = "bold", size = 11),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

idw_map

ggsave(output_idw_map, idw_map, width = 9, height = 4.5, dpi = 300, bg = "white")
```

```{r adult-breeding-radius-idw-map, fig.width=5.5, fig.height=7}
adult_breeding_radius_summary <- adult_breeding_radius_prediction %>%
  st_drop_geometry() %>%
  summarize(
    breeding_activity_sites = nrow(breeding_activity_sites_sf_utm),
    radius_m = breeding_activity_radius_m,
    area_m2 = sum(clipped_area_m2, na.rm = TRUE),
    area_ha = area_m2 / 10000,
    estimated_adults = sum(adult_total_clipped, na.rm = TRUE),
    adult_density_per_ha = estimated_adults / area_ha
  )

knitr::kable(adult_breeding_radius_summary, digits = 2)

adult_breeding_radius_map <- ggplot() +
  geom_sf(
    data = adult_breeding_radius_prediction,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  geom_sf(
    data = breeding_activity_sites_sf_utm,
    shape = 21,
    fill = "#7B3294",
    color = "white",
    size = 2.2,
    stroke = 0.35
  ) +
  scale_fill_manual(
    values = c(
      "0" = "white",
      "1-10" = "#F0F0F0",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525"
    ),
    drop = FALSE,
    name = "Adult conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]]
    ),
    name = "Habitat"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = idw_plot_bbox[c("xmin", "xmax")],
    ylim = idw_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  labs(title = paste0("Adult conch density within ", breeding_activity_radius_m, " m of breeding activity")) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

adult_breeding_radius_map

ggsave(
  output_adult_breeding_radius_idw_map,
  adult_breeding_radius_map,
  width = 5.5,
  height = 7,
  dpi = 300,
  bg = "white"
)
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
