---
title: "Queen Conch Size and Lip Thickness"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

find_project_root <- function(start = getwd()) {
  start <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- c(start, normalizePath(file.path(start, ".."), winslash = "/", mustWork = TRUE))

  while (length(candidates) > 0) {
    candidate <- candidates[[1]]
    if (file.exists(file.path(candidate, "data/conch/TaggingData_June2026.txt"))) {
      return(candidate)
    }

    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not find the PENIP project root from ", start, call. = FALSE)
    }

    candidates <- c(candidates[-1], parent)
  }
}

knitr::opts_knit$set(root.dir = find_project_root())
```

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(purrr)
library(readxl)
library(sf)
library(tidyr)
```

```{r paths}
molasses_tagging_file <- "data/conch/TaggingData_June2026.txt"
port_everglades_workbook <- "data/conch/FLK_QueenConchSurveys_2023-2025_forNOAA.xlsx"
polygons_rds <- "data/processed/polygons_final.rds"

age_class_levels <- c("Juvenile", "Sub-adult", "Adult")
age_class_cols <- c("Juvenile" = "#2B8CBE", "Sub-adult" = "#31A354", "Adult" = "#D95F0E")
keys_comparison_location_levels <- c("Port Everglades", "Rest of Florida Keys")

output_length_lip_histograms <- "outputs/conch_molasses_length_lip_histograms.png"
output_length_by_ageclass <- "outputs/conch_molasses_length_by_lip_ageclass.png"
output_length_lip_scatter <- "outputs/conch_molasses_length_lip_scatter.png"
output_lip_ageclass_comparison <- "outputs/conch_molasses_lip_ageclass_comparison.png"
output_port_length_lip_histograms <- "outputs/conch_port_everglades_length_lip_histograms.png"
output_port_length_lip_scatter <- "outputs/conch_port_everglades_length_lip_scatter.png"
output_location_ageclass_counts <- "outputs/conch_size_lip_ageclass_counts_by_location.png"
output_location_ageclass_summary <- "outputs/conch_size_lip_summary_by_location_ageclass.png"
output_location_metric_distributions <- "outputs/conch_size_lip_distributions_by_location.png"
output_location_statistical_effects <- "outputs/conch_size_lip_port_vs_rest_keys_effects.png"
```

```{r helper-functions}
assign_lip_age_class <- function(lip_mm) {
  case_when(
    is.na(lip_mm) ~ "Juvenile",
    lip_mm < 15 ~ "Sub-adult",
    lip_mm >= 15 ~ "Adult"
  ) %>%
    factor(levels = age_class_levels)
}

mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

median_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else median(x, na.rm = TRUE)
}

sd_or_na <- function(x) {
  if (sum(!is.na(x)) < 2) NA_real_ else sd(x, na.rm = TRUE)
}

format_p_value <- function(p) {
  case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "<0.001",
    TRUE ~ format(round(p, 3), nsmall = 3)
  )
}

summarize_metric <- function(x) {
  if (all(is.na(x))) {
    return(tibble(
      n = 0,
      mean = NA_real_,
      sd = NA_real_,
      min = NA_real_,
      q25 = NA_real_,
      median = NA_real_,
      q75 = NA_real_,
      max = NA_real_
    ))
  }

  tibble(
    n = sum(!is.na(x)),
    mean = mean(x, na.rm = TRUE),
    sd = sd_or_na(x),
    min = min(x, na.rm = TRUE),
    q25 = as.numeric(quantile(x, 0.25, na.rm = TRUE)),
    median = median(x, na.rm = TRUE),
    q75 = as.numeric(quantile(x, 0.75, na.rm = TRUE)),
    max = max(x, na.rm = TRUE)
  )
}
```

# Molasses Reef Tagging Data

These data are the June 2026 Molasses Reef tagging measurements provided by Dr. Cao. Juveniles are defined here as animals with unmeasured lip thickness because the lip was not yet flared, sub-adults are animals with measured shell lip thickness `< 15` mm, and adults are animals with shell lip thickness `>= 15` mm.

```{r import-molasses}
molasses_raw <- read.delim(
  molasses_tagging_file,
  sep = "\t",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

molasses_conch <- molasses_raw %>%
  transmute(
    og_order = as.integer(`OG order`),
    date = as.Date(Date, format = "%m/%d/%Y"),
    gps = GPS,
    waypoint = Waypoint,
    tag = Tag,
    color = Color,
    length_mm = suppressWarnings(as.numeric(`Length mm`)),
    lip_mm = suppressWarnings(as.numeric(`Lip mm`)),
    source_ageclass = trimws(Ageclass),
    notes = Notes,
    quick_check = QuickCheck,
    offset_incorporated = `offset (incorporated now)`,
    has_picture = `pictures (1y0no)`,
    lip_measured = !is.na(lip_mm),
    lip_age_class = assign_lip_age_class(lip_mm),
    location = "Molasses Reef"
  )

molasses_import_summary <- tibble(
  records = nrow(molasses_conch),
  measured_length = sum(!is.na(molasses_conch$length_mm)),
  missing_length = sum(is.na(molasses_conch$length_mm)),
  measured_lip = sum(!is.na(molasses_conch$lip_mm)),
  missing_lip = sum(is.na(molasses_conch$lip_mm)),
  juveniles_by_lip = sum(molasses_conch$lip_age_class == "Juvenile", na.rm = TRUE),
  subadults_by_lip = sum(molasses_conch$lip_age_class == "Sub-adult", na.rm = TRUE),
  adults_by_lip = sum(molasses_conch$lip_age_class == "Adult", na.rm = TRUE)
)

knitr::kable(molasses_import_summary, digits = 0)
```

```{r ageclass-counts}
molasses_lip_age_counts <- molasses_conch %>%
  count(lip_age_class, name = "n") %>%
  mutate(percent = n / sum(n) * 100)

source_vs_lip_ageclass <- molasses_conch %>%
  mutate(source_ageclass = if_else(source_ageclass == "", NA_character_, source_ageclass)) %>%
  count(source_ageclass, lip_age_class, name = "n") %>%
  arrange(source_ageclass, lip_age_class)

knitr::kable(molasses_lip_age_counts, digits = 1)
knitr::kable(source_vs_lip_ageclass, digits = 0)
```

# Port Everglades Measurements

Port Everglades surveys are selected from the circular survey worksheet using latitude/longitude points that intersect the Port Everglades habitat polygon footprint. Individual shell measurements are then joined from the corresponding survey IDs in the circle conch measurement worksheet.

```{r import-port-everglades}
old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

polygons_final <- readRDS(polygons_rds) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)

port_circular_surveys <- readxl::read_xlsx(
  port_everglades_workbook,
  sheet = "CircularSurveys (10m rad)"
) %>%
  mutate(
    survey_id = as.numeric(Survey),
    lat = as.numeric(Lat),
    lon = as.numeric(Lon)
  ) %>%
  filter(!is.na(survey_id), !is.na(lat), !is.na(lon))

port_circular_surveys_sf <- port_circular_surveys %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

port_aoi <- st_union(polygons_final)

port_everglades_surveys <- port_circular_surveys_sf[
  lengths(st_intersects(port_circular_surveys_sf, port_aoi)) > 0,
] %>%
  st_drop_geometry()

port_survey_ids <- sort(unique(port_everglades_surveys$survey_id))

port_measurements_raw <- readxl::read_xlsx(
  port_everglades_workbook,
  sheet = "CircleConchMeasurements"
)

port_everglades_conch <- port_measurements_raw %>%
  mutate(survey_id = as.numeric(Survey)) %>%
  filter(survey_id %in% port_survey_ids) %>%
  transmute(
    year = Year,
    survey_id = survey_id,
    location_label = Location,
    diver = Diver,
    site = Site,
    depth_ft = `depth(ft)`,
    quad = quad,
    radius_m = `radius(m)`,
    length_mm = suppressWarnings(as.numeric(`length(mm)`)),
    lip_mm = suppressWarnings(as.numeric(`lip(mm)`)),
    lip_measured = !is.na(lip_mm),
    lip_age_class = assign_lip_age_class(lip_mm),
    location = "Port Everglades"
  )

florida_keys_conch <- port_measurements_raw %>%
  mutate(survey_id = as.numeric(Survey)) %>%
  filter(!survey_id %in% port_survey_ids) %>%
  transmute(
    year = Year,
    survey_id = survey_id,
    location_label = Location,
    diver = Diver,
    site = Site,
    depth_ft = `depth(ft)`,
    quad = quad,
    radius_m = `radius(m)`,
    length_mm = suppressWarnings(as.numeric(`length(mm)`)),
    lip_mm = suppressWarnings(as.numeric(`lip(mm)`)),
    lip_measured = !is.na(lip_mm),
    lip_age_class = assign_lip_age_class(lip_mm),
    location = "Rest of Florida Keys"
  )

port_import_summary <- tibble(
  circular_surveys_total = nrow(port_circular_surveys),
  port_everglades_surveys = length(port_survey_ids),
  records = nrow(port_everglades_conch),
  measured_length = sum(!is.na(port_everglades_conch$length_mm)),
  missing_length = sum(is.na(port_everglades_conch$length_mm)),
  measured_lip = sum(!is.na(port_everglades_conch$lip_mm)),
  missing_lip = sum(is.na(port_everglades_conch$lip_mm)),
  juveniles_by_lip = sum(port_everglades_conch$lip_age_class == "Juvenile", na.rm = TRUE),
  subadults_by_lip = sum(port_everglades_conch$lip_age_class == "Sub-adult", na.rm = TRUE),
  adults_by_lip = sum(port_everglades_conch$lip_age_class == "Adult", na.rm = TRUE)
)

port_lip_age_counts <- port_everglades_conch %>%
  count(lip_age_class, name = "n") %>%
  mutate(percent = n / sum(n) * 100)

knitr::kable(port_import_summary, digits = 0)
knitr::kable(port_lip_age_counts, digits = 1)
```

```{r import-florida-keys}
florida_keys_import_summary <- tibble(
  records = nrow(florida_keys_conch),
  measured_length = sum(!is.na(florida_keys_conch$length_mm)),
  missing_length = sum(is.na(florida_keys_conch$length_mm)),
  measured_lip = sum(!is.na(florida_keys_conch$lip_mm)),
  missing_lip = sum(is.na(florida_keys_conch$lip_mm)),
  juveniles_by_lip = sum(florida_keys_conch$lip_age_class == "Juvenile", na.rm = TRUE),
  subadults_by_lip = sum(florida_keys_conch$lip_age_class == "Sub-adult", na.rm = TRUE),
  adults_by_lip = sum(florida_keys_conch$lip_age_class == "Adult", na.rm = TRUE)
)

florida_keys_lip_age_counts <- florida_keys_conch %>%
  count(lip_age_class, name = "n") %>%
  mutate(percent = n / sum(n) * 100)

knitr::kable(florida_keys_import_summary, digits = 0)
knitr::kable(florida_keys_lip_age_counts, digits = 1)
```

# Summary Statistics

```{r summary-statistics}
molasses_overall_summary <- bind_rows(
  summarize_metric(molasses_conch$length_mm) %>% mutate(metric = "Shell length (mm)"),
  summarize_metric(molasses_conch$lip_mm) %>% mutate(metric = "Lip thickness (mm)")
) %>%
  select(metric, everything())

molasses_ageclass_summary <- molasses_conch %>%
  group_by(lip_age_class) %>%
  summarize(
    n = n(),
    measured_lip = sum(lip_measured),
    unflared_missing_lip = sum(!lip_measured),
    length_mean_mm = mean_or_na(length_mm),
    length_sd_mm = sd_or_na(length_mm),
    length_median_mm = median_or_na(length_mm),
    lip_mean_mm = mean_or_na(lip_mm),
    lip_sd_mm = sd_or_na(lip_mm),
    lip_median_mm = median_or_na(lip_mm),
    .groups = "drop"
  )

port_overall_summary <- bind_rows(
  summarize_metric(port_everglades_conch$length_mm) %>% mutate(metric = "Shell length (mm)"),
  summarize_metric(port_everglades_conch$lip_mm) %>% mutate(metric = "Lip thickness (mm)")
) %>%
  select(metric, everything())

port_ageclass_summary <- port_everglades_conch %>%
  group_by(lip_age_class) %>%
  summarize(
    n = n(),
    measured_lip = sum(lip_measured),
    unflared_missing_lip = sum(!lip_measured),
    length_mean_mm = mean_or_na(length_mm),
    length_sd_mm = sd_or_na(length_mm),
    length_median_mm = median_or_na(length_mm),
    lip_mean_mm = mean_or_na(lip_mm),
    lip_sd_mm = sd_or_na(lip_mm),
    lip_median_mm = median_or_na(lip_mm),
    .groups = "drop"
  )

florida_keys_overall_summary <- bind_rows(
  summarize_metric(florida_keys_conch$length_mm) %>% mutate(metric = "Shell length (mm)"),
  summarize_metric(florida_keys_conch$lip_mm) %>% mutate(metric = "Lip thickness (mm)")
) %>%
  select(metric, everything())

florida_keys_ageclass_summary <- florida_keys_conch %>%
  group_by(lip_age_class) %>%
  summarize(
    n = n(),
    measured_lip = sum(lip_measured),
    unflared_missing_lip = sum(!lip_measured),
    length_mean_mm = mean_or_na(length_mm),
    length_sd_mm = sd_or_na(length_mm),
    length_median_mm = median_or_na(length_mm),
    lip_mean_mm = mean_or_na(lip_mm),
    lip_sd_mm = sd_or_na(lip_mm),
    lip_median_mm = median_or_na(lip_mm),
    .groups = "drop"
  )

keys_comparison_conch <- bind_rows(
  port_everglades_conch %>%
    select(location, length_mm, lip_mm, lip_measured, lip_age_class),
  florida_keys_conch %>%
    select(location, length_mm, lip_mm, lip_measured, lip_age_class)
  ) %>%
  mutate(
    location = factor(location, levels = keys_comparison_location_levels),
    lip_age_class = factor(lip_age_class, levels = age_class_levels)
  )

location_age_counts <- keys_comparison_conch %>%
  count(location, lip_age_class, name = "n") %>%
  group_by(location) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup()

location_ageclass_summary <- keys_comparison_conch %>%
  group_by(location, lip_age_class) %>%
  summarize(
    n = n(),
    measured_lip = sum(lip_measured),
    unflared_missing_lip = sum(!lip_measured),
    length_mean_mm = mean_or_na(length_mm),
    length_median_mm = median_or_na(length_mm),
    lip_mean_mm = mean_or_na(lip_mm),
    lip_median_mm = median_or_na(lip_mm),
    .groups = "drop"
  )

knitr::kable(molasses_overall_summary, digits = 1)
knitr::kable(molasses_ageclass_summary, digits = 1)
knitr::kable(port_overall_summary, digits = 1)
knitr::kable(port_ageclass_summary, digits = 1)
knitr::kable(florida_keys_overall_summary, digits = 1)
knitr::kable(florida_keys_ageclass_summary, digits = 1)
knitr::kable(location_ageclass_summary, digits = 1)
```

# Distributions

```{r length-lip-histograms, fig.width=8, fig.height=5}
molasses_metric_long <- molasses_conch %>%
  select(lip_age_class, length_mm, lip_mm) %>%
  pivot_longer(
    cols = c(length_mm, lip_mm),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      length_mm = "Shell length (mm)",
      lip_mm = "Lip thickness (mm)"
    )
  ) %>%
  filter(!is.na(value))

length_lip_histograms <- ggplot(
  molasses_metric_long,
  aes(x = value)
) +
  geom_histogram(aes(fill = lip_age_class), color = "white", bins = 35, linewidth = 0.2) +
  geom_vline(
    data = tibble(metric = "Lip thickness (mm)", threshold = 15),
    aes(xintercept = threshold),
    linetype = "dashed",
    color = "black",
    linewidth = 0.6
  ) +
  facet_wrap(~ metric, scales = "free_x") +
  scale_fill_manual(values = age_class_cols, drop = FALSE) +
  labs(x = NULL, y = "Number of animals", fill = "Lip age class") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

length_lip_histograms

ggsave(output_length_lip_histograms, length_lip_histograms, width = 8, height = 5, dpi = 300)
```

```{r length-by-ageclass, fig.width=7, fig.height=5}
length_by_ageclass <- ggplot(
  molasses_conch,
  aes(x = lip_age_class, y = length_mm, fill = lip_age_class)
) +
  geom_violin(width = 0.8, alpha = 0.5, color = NA, trim = FALSE) +
  geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.85, color = "grey25") +
  geom_jitter(width = 0.12, height = 0, alpha = 0.25, size = 0.8, color = "grey20") +
  scale_fill_manual(values = age_class_cols) +
  labs(x = "Lip age class", y = "Shell length (mm)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

length_by_ageclass

ggsave(output_length_by_ageclass, length_by_ageclass, width = 7, height = 5, dpi = 300)
```

```{r length-lip-scatter, fig.width=7, fig.height=5}
length_lip_scatter <- ggplot(
  molasses_conch %>% filter(!is.na(length_mm), !is.na(lip_mm)),
  aes(x = length_mm, y = lip_mm, color = lip_age_class)
) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "grey25", linewidth = 0.6) +
  geom_point(alpha = 0.55, size = 1.5) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.8) +
  scale_color_manual(values = age_class_cols) +
  labs(
    x = "Shell length (mm)",
    y = "Lip thickness (mm)",
    color = "Lip age class"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

length_lip_scatter

ggsave(output_length_lip_scatter, length_lip_scatter, width = 7, height = 5, dpi = 300)
```

```{r lip-ageclass-comparison, fig.width=7, fig.height=5}
lip_ageclass_comparison <- ggplot(
  molasses_lip_age_counts,
  aes(x = lip_age_class, y = n, fill = lip_age_class)
) +
  geom_col(color = "grey30", linewidth = 0.2) +
  geom_text(aes(label = paste0(n, " (", round(percent, 1), "%)")), vjust = -0.35, size = 3.5) +
  scale_fill_manual(values = age_class_cols, drop = FALSE) +
  labs(x = "Lip age class", y = "Number of animals") +
  coord_cartesian(ylim = c(0, max(molasses_lip_age_counts$n) * 1.12)) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

lip_ageclass_comparison

ggsave(output_lip_ageclass_comparison, lip_ageclass_comparison, width = 7, height = 5, dpi = 300)
```

# Port Everglades Distributions

```{r port-length-lip-histograms, fig.width=8, fig.height=5}
port_metric_long <- port_everglades_conch %>%
  select(lip_age_class, length_mm, lip_mm) %>%
  pivot_longer(
    cols = c(length_mm, lip_mm),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      length_mm = "Shell length (mm)",
      lip_mm = "Lip thickness (mm)"
    )
  ) %>%
  filter(!is.na(value))

port_length_lip_histograms <- ggplot(
  port_metric_long,
  aes(x = value)
) +
  geom_histogram(aes(fill = lip_age_class), color = "white", bins = 30, linewidth = 0.2) +
  geom_vline(
    data = tibble(metric = "Lip thickness (mm)", threshold = 15),
    aes(xintercept = threshold),
    linetype = "dashed",
    color = "black",
    linewidth = 0.6
  ) +
  facet_wrap(~ metric, scales = "free_x") +
  scale_fill_manual(values = age_class_cols, drop = FALSE) +
  labs(x = NULL, y = "Number of animals", fill = "Lip age class") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

port_length_lip_histograms

ggsave(output_port_length_lip_histograms, port_length_lip_histograms, width = 8, height = 5, dpi = 300)
```

```{r port-length-lip-scatter, fig.width=7, fig.height=5}
port_length_lip_scatter <- ggplot(
  port_everglades_conch %>% filter(!is.na(length_mm), !is.na(lip_mm)),
  aes(x = length_mm, y = lip_mm, color = lip_age_class)
) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "grey25", linewidth = 0.6) +
  geom_point(alpha = 0.6, size = 1.7) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.8) +
  scale_color_manual(values = age_class_cols) +
  labs(
    x = "Shell length (mm)",
    y = "Lip thickness (mm)",
    color = "Lip age class"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

port_length_lip_scatter

ggsave(output_port_length_lip_scatter, port_length_lip_scatter, width = 7, height = 5, dpi = 300)
```

# Location Comparisons

```{r location-ageclass-counts, fig.width=8, fig.height=5}
location_ageclass_counts_plot <- ggplot(
  location_age_counts,
  aes(x = lip_age_class, y = n, fill = lip_age_class)
) +
  geom_col(color = "grey30", linewidth = 0.2) +
  geom_text(aes(label = paste0(n, " (", round(percent, 1), "%)")), vjust = -0.35, size = 3.2) +
  facet_wrap(~ location) +
  scale_fill_manual(values = age_class_cols, drop = FALSE) +
  labs(x = "Lip age class", y = "Number of animals") +
  coord_cartesian(ylim = c(0, max(location_age_counts$n) * 1.15)) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

location_ageclass_counts_plot

ggsave(output_location_ageclass_counts, location_ageclass_counts_plot, width = 8, height = 5, dpi = 300)
```

```{r location-metric-distributions, fig.width=9, fig.height=6}
combined_metric_long <- keys_comparison_conch %>%
  select(location, lip_age_class, length_mm, lip_mm) %>%
  pivot_longer(
    cols = c(length_mm, lip_mm),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      length_mm = "Shell length (mm)",
      lip_mm = "Lip thickness (mm)"
    )
  ) %>%
  filter(!is.na(value))

location_metric_distributions <- ggplot(
  combined_metric_long,
  aes(x = lip_age_class, y = value, fill = location)
) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.65, position = position_dodge(width = 0.75)) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = "Lip age class", y = NULL, fill = "Location") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

location_metric_distributions

ggsave(output_location_metric_distributions, location_metric_distributions, width = 9, height = 6, dpi = 300)
```

```{r location-summary-plot, fig.width=9, fig.height=6}
location_ageclass_summary_long <- location_ageclass_summary %>%
  select(location, lip_age_class, length_mean_mm, length_median_mm, lip_mean_mm, lip_median_mm) %>%
  pivot_longer(
    cols = c(length_mean_mm, length_median_mm, lip_mean_mm, lip_median_mm),
    names_to = "summary_metric",
    values_to = "value"
  ) %>%
  separate(summary_metric, into = c("measurement", "statistic", "unit"), sep = "_") %>%
  mutate(
    measurement = recode(
      measurement,
      length = "Shell length",
      lip = "Lip thickness"
    ),
    statistic = recode(
      statistic,
      mean = "Mean",
      median = "Median"
    )
  )

location_ageclass_summary_plot <- ggplot(
  location_ageclass_summary_long,
  aes(x = lip_age_class, y = value, color = location, shape = statistic, group = interaction(location, statistic))
) +
  geom_point(position = position_dodge(width = 0.45), size = 2.6, na.rm = TRUE) +
  facet_wrap(~ measurement, scales = "free_y") +
  labs(x = "Lip age class", y = "Measurement (mm)", color = "Location", shape = "Statistic") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

location_ageclass_summary_plot

ggsave(output_location_ageclass_summary, location_ageclass_summary_plot, width = 9, height = 6, dpi = 300)
```

# Statistical Comparisons

Mean shell length and mean lip thickness are compared between Port Everglades and the rest of the Florida Keys within each age class using Welch two-sample t-tests. Juveniles have unflared, unmeasured lips by definition, so lip thickness is not tested for juveniles.

```{r statistical-comparisons}
welch_test_port_vs_rest_keys <- function(dat) {
  port_values <- dat %>%
    filter(location == "Port Everglades") %>%
    pull(value)

  rest_keys_values <- dat %>%
    filter(location == "Rest of Florida Keys") %>%
    pull(value)

  if (length(port_values) < 2 || length(rest_keys_values) < 2) {
    return(tibble())
  }

  test_result <- t.test(port_values, rest_keys_values)

  tibble(
    n_port_everglades = length(port_values),
    n_rest_keys = length(rest_keys_values),
    mean_port_everglades = mean_or_na(port_values),
    mean_rest_keys = mean_or_na(rest_keys_values),
    mean_difference_port_minus_rest_keys = mean_port_everglades - mean_rest_keys,
    median_port_everglades = median_or_na(port_values),
    median_rest_keys = median_or_na(rest_keys_values),
    t_statistic = unname(test_result$statistic),
    df = unname(test_result$parameter),
    p_value = test_result$p.value
  )
}

measurement_location_tests <- combined_metric_long %>%
  group_by(metric, lip_age_class) %>%
  group_modify(~ welch_test_port_vs_rest_keys(.x)) %>%
  ungroup() %>%
  mutate(
    p_adjusted_bh = p.adjust(p_value, method = "BH"),
    p_value_formatted = format_p_value(p_value),
    p_adjusted_bh_formatted = format_p_value(p_adjusted_bh)
  ) %>%
  arrange(metric, lip_age_class)

knitr::kable(measurement_location_tests, digits = 3)
```

```{r statistical-effect-plot, fig.width=8, fig.height=5}
location_statistical_effects <- measurement_location_tests %>%
  mutate(
    difference = mean_difference_port_minus_rest_keys,
    comparison_label = paste(metric, lip_age_class, sep = "\n")
  )

location_statistical_effect_plot <- ggplot(
  location_statistical_effects,
  aes(x = reorder(comparison_label, difference), y = difference, fill = metric)
) +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.5) +
  geom_col(color = "grey30", linewidth = 0.2) +
  geom_text(
    aes(label = paste0("BH p=", p_adjusted_bh_formatted)),
    hjust = if_else(location_statistical_effects$difference >= 0, -0.08, 1.08),
    size = 3
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mean difference (Port Everglades - rest of Florida Keys, mm)",
    fill = "Measurement"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

location_statistical_effect_plot

ggsave(output_location_statistical_effects, location_statistical_effect_plot, width = 8, height = 5, dpi = 300)
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
