---
title: "PENIP Queen Conch Analysis"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

find_project_root <- function(start = getwd()) {
  start <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- c(start, normalizePath(file.path(start, ".."), winslash = "/", mustWork = TRUE))

  while (length(candidates) > 0) {
    candidate <- candidates[[1]]
    if (file.exists(file.path(candidate, "data/processed/polygons_final.rds"))) {
      return(candidate)
    }

    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not find the PENIP project root from ", start, call. = FALSE)
    }

    candidates <- c(candidates[-1], parent)
  }
}

knitr::opts_knit$set(root.dir = find_project_root())
```

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
library(readxl)
library(sf)
library(stringr)
library(tibble)
library(tidyr)
library(xml2)
```

```{r paths-and-constants}
conch_xlsx <- "data/conch/PENIP_QueenConchSurveys.xlsx"
radial_counts_csv <- "data/conch/PENIP_circular_counts.csv"
berry_broad_xlsx <- "data/conch/Berry_Conch_per_site.xlsx"
roving_xlsx <- "data/conch/Roving Diver Survey Data Sept 2025.xlsx"
port_everglades_workbook <- "data/conch/FLK_QueenConchSurveys_2023-2025_forNOAA.xlsx"
polygons_rds <- "data/processed/polygons_final.rds"
impact_kml <- "data/spatial/Impact_zones_Scenario 2.kml"
nmfs_kml <- "data/spatial/Scenario2_NMFSConsiderationAreas.kml"

radial_sheet <- "PENIP_Conch_10mRadial"
drift_sheet <- "PENIP_DriftTransects"
drift_position_sheet <- "PENIP_ConchDriftPosition"

utm_crs <- 32617
radial_survey_radius_m <- 10
radial_survey_area_m2 <- pi * radial_survey_radius_m^2
drift_half_width_m <- 2
drift_segment_length_m <- 20
berry_broad_site_area_m2 <- 2 * 30 * 4
prediction_cellsize_m <- 25
kriging_nmax <- 20
kriging_maxdist_m <- Inf
kriging_nmin <- 1
kriging_sensitivity_nmax <- c(8, 12, 20)
kriging_sensitivity_maxdist_m <- c(250, 330, 500, 1000)
kriging_sensitivity_nmin <- c(3, 5)
kriging_cv_cores <- 4
breeding_activity_radius_m <- 330

age_class_levels <- c("Juvenile", "Sub-adult", "Adult")
age_class_cols <- c("Juvenile" = "#2B8CBE", "Sub-adult" = "#31A354", "Adult" = "#D95F0E")
keys_comparison_location_levels <- c("Port Everglades", "Rest of Florida Keys")

output_kriging_map <- "outputs/conch_kriging_density_map.png"
output_kriging_loocv_plot <- "outputs/conch_kriging_loocv.png"
output_port_length_lip_histograms <- "outputs/conch_port_everglades_length_lip_histograms.png"
output_location_metric_distributions <- "outputs/conch_size_lip_distributions_by_location.png"
kriging_loocv_cache_rds <- "outputs/cache/conch_kriging_loocv_results.rds"

survey_effort_xlim <- c(-80.1116, -80.078)
survey_effort_ylim <- c(26.058, 26.112)
```

```{r size-lip-helper-functions}
assign_lip_age_class <- function(lip_mm) {
  case_when(
    is.na(lip_mm) ~ "Juvenile",
    lip_mm < 15 ~ "Sub-adult",
    lip_mm >= 15 ~ "Adult"
  ) %>%
    factor(levels = age_class_levels)
}

mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

median_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else median(x, na.rm = TRUE)
}

sd_or_na <- function(x) {
  if (sum(!is.na(x)) < 2) NA_real_ else sd(x, na.rm = TRUE)
}

format_p_value <- function(p) {
  case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "<0.001",
    TRUE ~ format(round(p, 3), nsmall = 3)
  )
}

summarize_metric <- function(x) {
  if (all(is.na(x))) {
    return(tibble(
      n = 0,
      mean = NA_real_,
      sd = NA_real_,
      min = NA_real_,
      q25 = NA_real_,
      median = NA_real_,
      q75 = NA_real_,
      max = NA_real_
    ))
  }

  tibble(
    n = sum(!is.na(x)),
    mean = mean(x, na.rm = TRUE),
    sd = sd_or_na(x),
    min = min(x, na.rm = TRUE),
    q25 = as.numeric(quantile(x, 0.25, na.rm = TRUE)),
    median = median(x, na.rm = TRUE),
    q75 = as.numeric(quantile(x, 0.75, na.rm = TRUE)),
    max = max(x, na.rm = TRUE)
  )
}
```

# Input Data

```{r import-data}
polygons_final <- readRDS(polygons_rds) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)

type_cols <- RColorBrewer::brewer.pal(nlevels(factor(polygons_final$Type)) + 1, "Set3")[-1]
names(type_cols) <- levels(factor(polygons_final$Type))

old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

polygons_utm <- polygons_final %>%
  st_transform(utm_crs) %>%
  st_make_valid()

conch_habitat_bbox <- st_bbox(polygons_utm)
sand_scattered_inshore_strip_height_m <- prediction_cellsize_m

make_inner_reef_inshore_mask <- function(inner_reef_utm, mask_bbox, strip_height_m) {
  y_breaks <- seq(mask_bbox[["ymin"]], mask_bbox[["ymax"]], by = strip_height_m)
  if (tail(y_breaks, 1) < mask_bbox[["ymax"]]) {
    y_breaks <- c(y_breaks, mask_bbox[["ymax"]])
  }

  inner_reef_boundary_xy <- inner_reef_utm %>%
    st_segmentize(dfMaxLength = strip_height_m) %>%
    st_boundary() %>%
    st_coordinates() %>%
    as_tibble()

  strip_east_edges <- tibble(
    strip_index = seq_len(length(y_breaks) - 1),
    y_min = head(y_breaks, -1),
    y_max = tail(y_breaks, -1)
  ) %>%
    mutate(east_x = map2_dbl(y_min, y_max, function(y_min_i, y_max_i) {
      strip_points <- inner_reef_boundary_xy %>%
        filter(Y >= y_min_i, Y < y_max_i)

      if (nrow(strip_points) == 0) {
        return(NA_real_)
      }

      max(strip_points$X, na.rm = TRUE)
    }))

  valid_strips <- which(is.finite(strip_east_edges$east_x))
  if (length(valid_strips) == 0) {
    stop("Could not build inshore Sand/Scattered Rock mask from Inner Reef polygons.", call. = FALSE)
  }

  run_id <- cumsum(c(TRUE, diff(valid_strips) != 1))
  mask_polygons <- map(split(valid_strips, run_id), function(strip_rows) {
    strip_run <- strip_east_edges[strip_rows, ]
    east_edge_coords <- map(seq(nrow(strip_run), 1), function(j) {
      rbind(
        c(strip_run$east_x[[j]], strip_run$y_max[[j]]),
        c(strip_run$east_x[[j]], strip_run$y_min[[j]])
      )
    }) %>%
      do.call(rbind, .)

    coords <- rbind(
      c(mask_bbox[["xmin"]], strip_run$y_min[[1]]),
      c(mask_bbox[["xmin"]], strip_run$y_max[[nrow(strip_run)]]),
      east_edge_coords,
      c(mask_bbox[["xmin"]], strip_run$y_min[[1]])
    )

    st_polygon(list(coords))
  })

  st_sfc(mask_polygons, crs = st_crs(inner_reef_utm)) %>%
    st_make_valid() %>%
    st_union() %>%
    st_make_valid()
}

inner_reef_inshore_mask_utm <- make_inner_reef_inshore_mask(
  polygons_utm %>%
    filter(Type == "Inner Reef") %>%
    st_union() %>%
    st_make_valid(),
  conch_habitat_bbox,
  sand_scattered_inshore_strip_height_m
)

polygons_conch_habitat_base_utm <- polygons_utm %>%
  mutate(
    conch_habitat = case_when(
      Type %in% c("Artificial", "Nearshore Ridge Complex") ~ "Nearshore Ridge Complex",
      Type == "Inner Reef" ~ "Inner Reef",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(conch_habitat))

sand_scattered_conch_habitat_utm <- polygons_utm %>%
  filter(Type %in% c("Sand", "Scattered Rock in Unconsolidated Sediment")) %>%
  st_intersection(inner_reef_inshore_mask_utm) %>%
  mutate(conch_habitat = "Sand/Scattered Rock") %>%
  filter(!st_is_empty(.))

polygons_conch_habitat_utm <- bind_rows(
  polygons_conch_habitat_base_utm,
  sand_scattered_conch_habitat_utm
)

target_habitats_utm <- polygons_conch_habitat_utm %>%
  filter(!is.na(conch_habitat)) %>%
  group_by(conch_habitat) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

extract_placemark_info <- function(kml_file, source_label) {
  ns <- c(kml = "http://www.opengis.net/kml/2.2")
  kml <- read_xml(kml_file)
  placemarks <- xml_find_all(kml, ".//kml:Placemark", ns)

  map(placemarks, function(pm) {
    name <- xml_text(xml_find_first(pm, ".//kml:name", ns))
    folder_node <- xml_parent(xml_parent(pm))
    folder_name <- xml_text(xml_find_first(folder_node, ".//kml:name", ns))
    if (is.na(folder_name) || folder_name == "") folder_name <- "Top Level"

    coord_nodes <- xml_find_all(pm, ".//kml:coordinates", ns)
    rings <- map(coord_nodes, function(cn) {
      coords <- strsplit(trimws(xml_text(cn)), "\\s+")[[1]]
      coords_mat <- do.call(rbind, lapply(coords, function(x) {
        as.numeric(strsplit(x, ",")[[1]][1:2])
      }))
      if (nrow(coords_mat) >= 3) coords_mat else NULL
    }) %>%
      compact()

    if (length(rings) == 0) return(NULL)

    tibble(
      folder = folder_name,
      name = name,
      geometry = st_sfc(st_multipolygon(lapply(rings, function(r) list(r))), crs = 4326),
      source = source_label
    )
  }) %>%
    compact() %>%
    bind_rows() %>%
    st_as_sf()
}

impact_zones_sf <- extract_placemark_info(impact_kml, "Impact Zone")
nmfs_sf <- extract_placemark_info(nmfs_kml, "NMFS Consideration Area")

impact_zones_utm <- bind_rows(impact_zones_sf, nmfs_sf) %>%
  mutate(
    name = if_else(
      is.na(name) | name == "",
      str_remove(str_remove(folder, "^Scenario_2_"), "\\.shp$"),
      name
    ),
    scenario = if_else(
      name %in% c("North", "South", "East", "Nearshore"),
      "NMFS Consideration",
      "Scenario 2"
    )
  ) %>%
  st_make_valid() %>%
  st_transform(utm_crs) %>%
  group_by(name, scenario, source) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

channel_poly_utm <- impact_zones_utm %>%
  filter(str_detect(tolower(as.character(name)), "channel"))

channel_bbox <- st_bbox(channel_poly_utm)
habitat_bbox <- st_bbox(polygons_utm)

impact_monitoring_box <- st_as_sfc(st_bbox(
  c(
    xmin = habitat_bbox[["xmin"]],
    xmax = habitat_bbox[["xmax"]],
    ymin = channel_bbox[["ymin"]] - 1200,
    ymax = channel_bbox[["ymax"]] + 1200
  ),
  crs = st_crs(polygons_utm)
))

impact_monitoring_habitat_utm <- st_intersection(
  impact_monitoring_box,
  st_union(polygons_utm)
) %>%
  st_union() %>%
  st_make_valid()

existing_impact_zone_hull_utm <- impact_zones_utm %>%
  st_union() %>%
  st_buffer(20) %>%
  st_union() %>%
  st_buffer(-20) %>%
  st_make_valid()

impact_monitoring_area_utm <- st_sf(
  name = "Impact Monitoring Area",
  scenario = "Impact Monitoring Area",
  source = "Generated",
  geometry = st_sfc(
    st_difference(impact_monitoring_habitat_utm, existing_impact_zone_hull_utm),
    crs = st_crs(polygons_utm)
  )
) %>%
  st_make_valid()

impact_zones_utm <- bind_rows(
  impact_zones_utm,
  impact_monitoring_area_utm
)

impact_zone_area_utm <- impact_zones_utm %>%
  st_union() %>%
  st_make_valid()

make_exterior_lines <- function(x) {
  polygon_parts <- x %>%
    st_collection_extract("POLYGON") %>%
    st_cast("POLYGON", warn = FALSE)

  st_sf(
    name = polygon_parts$name,
    scenario = polygon_parts$scenario,
    source = polygon_parts$source,
    geometry = st_sfc(
      map(st_geometry(polygon_parts), ~ st_linestring(.x[[1]])),
      crs = st_crs(polygon_parts)
    )
  )
}

impact_zone_plot_outlines_utm <- bind_rows(
  impact_zones_utm %>%
    filter(name != "Impact Monitoring Area") %>%
    st_boundary() %>%
    st_as_sf(),
  impact_monitoring_area_utm %>%
    make_exterior_lines()
) %>%
  st_make_valid()

radial_sites <- readxl::read_xlsx(conch_xlsx, sheet = radial_sheet)

radial_counts <- read.csv(
  radial_counts_csv,
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
) %>%
  filter(!is.na(`Site Code`), `Site Code` != "") %>%
  mutate(
    juvenile_count = as.numeric(`# Juv`),
    adult_count = as.numeric(`# Adult`),
    breeding_activity_count = as.numeric(ReproductiveActivity)
  )

radial_abundance <- radial_sites %>%
  select(`Site Code`, Lat, Lon, `Depth (ft)`) %>%
  left_join(
    radial_counts %>%
      select(`Site Code`, juvenile_count, adult_count, breeding_activity_count),
    by = "Site Code"
  )

stopifnot(
  nrow(radial_abundance) == nrow(radial_sites),
  sum(is.na(radial_abundance$adult_count)) == 0,
  sum(is.na(radial_abundance$juvenile_count)) == 0
)

radial_observations_sf_utm <- radial_abundance %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  mutate(
    observation_source = "PENIP radial survey",
    survey_area_m2 = radial_survey_area_m2
  ) %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

radial_breeding_activity_sites_sf_utm <- radial_abundance %>%
  filter(breeding_activity_count > 0) %>%
  mutate(
    breeding_activity_source = "PENIP radial survey",
    breeding_activity_type = "Breeding activity"
  ) %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  select(breeding_activity_source, breeding_activity_type, geometry)

roving_breeding_activity_sites_sf_utm <- readxl::read_xlsx(roving_xlsx, sheet = 1) %>%
  mutate(
    latitude = as.numeric(`Lat (dec deg)`),
    longitude = as.numeric(`Long (dec deg)`),
    comments_clean = str_squish(str_to_lower(replace_na(Comments, ""))),
    eggs_observed = str_detect(comments_clean, "egg"),
    mating_observed = str_detect(comments_clean, "mating|mate"),
    breeding_activity_type = case_when(
      eggs_observed & mating_observed ~ "Eggs and mating",
      eggs_observed ~ "Eggs",
      mating_observed ~ "Mating",
      TRUE ~ NA_character_
    ),
    breeding_activity_source = "Roving diver survey"
  ) %>%
  filter(!is.na(latitude), !is.na(longitude), eggs_observed | mating_observed) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  select(breeding_activity_source, breeding_activity_type, geometry)

breeding_activity_sites_sf_utm <- bind_rows(
  radial_breeding_activity_sites_sf_utm,
  roving_breeding_activity_sites_sf_utm
)

breeding_activity_summary <- breeding_activity_sites_sf_utm %>%
  st_drop_geometry() %>%
  count(breeding_activity_source, breeding_activity_type, name = "observations")

radial_summary <- radial_observations_sf_utm %>%
  st_drop_geometry() %>%
  summarize(
    radial_surveys = n(),
    radial_area_m2 = sum(survey_area_m2),
    adult_count = sum(adult_count),
    juvenile_count = sum(juvenile_count)
  )

knitr::kable(radial_summary, digits = 2)
knitr::kable(breeding_activity_summary)
```

# Drift Transects

Drift transects are converted to 4 m-wide rectangles and then divided into `r drift_segment_length_m` m segments. Observed conch from the drift survey point sheet are assigned to the nearest segment within the same transect.

```{r drift-segments}
make_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(st_polygon())
  }

  unit_direction <- direction / transect_length
  perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

  corners <- rbind(
    start_point + perpendicular,
    end_point + perpendicular,
    end_point - perpendicular,
    start_point - perpendicular,
    start_point + perpendicular
  )

  st_polygon(list(corners))
}

make_segment_rectangles <- function(start_point, end_point, segment_length_m, half_width_m) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) {
    return(tibble())
  }

  unit_direction <- direction / transect_length
  segment_starts <- seq(0, transect_length, by = segment_length_m)
  segment_ends <- pmin(segment_starts + segment_length_m, transect_length)
  keep_segments <- segment_ends > segment_starts

  tibble(
    segment_index = seq_along(segment_starts)[keep_segments],
    segment_start_m = segment_starts[keep_segments],
    segment_end_m = segment_ends[keep_segments],
    geometry = map2(
      segment_start_m,
      segment_end_m,
      ~ make_transect_rectangle(
        start_point + unit_direction * .x,
        start_point + unit_direction * .y,
        half_width_m = half_width_m
      )
    )
  )
}

drift_surveys <- readxl::read_xlsx(conch_xlsx, sheet = drift_sheet) %>%
  mutate(
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon), !is.na(end_lat), !is.na(end_lon)) %>%
  mutate(
    .record = row_number(),
    transect_group = sub(" cont\\.$", "", Transect)
  )

drift_start_sf <- drift_surveys %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

drift_end_sf <- drift_surveys %>%
  st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

start_xy <- st_coordinates(drift_start_sf)
end_xy <- st_coordinates(drift_end_sf)

drift_rectangles <- st_sf(
  drift_surveys %>% select(-start_lat, -start_lon, -end_lat, -end_lon),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ make_transect_rectangle(start_xy[.x, ], end_xy[.y, ], half_width_m = drift_half_width_m)
    ),
    crs = utm_crs
  )
) %>%
  filter(!st_is_empty(.)) %>%
  mutate(transect_area_m2 = as.numeric(st_area(geometry)))

drift_transect_lines <- st_sf(
  drift_surveys %>% select(.record, Transect, transect_group, Line),
  geometry = st_sfc(
    map2(
      seq_len(nrow(start_xy)),
      seq_len(nrow(end_xy)),
      ~ st_linestring(rbind(start_xy[.x, ], end_xy[.y, ]))
    ),
    crs = utm_crs
  )
)

drift_segment_table <- map_dfr(seq_len(nrow(drift_surveys)), function(i) {
  make_segment_rectangles(
    start_xy[i, ],
    end_xy[i, ],
    segment_length_m = drift_segment_length_m,
    half_width_m = drift_half_width_m
  ) %>%
    mutate(
      .record = drift_surveys$.record[[i]],
      Transect = drift_surveys$Transect[[i]],
      transect_group = drift_surveys$transect_group[[i]],
      Line = drift_surveys$Line[[i]],
      segment_id = paste(.record, segment_index, sep = "_")
    )
})

drift_segments <- st_as_sf(
  drift_segment_table,
  sf_column_name = "geometry",
  crs = utm_crs
) %>%
  select(segment_id, .record, Transect, transect_group, Line, segment_index, segment_start_m, segment_end_m) %>%
  mutate(segment_area_m2 = as.numeric(st_area(geometry)))

drift_segment_summary <- tibble(
  drift_transects = nrow(drift_rectangles),
  drift_segment_count = nrow(drift_segments),
  drift_area_m2 = sum(drift_rectangles$transect_area_m2),
  drift_segment_area_m2 = sum(drift_segments$segment_area_m2)
)

knitr::kable(drift_segment_summary, digits = 2)
```

```{r drift-position-counts}
conch_drift_positions <- readxl::read_xlsx(conch_xlsx, sheet = drift_position_sheet) %>%
  mutate(
    .record = row_number(),
    lat = as.numeric(`Lat (dec deg)`),
    lon = as.numeric(`Long (dec deg)`),
    animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`),
    animal_stage = case_when(
      grepl("^A", animal_id) ~ "Adult",
      grepl("^J", animal_id) ~ "Juvenile",
      TRUE ~ NA_character_
    ),
    transect_group = sub("([AJ][0-9]+)$", "", `WPT (B/E)`)
  ) %>%
  filter(!is.na(lat), !is.na(lon), !is.na(animal_stage))

conch_drift_positions_sf_utm <- conch_drift_positions %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

assign_to_nearest_segment <- function(points_sf, segments_sf) {
  map_dfr(seq_len(nrow(points_sf)), function(i) {
    point_i <- points_sf[i, ]
    point_dat <- point_i %>% st_drop_geometry()
    candidate_segments <- segments_sf %>%
      filter(transect_group == point_dat$transect_group[[1]])

    if (nrow(candidate_segments) == 0) {
      return(tibble(
        .record = point_dat$.record[[1]],
        animal_stage = point_dat$animal_stage[[1]],
        transect_group = point_dat$transect_group[[1]],
        segment_id = NA_character_,
        segment_index = NA_integer_,
        segment_area_m2 = NA_real_,
        distance_to_segment_m = NA_real_
      ))
    }

    distances_m <- as.numeric(st_distance(point_i, candidate_segments))
    nearest_i <- which.min(distances_m)
    nearest_segment <- candidate_segments[nearest_i, ] %>% st_drop_geometry()

    tibble(
      .record = point_dat$.record[[1]],
      animal_stage = point_dat$animal_stage[[1]],
      transect_group = point_dat$transect_group[[1]],
      segment_id = nearest_segment$segment_id[[1]],
      segment_index = nearest_segment$segment_index[[1]],
      segment_area_m2 = nearest_segment$segment_area_m2[[1]],
      distance_to_segment_m = distances_m[[nearest_i]]
    )
  })
}

drift_observation_segment_assignments <- assign_to_nearest_segment(
  conch_drift_positions_sf_utm,
  drift_segments
)

drift_segment_counts <- drift_observation_segment_assignments %>%
  filter(!is.na(segment_id)) %>%
  count(segment_id, animal_stage, name = "n_animals") %>%
  pivot_wider(
    names_from = animal_stage,
    values_from = n_animals,
    values_fill = 0
  )

drift_segment_observations_sf_utm <- drift_segments %>%
  st_point_on_surface() %>%
  left_join(drift_segment_counts, by = "segment_id") %>%
  mutate(
    Adult = replace_na(Adult, 0),
    Juvenile = replace_na(Juvenile, 0),
    observation_source = "PENIP drift segment",
    survey_area_m2 = segment_area_m2,
    adult_count = Adult,
    juvenile_count = Juvenile
  ) %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

drift_observation_assignment_summary <- tibble(
  observed_conch_total = nrow(conch_drift_positions),
  assigned_to_segment = sum(!is.na(drift_observation_segment_assignments$segment_id)),
  unassigned_to_segment = observed_conch_total - assigned_to_segment,
  median_distance_to_assigned_segment_m = median(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  ),
  max_distance_to_assigned_segment_m = max(
    drift_observation_segment_assignments$distance_to_segment_m,
    na.rm = TRUE
  )
)

knitr::kable(drift_observation_assignment_summary, digits = 2)
```

```{r drift-observation-locations-map, fig.width=7, fig.height=7}
drift_observation_map_padding_m <- 150
drift_observation_map_bbox <- st_bbox(c(
  st_geometry(drift_transect_lines),
  st_geometry(conch_drift_positions_sf_utm)
))

drift_observation_map_xlim <- drift_observation_map_bbox[c("xmin", "xmax")] +
  c(-drift_observation_map_padding_m, drift_observation_map_padding_m)
drift_observation_map_ylim <- drift_observation_map_bbox[c("ymin", "ymax")] +
  c(-drift_observation_map_padding_m, drift_observation_map_padding_m)

drift_observation_map_extent_utm <- st_as_sfc(st_bbox(
  c(
    xmin = drift_observation_map_xlim[[1]],
    xmax = drift_observation_map_xlim[[2]],
    ymin = drift_observation_map_ylim[[1]],
    ymax = drift_observation_map_ylim[[2]]
  ),
  crs = st_crs(polygons_utm)
))

drift_observation_habitats_utm <- polygons_utm %>%
  st_filter(drift_observation_map_extent_utm, .predicate = st_intersects)

drift_observation_map <- ggplot() +
  geom_sf(
    data = drift_observation_habitats_utm,
    aes(fill = Type),
    color = "white",
    linewidth = 0.05,
    alpha = 0.55
  ) +
  geom_sf(
    data = drift_transect_lines,
    color = "white",
    linewidth = 1.1,
    alpha = 0.8
  ) +
  geom_sf(
    data = drift_transect_lines,
    color = "black",
    linewidth = 0.35,
    alpha = 0.9
  ) +
  geom_sf(
    data = conch_drift_positions_sf_utm,
    aes(color = animal_stage),
    size = 2,
    alpha = 0.45
  ) +
  scale_fill_manual(values = type_cols, na.value = "gray80", name = "Habitat") +
  scale_color_manual(values = age_class_cols[c("Adult", "Juvenile")], name = "Observed conch") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = drift_observation_map_xlim,
    ylim = drift_observation_map_ylim,
    expand = FALSE,
    datum = NA
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

drift_observation_map
```

# Broad-Scale Survey Counts

The Berry broad-scale survey counts are classified as adult = `Adult + Old` and juvenile/non-adult = `Juv + Sub + Young`. Each broad-scale site is treated as 240 m2 of surveyed area, from two 30 m transects surveyed 2 m on each side of the tape.

```{r berry-broad-data}
berry_broad_raw <- readxl::read_xlsx(berry_broad_xlsx, sheet = 1)

berry_broad_sites <- berry_broad_raw %>%
  mutate(
    .record = row_number(),
    adult_count = as.numeric(Adult) + as.numeric(Old),
    juvenile_count = as.numeric(Juv) + as.numeric(Sub) + as.numeric(Young),
    total_count = adult_count + juvenile_count,
    survey_area_m2 = berry_broad_site_area_m2
  ) %>%
  select(.record, Site, LatDD, LonDD, Adult, Juv, Old, Sub, Young, adult_count, juvenile_count, total_count, survey_area_m2)

berry_observations_sf_utm <- berry_broad_sites %>%
  st_as_sf(coords = c("LonDD", "LatDD"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs) %>%
  mutate(observation_source = "Berry broad-scale site") %>%
  select(observation_source, survey_area_m2, adult_count, juvenile_count, geometry)

berry_broad_summary <- berry_broad_sites %>%
  summarize(
    berry_sites = n(),
    berry_area_m2 = sum(survey_area_m2, na.rm = TRUE),
    berry_adults = sum(adult_count, na.rm = TRUE),
    berry_juveniles = sum(juvenile_count, na.rm = TRUE),
    berry_total = sum(total_count, na.rm = TRUE)
  )

knitr::kable(berry_broad_summary, digits = 2)
```

# Data Collection Overview

This figure shows the spatial distribution of the three conch survey inputs using the same extent, habitat colors, and survey-point style as Figure 1B in the coral report.

```{r data-collection-overview-map, fig.width=7, fig.height=3.5}
survey_map_theme <- theme_minimal(base_size = 8) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.size = unit(0.3, "cm"),
    legend.text = element_text(size = 6),
    legend.title = element_blank(),
    strip.text = element_blank(),
    strip.background = element_blank(),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(2, 0, 0, 0),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

survey_effort_levels <- c(
  "Berry broad-scale survey",
  "PENIP radial survey",
  "PENIP drift survey"
)

survey_effort_panel_labels <- c(
  "A",
  "B",
  "C"
)

survey_effort_panel_annotations <- tibble(
  survey_effort = factor(survey_effort_levels, levels = survey_effort_levels),
  panel_label = survey_effort_panel_labels,
  x = survey_effort_xlim[[1]] + 0.001,
  y = survey_effort_ylim[[2]] - 0.0015
)

survey_effort_polygons <- crossing(survey_effort = survey_effort_levels) %>%
  pull(survey_effort) %>%
  map_dfr(function(panel_name) {
    polygons_final %>%
      mutate(survey_effort = panel_name)
  }) %>%
  st_as_sf() %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels))

berry_effort_points <- berry_observations_sf_utm %>%
  st_transform(4326) %>%
  mutate(survey_effort = "Berry broad-scale survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

radial_effort_points <- radial_observations_sf_utm %>%
  st_transform(4326) %>%
  mutate(survey_effort = "PENIP radial survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

drift_effort_lines <- drift_transect_lines %>%
  st_transform(4326) %>%
  mutate(survey_effort = "PENIP drift survey") %>%
  mutate(survey_effort = factor(survey_effort, levels = survey_effort_levels)) %>%
  select(survey_effort, geometry)

data_collection_map <- ggplot() +
  geom_sf(
    data = survey_effort_polygons,
    aes(fill = Type),
    color = NA,
    linewidth = 0.1,
    alpha = 0.5
  ) +
  geom_sf(
    data = drift_effort_lines,
    color = "black",
    linewidth = 0.25,
    alpha = 0.7
  ) +
  geom_sf(
    data = berry_effort_points,
    pch = 4,
    alpha = 0.6,
    size = 0.65,
    stroke = 0.3
  ) +
  geom_sf(
    data = radial_effort_points,
    pch = 4,
    alpha = 0.6,
    size = 0.65,
    stroke = 0.3
  ) +
  geom_text(
    data = survey_effort_panel_annotations,
    aes(x = x, y = y, label = panel_label),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3
  ) +
  scale_fill_manual(values = type_cols, na.value = "gray80") +
  facet_wrap(~ survey_effort, ncol = 3) +
  coord_sf(
    xlim = survey_effort_xlim,
    ylim = survey_effort_ylim,
    expand = FALSE
  ) +
  guides(fill = guide_legend(nrow = 2, override.aes = list(alpha = 0.5))) +
  survey_map_theme

data_collection_map
```

# Analysis Units and Prediction Grid

The abundance model combines radial surveys, segmented drift transects, and broad-scale survey sites as observation units. Adult and juvenile densities are modeled separately. Prediction is limited to Inner Reef, Nearshore Ridge Complex, and the inshore portion of Sand/Scattered Rock habitat within the survey-effort map extent; Artificial habitat is included with Nearshore Ridge Complex, and Sand/Scattered Rock is clipped strip-wise to the area west of the local eastern edge of the mapped Inner Reef polygons.

```{r conch-observations}
conch_observations_sf_utm <- bind_rows(
  radial_observations_sf_utm,
  drift_segment_observations_sf_utm,
  berry_observations_sf_utm
) %>%
  mutate(
    adult_density_m2 = adult_count / survey_area_m2,
    juvenile_density_m2 = juvenile_count / survey_area_m2
  )

observation_summary <- conch_observations_sf_utm %>%
  st_drop_geometry() %>%
  group_by(observation_source) %>%
  summarize(
    observations = n(),
    area_m2 = sum(survey_area_m2, na.rm = TRUE),
    adult_count = sum(adult_count, na.rm = TRUE),
    juvenile_count = sum(juvenile_count, na.rm = TRUE),
    adult_density_per_ha = adult_count / area_m2 * 10000,
    juvenile_density_per_ha = juvenile_count / area_m2 * 10000,
    .groups = "drop"
  )

knitr::kable(observation_summary, digits = 2)
```

```{r prediction-grid}
survey_effort_extent_utm <- st_as_sfc(st_bbox(
  c(
    xmin = survey_effort_xlim[[1]],
    xmax = survey_effort_xlim[[2]],
    ymin = survey_effort_ylim[[1]],
    ymax = survey_effort_ylim[[2]]
  ),
  crs = st_crs(4326)
)) %>%
  st_transform(utm_crs)

prediction_area <- st_intersection(
  target_habitats_utm %>% st_union() %>% st_make_valid(),
  survey_effort_extent_utm
) %>%
  st_make_valid()

prediction_bbox <- st_bbox(survey_effort_extent_utm)

conch_data_bbox <- st_bbox(conch_observations_sf_utm)
prediction_plot_bbox <- prediction_bbox
prediction_plot_bbox[c("ymin", "ymax")] <- conch_data_bbox[c("ymin", "ymax")] + c(-150, 150)

prediction_grid <- st_make_grid(
  prediction_area,
  cellsize = prediction_cellsize_m,
  what = "polygons",
  square = TRUE
) %>%
  st_sf(geometry = .) %>%
  st_intersection(st_sf(geometry = prediction_area)) %>%
  mutate(cell_area_m2 = as.numeric(st_area(geometry))) %>%
  filter(cell_area_m2 > 0)

impact_zone_order <- c(
  "Channel",
  "Side Slopes",
  "Depth_10cm",
  "Depth_5cm",
  "Depth_1cm",
  "Depth_0_5cm",
  "Depth_0_1cm",
  "Nearshore",
  "North",
  "East",
  "South",
  "Impact Monitoring Area"
)

impact_zone_labels <- c(
  "Channel" = "Channel",
  "Side Slopes" = "Side Slopes",
  "Depth_10cm" = "Sediment deposition >10 cm",
  "Depth_5cm" = "Sediment deposition 5-10 cm",
  "Depth_1cm" = "Sediment deposition 1-5 cm",
  "Depth_0_5cm" = "Sediment deposition 0.5-1 cm",
  "Depth_0_1cm" = "Sediment deposition 0.1-0.5 cm",
  "Nearshore" = "Nearshore",
  "North" = "North",
  "East" = "East",
  "South" = "South",
  "Impact Monitoring Area" = "Rest of impact monitoring area"
)

impact_zone_cell_area <- st_intersection(
  prediction_grid %>%
    mutate(cell_id = row_number()) %>%
    select(cell_id),
  impact_zones_utm %>% select(name)
) %>%
  mutate(zone_cell_area_m2 = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  filter(name %in% impact_zone_order)

breeding_activity_buffer_utm <- breeding_activity_sites_sf_utm %>%
  st_buffer(breeding_activity_radius_m) %>%
  st_union() %>%
  st_make_valid()

target_habitat_outline <- st_intersection(
  target_habitats_utm,
  st_sf(geometry = survey_effort_extent_utm)
) %>%
  st_make_valid()

make_inner_boundary_lines <- function(x, offset_m = 15) {
  map_dfr(seq_len(nrow(x)), function(i) {
    feature_i <- x[i, ]
    inner_i <- st_buffer(feature_i, -offset_m)

    if (all(st_is_empty(inner_i))) {
      inner_i <- feature_i
    }

    inner_i %>%
      st_boundary() %>%
      st_collection_extract("LINESTRING")
  }) %>%
    st_as_sf() %>%
    st_make_valid()
}

target_habitat_inner_lines <- make_inner_boundary_lines(target_habitat_outline, offset_m = 15)
```

# Kriging Analysis

Ordinary kriging is used for the final density and abundance estimates. The model is fit separately for adults and juveniles, using the same radial-survey, drift-segment, and broad-scale survey observation units defined above.
Density is modeled as conch per hectare, local-neighborhood parameters are selected by leave-one-out cross-validation, and negative kriging predictions are clipped to zero before abundance totals are calculated.

```{r kriging-functions}
prepare_kriging_observations <- function(observations_sf, density_column) {
  observation_points <- observations_sf %>%
    st_point_on_surface() %>%
    mutate(
      density_per_ha = .data[[density_column]] * 10000
    ) %>%
    filter(is.finite(density_per_ha))

  observation_xy <- st_coordinates(observation_points)

  observation_points %>%
    st_drop_geometry() %>%
    mutate(
      kriging_x = observation_xy[, "X"],
      kriging_y = observation_xy[, "Y"]
    ) %>%
    group_by(kriging_x, kriging_y) %>%
    summarize(
      density_per_ha = sum(density_per_ha * survey_area_m2, na.rm = TRUE) / sum(survey_area_m2, na.rm = TRUE),
      survey_area_m2 = sum(survey_area_m2, na.rm = TRUE),
      observation_units = n(),
      .groups = "drop"
    ) %>%
    st_as_sf(coords = c("kriging_x", "kriging_y"), crs = st_crs(observations_sf))
}

fit_best_variogram <- function(observations_sp) {
  empirical_variogram <- gstat::variogram(
    density_per_ha ~ 1,
    observations_sp
  )

  initial_partial_sill <- max(stats::var(observations_sp$density_per_ha, na.rm = TRUE) * 0.75, 0.001)
  initial_nugget <- max(stats::var(observations_sp$density_per_ha, na.rm = TRUE) * 0.25, 0.001)
  initial_range <- max(max(empirical_variogram$dist, na.rm = TRUE) / 3, 1)

  candidate_models <- c("Exp", "Sph", "Gau")

  fitted_models <- map(candidate_models, function(model_name) {
    candidate <- gstat::vgm(
      psill = initial_partial_sill,
      model = model_name,
      range = initial_range,
      nugget = initial_nugget
    )

    tryCatch(
      gstat::fit.variogram(empirical_variogram, candidate),
      error = function(e) NULL,
      warning = function(w) suppressWarnings(gstat::fit.variogram(empirical_variogram, candidate))
    )
  })

  keep_models <- !map_lgl(fitted_models, is.null)

  if (!any(keep_models)) {
    stop("No valid variogram model could be fit for kriging.")
  }

  fitted_models <- fitted_models[keep_models]
  model_scores <- map_dbl(fitted_models, ~ attr(.x, "SSErr"))
  fitted_models[[which.min(model_scores)]]
}

predict_ordinary_kriging <- function(
  observations_sf,
  prediction_grid_sf,
  density_column,
  age_class,
  nmax = kriging_nmax,
  maxdist = kriging_maxdist_m,
  nmin = kriging_nmin
) {
  observations_for_kriging <- prepare_kriging_observations(observations_sf, density_column)
  observations_sp <- as(observations_for_kriging, "Spatial")

  prediction_points <- prediction_grid_sf %>%
    mutate(cell_id = row_number()) %>%
    st_point_on_surface()

  prediction_points_sp <- as(prediction_points, "Spatial")
  best_variogram <- fit_best_variogram(observations_sp)

  kriging_prediction <- gstat::krige(
    density_per_ha ~ 1,
    observations_sp,
    prediction_points_sp,
    model = best_variogram,
    nmax = nmax,
    maxdist = maxdist,
    nmin = nmin,
    debug.level = 0
  )

  prediction_grid_sf %>%
    mutate(
      cell_id = row_number(),
      age_class = age_class,
      density_per_ha_raw = kriging_prediction$var1.pred,
      kriging_variance_per_ha = kriging_prediction$var1.var,
      density_per_ha = pmax(density_per_ha_raw, 0),
      density_m2 = density_per_ha / 10000,
      estimated_total = density_m2 * cell_area_m2,
      variogram_model = as.character(best_variogram$model[[2]]),
      variogram_nugget = best_variogram$psill[[1]],
      variogram_partial_sill = best_variogram$psill[[2]],
      variogram_range_m = best_variogram$range[[2]],
      kriging_nmax = nmax,
      kriging_maxdist_m = maxdist,
      kriging_nmin = nmin
    )
}
```

# Kriging Cross-Validation

Leave-one-out cross-validation is used here to evaluate the same local kriging parameters considered in the sensitivity analysis. Adult and juvenile densities are evaluated separately, so the preferred kriging settings can differ between age classes. Exact duplicate observation coordinates are collapsed before cross-validation using the same area-weighted density calculation used for prediction.

```{r kriging-loocv, cache=TRUE}
dir.create(dirname(kriging_loocv_cache_rds), recursive = TRUE, showWarnings = FALSE)

kriging_cv_grid <- tidyr::expand_grid(
  kriging_nmax = kriging_sensitivity_nmax,
  kriging_maxdist_m = kriging_sensitivity_maxdist_m,
  kriging_nmin = kriging_sensitivity_nmin
) %>%
  mutate(
    kriging_maxdist_label = paste0(kriging_maxdist_m, " m"),
    kriging_parameter_label = paste0(
      "nmax ", kriging_nmax,
      ", maxdist ", kriging_maxdist_label,
      ", nmin ", kriging_nmin
    )
  )

calculate_kriging_loocv <- function(observations_sf, density_column, age_class, parameter_grid) {
  observations_for_kriging <- prepare_kriging_observations(observations_sf, density_column)
  observations_sp <- as(observations_for_kriging, "Spatial")
  best_variogram <- fit_best_variogram(observations_sp)
  null_rmse <- sqrt(mean(
    (observations_for_kriging$density_per_ha - mean(observations_for_kriging$density_per_ha, na.rm = TRUE))^2,
    na.rm = TRUE
  ))

  run_parameter_set <- function(parameter_row) {
    kriging_nmax <- parameter_row$kriging_nmax
    kriging_maxdist_m <- parameter_row$kriging_maxdist_m
    kriging_nmin <- parameter_row$kriging_nmin
    kriging_maxdist_label <- parameter_row$kriging_maxdist_label
    kriging_parameter_label <- parameter_row$kriging_parameter_label

    cv_prediction <- tryCatch(
      gstat::krige.cv(
        density_per_ha ~ 1,
        observations_sp,
        model = best_variogram,
        nfold = nrow(observations_for_kriging),
        nmax = kriging_nmax,
        maxdist = kriging_maxdist_m,
        nmin = kriging_nmin,
        debug.level = 0
      ),
      error = function(e) NULL
    )

    if (is.null(cv_prediction)) {
      return(tibble(
        age_class = age_class,
        kriging_nmax = kriging_nmax,
        kriging_maxdist_m = kriging_maxdist_m,
        kriging_maxdist_label = kriging_maxdist_label,
        kriging_nmin = kriging_nmin,
        kriging_parameter_label = kriging_parameter_label,
        observation_locations = nrow(observations_for_kriging),
        predicted_locations = 0,
        prediction_coverage_percent = 0,
        rmse_per_ha = NA_real_,
        mae_per_ha = NA_real_,
        mean_error_per_ha = NA_real_,
        rms_standardized_error = NA_real_,
        rmse_skill_vs_mean = NA_real_,
        variogram_model = as.character(best_variogram$model[[2]]),
        variogram_nugget = best_variogram$psill[[1]],
        variogram_partial_sill = best_variogram$psill[[2]],
        variogram_range_m = best_variogram$range[[2]]
      ))
    }

    cv_output <- as.data.frame(cv_prediction)

    tibble(
      age_class = age_class,
      kriging_nmax = kriging_nmax,
      kriging_maxdist_m = kriging_maxdist_m,
      kriging_maxdist_label = kriging_maxdist_label,
      kriging_nmin = kriging_nmin,
      kriging_parameter_label = kriging_parameter_label,
      observation_locations = nrow(cv_output),
      predicted_locations = sum(is.finite(cv_output$var1.pred)),
      prediction_coverage_percent = predicted_locations / observation_locations * 100,
      rmse_per_ha = sqrt(mean(cv_output$residual^2, na.rm = TRUE)),
      mae_per_ha = mean(abs(cv_output$residual), na.rm = TRUE),
      mean_error_per_ha = mean(cv_output$residual, na.rm = TRUE),
      rms_standardized_error = sqrt(mean(cv_output$zscore^2, na.rm = TRUE)),
      rmse_skill_vs_mean = 1 - rmse_per_ha / null_rmse,
      variogram_model = as.character(best_variogram$model[[2]]),
      variogram_nugget = best_variogram$psill[[1]],
      variogram_partial_sill = best_variogram$psill[[2]],
      variogram_range_m = best_variogram$range[[2]]
    )
  }

  parameter_rows <- split(parameter_grid, seq_len(nrow(parameter_grid)))

  if (.Platform$OS.type == "windows" || kriging_cv_cores <= 1) {
    map_dfr(parameter_rows, run_parameter_set)
  } else {
    parallel::mclapply(
      parameter_rows,
      run_parameter_set,
      mc.cores = min(kriging_cv_cores, length(parameter_rows))
    ) %>%
      bind_rows()
  }
}

make_kriging_loocv_cache_metadata <- function(observations_sf, parameter_grid) {
  observation_xy <- st_coordinates(st_point_on_surface(observations_sf))

  list(
    cache_version = 1,
    parameter_grid = parameter_grid %>%
      arrange(kriging_nmax, kriging_maxdist_m, kriging_nmin),
    observations = observations_sf %>%
      st_drop_geometry() %>%
      mutate(
        x = round(observation_xy[, "X"], 3),
        y = round(observation_xy[, "Y"], 3)
      ) %>%
      select(
        observation_source,
        survey_area_m2,
        adult_count,
        juvenile_count,
        adult_density_m2,
        juvenile_density_m2,
        x,
        y
      ) %>%
      arrange(observation_source, x, y, survey_area_m2, adult_count, juvenile_count)
  )
}

kriging_loocv_cache_metadata <- make_kriging_loocv_cache_metadata(
  conch_observations_sf_utm,
  kriging_cv_grid
)

kriging_loocv_cache <- if (file.exists(kriging_loocv_cache_rds)) {
  readRDS(kriging_loocv_cache_rds)
} else {
  NULL
}

if (
  is.list(kriging_loocv_cache) &&
    identical(kriging_loocv_cache$metadata, kriging_loocv_cache_metadata)
) {
  kriging_loocv_cache_status <- paste("Loaded kriging LOOCV results from", kriging_loocv_cache_rds)
  kriging_loocv_results <- kriging_loocv_cache$kriging_loocv_results
} else {
  kriging_loocv_cache_status <- paste("Recomputed kriging LOOCV results and saved", kriging_loocv_cache_rds)
  kriging_loocv_results <- bind_rows(
    calculate_kriging_loocv(
      conch_observations_sf_utm,
      "adult_density_m2",
      "Adult",
      kriging_cv_grid
    ),
    calculate_kriging_loocv(
      conch_observations_sf_utm,
      "juvenile_density_m2",
      "Juvenile",
      kriging_cv_grid
    )
  ) %>%
    arrange(age_class, rmse_per_ha)

  saveRDS(
    list(
      metadata = kriging_loocv_cache_metadata,
      kriging_loocv_results = kriging_loocv_results
    ),
    kriging_loocv_cache_rds
  )
}

message(kriging_loocv_cache_status)

kriging_loocv_best <- kriging_loocv_results %>%
  filter(is.finite(rmse_per_ha)) %>%
  group_by(age_class) %>%
  slice_min(rmse_per_ha, n = 1, with_ties = FALSE) %>%
  ungroup()

knitr::kable(
  kriging_loocv_best %>%
    transmute(
      `Age class` = age_class,
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_label,
      nmin = kriging_nmin,
      `Observation locations` = observation_locations,
      `Predicted locations` = predicted_locations,
      `Coverage %` = round(prediction_coverage_percent, 1),
      `RMSE conch/ha` = round(rmse_per_ha, 2),
      `MAE conch/ha` = round(mae_per_ha, 2),
      `Mean error conch/ha` = round(mean_error_per_ha, 2),
      `RMSE skill vs mean` = round(rmse_skill_vs_mean, 3)
    ),
  caption = "Best local ordinary kriging settings by leave-one-out RMSE. Adult and juvenile density models are evaluated separately."
)

knitr::kable(
  kriging_loocv_results %>%
    transmute(
      `Age class` = age_class,
      nmax = kriging_nmax,
      maxdist = kriging_maxdist_label,
      nmin = kriging_nmin,
      `Coverage %` = round(prediction_coverage_percent, 1),
      `RMSE conch/ha` = round(rmse_per_ha, 2),
      `MAE conch/ha` = round(mae_per_ha, 2),
      `Mean error conch/ha` = round(mean_error_per_ha, 2),
      `RMS standardized error` = round(rms_standardized_error, 2),
      `RMSE skill vs mean` = round(rmse_skill_vs_mean, 3)
    ),
  caption = "Leave-one-out kriging cross-validation for all local-neighborhood parameter sets. Lower RMSE and MAE indicate better held-out prediction; coverage shows the percentage of observation locations that could be predicted under each neighborhood constraint."
)
```

```{r kriging-loocv-plot, fig.width=9, fig.height=6}
kriging_loocv_plot <- ggplot(
  kriging_loocv_results,
  aes(
    x = kriging_maxdist_m,
    y = rmse_per_ha,
    color = factor(kriging_nmax),
    group = kriging_nmax
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(aes(shape = prediction_coverage_percent < 100), size = 2) +
  facet_grid(age_class ~ kriging_nmin, labeller = label_both) +
  scale_x_continuous(breaks = kriging_sensitivity_maxdist_m) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2", name = "nmax") +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 1),
    labels = c(`FALSE` = "100%", `TRUE` = "<100%"),
    name = "CV coverage"
  ) +
  labs(
    x = "Maximum search distance for kriging (m)",
    y = "Leave-one-out RMSE (conch per ha)",
    title = "Kriging parameter cross-validation"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

kriging_loocv_plot

ggsave(
  output_kriging_loocv_plot,
  kriging_loocv_plot,
  width = 9,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

```{r kriging-prediction}
kriging_best_adult_params <- kriging_loocv_best %>%
  filter(age_class == "Adult")

kriging_best_juvenile_params <- kriging_loocv_best %>%
  filter(age_class == "Juvenile")

adult_kriging_prediction <- predict_ordinary_kriging(
  conch_observations_sf_utm,
  prediction_grid,
  "adult_density_m2",
  "Adult",
  nmax = kriging_best_adult_params$kriging_nmax[[1]],
  maxdist = kriging_best_adult_params$kriging_maxdist_m[[1]],
  nmin = kriging_best_adult_params$kriging_nmin[[1]]
)

juvenile_kriging_prediction <- predict_ordinary_kriging(
  conch_observations_sf_utm,
  prediction_grid,
  "juvenile_density_m2",
  "Juvenile",
  nmax = kriging_best_juvenile_params$kriging_nmax[[1]],
  maxdist = kriging_best_juvenile_params$kriging_maxdist_m[[1]],
  nmin = kriging_best_juvenile_params$kriging_nmin[[1]]
)

kriging_prediction <- bind_rows(
  adult_kriging_prediction,
  juvenile_kriging_prediction
) %>%
  mutate(
    density_class = case_when(
      is.na(density_per_ha) ~ "Unpredicted",
      density_per_ha <= 0 ~ "0",
      density_per_ha <= 10 ~ "1-10",
      density_per_ha <= 50 ~ "10-50",
      density_per_ha <= 100 ~ "50-100",
      TRUE ~ ">100"
    ),
    density_class = factor(density_class, levels = c("0", "1-10", "10-50", "50-100", ">100", "Unpredicted"))
  )

kriging_variogram_summary <- kriging_prediction %>%
  st_drop_geometry() %>%
  distinct(
    age_class,
    variogram_model,
    variogram_nugget,
    variogram_partial_sill,
    variogram_range_m,
    kriging_nmax,
    kriging_maxdist_m,
    kriging_nmin
  )

kriging_population_summary <- kriging_prediction %>%
  st_drop_geometry() %>%
  group_by(age_class) %>%
  summarize(
    modeled_area_m2 = sum(cell_area_m2, na.rm = TRUE),
    modeled_area_ha = modeled_area_m2 / 10000,
    predicted_area_m2 = sum(cell_area_m2[!is.na(density_per_ha)], na.rm = TRUE),
    predicted_area_ha = predicted_area_m2 / 10000,
    estimated_conch = sum(estimated_total, na.rm = TRUE),
    density_per_predicted_ha = estimated_conch / predicted_area_ha,
    .groups = "drop"
  )

kriging_prediction_summary_xy <- st_coordinates(st_point_on_surface(kriging_prediction))

kriging_prediction_for_summary <- kriging_prediction %>%
  mutate(
    in_map_extent = kriging_prediction_summary_xy[, "X"] >= prediction_plot_bbox[["xmin"]] &
      kriging_prediction_summary_xy[, "X"] <= prediction_plot_bbox[["xmax"]] &
      kriging_prediction_summary_xy[, "Y"] >= prediction_plot_bbox[["ymin"]] &
      kriging_prediction_summary_xy[, "Y"] <= prediction_plot_bbox[["ymax"]]
  )

summarize_kriging_prediction_area <- function(prediction_data, summary_area) {
  prediction_data %>%
    st_drop_geometry() %>%
    group_by(age_class) %>%
    summarize(
      summary_area = summary_area,
      modeled_area_ha = sum(cell_area_m2, na.rm = TRUE) / 10000,
      predicted_area_ha = sum(cell_area_m2[!is.na(density_per_ha)], na.rm = TRUE) / 10000,
      unpredicted_area_ha = sum(cell_area_m2[is.na(density_per_ha)], na.rm = TRUE) / 10000,
      zero_density_area_ha = sum(cell_area_m2[!is.na(density_per_ha) & density_per_ha <= 0], na.rm = TRUE) / 10000,
      positive_density_area_ha = sum(cell_area_m2[!is.na(density_per_ha) & density_per_ha > 0], na.rm = TRUE) / 10000,
      .groups = "drop"
    )
}

kriging_map_cell_summary <- bind_rows(
  summarize_kriging_prediction_area(kriging_prediction_for_summary, "Full prediction grid"),
  summarize_kriging_prediction_area(
    kriging_prediction_for_summary %>% filter(in_map_extent),
    "Cropped map extent"
  )
) %>%
  select(summary_area, everything()) %>%
  arrange(age_class, summary_area)

kriging_map_unpredicted_summary <- kriging_prediction_for_summary %>%
  st_drop_geometry() %>%
  group_by(age_class) %>%
  summarize(
    unpredicted_cells_in_cropped_map_extent = sum(is.na(density_per_ha) & in_map_extent),
    .groups = "drop"
  )

knitr::kable(
  kriging_variogram_summary,
  digits = 3,
  caption = "Fitted local ordinary kriging variogram parameters for conch density in conch per hectare."
)

knitr::kable(
  kriging_population_summary,
  digits = 2,
  caption = "Kriging-estimated conch abundance over the habitat-limited prediction grid."
)

knitr::kable(
  kriging_map_cell_summary,
  digits = 2,
  caption = "Prediction-grid area represented in the final kriging map. Unpredicted cells are habitat cells where the selected kriging neighborhood did not contain enough observations; zero-density cells are predicted cells with an estimated density of zero."
)
```

```{r kriging-impact-zone-table}
kriging_habitat_prediction_area <- st_intersection(
  kriging_prediction,
  target_habitats_utm
) %>%
  mutate(intersection_area_m2 = as.numeric(st_area(geometry)))

kriging_habitat_zone_prediction_area <- st_intersection(
  kriging_habitat_prediction_area,
  impact_zones_utm %>% select(name, scenario)
) %>%
  mutate(
    zone_area_m2 = as.numeric(st_area(geometry)),
    kriging_variance_m2 = kriging_variance_per_ha / 10000^2,
    total_variance = kriging_variance_m2 * zone_area_m2^2
  )

kriging_adult_breeding_activity_zone_totals <- st_intersection(
  kriging_habitat_zone_prediction_area %>%
    filter(age_class == "Adult") %>%
    select(name, density_m2, kriging_variance_per_ha),
  st_sf(geometry = breeding_activity_buffer_utm)
) %>%
  mutate(
    breeding_zone_area_m2 = as.numeric(st_area(geometry)),
    kriging_variance_m2 = kriging_variance_per_ha / 10000^2,
    total_variance = kriging_variance_m2 * breeding_zone_area_m2^2
  ) %>%
  st_drop_geometry() %>%
  group_by(name) %>%
  summarize(
    estimated_adults_near_breeding = sum(density_m2 * breeding_zone_area_m2, na.rm = TRUE),
    estimate_se_adults_near_breeding = sqrt(sum(total_variance, na.rm = TRUE)),
    lower_95_adults_near_breeding = pmax(estimated_adults_near_breeding - 1.96 * estimate_se_adults_near_breeding, 0),
    upper_95_adults_near_breeding = estimated_adults_near_breeding + 1.96 * estimate_se_adults_near_breeding,
    .groups = "drop"
  )

kriging_impact_zone_totals <- kriging_habitat_zone_prediction_area %>%
  st_drop_geometry() %>%
  group_by(name, age_class) %>%
  summarize(
    estimated_conch = sum(density_m2 * zone_area_m2, na.rm = TRUE),
    estimate_se = sqrt(sum(total_variance, na.rm = TRUE)),
    lower_95 = pmax(estimated_conch - 1.96 * estimate_se, 0),
    upper_95 = estimated_conch + 1.96 * estimate_se,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = age_class,
    values_from = c(estimated_conch, estimate_se, lower_95, upper_95),
    values_fill = 0
  ) %>%
  left_join(kriging_adult_breeding_activity_zone_totals, by = "name") %>%
  right_join(tibble(name = impact_zone_order), by = "name") %>%
  mutate(
    across(where(is.numeric), ~ replace_na(., 0)),
    estimated_total = estimated_conch_Adult + estimated_conch_Juvenile,
    total_se = sqrt(estimate_se_Adult^2 + estimate_se_Juvenile^2),
    total_lower_95 = pmax(estimated_total - 1.96 * total_se, 0),
    total_upper_95 = estimated_total + 1.96 * total_se,
    impact_area = impact_zone_labels[name],
    name = factor(name, levels = impact_zone_order)
  ) %>%
  arrange(name) %>%
  mutate(
    cumulative_adults = cumsum(estimated_conch_Adult),
    cumulative_juveniles = cumsum(estimated_conch_Juvenile),
    cumulative_total = cumsum(estimated_total)
  ) %>%
  select(
    impact_area,
    estimated_conch_Adult,
    estimate_se_Adult,
    lower_95_Adult,
    upper_95_Adult,
    estimated_adults_near_breeding,
    estimate_se_adults_near_breeding,
    lower_95_adults_near_breeding,
    upper_95_adults_near_breeding,
    estimated_conch_Juvenile,
    estimate_se_Juvenile,
    lower_95_Juvenile,
    upper_95_Juvenile,
    estimated_total,
    total_se,
    total_lower_95,
    total_upper_95,
    cumulative_adults,
    cumulative_juveniles,
    cumulative_total
  )

kriging_impact_zone_totals_with_total <- bind_rows(
  kriging_impact_zone_totals,
  kriging_impact_zone_totals %>%
    summarize(
      impact_area = "Total",
      estimated_conch_Adult = sum(estimated_conch_Adult, na.rm = TRUE),
      estimate_se_Adult = sqrt(sum(estimate_se_Adult^2, na.rm = TRUE)),
      lower_95_Adult = pmax(estimated_conch_Adult - 1.96 * estimate_se_Adult, 0),
      upper_95_Adult = estimated_conch_Adult + 1.96 * estimate_se_Adult,
      estimated_adults_near_breeding = sum(estimated_adults_near_breeding, na.rm = TRUE),
      estimate_se_adults_near_breeding = sqrt(sum(estimate_se_adults_near_breeding^2, na.rm = TRUE)),
      lower_95_adults_near_breeding = pmax(estimated_adults_near_breeding - 1.96 * estimate_se_adults_near_breeding, 0),
      upper_95_adults_near_breeding = estimated_adults_near_breeding + 1.96 * estimate_se_adults_near_breeding,
      estimated_conch_Juvenile = sum(estimated_conch_Juvenile, na.rm = TRUE),
      estimate_se_Juvenile = sqrt(sum(estimate_se_Juvenile^2, na.rm = TRUE)),
      lower_95_Juvenile = pmax(estimated_conch_Juvenile - 1.96 * estimate_se_Juvenile, 0),
      upper_95_Juvenile = estimated_conch_Juvenile + 1.96 * estimate_se_Juvenile,
      estimated_total = sum(estimated_total, na.rm = TRUE),
      total_se = sqrt(sum(total_se^2, na.rm = TRUE)),
      total_lower_95 = pmax(estimated_total - 1.96 * total_se, 0),
      total_upper_95 = estimated_total + 1.96 * total_se,
      cumulative_adults = sum(estimated_conch_Adult, na.rm = TRUE),
      cumulative_juveniles = sum(estimated_conch_Juvenile, na.rm = TRUE),
      cumulative_total = sum(estimated_total, na.rm = TRUE)
    )
)

format_interval <- function(estimate, lower, upper) {
  paste0(
    format(round(estimate), big.mark = ",", scientific = FALSE, trim = TRUE),
    " (",
    format(round(lower), big.mark = ",", scientific = FALSE, trim = TRUE),
    "-",
    format(round(upper), big.mark = ",", scientific = FALSE, trim = TRUE),
    ")"
  )
}

kriging_impact_zone_totals_display <- kriging_impact_zone_totals_with_total %>%
  transmute(
    `Impact area` = impact_area,
    `Adult conch (approx. 95% CI)` = format_interval(estimated_conch_Adult, lower_95_Adult, upper_95_Adult),
    `Adult conch within 330 m of breeding activity (approx. 95% CI)` = format_interval(
      estimated_adults_near_breeding,
      lower_95_adults_near_breeding,
      upper_95_adults_near_breeding
    ),
    `Juvenile conch (approx. 95% CI)` = format_interval(estimated_conch_Juvenile, lower_95_Juvenile, upper_95_Juvenile),
    `Total conch (approx. 95% CI)` = format_interval(estimated_total, total_lower_95, total_upper_95),
    `Cumulative adult conch` = format(round(cumulative_adults), big.mark = ",", scientific = FALSE, trim = TRUE),
    `Cumulative juvenile conch` = format(round(cumulative_juveniles), big.mark = ",", scientific = FALSE, trim = TRUE),
    `Cumulative total conch` = format(round(cumulative_total), big.mark = ",", scientific = FALSE, trim = TRUE)
  )

knitr::kable(
  kriging_impact_zone_totals_display,
  caption = "Recommended kriging-estimated adult and juvenile conch abundance by impact area. Estimates use the age-specific best leave-one-out cross-validation settings; values in parentheses are approximate 95% intervals propagated from cell-level kriging prediction variances. The Total row sums adults and juveniles across all listed impact and monitoring areas. Intervals do not include spatial covariance among cells."
)
```



```{r kriging-map, fig.width=8, fig.height=4.5}
kriging_prediction_for_map <- kriging_prediction_for_summary %>%
  filter(in_map_extent)

kriging_map_panel_labels <- tibble(
  age_class = c("Adult", "Juvenile"),
  label = c("A", "B"),
  x = prediction_plot_bbox[["xmin"]] + 60,
  y = prediction_plot_bbox[["ymax"]] - 60
)

adult_breeding_activity_buffer_outline_utm <- st_sf(
  age_class = "Adult",
  geometry = st_boundary(breeding_activity_buffer_utm)
) %>%
  st_collection_extract("LINESTRING")

kriging_map <- ggplot() +
  geom_sf(
    data = kriging_prediction_for_map,
    aes(fill = density_class),
    color = NA,
    linewidth = 0,
    alpha = 0.95
  ) +
  geom_sf(
    data = target_habitat_inner_lines,
    aes(color = conch_habitat),
    linewidth = 0.8
  ) +
  geom_sf(
    data = adult_breeding_activity_buffer_outline_utm,
    aes(color = "Within 330 m of breeding activity"),
    linewidth = 0.55
  ) +
  geom_sf(
    data = impact_zone_plot_outlines_utm,
    fill = NA,
    color = "grey20",
    linewidth = 0.25
  ) +
  geom_text(
    data = kriging_map_panel_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    fontface = "bold",
    size = 6
  ) +
  facet_wrap(~ age_class) +
  scale_fill_manual(
    values = c(
      "0" = "#F7F7F7",
      "1-10" = "#D9D9D9",
      "10-50" = "#BDBDBD",
      "50-100" = "#737373",
      ">100" = "#252525",
      "Unpredicted" = "#F28E2B"
    ),
    drop = TRUE,
    name = "Conch per ha"
  ) +
  scale_color_manual(
    values = c(
      "Inner Reef" = type_cols[["Inner Reef"]],
      "Nearshore Ridge Complex" = type_cols[["Nearshore Ridge Complex"]],
      "Sand/Scattered Rock" = type_cols[["Sand"]],
      "Within 330 m of breeding activity" = "#7B3294"
    ),
    name = "Map overlays"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = prediction_plot_bbox[c("xmin", "xmax")],
    ylim = prediction_plot_bbox[c("ymin", "ymax")],
    expand = FALSE,
    datum = NA
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    strip.text = element_blank(),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

kriging_map

ggsave(output_kriging_map, kriging_map, width = 9, height = 4.5, dpi = 300, bg = "white")
```

# Shell Size and Lip Thickness

## Port Everglades Measurements

Port Everglades surveys are selected from the circular survey worksheet using latitude/longitude points that intersect the Port Everglades habitat polygon footprint. Individual shell measurements are then joined from the corresponding survey IDs in the circle conch measurement worksheet.

```{r import-size-lip-port-everglades}
port_circular_surveys <- readxl::read_xlsx(
  port_everglades_workbook,
  sheet = "CircularSurveys (10m rad)"
) %>%
  mutate(
    survey_id = as.numeric(Survey),
    lat = as.numeric(Lat),
    lon = as.numeric(Lon)
  ) %>%
  filter(!is.na(survey_id), !is.na(lat), !is.na(lon))

port_circular_surveys_sf <- port_circular_surveys %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

port_aoi <- st_union(polygons_utm)

port_everglades_surveys <- port_circular_surveys_sf[
  lengths(st_intersects(port_circular_surveys_sf, port_aoi)) > 0,
] %>%
  st_drop_geometry()

port_survey_ids <- sort(unique(port_everglades_surveys$survey_id))

port_measurements_raw <- readxl::read_xlsx(
  port_everglades_workbook,
  sheet = "CircleConchMeasurements"
)

port_everglades_conch <- port_measurements_raw %>%
  mutate(survey_id = as.numeric(Survey)) %>%
  filter(survey_id %in% port_survey_ids) %>%
  transmute(
    year = Year,
    survey_id = survey_id,
    location_label = Location,
    diver = Diver,
    site = Site,
    depth_ft = `depth(ft)`,
    quad = quad,
    radius_m = `radius(m)`,
    length_mm = suppressWarnings(as.numeric(`length(mm)`)),
    lip_mm = suppressWarnings(as.numeric(`lip(mm)`)),
    lip_measured = !is.na(lip_mm),
    lip_age_class = assign_lip_age_class(lip_mm),
    location = "Port Everglades"
  )

florida_keys_conch <- port_measurements_raw %>%
  mutate(survey_id = as.numeric(Survey)) %>%
  filter(!survey_id %in% port_survey_ids) %>%
  transmute(
    year = Year,
    survey_id = survey_id,
    location_label = Location,
    diver = Diver,
    site = Site,
    depth_ft = `depth(ft)`,
    quad = quad,
    radius_m = `radius(m)`,
    length_mm = suppressWarnings(as.numeric(`length(mm)`)),
    lip_mm = suppressWarnings(as.numeric(`lip(mm)`)),
    lip_measured = !is.na(lip_mm),
    lip_age_class = assign_lip_age_class(lip_mm),
    location = "Rest of Florida Keys"
  )

port_import_summary <- tibble(
  circular_surveys_total = nrow(port_circular_surveys),
  port_everglades_surveys = length(port_survey_ids),
  records = nrow(port_everglades_conch),
  measured_length = sum(!is.na(port_everglades_conch$length_mm)),
  missing_length = sum(is.na(port_everglades_conch$length_mm)),
  measured_lip = sum(!is.na(port_everglades_conch$lip_mm)),
  missing_lip = sum(is.na(port_everglades_conch$lip_mm)),
  juveniles_by_lip = sum(port_everglades_conch$lip_age_class == "Juvenile", na.rm = TRUE),
  subadults_by_lip = sum(port_everglades_conch$lip_age_class == "Sub-adult", na.rm = TRUE),
  adults_by_lip = sum(port_everglades_conch$lip_age_class == "Adult", na.rm = TRUE)
)

port_lip_age_counts <- port_everglades_conch %>%
  count(lip_age_class, name = "n") %>%
  mutate(percent = n / sum(n) * 100)

knitr::kable(port_import_summary, digits = 0)
knitr::kable(port_lip_age_counts, digits = 1)
```

```{r import-size-lip-florida-keys}
florida_keys_import_summary <- tibble(
  records = nrow(florida_keys_conch),
  measured_length = sum(!is.na(florida_keys_conch$length_mm)),
  missing_length = sum(is.na(florida_keys_conch$length_mm)),
  measured_lip = sum(!is.na(florida_keys_conch$lip_mm)),
  missing_lip = sum(is.na(florida_keys_conch$lip_mm)),
  juveniles_by_lip = sum(florida_keys_conch$lip_age_class == "Juvenile", na.rm = TRUE),
  subadults_by_lip = sum(florida_keys_conch$lip_age_class == "Sub-adult", na.rm = TRUE),
  adults_by_lip = sum(florida_keys_conch$lip_age_class == "Adult", na.rm = TRUE)
)

florida_keys_lip_age_counts <- florida_keys_conch %>%
  count(lip_age_class, name = "n") %>%
  mutate(percent = n / sum(n) * 100)

knitr::kable(florida_keys_import_summary, digits = 0)
knitr::kable(florida_keys_lip_age_counts, digits = 1)
```

## Summary Statistics

```{r size-lip-summary-statistics}
keys_comparison_conch <- bind_rows(
  port_everglades_conch %>%
    select(location, length_mm, lip_mm, lip_measured, lip_age_class),
  florida_keys_conch %>%
    select(location, length_mm, lip_mm, lip_measured, lip_age_class)
) %>%
  mutate(
    location = factor(location, levels = keys_comparison_location_levels),
    lip_age_class = factor(lip_age_class, levels = age_class_levels)
  )

location_ageclass_summary <- keys_comparison_conch %>%
  group_by(location, lip_age_class) %>%
  summarize(
    n = n(),
    measured_lip = sum(lip_measured),
    unflared_missing_lip = sum(!lip_measured),
    length_mean_mm = mean_or_na(length_mm),
    length_median_mm = median_or_na(length_mm),
    lip_mean_mm = mean_or_na(lip_mm),
    lip_median_mm = median_or_na(lip_mm),
    .groups = "drop"
  )

knitr::kable(
  location_ageclass_summary,
  digits = 1,
  caption = "Port Everglades and rest-of-Florida-Keys shell measurement summary by lip-based age class."
)
```

## Port Everglades Distributions

```{r port-length-lip-histograms, fig.width=8, fig.height=5}
port_metric_long <- port_everglades_conch %>%
  select(lip_age_class, length_mm, lip_mm) %>%
  pivot_longer(
    cols = c(length_mm, lip_mm),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      length_mm = "Shell length (mm)",
      lip_mm = "Lip thickness (mm)"
    )
  ) %>%
  filter(!is.na(value))

histogram_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.tag = element_text(face = "bold", size = 18),
    legend.position = "bottom"
  )

lip_thickness_histogram <- ggplot(
  port_metric_long %>% filter(metric == "Lip thickness (mm)"),
  aes(x = value, fill = lip_age_class)
) +
  geom_histogram(color = "white", bins = 30, linewidth = 0.2) +
  geom_vline(
    xintercept = 15,
    linetype = "dashed",
    color = "black",
    linewidth = 0.6
  ) +
  scale_fill_manual(values = age_class_cols, drop = FALSE) +
  labs(x = "Lip thickness (mm)", y = "Number of animals", fill = "Age class") +
  histogram_theme +
  guides(fill = "none")

shell_length_histogram <- ggplot(
  port_metric_long %>% filter(metric == "Shell length (mm)"),
  aes(x = value, fill = lip_age_class)
) +
  geom_histogram(color = "white", bins = 30, linewidth = 0.2) +
  scale_fill_manual(values = age_class_cols, drop = FALSE) +
  labs(x = "Shell length (mm)", y = "Number of animals", fill = "Age class") +
  histogram_theme

port_length_lip_histograms <- (
  lip_thickness_histogram + shell_length_histogram +
    plot_layout(guides = "collect")
) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")

port_length_lip_histograms

ggsave(output_port_length_lip_histograms, port_length_lip_histograms, width = 8, height = 5, dpi = 300, bg = "white")
```



## Location Comparisons



```{r location-metric-distributions, fig.width=9, fig.height=6}
combined_metric_long <- keys_comparison_conch %>%
  select(location, lip_age_class, length_mm, lip_mm) %>%
  pivot_longer(
    cols = c(length_mm, lip_mm),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      length_mm = "Shell length (mm)",
      lip_mm = "Lip thickness (mm)"
    )
  ) %>%
  filter(!is.na(value))

p_to_stars <- function(p) {
  case_when(
    is.na(p) ~ "",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ ""
  )
}

plot_welch_test <- function(dat) {
  port_values <- dat %>%
    filter(location == "Port Everglades") %>%
    pull(value)

  rest_keys_values <- dat %>%
    filter(location == "Rest of Florida Keys") %>%
    pull(value)

  if (length(port_values) < 2 || length(rest_keys_values) < 2) {
    return(tibble(p_value = NA_real_))
  }

  tibble(p_value = t.test(port_values, rest_keys_values)$p.value)
}

location_metric_significance <- combined_metric_long %>%
  group_by(metric, lip_age_class) %>%
  group_modify(~ plot_welch_test(.x)) %>%
  ungroup() %>%
  mutate(
    p_adjusted_bh = p.adjust(p_value, method = "BH"),
    significance = p_to_stars(p_adjusted_bh)
  ) %>%
  filter(significance != "") %>%
  left_join(
    combined_metric_long %>%
      group_by(metric, lip_age_class) %>%
      summarize(y_position = max(value, na.rm = TRUE), .groups = "drop") %>%
      group_by(metric) %>%
      mutate(y_position = y_position + 0.08 * diff(range(y_position, na.rm = TRUE))) %>%
      ungroup(),
    by = c("metric", "lip_age_class")
  )

lip_metric_data <- combined_metric_long %>%
  filter(metric == "Lip thickness (mm)", lip_age_class != "Juvenile") %>%
  mutate(lip_age_class = factor(lip_age_class, levels = c("Sub-adult", "Adult")))

length_metric_data <- combined_metric_long %>%
  filter(metric == "Shell length (mm)")

lip_significance <- location_metric_significance %>%
  filter(metric == "Lip thickness (mm)", lip_age_class != "Juvenile") %>%
  mutate(lip_age_class = factor(lip_age_class, levels = c("Sub-adult", "Adult")))

length_significance <- location_metric_significance %>%
  filter(metric == "Shell length (mm)")

location_metric_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.tag = element_text(face = "bold", size = 18),
    legend.position = "bottom"
  )

lip_thickness_plot <- ggplot(
  lip_metric_data,
  aes(x = lip_age_class, y = value, fill = location)
) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.65, position = position_dodge(width = 0.75)) +
  geom_text(
    data = lip_significance,
    aes(x = lip_age_class, y = y_position, label = significance),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 6
  ) +
  scale_fill_manual(values = c("Port Everglades" = "#0072B2", "Rest of Florida Keys" = "#D55E00")) +
  labs(x = NULL, y = "Lip thickness (mm)", fill = "Location") +
  location_metric_theme

shell_length_plot <- ggplot(
  length_metric_data,
  aes(x = lip_age_class, y = value, fill = location)
) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.65, position = position_dodge(width = 0.75)) +
  geom_text(
    data = length_significance,
    aes(x = lip_age_class, y = y_position, label = significance),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 6
  ) +
  scale_fill_manual(values = c("Port Everglades" = "#0072B2", "Rest of Florida Keys" = "#D55E00")) +
  labs(x = NULL, y = "Shell length (mm)", fill = "Location") +
  location_metric_theme

location_metric_distributions <- (
  lip_thickness_plot + shell_length_plot +
    plot_layout(widths = c(2, 3), guides = "collect")
) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")

location_metric_distributions

ggsave(output_location_metric_distributions, location_metric_distributions, width = 9, height = 6, dpi = 300, bg = "white")
```



## Statistical Comparisons

Mean shell length and mean lip thickness are compared between Port Everglades and the rest of the Florida Keys within each age class using Welch two-sample t-tests. Juveniles have unflared, unmeasured lips by definition, so lip thickness is not tested for juveniles.

```{r size-lip-statistical-comparisons}
welch_test_port_vs_rest_keys <- function(dat) {
  port_values <- dat %>%
    filter(location == "Port Everglades") %>%
    filter(!is.na(value)) %>%
    pull(value)

  rest_keys_values <- dat %>%
    filter(location == "Rest of Florida Keys") %>%
    filter(!is.na(value)) %>%
    pull(value)

  if (length(port_values) < 2 || length(rest_keys_values) < 2) {
    return(tibble())
  }

  test_result <- t.test(port_values, rest_keys_values)

  tibble(
    n_port_everglades = length(port_values),
    n_rest_keys = length(rest_keys_values),
    mean_port_everglades = mean_or_na(port_values),
    sd_port_everglades = sd_or_na(port_values),
    mean_rest_keys = mean_or_na(rest_keys_values),
    sd_rest_keys = sd_or_na(rest_keys_values),
    mean_difference_port_minus_rest_keys = mean_port_everglades - mean_rest_keys,
    lower_95_difference = test_result$conf.int[[1]],
    upper_95_difference = test_result$conf.int[[2]],
    p_value = test_result$p.value
  )
}

measurement_location_tests <- combined_metric_long %>%
  group_by(metric, lip_age_class) %>%
  group_modify(~ welch_test_port_vs_rest_keys(.x)) %>%
  ungroup() %>%
  mutate(
    p_adjusted_bh = p.adjust(p_value, method = "BH"),
    p_value_formatted = format_p_value(p_value),
    p_adjusted_bh_formatted = format_p_value(p_adjusted_bh)
  ) %>%
  arrange(metric, lip_age_class)

measurement_location_tests_display <- measurement_location_tests %>%
  transmute(
    Metric = metric,
    `Age class` = lip_age_class,
    Test = "Welch two-sample t-test",
    `n PE` = n_port_everglades,
    `PE mean +/- SD` = paste0(round(mean_port_everglades, 1), " +/- ", round(sd_port_everglades, 1)),
    `n rest Keys` = n_rest_keys,
    `Rest Keys mean +/- SD` = paste0(round(mean_rest_keys, 1), " +/- ", round(sd_rest_keys, 1)),
    `Mean difference, PE - rest Keys (95% CI)` = paste0(
      round(mean_difference_port_minus_rest_keys, 1),
      " (",
      round(lower_95_difference, 1),
      ", ",
      round(upper_95_difference, 1),
      ")"
    ),
    `p value` = p_value_formatted,
    `BH-adjusted p value` = p_adjusted_bh_formatted
  )

knitr::kable(
  measurement_location_tests_display,
  caption = "Welch two-sample t-tests comparing mean shell length and lip thickness between Port Everglades and the rest of the Florida Keys within each lip-based age class."
)
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
