---
title: "PENIP Coral and Conch Figures and Tables"
author: "R. Cunning"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

find_project_root <- function(start = getwd()) {
  start <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- c(start, normalizePath(file.path(start, ".."), winslash = "/", mustWork = TRUE))

  while (length(candidates) > 0) {
    candidate <- candidates[[1]]
    if (file.exists(file.path(candidate, "data/processed/polygons_final.rds"))) {
      return(candidate)
    }

    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not find the PENIP project root from ", start, call. = FALSE)
    }

    candidates <- c(candidates[-1], parent)
  }
}

knitr::opts_knit$set(root.dir = find_project_root())
```

# Setup

```{r libraries}
library(cowplot)
library(dplyr)
library(ggnewscale)
library(ggplot2)
library(ggrepel)
library(janitor)
library(patchwork)
library(purrr)
library(readr)
library(readxl)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(stringr)
library(tibble)
library(tidyr)
library(xml2)
```

```{r paths-and-constants}
polygons_rds <- "data/processed/polygons_final.rds"
impact_kml <- "data/spatial/Impact_zones_Scenario 2.kml"
nmfs_kml <- "data/spatial/Scenario2_NMFSConsiderationAreas.kml"
kj_eca_gpkg <- "data/spatial/KJ_Coral_Aquatic_Preserve.gpkg"
gshhg_full_coast_shp <- "data/spatial/gshhg/GSHHS_shp/f/GSHHS_f_L1.shp"
gshhg_zip <- "data/spatial/gshhg.zip"
gshhg_url <- "http://www.soest.hawaii.edu/pwessel/gshhg/gshhg-shp-2.3.7.zip"
conch_xlsx <- "data/conch/PENIP_QueenConchSurveys.xlsx"
berry_broad_xlsx <- "data/conch/Berry_Conch_per_site.xlsx"

survey_effort_xlim <- c(-80.1116, -80.078)
survey_effort_ylim <- c(26.058, 26.112)
survey_effort_types <- c(
  "Aggregated Patch Reef",
  "Artificial",
  "Inlet Channel",
  "Inner Reef",
  "Middle Reef",
  "Nearshore Ridge Complex",
  "Outer Reef",
  "Ridge",
  "Sand",
  "Scattered Rock in Unconsolidated Sediment"
)

dataset_levels <- c(
  "nsu11_esa",
  "dca17_esa",
  "dca17",
  "tt21",
  "tt23",
  "drm24",
  "berry_conch",
  "penip_radial_conch",
  "penip_drift_conch"
)

dataset_labels <- c(
  "nsu11_esa" = "2011 - NSU ESA Survey",
  "dca17_esa" = "2017 - DCA ESA Survey",
  "dca17" = "2017 - DCA Recon Survey",
  "tt21" = "2021 - Tetra Tech Survey",
  "tt23" = "2023 - Tetra Tech Survey",
  "drm24" = "2024 - Shedd Survey",
  "berry_conch" = "2012 - Berry et al.",
  "penip_radial_conch" = "2024 - Radial Survey",
  "penip_drift_conch" = "2024 - Drift Survey"
)

output_combined_figure1 <- "outputs/Fig1_New.png"
output_table1_new <- "outputs/Table1_New.csv"
output_table2_new <- "outputs/Table2_New.csv"
coral_table2_intervals_csv <- "outputs/Table2_coral_intervals.csv"
conch_kriging_impact_zone_totals_rds <- "outputs/cache/conch_kriging_impact_zone_totals_with_total.rds"
original_coral_figure1 <- "outputs/Fig1.png"
original_coral_panel_c_png <- "outputs/combined_assets/coral_fig1_panel_c.png"
original_coral_panel_c_crop <- "outputs/coral_fig1_panel_c_crop.png"
original_coral_panel_c_rds <- "outputs/combined_assets/coral_fig1_panel_c_ggplot.rds"
use_original_coral_panel_c_rds <- FALSE
panel_c_crop_rows <- 55:1310
panel_c_crop_cols <- 1800:2585
panel_c_left_artifact_mask_px <- 35
panel_c_left_padding_px <- 0
panel_c_x_shift <- -0.02
panel_c_width_scale <- 0.96
panel_b_x_shift <- -0.01
```

# Shared Spatial Data

```{r spatial-data}
polygons_final <- readRDS(polygons_rds) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(4326)

type_cols <- RColorBrewer::brewer.pal(nlevels(factor(polygons_final$Type)) + 1, "Set3")[-1]
names(type_cols) <- levels(factor(polygons_final$Type))

old_s2 <- sf_use_s2()
sf_use_s2(FALSE)

extract_placemark_info <- function(kml_file, source_label) {
  ns <- c(kml = "http://www.opengis.net/kml/2.2")
  kml <- read_xml(kml_file)
  placemarks <- xml_find_all(kml, ".//kml:Placemark", ns)

  map(placemarks, function(pm) {
    name <- xml_text(xml_find_first(pm, ".//kml:name", ns))
    folder_node <- xml_parent(xml_parent(pm))
    folder_name <- xml_text(xml_find_first(folder_node, ".//kml:name", ns))
    if (is.na(folder_name) || folder_name == "") folder_name <- "Top Level"

    coord_nodes <- xml_find_all(pm, ".//kml:coordinates", ns)
    rings <- map(coord_nodes, function(cn) {
      coords <- strsplit(trimws(xml_text(cn)), "\\s+")[[1]]
      coords_mat <- do.call(rbind, lapply(coords, function(x) {
        as.numeric(strsplit(x, ",")[[1]][1:2])
      }))
      if (nrow(coords_mat) >= 3) coords_mat else NULL
    }) %>%
      compact()

    if (length(rings) == 0) return(NULL)

    tibble(
      folder = folder_name,
      name = name,
      geometry = st_sfc(st_multipolygon(lapply(rings, function(r) list(r))), crs = 4326),
      source = source_label
    )
  }) %>%
    compact() %>%
    bind_rows() %>%
    st_as_sf()
}

impact_zones_sf <- extract_placemark_info(impact_kml, "Impact Zone")
nmfs_sf <- extract_placemark_info(nmfs_kml, "NMFS Consideration Area")

zones_merged <- bind_rows(impact_zones_sf, nmfs_sf) %>%
  mutate(
    name = if_else(
      is.na(name) | name == "",
      str_remove(str_remove(folder, "^Scenario_2_"), "\\.shp$"),
      name
    ),
    scenario = if_else(
      name %in% c("North", "South", "East", "Nearshore"),
      "NMFS Consideration",
      "Scenario 2"
    )
  ) %>%
  st_make_valid() %>%
  group_by(name, scenario, source) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

utm_crs <- 32617
polygons_utm <- st_transform(polygons_final, utm_crs)
zones_utm <- st_transform(zones_merged, utm_crs)
channel_utm <- zones_utm %>% filter(str_detect(tolower(as.character(name)), "channel"))

channel_bbox <- st_bbox(channel_utm)
habitat_bbox <- st_bbox(polygons_utm)

impact_monitoring_box <- st_as_sfc(st_bbox(
  c(
    xmin = habitat_bbox[["xmin"]],
    xmax = habitat_bbox[["xmax"]],
    ymin = channel_bbox[["ymin"]] - 1200,
    ymax = channel_bbox[["ymax"]] + 1200
  ),
  crs = st_crs(polygons_utm)
))

impact_monitoring_habitat_utm <- st_intersection(
  impact_monitoring_box,
  st_union(polygons_utm)
) %>%
  st_union() %>%
  st_make_valid()

existing_impact_zone_hull_utm <- zones_utm %>%
  st_union() %>%
  st_buffer(20) %>%
  st_union() %>%
  st_buffer(-20) %>%
  st_make_valid()

impact_monitoring_area_utm <- st_sf(
  name = "Rest of Monitoring Area",
  scenario = "Impact Monitoring Area",
  source = "Generated",
  geometry = st_sfc(
    st_difference(impact_monitoring_habitat_utm, existing_impact_zone_hull_utm),
    crs = st_crs(polygons_utm)
  )
) %>%
  st_make_valid()

all_zones_utm <- bind_rows(zones_utm, impact_monitoring_area_utm) %>%
  st_collection_extract("POLYGON") %>%
  st_cast("POLYGON", warn = FALSE)

channel_centroid_y <- st_coordinates(st_centroid(channel_utm))[, "Y"]
zones_with_dist <- all_zones_utm %>%
  mutate(
    is_channel = name == "Channel",
    centroid = st_centroid(geometry),
    centroid_y = st_coordinates(centroid)[, "Y"],
    dist_from_channel = if_else(is_channel, 0, abs(centroid_y - channel_centroid_y))
  ) %>%
  st_transform(4326)

distance_palette <- c("#67001F", "#B2182B", "#EF8A62", "#FDB863", "#FFFFBF")
```

# Survey Effort Data

```{r coral-survey-effort-data}
nsu11_esa_sitemd <- readxl::read_xlsx(
  "data/2011_nsu_esa/Port Everglades_NSU 2011and DCA 2017_ESA surveys.xlsx",
  sheet = "2011_NSU_ESA survey"
) %>%
  clean_names() %>%
  transmute(dataset = "nsu11_esa", site = as.character(ident), latitude = as.numeric(lat), longitude = as.numeric(long))

dca17_esa_sitemd <- readxl::read_xlsx(
  "data/2017_dca_esa/Port Everglades_NSU 2011and DCA 2017_ESA surveys.xlsx",
  sheet = "2017_DCA_ESA survey"
) %>%
  clean_names() %>%
  transmute(dataset = "dca17_esa", site = as.character(id), latitude = as.numeric(latitude), longitude = as.numeric(longitude))

dca17_sitemd <- readxl::read_xlsx("data/2017_dca_recon/Recon_Site_Coordinates_Extracted.xlsx") %>%
  clean_names() %>%
  mutate(across(c(latitude, longitude), as.numeric)) %>%
  group_by(site = transect) %>%
  summarize(latitude = mean(latitude, na.rm = TRUE), longitude = mean(longitude, na.rm = TRUE), .groups = "drop") %>%
  mutate(dataset = "dca17", site = as.character(site)) %>%
  select(dataset, site, latitude, longitude)

tt21_sitemd <- readr::read_csv("data/2021_tt_recon_esa/midpoints_latlon.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  mutate(site = str_remove(name, "A$")) %>%
  transmute(dataset = "tt21", site = as.character(site), latitude = as.numeric(lat), longitude = as.numeric(lon))

tt21_sand <- readr::read_csv("data/2021_tt_recon_esa/Table_2_Port_Everglades_RECON.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  mutate(site = as.character(site))

tt21_sitemd <- tt21_sitemd %>%
  filter(!site %in% (tt21_sand %>% filter(percent_cover_sc_sp == "100%") %>% pull(site)))

tt23_sitemd <- readxl::read_xlsx("data/2023_tt_impact/Impact site tracking.xlsx", skip = 1) %>%
  clean_names() %>%
  transmute(
    dataset = "tt23",
    site = tolower(str_remove_all(transect_name, "[[:space:][:punct:]]")),
    latitude = as.numeric(actual_start_y),
    longitude = as.numeric(actual_start_x)
  ) %>%
  drop_na(site, latitude, longitude)

tt23v_sitemd <- readxl::read_xlsx(
  "data/2023_tt_impact/Impact site tracking.xlsx",
  sheet = "Vertical Wall"
) %>%
  clean_names() %>%
  transmute(
    dataset = "tt23",
    site = tolower(str_remove_all(name, "[[:space:][:punct:]]")),
    latitude = (as.numeric(actual_start_y) + as.numeric(actual_end_y)) / 2,
    longitude = (as.numeric(actual_start_x) + as.numeric(actual_end_x)) / 2
  ) %>%
  mutate(
    site = gsub("cpss", "cpsse", site),
    site = gsub("lrnw", "lrmnw", site),
    site = gsub("lrsw", "lrmsw", site)
  ) %>%
  drop_na(site, latitude, longitude)

tt23_sitemd <- bind_rows(tt23_sitemd, tt23v_sitemd)

drm24_sitemd <- readxl::read_xlsx("data/2024_shedd_drm/site_metadata.xlsx") %>%
  clean_names() %>%
  transmute(dataset = "drm24", site = as.character(drm_site_id), latitude = as.numeric(lat), longitude = as.numeric(lon)) %>%
  drop_na(site, latitude, longitude)

coral_effort_points <- bind_rows(
  nsu11_esa_sitemd,
  dca17_esa_sitemd,
  dca17_sitemd,
  tt21_sitemd,
  tt23_sitemd,
  drm24_sitemd
) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_join(polygons_final, join = st_within) %>%
  st_drop_geometry() %>%
  filter(Type %in% survey_effort_types) %>%
  mutate(dataset = factor(dataset, levels = dataset_levels))
```

```{r conch-survey-effort-data}
berry_conch_points <- readxl::read_xlsx(berry_broad_xlsx, sheet = 1) %>%
  transmute(dataset = "berry_conch", site = as.character(Site), latitude = as.numeric(LatDD), longitude = as.numeric(LonDD)) %>%
  filter(!is.na(latitude), !is.na(longitude))

penip_radial_conch_points <- readxl::read_xlsx(conch_xlsx, sheet = "PENIP_Conch_10mRadial") %>%
  transmute(dataset = "penip_radial_conch", site = as.character(`Site Code`), latitude = as.numeric(Lat), longitude = as.numeric(Lon)) %>%
  filter(!is.na(latitude), !is.na(longitude))

penip_drift_conch_lines <- readxl::read_xlsx(conch_xlsx, sheet = "PENIP_DriftTransects") %>%
  mutate(
    dataset = "penip_drift_conch",
    site = as.character(Transect),
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon), !is.na(end_lat), !is.na(end_lon))

penip_drift_conch_lines <- st_sf(
  penip_drift_conch_lines %>% select(dataset, site),
  geometry = st_sfc(
    map2(
      seq_len(nrow(penip_drift_conch_lines)),
      seq_len(nrow(penip_drift_conch_lines)),
      ~ st_linestring(rbind(
        c(penip_drift_conch_lines$start_lon[[.x]], penip_drift_conch_lines$start_lat[[.x]]),
        c(penip_drift_conch_lines$end_lon[[.y]], penip_drift_conch_lines$end_lat[[.y]])
      ))
    ),
    crs = 4326
  )
) %>%
  mutate(dataset = factor(dataset, levels = dataset_levels))

conch_effort_points <- bind_rows(berry_conch_points, penip_radial_conch_points) %>%
  mutate(dataset = factor(dataset, levels = dataset_levels))

all_effort_panel_polygons <- crossing(dataset = factor(dataset_levels, levels = dataset_levels)) %>%
  pull(dataset) %>%
  map_dfr(function(dataset_i) {
    polygons_final %>% mutate(dataset = dataset_i)
  }) %>%
  st_as_sf() %>%
  mutate(dataset = factor(dataset, levels = dataset_levels))
```

# Figure 1

```{r figure-1, fig.width=4.72, fig.height=5.71}
panel_a_bbox_ymin <- 25.60
panel_a_bbox_ymax <- 26.30
panel_a_original_bbox_ymax <- 26.50
panel_a_original_rel_height <- 0.48
panel_c_rel_height <- 0.52
panel_a_rel_height <- panel_a_original_rel_height *
  (panel_a_bbox_ymax - panel_a_bbox_ymin) /
  (panel_a_original_bbox_ymax - panel_a_bbox_ymin)

bbox_view <- st_bbox(
  c(xmin = -80.4, ymin = panel_a_bbox_ymin, xmax = -79.72, ymax = panel_a_bbox_ymax),
  crs = st_crs(4326)
)

fl_bbox <- st_bbox(
  c(xmin = -87.8, ymin = 24.0, xmax = -79.5, ymax = 31.0),
  crs = st_crs(4326)
)

usa_states <- rnaturalearth::ne_states(country = "United States of America", returnclass = "sf")
fl_outline <- usa_states %>% filter(name == "Florida")
kj_eca <- st_read(kj_eca_gpkg, quiet = TRUE) %>% st_transform(4326)

if (!file.exists(gshhg_full_coast_shp)) {
  if (!file.exists(gshhg_zip)) utils::download.file(gshhg_url, gshhg_zip, mode = "wb")
  unzip(gshhg_zip, exdir = dirname(dirname(dirname(gshhg_full_coast_shp))))
}

panel_a_bbox_sfc <- st_as_sfc(bbox_view)
panel_a_land_selected <- st_read(gshhg_full_coast_shp, quiet = TRUE) %>%
  st_transform(4326) %>%
  st_filter(panel_a_bbox_sfc, .predicate = st_intersects) %>%
  st_make_valid()

panel_a_land <- panel_a_land_selected %>%
  suppressWarnings(st_intersection(panel_a_bbox_sfc)) %>%
  st_collection_extract("POLYGON")

panel_a_coastline <- panel_a_land_selected %>%
  st_boundary() %>%
  suppressWarnings(st_intersection(panel_a_bbox_sfc)) %>%
  st_collection_extract("LINESTRING")

ports <- tibble(
  name = c("Port Everglades", "Port of Miami"),
  lon = c(-80.118, -80.133),
  lat = c(26.090, 25.770)
)

cities <- tibble(
  name = c("Miami", "Fort Lauderdale"),
  lon = c(-80.1918, -80.1373),
  lat = c(25.7617, 26.1224)
)

study_box <- st_as_sfc(st_bbox(
  c(xmin = survey_effort_xlim[[1]], xmax = survey_effort_xlim[[2]], ymin = survey_effort_ylim[[1]], ymax = survey_effort_ylim[[2]]),
  crs = 4326
))

panel_a_inset <- ggplot() +
  geom_sf(data = fl_outline, fill = "grey92", color = "black", linewidth = 0.1) +
  geom_sf(data = st_as_sfc(bbox_view), fill = NA, color = "#D7301F", linewidth = 0.6) +
  coord_sf(
    xlim = c(fl_bbox[["xmin"]], fl_bbox[["xmax"]]),
    ylim = c(fl_bbox[["ymin"]], fl_bbox[["ymax"]]),
    expand = FALSE
  ) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.25),
    panel.background = element_rect(fill = "white", color = NA)
  )

panel_a <- ggplot() +
  geom_sf(data = panel_a_land, fill = "grey90", color = NA) +
  geom_sf(data = panel_a_coastline, color = "black", linewidth = 0.1) +
  geom_sf(data = kj_eca, fill = NA, color = "#4B6CB7", linewidth = 0.4) +
  geom_sf(data = study_box, fill = NA, color = "#D7301F", linewidth = 0.8) +
  geom_text_repel(
    data = ports,
    aes(lon, lat, label = name),
    hjust = 0,
    size = 1.8,
    nudge_y = -0.06,
    xlim = c(-80.04, NA),
    segment.size = 0.2
  ) +
  geom_point(data = cities, aes(lon, lat), size = 0.8) +
  geom_text(data = cities, aes(lon, lat, label = name), hjust = 1, vjust = 0, size = 2, nudge_x = -0.01) +
  annotate(
    "text",
    x = -80.05,
    y = 25.9,
    label = "Kristin Jacobs\nCoral Reef Ecosystem\nConservation Area",
    color = "#4B6CB7",
    hjust = 0,
    size = 1.8
  ) +
  coord_sf(
    xlim = c(bbox_view[["xmin"]], bbox_view[["xmax"]]),
    ylim = c(bbox_view[["ymin"]], bbox_view[["ymax"]]),
    expand = FALSE
  ) +
  theme_void() +
  theme(plot.margin = margin(0, 2, 2, 2)) +
  inset_element(panel_a_inset, left = 0.64, bottom = 0.74, right = 0.99, top = 0.99, align_to = "panel")

survey_map_theme <- theme_minimal(base_size = 7) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    strip.text = element_text(size = 5.1, lineheight = 0.92),
    panel.spacing = unit(0.35, "lines"),
    plot.margin = margin(0, 0, 0, 0),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

make_effort_panel <- function(dataset_i) {
  ggplot() +
    geom_sf(
      data = all_effort_panel_polygons %>% filter(dataset == dataset_i),
      aes(fill = Type),
      color = NA,
      linewidth = 0.1,
      alpha = 0.5
    ) +
    geom_point(
      data = coral_effort_points %>% filter(dataset == dataset_i),
      aes(x = longitude, y = latitude),
      pch = 4,
      alpha = 0.6,
      size = 0.55,
      stroke = 0.25,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = conch_effort_points %>% filter(dataset == dataset_i),
      aes(x = longitude, y = latitude),
      pch = 4,
      alpha = 0.6,
      size = 0.55,
      stroke = 0.25,
      inherit.aes = FALSE
    ) +
    geom_sf(
      data = penip_drift_conch_lines %>% filter(dataset == dataset_i),
      color = "black",
      linewidth = 0.22,
      alpha = 0.7
    ) +
    scale_fill_manual(values = type_cols, na.value = "gray80") +
    coord_sf(xlim = survey_effort_xlim, ylim = survey_effort_ylim, expand = FALSE) +
    labs(title = dataset_labels[[dataset_i]]) +
    survey_map_theme +
    theme(
      plot.title = element_text(size = 4.9, hjust = 0.5, lineheight = 0.92, margin = margin(0, 0, 1, 0)),
      plot.margin = margin(0, 0, 0, 0)
    )
}

panel_b_maps <- map(dataset_levels, make_effort_panel)

panel_b <- plot_grid(
  plotlist = panel_b_maps,
  ncol = 3,
  align = "hv",
  axis = "tblr"
)

zone_labels <- tribble(
  ~name, ~label_x, ~label_y,
  "Channel", -80.0848, 26.0938,
  "Side Slopes", -80.1024, 26.08,
  ">10 cm", -80.1018, 26.078,
  "5-10 cm", -80.1012, 26.076,
  "1-5 cm", -80.1006, 26.074,
  "0.5-1 cm", -80.1000, 26.072,
  "0.1-0.5 cm", -80.0994, 26.070,
  "Nearshore", -80.1035, 26.094,
  "North", -80.0794, 26.097,
  "East", -80.080, 26.0938,
  "South", -80.0792, 26.091,
  "Impact Monitoring Area", -80.0815, 26.083
)

zone_segments <- tribble(
  ~name, ~segment_x, ~segment_y,
  "Channel", -80.085, 26.0938,
  "Side Slopes", -80.102, 26.0945,
  "Side Slopes", -80.1027, 26.093,
  ">10 cm", -80.1007, 26.0947,
  ">10 cm", -80.1014, 26.09285,
  "5-10 cm", -80.0995, 26.0949,
  "5-10 cm", -80.1003, 26.0926,
  "1-5 cm", -80.0983, 26.0957,
  "1-5 cm", -80.0992, 26.092,
  "0.5-1 cm", -80.097, 26.0968,
  "0.5-1 cm", -80.0981, 26.091,
  "0.1-0.5 cm", -80.0954, 26.098,
  "0.1-0.5 cm", -80.097, 26.0897,
  "Nearshore", -80.106, 26.09,
  "North", -80.0794, 26.097,
  "East", -80.080, 26.0938,
  "South", -80.0792, 26.091,
  "Impact Monitoring Area", -80.0815, 26.083
)

label_segments <- zone_segments %>%
  left_join(zone_labels, by = "name")

tick_offsets <- c(300, 600, 1200)
x_lon <- -80.077
channel_bbox_latlon <- st_bbox(st_transform(channel_utm, 4326))
x_pos_utm <- st_transform(
  st_sfc(st_point(c(x_lon, mean(c(channel_bbox_latlon[["ymin"]], channel_bbox_latlon[["ymax"]])))), crs = 4326),
  utm_crs
) %>%
  st_coordinates() %>%
  .[, "X"]

ticks_utm <- tibble(
  x = rep(x_pos_utm, 2 * length(tick_offsets) + 2),
  y = c(
    channel_bbox[["ymin"]] - rev(tick_offsets),
    channel_bbox[["ymin"]],
    channel_bbox[["ymax"]],
    channel_bbox[["ymax"]] + tick_offsets
  ),
  dist_m = c(-rev(tick_offsets), 0, 0, tick_offsets)
)

ticks_latlon <- st_as_sf(ticks_utm, coords = c("x", "y"), crs = utm_crs) %>%
  st_transform(4326) %>%
  st_coordinates() %>%
  as_tibble() %>%
  mutate(dist_m = ticks_utm$dist_m)

bbox_channel_latlon <- st_as_sf(
  tibble(
    x = rep(x_pos_utm, 2),
    y = c(channel_bbox[["ymin"]], channel_bbox[["ymax"]])
  ),
  coords = c("x", "y"),
  crs = utm_crs
) %>%
  st_transform(4326) %>%
  st_coordinates()

scale_lines <- tibble(
  x = c(x_lon, x_lon),
  xend = c(x_lon, x_lon),
  y = c(bbox_channel_latlon[2, "Y"], bbox_channel_latlon[1, "Y"]),
  yend = c(max(ticks_latlon$Y), min(ticks_latlon$Y))
)

tick_length <- 0.0005
ticks_segments <- tibble(
  x = x_lon - tick_length,
  xend = x_lon + tick_length,
  y = ticks_latlon$Y,
  yend = ticks_latlon$Y
)

tick_labels <- tibble(
  x = x_lon - 0.0004,
  y = ticks_latlon$Y,
  label = if_else(ticks_latlon$dist_m == 0, "", paste0(abs(ticks_latlon$dist_m), "m"))
)

panel_c_base <- ggplot() +
  geom_sf(data = polygons_final, aes(fill = Type), color = NA, linewidth = 0.1, alpha = 0.25, show.legend = FALSE) +
  scale_fill_manual(values = type_cols, na.value = "white") +
  ggnewscale::new_scale_fill() +
  geom_sf(
    data = zones_with_dist %>% filter(!is_channel),
    aes(fill = dist_from_channel),
    color = "black",
    alpha = 0.5,
    linewidth = 0.075
  ) +
  geom_sf(
    data = zones_with_dist %>% filter(is_channel),
    fill = "#67001F",
    color = "black",
    linewidth = 0.075,
    alpha = 0.5
  ) +
  scale_fill_gradientn(colors = distance_palette, name = "Distance from Channel (m)")

panel_c <- panel_c_base +
  geom_segment(
    data = label_segments,
    aes(x = segment_x, y = segment_y, xend = label_x, yend = label_y),
    color = "black",
    linewidth = 0.2
  ) +
  geom_text(data = zone_labels, aes(x = label_x, y = label_y, label = name), size = 1.6, hjust = 1) +
  geom_segment(
    data = scale_lines,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.35
  ) +
  geom_segment(
    data = ticks_segments,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "black",
    linewidth = 0.25
  ) +
  geom_text(
    data = tick_labels,
    aes(x = x, y = y, label = label),
    hjust = 1,
    vjust = c(rep(1.5, 4), rep(-0.5, 4)),
    size = 1.35
  ) +
  coord_sf(xlim = c(-80.1116, -80.077), ylim = c(26.058, 26.112), expand = FALSE, clip = "off") +
  theme_minimal(base_size = 7) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = margin(2, 2, 0, 2)
  )

if (use_original_coral_panel_c_rds && file.exists(original_coral_panel_c_rds)) {
  panel_c <- readRDS(original_coral_panel_c_rds) +
    coord_sf(xlim = c(-80.113, -80.077), ylim = c(26.058, 26.112), expand = FALSE, clip = "off") +
    theme(plot.margin = margin(0, 10, 0, 16))
} else {
  if (file.exists(original_coral_panel_c_png)) {
    panel_c_image_path <- original_coral_panel_c_png
    original_panel_c_image <- png::readPNG(panel_c_image_path)
  } else if (file.exists(original_coral_figure1)) {
    original_fig1_image <- png::readPNG(original_coral_figure1)
    original_panel_c_image <- original_fig1_image[panel_c_crop_rows, panel_c_crop_cols, , drop = FALSE]
    original_panel_c_image[, seq_len(panel_c_left_artifact_mask_px), ] <- 1

    if (panel_c_left_padding_px > 0) {
      panel_c_padding <- array(
        1,
        dim = c(dim(original_panel_c_image)[1], panel_c_left_padding_px, dim(original_panel_c_image)[3])
      )
      original_panel_c_image_padded <- array(
        1,
        dim = c(
          dim(original_panel_c_image)[1],
          dim(original_panel_c_image)[2] + dim(panel_c_padding)[2],
          dim(original_panel_c_image)[3]
        )
      )
      original_panel_c_image_padded[, seq_len(dim(panel_c_padding)[2]), ] <- panel_c_padding
      original_panel_c_image_padded[, (dim(panel_c_padding)[2] + 1):dim(original_panel_c_image_padded)[2], ] <-
        original_panel_c_image
      original_panel_c_image <- original_panel_c_image_padded
    }

    png::writePNG(original_panel_c_image, original_coral_panel_c_crop)
    panel_c_image_path <- original_coral_panel_c_crop
  } else {
    stop(
      "Could not find ", original_coral_panel_c_png, ", ", original_coral_panel_c_rds,
      ", or ", original_coral_figure1,
      ". Knit PENIP_analysis.Rmd first or restore outputs/Fig1.png.",
      call. = FALSE
    )
  }

  panel_c <- ggdraw() +
    draw_grob(
      grid::rasterGrob(
        original_panel_c_image,
        x = grid::unit(0, "npc"),
        y = grid::unit(0.5, "npc"),
        width = grid::unit(panel_c_width_scale, "npc"),
        height = grid::unit(1, "npc"),
        just = c("left", "center"),
        interpolate = TRUE
      ),
      x = panel_c_x_shift,
      y = 0,
      width = 1,
      height = 1
    )
}

panel_b_group_labels <- ggdraw() +
  draw_label("Coral surveys", x = 0.35, y = 0.68, angle = 90, size = 7) +
  draw_line(x = c(0.99, 0.99), y = c(0.36, 0.98), size = 0.35) +
  draw_label("Conch surveys", x = 0.35, y = 0.17, angle = 90, size = 7) +
  draw_line(x = c(0.99, 0.99), y = c(0.02, 0.31), size = 0.35)

panel_b_labeled <- plot_grid(
  panel_b_group_labels,
  panel_b,
  nrow = 1,
  rel_widths = c(0.055, 0.945)
)

panel_b_labeled_shifted <- ggdraw() +
  draw_plot(
    panel_b_labeled,
    x = panel_b_x_shift,
    y = 0,
    width = 1,
    height = 1
  )

left_column <- plot_grid(
  panel_a,
  panel_c,
  ncol = 1,
  labels = c("A", "C"),
  label_size = 11,
  hjust = 0,
  rel_heights = c(panel_a_rel_height, panel_c_rel_height)
)

figure_1_no_legend <- plot_grid(
  left_column,
  panel_b_labeled_shifted,
  nrow = 1,
  labels = c("", "B"),
  label_size = 11,
  hjust = 0,
  rel_widths = c(0.42, 0.58)
)

legend_items <- tibble(
  Type = factor(survey_effort_types, levels = survey_effort_types),
  x = seq_along(survey_effort_types),
  y = 1
)

legend_type_labels <- setNames(survey_effort_types, survey_effort_types)
legend_type_labels[["Scattered Rock in Unconsolidated Sediment"]] <- "Scattered Rock in\nUnconsolidated Sediment"

legend_plot <- ggplot(legend_items, aes(x, y, fill = Type)) +
  geom_tile(alpha = 0.5) +
  scale_fill_manual(
    values = type_cols[survey_effort_types],
    labels = legend_type_labels,
    na.value = "gray80"
  ) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_void(base_size = 7) +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(0.28, "cm"),
    legend.text = element_text(size = 5.5),
    legend.title = element_blank(),
    legend.box.margin = margin(0, 0, 0, 0),
    plot.margin = margin(0, 0, 0, 0)
  )

fill_legend <- gtable::gtable_filter(ggplotGrob(legend_plot), "guide-box-bottom")

combined_figure_1 <- plot_grid(
  figure_1_no_legend,
  fill_legend,
  ncol = 1,
  rel_heights = c(0.9, 0.1)
)

combined_figure_1

ggsave(
  output_combined_figure1,
  combined_figure_1,
  width = 120,
  height = 145,
  units = "mm",
  dpi = 300,
  bg = "white"
)
```

# Table 1

```{r table-1-new}
table1_habitat_levels <- c(
  "Nearshore Ridge Complex",
  "Inner Reef",
  "Middle Reef",
  "Outer Reef",
  "Sand/Scattered Rock"
)

point_habitat_tolerance_m <- 5
drift_half_width_m <- 2
drift_segment_length_m <- 20

format_area_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) return(NA_character_)

  area_range <- range(x)
  formatted <- format(round(area_range, 1), trim = TRUE, scientific = FALSE)
  if (isTRUE(all.equal(area_range[[1]], area_range[[2]]))) {
    formatted[[1]]
  } else {
    paste(formatted, collapse = " - ")
  }
}

recode_table1_habitat <- function(type) {
  case_when(
    type == "Artificial" ~ "Nearshore Ridge Complex",
    type == "Aggregated Patch Reef" ~ "Outer Reef",
    type %in% c("Sand", "Scattered Rock in Unconsolidated Sediment") ~ "Sand/Scattered Rock",
    TRUE ~ as.character(type)
  )
}

assign_point_habitat <- function(points_df, lon_col, lat_col, nearest_tolerance_m = point_habitat_tolerance_m) {
  points_sf <- points_df %>%
    filter(!is.na(.data[[lon_col]]), !is.na(.data[[lat_col]])) %>%
    st_as_sf(coords = c(lon_col, lat_col), crs = 4326, remove = FALSE)

  polygon_hits <- st_intersects(points_sf, polygons_final)
  point_types <- map_chr(polygon_hits, function(i) {
    if (length(i) == 0) NA_character_ else polygons_final$Type[[i[[1]]]]
  })

  unmatched <- which(is.na(point_types))
  if (length(unmatched) > 0 && is.finite(nearest_tolerance_m) && nearest_tolerance_m > 0) {
    unmatched_points_utm <- st_transform(points_sf[unmatched, ], utm_crs)
    nearest_idx <- st_nearest_feature(unmatched_points_utm, polygons_utm)
    nearest_dist_m <- as.numeric(st_distance(
      unmatched_points_utm,
      polygons_utm[nearest_idx, ],
      by_element = TRUE
    ))
    close_enough <- nearest_dist_m <= nearest_tolerance_m
    point_types[unmatched[close_enough]] <- polygons_utm$Type[nearest_idx[close_enough]]
  }

  points_sf %>%
    mutate(Type = point_types) %>%
    st_drop_geometry()
}

summarize_site_habitats <- function(site_types) {
  habitat_counts <- site_types %>%
    mutate(Type = recode_table1_habitat(Type)) %>%
    distinct(site, Type) %>%
    filter(Type %in% table1_habitat_levels) %>%
    count(Type, name = "n") %>%
    mutate(Type = factor(Type, levels = table1_habitat_levels)) %>%
    complete(Type, fill = list(n = 0)) %>%
    arrange(Type) %>%
    pivot_wider(names_from = Type, values_from = n)

  bind_cols(
    habitat_counts,
    tibble(`Total Sites` = n_distinct(site_types$site))
  )
}

make_conch_table_row <- function(dataset, dataset2, site_types, area_per_site_m2, total_n) {
  summarize_site_habitats(site_types) %>%
    mutate(
      dataset = dataset,
      area_per_site = format_area_range(area_per_site_m2),
      total_area = round(sum(area_per_site_m2, na.rm = TRUE), 0),
      total_n = total_n,
      dataset2 = dataset2,
      .before = 1
    )
}

table1_coral <- read_csv("outputs/Table1.csv", show_col_types = FALSE) %>%
  filter(dataset != "Total")

berry_broad_table <- readxl::read_xlsx(berry_broad_xlsx, sheet = 1) %>%
  transmute(
    site = as.character(Site),
    latitude = as.numeric(LatDD),
    longitude = as.numeric(LonDD),
    total_count = rowSums(
      across(c(Adult, Juv, Old, Sub, Young), ~ replace_na(as.numeric(.x), 0))
    ),
    survey_area_m2 = 2 * 30 * 4
  ) %>%
  filter(!is.na(site), !is.na(latitude), !is.na(longitude))

berry_site_types <- berry_broad_table %>%
  assign_point_habitat("longitude", "latitude")

berry_table_row <- make_conch_table_row(
  dataset = "berry_conch",
  dataset2 = "2012 / Berry et al.",
  site_types = berry_site_types,
  area_per_site_m2 = berry_broad_table$survey_area_m2,
  total_n = sum(berry_broad_table$total_count, na.rm = TRUE)
)

radial_sites_table <- readxl::read_xlsx(conch_xlsx, sheet = "PENIP_Conch_10mRadial") %>%
  transmute(
    site = as.character(`Site Code`),
    latitude = as.numeric(Lat),
    longitude = as.numeric(Lon),
    survey_area_m2 = pi * 10^2
  ) %>%
  filter(!is.na(site), !is.na(latitude), !is.na(longitude))

radial_counts_table <- read_csv("data/conch/PENIP_circular_counts.csv", show_col_types = FALSE) %>%
  transmute(
    site = as.character(`Site Code`),
    total_count = replace_na(as.numeric(`# Juv`), 0) + replace_na(as.numeric(`# Adult`), 0)
  )

radial_site_types <- radial_sites_table %>%
  assign_point_habitat("longitude", "latitude")

radial_table_row <- make_conch_table_row(
  dataset = "penip_radial_conch",
  dataset2 = "2024 / Radial Survey",
  site_types = radial_site_types,
  area_per_site_m2 = radial_sites_table$survey_area_m2,
  total_n = radial_sites_table %>%
    left_join(radial_counts_table, by = "site") %>%
    summarize(total_n = sum(total_count, na.rm = TRUE)) %>%
    pull(total_n)
)

make_table_transect_rectangle <- function(start_point, end_point, half_width_m = 2) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) return(st_polygon())

  unit_direction <- direction / transect_length
  perpendicular <- c(-unit_direction[[2]], unit_direction[[1]]) * half_width_m

  st_polygon(list(rbind(
    start_point + perpendicular,
    end_point + perpendicular,
    end_point - perpendicular,
    start_point - perpendicular,
    start_point + perpendicular
  )))
}

make_table_segment_rectangles <- function(start_point, end_point, segment_length_m, half_width_m) {
  direction <- end_point - start_point
  transect_length <- sqrt(sum(direction^2))

  if (transect_length == 0) return(tibble())

  unit_direction <- direction / transect_length
  segment_starts <- seq(0, transect_length, by = segment_length_m)
  segment_ends <- pmin(segment_starts + segment_length_m, transect_length)
  keep_segments <- segment_ends > segment_starts

  tibble(
    segment_index = seq_along(segment_starts)[keep_segments],
    segment_start_m = segment_starts[keep_segments],
    segment_end_m = segment_ends[keep_segments],
    geometry = map2(
      segment_start_m,
      segment_end_m,
      ~ make_table_transect_rectangle(
        start_point + unit_direction * .x,
        start_point + unit_direction * .y,
        half_width_m = half_width_m
      )
    )
  )
}

drift_surveys_table <- readxl::read_xlsx(conch_xlsx, sheet = "PENIP_DriftTransects") %>%
  mutate(
    site = paste(Line, Transect, sep = "_"),
    start_lat = as.numeric(`StartLat (dec deg)`),
    start_lon = as.numeric(`StartLong (dec deg)`),
    end_lat = as.numeric(`EndLat (dec deg)`),
    end_lon = as.numeric(`EndLon (dec deg)`)
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon), !is.na(end_lat), !is.na(end_lon))

drift_start_table <- drift_surveys_table %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

drift_end_table <- drift_surveys_table %>%
  st_as_sf(coords = c("end_lon", "end_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(utm_crs)

drift_start_xy <- st_coordinates(drift_start_table)
drift_end_xy <- st_coordinates(drift_end_table)

drift_segment_table <- map_dfr(seq_len(nrow(drift_surveys_table)), function(i) {
  make_table_segment_rectangles(
    drift_start_xy[i, ],
    drift_end_xy[i, ],
    segment_length_m = drift_segment_length_m,
    half_width_m = drift_half_width_m
  ) %>%
    mutate(
      transect_site = drift_surveys_table$site[[i]],
      segment_id = paste(transect_site, segment_index, sep = "_seg")
    )
})

drift_segments_table <- st_as_sf(
  drift_segment_table,
  sf_column_name = "geometry",
  crs = utm_crs
) %>%
  select(site = segment_id, transect_site, segment_index, segment_start_m, segment_end_m) %>%
  mutate(survey_area_m2 = as.numeric(st_area(geometry)))

drift_site_types <- st_intersection(drift_segments_table, polygons_utm) %>%
  mutate(
    Type = recode_table1_habitat(Type),
    habitat_overlap_m2 = as.numeric(st_area(geometry))
  ) %>%
  st_drop_geometry() %>%
  group_by(site, Type) %>%
  summarize(habitat_overlap_m2 = sum(habitat_overlap_m2, na.rm = TRUE), .groups = "drop") %>%
  group_by(site) %>%
  slice_max(habitat_overlap_m2, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(site, Type) %>%
  right_join(
    drift_segments_table %>% st_drop_geometry() %>% distinct(site),
    by = "site"
  )

drift_observation_count <- readxl::read_xlsx(conch_xlsx, sheet = "PENIP_ConchDriftPosition") %>%
  mutate(
    lat = as.numeric(`Lat (dec deg)`),
    lon = as.numeric(`Long (dec deg)`),
    animal_id = sub("^.*([AJ][0-9]+)$", "\\1", `WPT (B/E)`)
  ) %>%
  filter(!is.na(lat), !is.na(lon), grepl("^[AJ]", animal_id)) %>%
  nrow()

drift_table_row <- make_conch_table_row(
  dataset = "penip_drift_conch",
  dataset2 = "2024 / Drift Survey",
  site_types = drift_site_types,
  area_per_site_m2 = drift_segments_table$survey_area_m2,
  total_n = drift_observation_count
)

table1_new <- bind_rows(
  table1_coral,
  berry_table_row,
  radial_table_row,
  drift_table_row
) %>%
  mutate(
    across(all_of(table1_habitat_levels), ~ replace_na(as.numeric(.x), 0)),
    across(all_of(c("Total Sites", "total_area", "total_n")), as.numeric)
  )

table1_total <- table1_new %>%
  summarize(
    dataset = "Total",
    across(all_of(c(table1_habitat_levels, "Total Sites")), ~ sum(.x, na.rm = TRUE)),
    area_per_site = "-",
    total_area = sum(total_area, na.rm = TRUE),
    total_n = sum(total_n, na.rm = TRUE),
    dataset2 = "Total"
  )

table1_new <- bind_rows(table1_new, table1_total) %>%
  select(
    dataset,
    all_of(table1_habitat_levels),
    `Total Sites`,
    area_per_site,
    total_area,
    total_n,
    dataset2
  )

write_csv(table1_new, output_table1_new)

table1_new %>%
  knitr::kable()
```

# Table 2

The combined Table 2 is streamlined relative to the coral-only table by removing the cumulative columns. This keeps the main table focused on estimated abundance within each impact area for the coral groups and conch age classes. If cumulative totals are needed, they are better handled as a companion appendix table.

```{r table-2-new}
format_count <- function(x) {
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

format_interval_count <- function(estimate, lower, upper) {
  paste0(
    format_count(estimate),
    " (",
    format_count(lower),
    "-",
    format_count(upper),
    ")"
  )
}

coral_table2_area_labels <- c(
  "Channel" = "Channel",
  "Side Slopes" = "Side Slopes",
  "Depth_10cm" = "Sediment deposition >10 cm",
  "Depth_5cm" = "Sediment deposition 5-10 cm",
  "Depth_1cm" = "Sediment deposition 1-5 cm",
  "Depth_0_5cm" = "Sediment deposition 0.5-1 cm",
  "Depth_0_1cm" = "Sediment deposition 0.1-0.5 cm",
  "Nearshore" = "Nearshore",
  "North" = "North",
  "East" = "East",
  "South" = "South",
  "Rest of Monitoring Area" = "Rest of impact monitoring area"
)

table2_area_order <- unname(coral_table2_area_labels)

if (!file.exists(coral_table2_intervals_csv)) {
  stop(
    "Could not find ", coral_table2_intervals_csv,
    ". Run analysis/PENIP_analysis.Rmd first to generate coral Table 2 intervals.",
    call. = FALSE
  )
}

table2_coral_intervals <- read_csv(coral_table2_intervals_csv, show_col_types = FALSE)

table2_coral <- table2_coral_intervals %>%
  filter(name != "Total") %>%
  transmute(
    `Impact area` = unname(coral_table2_area_labels[name]),
    ACER = format_interval_count(ACER, ACER_lower_95, ACER_upper_95),
    ORBI = format_interval_count(ORBI, ORBI_lower_95, ORBI_upper_95),
    `Non-ESA corals` = format_interval_count(`non-ESA`, `non-ESA_lower_95`, `non-ESA_upper_95`),
    `Total corals` = format_interval_count(Total, Total_lower_95, Total_upper_95)
  )

table2_coral_total <- table2_coral_intervals %>%
  filter(name == "Total") %>%
  transmute(
    `Impact area` = "Total",
    ACER = format_interval_count(ACER, ACER_lower_95, ACER_upper_95),
    ORBI = format_interval_count(ORBI, ORBI_lower_95, ORBI_upper_95),
    `Non-ESA corals` = format_interval_count(`non-ESA`, `non-ESA_lower_95`, `non-ESA_upper_95`),
    `Total corals` = format_interval_count(Total, Total_lower_95, Total_upper_95)
  )

if (!file.exists(conch_kriging_impact_zone_totals_rds)) {
  dir.create(dirname(conch_kriging_impact_zone_totals_rds), recursive = TRUE, showWarnings = FALSE)

  conch_integrated_script <- tempfile(fileext = ".R")
  knitr::purl("analysis/conch_integrated_analysis.Rmd", output = conch_integrated_script, quiet = TRUE)
  source(conch_integrated_script)

  saveRDS(
    kriging_impact_zone_totals_with_total,
    conch_kriging_impact_zone_totals_rds
  )
}

table2_conch <- readRDS(conch_kriging_impact_zone_totals_rds) %>%
  filter(impact_area != "Total") %>%
  transmute(
    `Impact area` = impact_area,
    `Adult conch` = format_interval_count(estimated_conch_Adult, lower_95_Adult, upper_95_Adult),
    `Adult conch within 330 m of breeding activity` = format_interval_count(
      estimated_adults_near_breeding,
      lower_95_adults_near_breeding,
      upper_95_adults_near_breeding
    ),
    `Juvenile conch` = format_interval_count(estimated_conch_Juvenile, lower_95_Juvenile, upper_95_Juvenile),
    `Total conch` = format_interval_count(estimated_total, total_lower_95, total_upper_95)
  )

table2_conch_total <- readRDS(conch_kriging_impact_zone_totals_rds) %>%
  filter(impact_area == "Total") %>%
  transmute(
    `Impact area` = "Total",
    `Adult conch` = format_interval_count(estimated_conch_Adult, lower_95_Adult, upper_95_Adult),
    `Adult conch within 330 m of breeding activity` = format_interval_count(
      estimated_adults_near_breeding,
      lower_95_adults_near_breeding,
      upper_95_adults_near_breeding
    ),
    `Juvenile conch` = format_interval_count(estimated_conch_Juvenile, lower_95_Juvenile, upper_95_Juvenile),
    `Total conch` = format_interval_count(estimated_total, total_lower_95, total_upper_95)
  )

table2_new <- full_join(table2_coral, table2_conch, by = "Impact area") %>%
  mutate(`Impact area` = factor(`Impact area`, levels = table2_area_order)) %>%
  mutate(
    `Impact area` = as.character(`Impact area`)
  ) %>%
  bind_rows(
    full_join(table2_coral_total, table2_conch_total, by = "Impact area")
  )

write_csv(table2_new, output_table2_new)

table2_new %>%
  knitr::kable()
```

```{r cleanup, include=FALSE}
sf_use_s2(old_s2)
```
