---
title: "Port Everglades Coral Survey Data Analysis"
author: "R. Cunning"
date: "2025-04-21"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2      
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Setup
```{r libraries}
# Load packages
library(sf)
library(xml2)
library(tidyverse)
library(tidybayes)
library(brms)
library(vegan)
```

```{r custom}
# Define genus level taxon groups (plus one family FAVI)
taxon_groups <- list(
  PORI = c("PPOR", "PFUR", "PDIV", "PAST", "PORI"),
  ORBI = c("OFAV", "OANN", "OFRA", "ORBI"),
  FAVI = c("CNAT", "DLAB", "PSTR", "PCLI", "MARE", "FAVI"),
  AGAR = c("AFRA", "AAGA", "AHUM", "ALAM", "AGAR"),
  MADR = c("MAUR", "MSEN", "MDEC", "MPHA", "MADR"),
  SOLE = c("SHYA", "SBOU", "SOLE"),
  SCOL = c("SLAC", "SCUB", "SCOL"),
  SIDE = c("SSID", "SRAD", "SIDE"),
  MYCE = c("MFER", "MLAM", "MALI", "MYCE"),
  OCUL = c("OROB", "ODIF", "OCUL")
)

# Convert to lookup tibble
taxon_lookup <- enframe(taxon_groups, name = "taxon_group", value = "taxon") %>%
  unnest(taxon)



# Define juvenile family level taxon groups (following DRM survey convention)
taxon_groups_juv <- list(
  MUSS = c("ISIN", "ISOP", "MANG", "MYCE", "SCOL", "MUSS"),
  FAVI = c("FAVI", "FFRA", "MARE"),
  MEAN = c("MMEA", "MEAN", "DCYL", "DSTO", "EFAS")
)
# Convert to lookup tibble
taxon_lookup_juv <- enframe(taxon_groups_juv, name = "taxon_group", value = "taxon") %>%
  unnest(taxon)


# Order levels of Habitat Types
type_levels <- c("Nearshore Ridge Complex", "Inner Reef", "Middle Reef",
                 "Outer Reef", "Aggregated Patch Reef")
type_labels <- c("NRC", "IR", "MR", "OR", "APR")
names(type_labels) <- type_levels

# Order survey datasets
dataset_levels <- c("dca17", "tt21", "tt23", "drm24")
dataset_labels <- c("2017—DCA", "2021—TT", "2023—TT", "2024—Shedd")
names(dataset_labels) <- dataset_levels
```

# Import data

## 2011 NSU ESA Survey
```{r}
# Import data
nsu11_esa0 <- readxl::read_xlsx("data/2011_nsu_esa/Port Everglades_NSU 2011and DCA 2017_ESA surveys.xlsx",
                           sheet = "2011_NSU_ESA survey") %>%
  janitor::clean_names()

# Site metadata
nsu11_esa_sitemd <- nsu11_esa0 %>%
  select(site = ident, latitude = lat, longitude = long) %>%
  mutate(site = as.character(site))



# ESA coral count data
nsu11_esa_counts <- nsu11_esa0 %>%
  select(site = ident,
         ACER = a_cervic_1, OANN = m_annula_1, OFAV = m_faveol_1, OFRA = m_franks_1, MFER = m_ferox_1) %>%
  pivot_longer(-site, names_to = "taxon", values_to = "n") %>%
  # Assume all were >4cm since no sizes are reported
  mutate(class = ">4cm") %>%
  mutate(site = as.character(site))

# Aggregate taxa (multiple orbicella observed --> ORBI)
nsu11_esa_counts_ag <- nsu11_esa_counts %>%
  left_join(taxon_lookup, by = "taxon") %>%
  mutate(taxon = coalesce(taxon_group, taxon)) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")
  
```

## 2017 DCA ESA Survey
```{r}
library(sf)

# Read KML file
dca17_esa_sitemd0 <- st_read("data/2017_dca_esa/Listed_Coral_Survey_2017.kml")

# Extract coordinates and save as CSV
dca17_esa_sitemd <- dca17_esa_sitemd0 %>%
  mutate(site = Name,
         longitude = st_coordinates(.)[, 1],
         latitude = st_coordinates(.)[, 2]) %>%
  st_drop_geometry() %>%
  select(site, longitude, latitude) %>%
  as_tibble() %>%
  mutate(site = as.character(site))




# Get coral data
# Appendix C -- does not contain individual sizes
# dca17_esa0 <- readxl::read_xlsx("data/2017_dca_esa/Appendix C_DCA2017_survey_results_FINAL.xlsx") %>%
#   janitor::clean_names()
# dca17_esa0 %>%
#   select(site = esa_site, ACER = a_cervicornis_n, OFAV = o_faveolata_n, 
#          area_m2 = habitat_surveyed_m2_based_on_walker_and_klug_2014)

# Appendix D -- contains individual colony sizes
# what is the "NUMBER" column? ignore for now.
dca17_esa0 <- readxl::read_xlsx("data/2017_dca_esa/Appendix D_ESA_listed_coral_Plotted_Locations.xlsx") %>%
  janitor::clean_names() %>%
  select(site = site_id, taxon = species, length_cm, width_cm, height_cm, m2_of_habi, density) %>%
  mutate(across(ends_with("cm"), as.numeric))

# Assign size classes
dca17_esa <- dca17_esa0 %>%
  mutate(max_dim_cm = pmax(length_cm, width_cm, height_cm, na.rm = TRUE),
         class = if_else(max_dim_cm >= 4, ">4cm", "<4cm"))

# Count taxon and size class per site
dca17_esa_counts <- dca17_esa %>%
  count(site, taxon, class) %>%
  mutate(site = as.character(site))

# Total area surveyed per site = 784 m2 (crossed 100x4m belt transects with 16m2 overlap)

# Agreggate taxa (OFAV -> ORBI)
dca17_esa_counts_ag <- dca17_esa_counts %>%
  left_join(taxon_lookup, by = "taxon") %>%
  mutate(taxon = coalesce(taxon_group, taxon)) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")
```


## 2017 DCA Recon Survey

DCA site metadata
```{r 2017_DCA_site_metadata}
# Site metadata
# Read in site coordinates
dca17_sitemd0 <- readxl::read_xlsx("data/2017_dca_recon/Recon_Site_Coordinates_Extracted.xlsx") %>%
  janitor::clean_names()

# All sites have start and end coordinates...

# Tidy and Calculate midpoint per transect
dca17_sitemd <- dca17_sitemd0 %>%
  mutate(
    depth = abs(as.numeric(depth)),
    across(c(latitude, longitude), as.numeric)
  ) %>%
  group_by(site = transect) %>%
  summarize(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    .groups = "drop"
  )
```

DCA coral data
```{r 2017_DCA_coral_data}
# Read in survey data
dca170 <- readxl::read_xlsx("data/2017_dca_recon/Compiled_DCA_RECON_Belt_data.xlsx") %>%
  janitor::clean_names()

dca17 <- dca170 %>%
  select(1:18) %>%
  rename(site = site_name) %>%
  mutate(site = factor(site)) %>%
  select(site, taxon = coral_species, max_width_cm = max_size_cm)

# Adjust/corrects species IDs
sort(unique(dca17$taxon))
dca17 <- dca17 %>%
  mutate(taxon = case_when(
    taxon == "AGA SP" ~ "AGAR",
    taxon == "LCUC" ~ "HCUC",
    taxon %in% c("MYCSP", "Mycetophyllia spp.") ~ "MYCE",
    taxon == "OFAV\\" ~ "OFAV",
    taxon %in% c("MAD SP", "MADSP") ~ "MADR",
    taxon == "Scolymia Spp" ~ "SCOL",
    taxon %in% c("SIDSP", "Sid SP", "SID SP.", "SID SP") ~ "SIDE",
    TRUE ~ taxon
  ))
sort(unique(dca17$taxon))

# Filter out unidentified corals
dca17 <- dca17 %>%
  filter(!taxon %in% c("CORAL", "Cup Coral"))

# Write long data to file
write_csv(dca17, file = "data/processed/dca_2017_long.csv")


# Convert to count data
# Add explicit zeros for any taxon/size class missing at any site
dca17_counts <- dca17 %>%
  mutate(class = ifelse(max_width_cm >= 4, ">4cm", "<4cm")) %>%
  count(site, taxon, class) %>%
  complete(site, taxon, class = c(">4cm", "<4cm"), fill = list(n = 0)) %>%
  # # Don't create zeros for MEAN/MUSS/FAVI adults, since these IDs only applied to juv
  filter(!(taxon %in% c("MEAN", "MUSS", "FAVI") & class == ">4cm" & n == 0))

write_csv(dca17_counts, file = "data/processed/dca_2017_counts.csv")

# Aggregate count data based on taxonomic groups defined above
dca17_counts_ag <- dca17_counts %>%
  left_join(taxon_lookup, by = "taxon") %>%
  mutate(taxon = coalesce(taxon_group, taxon)) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")



# Further aggregate juvenile counts to family (following DRM methods)
dca17_counts_ag <- dca17_counts_ag %>%
  left_join(taxon_lookup_juv, by = "taxon") %>%
  mutate(
    taxon = if_else(class == "<4cm" & !is.na(taxon_group), taxon_group, taxon)
  ) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")


write_csv(dca17_counts_ag, file = "data/processed/dca_2017_counts_ag.csv")
```

## 2021 TT Recon & ESA Survey

2021 TT site metadata
```{r tt21_site_metadata}
# site metadata
tt21_sitemd0 <- read_csv("data/2021_tt_recon_esa/midpoints_latlon.csv") %>%
  janitor::clean_names()

tt21_sitemd <- tt21_sitemd0 %>%
  mutate(name = str_remove(name, "A$")) %>%
  select(site = name, longitude = lon, latitude = lat)

# There was a lot of sand in these transects that was quantified in final RECON report (though these were the same transects for the ESA and RECON datasets). Extracted this data from RECON report, Table 2:
tt21_sand <- read_csv("data/2021_tt_recon_esa/Table_2_Port_Everglades_RECON.csv") %>%
  janitor::clean_names()

tt21_m_nosand <- tt21_sand %>%
  mutate(area_m2 = 30 - meters_of_sc_sp) %>%
  mutate(site = as.character(site))

```

2021 TT coral data
```{r tt21_coral_data}
# coral data - recon belt transects
tt21recon0 <- readxl::read_xlsx("data/2021_tt_recon_esa/Recon 30x1m Coral Belt Transect.xlsx") %>%
  janitor::clean_names()

tt21recon <- tt21recon0 %>%
  select(site, taxon = id_abbrev, coral_length_cm, coral_width_cm) %>%
  mutate(taxon = toupper(taxon), site = factor(site)) %>%
  mutate(across(c(coral_length_cm, coral_width_cm), as.numeric)) %>%
  mutate(max_width_cm = pmax(coral_length_cm, coral_width_cm)) %>%
  select(site, taxon, max_width_cm)

# ESA survey data
tt21esa0 <- readxl::read_xlsx("data/2021_tt_recon_esa/ESA Coral Belt Transect.xlsx") %>%
  janitor::clean_names()

sort(unique(tt21esa0$esa_id))

tt21esa <- tt21esa0 %>%
  mutate(site = factor(site),
         taxon = case_when(
           esa_id == "Orbicella franksi" ~ "OFRA",
           esa_id == "Orbicella faveolata" ~ "OFAV",
           esa_id == "Acropora cervicornis" ~ "ACER")) %>%
  filter(!is.na(taxon)) %>%
  mutate(max_width_cm = pmax(coral_length_cm, coral_width_cm)) %>%
  select(site, taxon, max_width_cm)

# Combine Recon and ESA survey data
tt21 <- bind_rows(tt21recon, tt21esa)


# Check taxa names
sort(unique(tt21recon$taxon))

# Filter out unidentified OR NON-CORAL taxa
tt21 <- tt21 %>%
  filter(!taxon %in% c("0", "JUVENILE-UNIDENTIFIABLE", "XESTO", "MALC"))

# Adjust/corrects species IDs
tt21 <- tt21 %>%
  mutate(taxon = case_when(
    taxon == "AFRAG" ~ "AFRA",
    taxon == "FFRAG" ~ "FFRA",
    taxon == "MYALI" ~ "MALI",
    taxon == "MYFER" ~ "MFER",
    taxon == "MYLAM" ~ "MLAM",
    taxon == "ODIF/OROB" ~ "OCUL",
    taxon == "PDCLIV" ~ "PCLI",
    taxon == "PDSTR" ~ "PSTR",
    taxon == "PHYLLANGIA AMERICANA" ~ "PAME",
    taxon == "SCOLYMIA CUBENSIS" ~ "SCUB",
    taxon == "SCOLYMIA LACERA" ~ "SLAC",
    TRUE ~ taxon
  ))
sort(unique(tt21$taxon))

# Write long data to file
write_csv(tt21, file = "data/processed/tt_2021_long.csv")





# Count
# Add explicit zeros for any taxon/size class missing at any site
tt21_counts <- tt21 %>%
  mutate(class = ifelse(max_width_cm >= 4, ">4cm", "<4cm")) %>%
  count(site, taxon, class) %>%
  complete(site, taxon, class = c(">4cm", "<4cm"), fill = list(n = 0)) %>%
  # Don't create zeros for MEAN/MUSS/FAVI adults, since these IDs only applied to juv
  filter(!(taxon %in% c("MEAN", "MUSS", "FAVI") & class == ">4cm" & n == 0))

write_csv(tt21_counts, file = "data/processed/tt_2021_counts.csv")




# Aggregate count data based on taxonomic groups defined above
tt21_counts_ag <- tt21_counts %>%
  left_join(taxon_lookup, by = "taxon") %>%
  mutate(taxon = coalesce(taxon_group, taxon)) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")

# Further aggregate juvenile counts to family (following DRM methods)
tt21_counts_ag <- tt21_counts_ag %>%
  left_join(taxon_lookup_juv, by = "taxon") %>%
  mutate(
    taxon = if_else(class == "<4cm" & !is.na(taxon_group), taxon_group, taxon)
  ) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")

write_csv(tt21_counts_ag, file = "data/processed/tt21_counts_ag.csv")
```

## 2023 TT Impact Survey

TT site metadata
```{r 2023_tt_site_metadata}
# Site metadata
tt23_sitemd <- readxl::read_xlsx("data/2023_tt_impact/Impact site tracking.xlsx", skip = 1) %>%
  janitor::clean_names()

tt23_sitemd <- tt23_sitemd %>%
  mutate(site = transect_name,
         latitude = as.numeric(actual_start_y),
         longitude = as.numeric(actual_start_x)) %>%
  select(site, latitude, longitude) 

# Many sites missing coords in sheet.... whats up with that
tt23_sitemd <- drop_na(tt23_sitemd, longitude)
```

TT coral data
```{r 2023_tt_coral_data}
tt230 <- readxl::read_xlsx("data/2023_tt_impact/Impact Raw Data 05 31 2024.xlsx") %>%
  janitor::clean_names() 

tt23 <- tt230 %>%
  select(site = transect_name, depth_ft_start,
         taxon = id_abbrev, coral_length_cm, coral_width_cm) %>%
  filter(taxon != "Xesto") %>%
  mutate(taxon = toupper(taxon)) %>%
  mutate(across(c(coral_length_cm, coral_width_cm), as.numeric)) %>%
  mutate(site = factor(site))

tt23 <- tt23 %>%
  mutate(max_width_cm = pmax(coral_length_cm, coral_width_cm)) %>%
  select(site, taxon, max_width_cm)

# Check taxa names
sort(unique(tt23$taxon))

# Filter out unidentified taxa
tt23 <- tt23 %>%
  filter(!taxon %in% c("?", "ID-ABBREV", "NONE", "MHEARD"))

# Adjust/corrects species IDs
tt23 <- tt23 %>%
  mutate(taxon = case_when(
    taxon == "AFRAG" ~ "AFRA",
    taxon %in% c("ASP", "ASP.") ~ "AGAR",
    taxon == "MCAV?" ~ "MCAV",
    taxon == "MSP." ~ "MADR",
    taxon == "MUSSID" ~ "MUSS",
    taxon == "MYALI" ~ "MALI",
    taxon == "MYFER" ~ "MFER",
    taxon == "MYLAM" ~ "MLAM",
    taxon == "OFR" ~ "OFRA",
    taxon == "PCLI?" ~ "PCLI",
    taxon %in% c("PSP", "PSP.") ~ "PORI",
    taxon == "SSP." ~ "SIDE",
    taxon == "STOK" ~ "DSTO",
    TRUE ~ taxon
  ))
sort(unique(tt23$taxon))

# Write long data to file
write_csv(tt23, file = "data/processed/tt_2024_long.csv")





# Count
# Add explicit zeros for any taxon/size class missing at any site
tt23_counts <- tt23 %>%
  mutate(class = ifelse(max_width_cm >= 4, ">4cm", "<4cm")) %>%
  count(site, taxon, class) %>%
  complete(site, taxon, class = c(">4cm", "<4cm"), fill = list(n = 0)) %>%
  # Don't create zeros for MEAN/MUSS/FAVI adults, since these IDs only applied to juv
  filter(!(taxon %in% c("MEAN", "MUSS", "FAVI") & class == ">4cm" & n == 0))

write_csv(tt23_counts, file = "data/processed/tt_2024_counts.csv")




# Aggregate count data based on taxonomic groups defined above
tt23_counts_ag <- tt23_counts %>%
  left_join(taxon_lookup, by = "taxon") %>%
  mutate(taxon = coalesce(taxon_group, taxon)) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")

# Further aggregate juvenile counts to family (following DRM methods)
tt23_counts_ag <- tt23_counts_ag %>%
  left_join(taxon_lookup_juv, by = "taxon") %>%
  mutate(
    taxon = if_else(class == "<4cm" & !is.na(taxon_group), taxon_group, taxon)
  ) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")

write_csv(tt23_counts_ag, file = "data/processed/tt23_counts_ag.csv")
```

## 2024 DRM/Shedd Survey

Shedd site metadata
```{r 2024_shedd_site_metadata}
# Site metadata
drm24_sitemd <- readxl::read_xlsx("data/2024_shedd_drm/site_metadata.xlsx") %>%
  janitor::clean_names() %>%
  mutate(site = as.character(drm_site_id)) %>%
  select(site, latitude = lat, longitude = lon) %>%
  mutate(depth = NA)
```

Shedd coral data
```{r 2024_shedd_coral_data}
# Adult coral data from main DRM surveys
#adults0 <- read_csv("data/2024_shedd_drm/DRM_broward_corals.csv")
#adults0 %>% filter(team == "Shedd Aquarium")
alldrm2024 <- readxl::read_xlsx("data/2024_shedd_drm/2024ANU_RawCoralDataTransect1and2_Shedd.xlsx") %>%
  janitor::clean_names() %>%
  distinct(site, team, date, subregion)

shedddrm2024 <- alldrm2024 %>% filter(team == "Shedd Aquarium")

shedddrm2024 %>%
  count(subregion, date)

# Most sites were included in main DRM database for 2024 -- Import these
adultst1t2 <- readxl::read_xlsx("data/2024_shedd_drm/2024ANU_RawCoralDataTransect1and2_Shedd.xlsx") %>%
  janitor::clean_names() %>%
  filter(subregion == "Broward-Miami", team == "Shedd Aquarium") %>%
  select(site, transect_num, species, width, height)

adultst3t4 <- readxl::read_xlsx("data/2024_shedd_drm/2024ANU_RawCoralDataTransect3and4_Shedd.xlsx") %>%
  janitor::clean_names() %>%
  filter(subregion == "Broward-Miami", team == "Shedd Aquarium") %>%
  select(site, transect_num, species, width, height)

# 9 of our PEV sites were removed from DRM database to avoid oversaturating the ares -- Import these separately
removedt1t2 <- readxl::read_xlsx(
  "data/2024_shedd_drm/2024_DRM_Broward_RemovedSites_T1-T4_Shedd.xlsx", sheet = "Removed Sites T1-T2") %>%
  janitor::clean_names() %>%
  select(site, transect_num, species, width, height)
removedt3t4 <- readxl::read_xlsx(
  "data/2024_shedd_drm/2024_DRM_Broward_RemovedSites_T1-T4_Shedd.xlsx", sheet = "Removed Sites T3-T4") %>%
  janitor::clean_names() %>%
  select(site, transect_num, species, width, height)

# Combine all adult coral data for Shedd DRM surveys at PEV
adults0 <- bind_rows(adultst1t2, adultst3t4, removedt1t2, removedt3t4)
# Convert adult data to long format
adults_long <- adults0 %>%
  mutate(max_width_cm = pmax(width, height, na.rm = TRUE)) %>%
  mutate(max_width_cm = as.character(max_width_cm)) %>%
  mutate(site = str_remove(site, "^AA")) %>%
  select(site, transect_num, taxon = species, max_width_cm) %>%
  drop_na(taxon)     # DROPS when taxon is blank, this is when no corals >4cm were observed

# Import juvenile counts from main DRM dataset
juv <- readxl::read_xlsx("data/2024_shedd_drm/2024ANU_JuvenileCoralData_Shedd.xlsx") %>%
  janitor::clean_names() %>%
  filter(subregion == "Broward-Miami", team == "Shedd Aquarium")

# Import juvenile counts from sites that were removed from main DRM dataset
removed_juv <- readxl::read_xlsx("data/2024_shedd_drm/Shedd_removed_sites_Juveniles_2024.xlsx") %>%
  janitor::clean_names() %>%
  # Missing values in count data should be zero counts (unique to this datasheet from FWC)
  mutate(across(ends_with("_ct"), ~replace_na(., 0)))

# Combine juvenile data
juv0 <- bind_rows(juv, removed_juv) %>%
  mutate(site = str_remove(site, "^AA")) %>%
  select(site, transect_num, ends_with("ct")) %>%
  rename(MCAV = montastraea_ct, MUSS = mussinae_ct, FAVI = faviinae_ct, MEAN = meandrinidae_ct)

# Convert juvenile data to long format
juv_long <- juv0 %>%
  pivot_longer(c(MUSS, FAVI, MEAN, MCAV), names_to = "taxon", values_to = "n") %>%
  mutate(max_width_cm = "<4") %>%
  uncount(n)



# Other juvenile taxa counts from Transects 1 and 2 (DRM 'bonus data')
t1t2bonus <- read_csv("data/2024_shedd_drm/T1_T2_bonus_data.csv") %>%
  janitor::clean_names() %>%
  mutate(site = replace_na(site, "NA")) %>%    # Because one site is called "NA"
  mutate(transect_num = parse_number(transect))
t1t2juv <- t1t2bonus %>%
  select(site, transect_num, starts_with("small")) %>%
  rename_with(~ toupper(gsub("^small_", "", .x)), starts_with("small_"))
t1t2juv_long <- t1t2juv %>%
  pivot_longer(3:10, names_to = "taxon", values_to = "n") %>%
  mutate(max_width_cm = "<4") %>%
  uncount(n)
# Replace site names in t1t2 bonus data with the correct DRM site ID
penipsites <- readxl::read_xlsx("data/2024_shedd_drm/site_metadata.xlsx") %>%
  janitor::clean_names()
t1t2juv_long_updated <- t1t2juv_long %>%
  left_join(penipsites %>% select(site, drm_site_id), by = "site") %>%
  mutate(site = as.character(drm_site_id)) %>%
  select(-drm_site_id)


# Combine all data
drm24_long <- bind_rows(adults_long, juv_long, t1t2juv_long_updated) %>%
  mutate(team = "Shedd Aquarium")

# Check species names
sort(unique(drm24_long$taxon))

write_csv(drm24_long, file = "data/processed/drm_2024_long.csv")


# COUNT based on rules
# ✅ Updated Rules Summary for counting from DRM/Shedd data:
# Juvenile taxa (searched for in <4cm size class only):
#    "MEAN", "MUSS", "FAVI"
#    → these should only ever appear in <4cm, never >4cm, and should not be zero-filled for adults.
# Other juvenile-capable taxa:
#    "MCAV", "SSID", "SRAD", "PAST", "PPOR", "SINT", "SBOU", "AAGA", "MAUR"
#    → these can be counted in both >4cm and <4cm, but only in <4cm if juveniles were searched on that transect and team.
# Transect-based search rules still apply:
# Transects 1 & 2: all adult taxa always searched. Juvenile search depends on team:
#    "Shedd Aquarium" → all juvenile taxa above searched
#    others → only MEAN, MUSS, FAVI, MCAV
# Transects 3 & 4:
#    only subset of adult taxa searched (adult_taxa_t3t4)
#    only juveniles: MEAN, MUSS, FAVI, MCAV

# Step 1: Define size classes
drm24_classed <- drm24_long %>%
  mutate(class = case_when(as.numeric(max_width_cm) >= 4 ~ ">4cm",
                           max_width_cm == "<4" ~ "<4cm"))

# Step 2: Define species sets
all_taxa <- unique(drm24_classed$taxon)
adult_taxa_t3t4 <- c("CNAT", "DSTO", "DLAB", "MMEA", "MANG", "MALI", 
                     "MFER", "MLAM", "PCLI", "PSTR")
juv_only_taxa <- c("MEAN", "MUSS", "FAVI")
juv_both_taxa <- c("MCAV", "SSID", "SRAD", "PAST", "PPOR", "SINT", "SBOU", "AAGA", "MAUR")
all_juv_taxa <- c(juv_only_taxa, juv_both_taxa)

# Step 3: Build search grid per site × transect × team
search_grid <- drm24_classed %>%
  distinct(site, transect_num, team) %>%    # if multiple teams in data, remove value for team
  mutate(
    searched_taxa_class = pmap(list(transect_num, team), function(transect, team) {
      # Helper: define juv taxa allowed for this transect/team
      juv_taxa <- if (transect %in% c(1, 2)) {
        if (team == "Shedd Aquarium") {      # Shedd searched for other juv taxa on T1 and T2, other DRM survey teams did not
          all_juv_taxa
        } else {
          c(juv_only_taxa, "MCAV")
        }
      } else {
        c(juv_only_taxa, "MCAV")
      }
      
      # Adults always searched in 1 & 2, subset in 3 & 4
      adult_taxa <- if (transect %in% c(1, 2)) {
        setdiff(all_taxa, juv_only_taxa)  # exclude juv-only taxa
      } else {
        adult_taxa_t3t4
      }

      # Build grid
      bind_rows(
        expand_grid(taxon = adult_taxa, class = ">4cm"),
        expand_grid(taxon = juv_taxa, class = "<4cm")
      )
    })
  ) %>%
  unnest(searched_taxa_class)

# Step 4: Count observations
counts <- drm24_classed %>%
  group_by(site, transect_num, team, taxon, class) %>%
  summarize(n = n(), .groups = "drop")

# Step 5: Join with grid and fill in zeros where appropriate
final_counts <- search_grid %>%
  left_join(counts, by = c("site", "transect_num", "team", "taxon", "class")) %>%
  mutate(n = replace_na(n, 0))

write_csv(final_counts, file = "data/processed/drm_2024_counts.csv")




# AGGREGATE COUNT DATA
# Aggregate taxa
drm24_counts_ag <- final_counts %>%
  left_join(taxon_lookup, by = "taxon") %>%
  mutate(taxon = coalesce(taxon_group, taxon)) %>%
  select(-taxon_group) %>%
  group_by(across(-n)) %>%
  summarize(n = sum(n), .groups = "drop")

write_csv(drm24_counts_ag, file = "data/processed/drm_2024_counts_ag.csv")
```

## Habitat maps
```{r habitat_maps, fig.width = 10, fig.height = 10}
# --- STEP 1: Load KML Polygons ---
polygons <- st_read("data/Habitat classifications.kml")  # update path as needed

# --- STEP 2: Extract Attributes from HTML Description ---
extract_attrs <- function(desc) {
  if (is.na(desc) || desc == "") {
    return(tibble(Habitat = NA, Type = NA, Modifier = NA, Region = NA, Type2 = NA))
  }
  html <- read_html(desc)
  rows <- xml_find_all(html, "//table//table//tr")
  keys <- rows %>% xml_find_all(".//td[1]") %>% xml_text(trim = TRUE)
  vals <- rows %>% xml_find_all(".//td[2]") %>% xml_text(trim = TRUE)
  n <- min(length(keys), length(vals))
  named_vals <- set_names(vals[1:n], keys[1:n])
  tibble(
    Habitat  = named_vals[["Habitat"]],
    Type     = named_vals[["Type"]],
    Modifier = named_vals[["Modifier"]],
    Region   = named_vals[["Region"]],
    Type2    = named_vals[["Type2"]]
  )
}

# Apply function and combine with spatial geometries
attrs <- map_dfr(polygons$Description, extract_attrs)
polygons_clean <- bind_cols(polygons %>% select(-Description), attrs)

# --- STEP 3: Prepare Site Coordinate Data ---
# Get all site coordinates, and assign north and south
# Combine site metadata
allsitemd <- bind_rows(.id = "dataset",
  nsu11_esa = nsu11_esa_sitemd,
  dca17_esa = dca17_esa_sitemd,
  dca17 = dca17_sitemd,
  tt21 = tt21_sitemd,
  tt23 = tt23_sitemd,
  drm24 = drm24_sitemd
) %>%
  mutate(dir = if_else(latitude > 26.093570, "N", "S"))
points <- st_as_sf(allsitemd, coords = c("longitude", "latitude"), crs = 4326)

# --- STEP 4: Validate Geometry and Match CRS ---
polygons_clean <- polygons_clean %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(st_crs(points))

sf_use_s2(FALSE)  # prevent s2 geometry issues

# --- STEP 5: Spatial Join ---
joined <- st_join(points, polygons_clean, join = st_within)

joined_df <- joined %>%
  mutate(longitude = st_coordinates(.)[,1],
         latitude = st_coordinates(.)[,2]) %>%
  st_drop_geometry()

allsitemd <- joined_df

# 'Sand' overlaps some of the other polygons instead of just surrounding them...
# If a point is classified as Sand AND something else, remove Sand...
multiclass <- allsitemd %>%
  group_by(site) %>%
  filter(n() > 1)

allsitemd_clean <- allsitemd %>%
  group_by(site) %>%
  filter(!(Type == "Sand" & n() > 1)) %>%
  ungroup()

# Add factor if survey was ESA coral species only
allsitemd_clean <- allsitemd_clean %>%
  mutate(ESAonly = if_else(dataset %in% c("nsu11_esa", "dca17_esa"), "ESA only", "All corals"))

# Visualize habitat classifications
polyplot <- polygons_clean %>% 
  #filter(Type != "Sand") %>%
  ggplot() +
  geom_sf(aes(fill = Type), color = "black", size = 0.2, alpha = 0.6) +
  scale_fill_brewer(palette = "Set3", na.value = "gray80") +
  facet_wrap(~ESAonly) +
  theme_minimal() +
  labs(title = "Habitat Polygons Colored by Type", fill = "Type") +
  theme(legend.position = "right") +
  xlim(-80.11, -80.079) +
  ylim(26.0675, 26.11)

# Plot all surveyed sites
polyplot + 
  geom_point(data = allsitemd_clean, 
             aes(x = longitude, y = latitude, shape = dataset), 
             inherit.aes = FALSE, alpha = 0.6) +
  scale_shape_manual(values = c(2, 3, 4, 5, 6, 7)) +
  labs(title = "All sites - all surveys")
```

# Combine and filter data

## Combine surveys datasets
```{r combine_data}
# Combine all aggregated count data
df <- bind_rows(.id = "dataset",
  nsu11_esa = nsu11_esa_counts_ag,
  dca17_esa = dca17_esa_counts_ag,
  dca17 = dca17_counts_ag,
  tt21 = tt21_counts_ag,
  tt23 = tt23_counts_ag,
  drm24 = select(drm24_counts_ag, -team)
) %>%
  mutate(transect_num = if_else(is.na(transect_num), 1, transect_num))

# Add transect area information
## Site-specific areas for tt21 (since there was sand in transects)
tt21_areas <- tt21_m_nosand %>%
  mutate(dataset = "tt21") %>%
  distinct(dataset, site, transect_area_m2 = area_m2)
df <- df %>% left_join(tt21_areas)

## Other datasets had fixed transect areas
df <- df %>%
  mutate(transect_area_m2 = case_when(
     # Count-specific areas for nsu11 (since they did tier 1 surveys and tier 2 only if n > 5 for a taxon)
    dataset == "nsu11_esa" & n >= 5 ~ 600,  
    dataset == "nsu11_esa" & n < 5 ~ 3600,
    dataset == "dca17_esa" ~ 784,
    dataset == "dca17" ~ 30,     # DCA belt transects were 30m
    dataset == "tt23" ~ 20,     # TT23 belt transects were 20m
    dataset == "drm24" ~ 10,     # DRM belt transects were 10m)
    TRUE ~ transect_area_m2))

# Remove surveys with transect_area_m2 == 0    (a couple of TT21 surveys were all sand)
df <- df %>% filter(transect_area_m2 > 0)
```


## Filter by habitat

```{r select_habitats, fig.width = 10, fig.height = 10}
# Select sites in Nearshore Ridge Complex, Inner Reef, and Middle Reef, Outer Reef and Aggregated Patch Reef
selected <- allsitemd_clean %>%
  filter(Type %in% c("Nearshore Ridge Complex", "Inner Reef", "Middle Reef", 
                     "Artificial", "Outer Reef", "Aggregated Patch Reef"))

# Plot all selected sites
polyplot + 
  geom_point(data = selected, 
             aes(x = longitude, y = latitude, shape = dataset), 
             inherit.aes = FALSE, alpha = 0.6) +
  scale_shape_manual(values = c(2, 3, 4, 5, 6, 7)) +
  facet_wrap(~ESAonly) +
  labs(title = "Selected sites - all surveys")


# Filter dataset to just selected sites
dff <- df %>% inner_join(selected)

```

###### Shedd dataset has some sites in the NRC south that are further from the channel relative to other datasets. Are these similar enough to the NRC south sites that are closer to the channel that they can be reasonably included?

First, look at total coral density between the nearer vs. farther sites

```{r NRCS_density}
drm24_NRCS <- drm24_counts_ag %>%
  left_join(allsitemd_clean) %>%
  filter(Type == "Nearshore Ridge Complex", dir == "S")

drm24_NRCS <- drm24_NRCS %>%
  mutate(grp = if_else(latitude > 26.08, "near", "far"))

# Fit a Negative Binomial GLM
mod_nb <- MASS::glm.nb(n ~ grp, data = drm24_NRCS)

# Generate new data only for existing taxon-size class combinations
newdata_1 <- drm24_NRCS %>%
  distinct(grp)

# Get predicted values & standard errors (log scale)
preds_nb <- predict(mod_nb, newdata_1, type = "link", se.fit = TRUE)

# Compute total coral density 
results_nb <- newdata_1 %>%
  mutate(
    fit = exp(preds_nb$fit),                    # Convert fitted values to response scale
    fit_se = exp(preds_nb$fit) * preds_nb$se.fit,   # Convert SE using the Delta Method
    fit_var = (fit * preds_nb$se.fit)^2,         # Variance propagation
    fit_lower = exp(preds_nb$fit - 1.96 * preds_nb$se.fit),  # Lower CI
    fit_upper = exp(preds_nb$fit + 1.96 * preds_nb$se.fit)   # Upper CI
  )

# Compute total coral density + confidence intervals
total_ci_nb <- results_nb %>%
  group_by(grp) %>%
  summarize(
    total_density = sum(fit),
    total_se = sqrt(sum(fit_var)),
    lower_95CI = exp(log(total_density) - 1.96 * (total_se / total_density)),
    upper_95CI = exp(log(total_density) + 1.96 * (total_se / total_density))
  )

knitr::kable(total_ci_nb)

```

Next, look at community composition between the nearer vs. farther sites
```{r NRCS_community_composition}

# 1. Create a wide community matrix: rows = site x grp, columns = taxa
comm_matrix <- drm24_NRCS %>%
  group_by(site, grp, taxon) %>%
  summarize(total_n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = taxon, values_from = total_n, values_fill = 0)

# 2. Extract community matrix and metadata
comm_data <- comm_matrix %>% select(-site, -grp)
site_info <- comm_matrix %>% select(site, grp)

# 3. Run NMDS (k = 2 dimensions is standard)
set.seed(123)  # for reproducibility
nmds <- metaMDS(comm_data, distance = "bray", k = 2, trymax = 100)

# 4. Prepare data for plotting
scores_df <- scores(nmds)$sites %>%
  bind_cols(site_info)

# 5. Plot NMDS with ggplot2
ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = grp)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "NMDS of Coral Community Structure\nby Dist from Channel in NRC South",
       color = "Dist from channel")
```

Density and community composition are not different in the NRC S sites that are nearer vs. farther from the channel. So, no need to exclude the farther sites.


###### DCA dataset has the most sites in the "Artificial" habitat classification. (Some in other datasets too, but DCA has most). Can 'Artificial' be aggregated with 'Nearshore Ridge Complex'?

'Artificial' is only present in DCA and minorly in TT -- but absent from Shedd.   
It is in close spatial proximity to Nearshore Ridge Complex -- can these be combined?

Test for differences in coral density in DCA survey between habitat types
```{r artificial_vs_nrc_density}
# Subset DCA data
dcadf <- dca17_counts_ag %>%
  left_join(allsitemd_clean) %>%
  filter(Type %in% c("Nearshore Ridge Complex", "Inner Reef", "Middle Reef", "Artificial", "Outer Reef", "Aggregated Patch Reef"))

# Fit a Negative Binomial GLM
mod_nb <- MASS::glm.nb(n ~ Type, data = dcadf)

# Generate new data only for existing taxon-size class combinations
newdata_1 <- dcadf %>%
  distinct(Type)

# Get predicted values & standard errors (log scale)
preds_nb <- predict(mod_nb, newdata_1, type = "link", se.fit = TRUE)

# Compute both total coral density & taxon-size class-specific densities in one step
results_nb <- newdata_1 %>%
  mutate(
    fit = exp(preds_nb$fit),                    # Convert fitted values to response scale
    fit_se = exp(preds_nb$fit) * preds_nb$se.fit,   # Convert SE using the Delta Method
    fit_var = (fit * preds_nb$se.fit)^2,         # Variance propagation
    fit_lower = exp(preds_nb$fit - 1.96 * preds_nb$se.fit),  # Lower CI
    fit_upper = exp(preds_nb$fit + 1.96 * preds_nb$se.fit)   # Upper CI
  )

# Compute total coral density + confidence intervals
total_ci_nb <- results_nb %>%
  group_by(Type) %>%
  summarize(
    total_density = sum(fit),
    total_se = sqrt(sum(fit_var)),
    lower_95CI = exp(log(total_density) - 1.96 * (total_se / total_density)),
    upper_95CI = exp(log(total_density) + 1.96 * (total_se / total_density))
  )

knitr::kable(total_ci_nb)
```
Highly overlapping confidence intervals for Artifical and Nearshore Ridge Complex, so no difference in coral density.

Test for differences in community composition in DCA survey
```{r artificial_vs_nrc_nmds}
library(vegan)

# 1. Create a wide community matrix: rows = site x Type, columns = taxa
comm_matrix <- dcadf %>%
  group_by(site, Type, taxon) %>%
  summarize(total_n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = taxon, values_from = total_n, values_fill = 0)

# 2. Extract community matrix and metadata
comm_data <- comm_matrix %>% select(-site, -Type)
site_info <- comm_matrix %>% select(site, Type)

# 3. Run NMDS (k = 2 dimensions is standard)
set.seed(123)  # for reproducibility
nmds <- metaMDS(comm_data, distance = "bray", k = 2, trymax = 100)

# 4. Prepare data for plotting
scores_df <- scores(nmds)$sites %>%
  bind_cols(site_info)

# 5. Plot NMDS with ggplot2
ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = Type)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "NMDS of Coral Community Structure by Habitat Type",
       color = "Habitat Type")
```

High degree of similarity between Artificial and Nearshore Ridge Complex coral communities.

Combine 'Artificial' with 'Nearshore Ridge Complex'
```{r combine_artificial_nrc}
# Based on these results, combine "Artificial" with "Nearshore Ridge Complex"
dff[dff$Type == "Artificial", "Type"] <- "Nearshore Ridge Complex"
```

## Filter by coral taxa

Remove lowest abundance coral taxa

These will be problematic for statistical models

```{r filter_low_taxa}
# Drop taxa with very few observations
sppcounts <- dff %>%
  group_by(taxon) %>%
  summarize(total = sum(n), .groups = "drop") %>%
  arrange(total) 
sppcounts

dff <- dff %>%
  filter(taxon %in% filter(sppcounts, total >= 5)$taxon) %>%
  mutate(Type = factor(Type, levels = type_levels))
```

Taxa with less than 5 total observations were filtered out, which included: HCUC, MANG, PAME, FFRA, OCUL, SCOL

# Map datasets
```{r plot_maps, fig.width = 10, fig.height = 10}
(nsu11_esa_plot <- polyplot + 
  geom_point(
    data = filter(selected, dataset == "nsu11_esa"),
    aes(x = longitude, y = latitude), pch = 13, inherit.aes = FALSE, alpha = 0.6) +
  labs(title = "2011 — NSU ESA Survey"))

(dca17_esa_plot <- polyplot + 
  geom_point(
    data = filter(selected, dataset == "dca17_esa"),
    aes(x = longitude, y = latitude), pch = 13, inherit.aes = FALSE, alpha = 0.6) +
  labs(title = "2017 — DCA ESA Survey"))

(dca_plot <- polyplot + 
  geom_point(
    data = filter(selected, dataset == "dca17"),
    aes(x = longitude, y = latitude), pch = 13, inherit.aes = FALSE, alpha = 0.6) +
  labs(title = "2017 — DCA RECON Survey"))

(tt21_plot <- polyplot + 
  geom_point(
    data = filter(selected, dataset == "tt21"),
    aes(x = longitude, y = latitude), pch = 13, inherit.aes = FALSE, alpha = 0.6) +
  labs(title = "2021 — Tetra Tech Survey"))

(tt23_plot <- polyplot + 
  geom_point(
    data = filter(selected, dataset == "tt23"),
    aes(x = longitude, y = latitude), pch = 13, inherit.aes = FALSE, alpha = 0.6) +
  labs(title = "2023 — Tetra Tech Survey"))

(drm24_plot <- polyplot + 
  geom_point(
    data = filter(selected, dataset == "drm24"),
    aes(x = longitude, y = latitude), pch = 13, inherit.aes = FALSE, alpha = 0.6) +
  labs(title = "2024 — Shedd Survey"))
```


# Coral densities - by dataset

## Fit Bayesian negative binomial model

Predictors: dataset, habitat Type, direction from channel, taxon

```{r brms}
# First, analyze surveys that included ALL corals (exclude ESA ONLY surveys)
dff_alltaxa <- dff %>%
  filter(!dataset %in% c("nsu11_esa", "dca17_esa"))

# Get total counts for each taxon at each site
dfftaxon <- dff_alltaxa %>% 
  group_by(dataset, Type, dir, site, transect_num, transect_area_m2, taxon, class) %>%
  summarize(n = sum(n)) %>%
  ungroup()

# Remove any taxa not observed in a given dataset / habitat Type ?????
dfftaxon_trimmed <- dfftaxon %>%
  group_by(dataset, Type, dir, taxon, class) %>%
  filter(any(n > 0)) %>%
  ungroup()



# SUPER MODEL

mod_nb <- brm(
  bf(n ~ dataset * taxon * class * Type + offset(log(transect_area_m2)),
     zi ~ taxon * Type),
  family = zero_inflated_negbinomial(),
  data = dfftaxon,    # choose dataset with or without all structural zeros
  prior = c(prior(normal(0, 2), class = "b"),          # Weak prior on coefficients
            prior(normal(0, 5), class = "Intercept"),  # Weak prior on intercept
            prior(exponential(1), class = "shape")),     # Reasonable prior for NB dispersion
  chains = 4,
  cores = 4,
  threads = threading(5),
  iter = 400, warmup = 100,    #iter = 1000, warmup = 500,
  thin = 2,
  control = list(adapt_delta = 0.9, max_treedepth = 12),
  backend = "cmdstanr"
)
saveRDS(mod_nb, file = "data/processed/mod_nb.rds")
mod_nb <- readRDS("data/processed/mod_nb.rds")

mod_nb

# 1. Create newdata grid (1 m² for standardization)
newdata <- dfftaxon_trimmed %>%
  distinct(dataset, Type, dir, taxon, class) %>%
  mutate(transect_area_m2 = 1)

# 2. Get fitted values manually by computing summary statistics across all draws
posterior_draws <- add_epred_draws(mod_nb, newdata = newdata) %>%
  mutate(Type = factor(Type, levels = type_levels),
         dataset = factor(dataset, levels = dataset_levels))
```

## Totals by survey, habitat, direction

```{r total_density}
# Compute and plot total coral abundance by dataset, habitat Type, and direction from channel
total_abund <- posterior_draws %>%
  group_by(.draw, dataset, Type, dir) %>%
  summarize(total_epred = sum(.epred), .groups = "drop") %>%
  group_by(dataset, Type, dir) %>%
  summarize(
    fit_mean = mean(total_epred),
    fit_sd = sd(total_epred),
    fit_lower = quantile(total_epred, 0.025),
    fit_upper = quantile(total_epred, 0.975),
    .groups = "drop"
  )

ggplot(total_abund, aes(x = dataset, y = fit_mean, color = dir)) +
  facet_grid(~ Type, labeller = as_labeller(type_labels)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = fit_lower, ymax = fit_upper), width = 0,
                position = position_dodge(width = 0.5)) +
  scale_x_discrete(labels = dataset_labels) +
  theme(axis.text.x = element_text(angle = 90))
```

### North-south differences?
```{r total_density_ns_diffs}
# Any N-S differences in total coral abundance?
total_NS_diff <- posterior_draws %>%
  ungroup() %>%
  group_by(.draw, dataset, Type, dir) %>%
  summarize(total_epred = sum(.epred), .groups = "drop") %>%
  pivot_wider(names_from = dir, values_from = total_epred) %>%
  filter(!is.na(N), !is.na(S)) %>%  # make sure both are present
  mutate(diff = N - S)  # or S - N depending on desired contrast

total_NS_summ <- total_NS_diff %>%
  group_by(dataset, Type) %>%
  summarize(
    mean_N = mean(N),
    mean_S = mean(S),
    mean_diff = mean(diff),
    lower_95 = quantile(diff, 0.025),
    upper_95 = quantile(diff, 0.975),
    p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),  # <-- two-sided p-value
    .groups = "drop"
  )

total_NS_summ %>%
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni")) %>%
  filter(p_adj < 0.01) %>%
  knitr::kable()
```

North-south differences within a habitat type within a survey show almost no differences. TT21 NRC is different N vs. S but barely had any N sites so we can discount that. Otherwise, DCA17 showed a slight N-S difference on the Outer Reef, and TT23 should a slight N-S difference on the NRC. All other habitats in all surveys showed no N-S differences in coral density.

### Survey differences?
```{r total_density_survey_diffs}
# 1. Sum predicted values across taxa (total coral density per draw × dataset × Type × dir)
draws_total <- posterior_draws %>%
  mutate(dataset = as.character(dataset)) %>%
  group_by(.draw, dataset, Type, dir) %>%
  summarize(total_epred = sum(.epred), .groups = "drop")

# 2. Self-join to compare dataset pairs within each draw × Type × dir
pairwise_total <- draws_total %>%
  rename(dataset1 = dataset, epred1 = total_epred) %>%
  inner_join(
    draws_total %>% rename(dataset2 = dataset, epred2 = total_epred),
    by = c(".draw", "Type", "dir")
  ) %>%
  filter(dataset1 < dataset2) %>%
  mutate(diff = epred1 - epred2)

# 3. Summarize posterior contrasts
summary_total_contrasts <- pairwise_total %>%
  group_by(Type, dir, dataset1, dataset2) %>%
  summarize(
    mean_diff = mean(diff),
    lower = quantile(diff, 0.025),
    upper = quantile(diff, 0.975),
    p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),
    .groups = "drop"
  )

# 4. Filter significant differences
summary_total_contrasts %>%
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni")) %>%
  filter(p_adj < 0.01) %>%
  arrange(Type, dir) %>%
  knitr::kable()
```

## Taxon totals by survey, habitat, direction from channel
```{r taxon_density, fig.width = 10, fig.height = 10}
fitted_taxon <- posterior_draws %>%
  group_by(dataset, Type, dir, taxon) %>%
  summarize(fit_mean = mean(.epred),  # Posterior mean (expected value)
            fit_sd = sd(.epred),  # Posterior standard deviation
            fit_lower = quantile(.epred, 0.025),        # 2.5% quantile (lower CI)
            fit_upper = quantile(.epred, 0.975))    # 97.5% quantile (upper CI)

# Plot
ggplot(fitted_taxon, aes(y = fit_mean, x = Type, color = dir, shape = dataset,
                          group = interaction(dataset, dir))) +
  facet_wrap(taxon ~ ., scales = "free_x") +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_line(position = position_dodge(width = 0.5), alpha = 0.5) +
  geom_errorbar(aes(ymin = fit_lower, ymax = fit_upper), width = 0,
                position = position_dodge(width = 0.5), lwd = 0.25) +
  scale_y_log10() +
  scale_x_discrete(labels = type_labels) +
  labs(x = "Habitat Type", y = "Density (per m2)")
```

### North-south differences?
```{r taxon_density_ns_diffs}
# Any N-S differences in taxon abundance within habitat Types/datasets?
#### Pivot wide so you can calculate N vs S difference per draw
taxon_NS_diff <- posterior_draws %>%
  ungroup() %>%
  select(.draw, dataset, Type, dir, taxon, class, .epred) %>%
  pivot_wider(names_from = dir, values_from = .epred) %>%
  filter(!is.na(N), !is.na(S)) %>%  # ensure both directions exist for the draw
  mutate(diff = S - N)  # or N - S depending on interpretation

taxon_NS_summ <- taxon_NS_diff %>%
  group_by(dataset, Type, taxon, class) %>%
  summarize(
    mean_diff = mean(diff),
    lower_95 = quantile(diff, 0.025),
    upper_95 = quantile(diff, 0.975),
    p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),
    .groups = "drop"
  )

taxon_NS_summ %>%
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni")) %>%
  filter(p_adj < 0.01) %>%
  mutate(greater = if_else(mean_diff > 0, "N", "S")) %>%
  group_by(dataset, Type, greater) %>%
  summarize(taxa = paste(taxon, class, collapse = ",")) %>%
  arrange(Type, greater) %>%
  knitr::kable()
```

### Survey differences?
```{r taxon_survey_diffs}
# Ensure dataset is character so we can do < comparison
draws_clean <- posterior_draws %>%
  mutate(dataset = as.character(dataset))

# One self-join to get all unique dataset pairs
pairwise_contrasts <- draws_clean %>%
  rename(dataset1 = dataset, epred1 = .epred) %>%
  inner_join(
    draws_clean %>% rename(dataset2 = dataset, epred2 = .epred),
    by = c(".draw", "taxon", "class", "dir", "Type")
  ) %>%
  filter(dataset1 < dataset2) %>%  # Avoid duplicates and self-pairs
  mutate(diff = epred1 - epred2)

# Step: summarize the posterior differences
summary_contrasts <- pairwise_contrasts %>%
  group_by(taxon, class, dir, Type, dataset1, dataset2) %>%
  summarize(mean_diff = mean(diff),
            lower = quantile(diff, 0.025),
            upper = quantile(diff, 0.975),
            p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),
            .groups = "drop")

summary_contrasts %>% 
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni")) %>%
  filter(p_adj < 0.01) %>%
  arrange(taxon, Type, dir) %>%
  knitr::kable()
```

COMBINE N AND S FOR THIS ANALYSIS BASED ON FEW DIFFS DETECTED EARLIER TO GAIN STATISTICAL POWER?
```{r}
# Collapse across direction (North/South) by averaging per draw
draws_collapsed <- posterior_draws %>%
  group_by(.draw, dataset, Type, taxon, class) %>%
  summarize(epred = mean(.epred), .groups = "drop") %>%
  mutate(dataset = as.character(dataset))

pairwise_taxon_diffs <- draws_collapsed %>%
  rename(dataset1 = dataset, epred1 = epred) %>%
  inner_join(
    draws_collapsed %>% rename(dataset2 = dataset, epred2 = epred),
    by = c(".draw", "Type", "taxon", "class")
  ) %>%
  filter(dataset1 < dataset2) %>%
  mutate(diff = epred1 - epred2)

summary_taxon_diffs <- pairwise_taxon_diffs %>%
  group_by(Type, taxon, class, dataset1, dataset2) %>%
  summarize(
    mean_diff = mean(diff),
    lower = quantile(diff, 0.025),
    upper = quantile(diff, 0.975),
    p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),
    .groups = "drop"
  )

summary_taxon_diffs %>% 
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni")) %>%
  filter(p_adj < 0.01) %>%
  arrange(taxon, Type) %>%
  knitr::kable()
```


TEST IF SCTLD-SUSCEPTIBLE CORALS AGGREGATED SHOW SIG DIFFS ACROSS SURVEYS?
```{r}
# susceptibility from Papke et al. 2024
group_definitions <- tibble(
  taxon = unique(posterior_draws$taxon),
  taxon_group = case_when(
    taxon %in% c("FAVI", "MUSS", "MEAN", "DSTO", "EFAS", "MMEA") ~ "high_sus",
    taxon %in% c("MCAV", "ORBI", "SIDE", "SINT", "SOLE") ~ "med_sus",
    taxon %in% c("ACER", "PORI", "AGAR", "MYCE", "MADR") ~ "low_sus",
    TRUE ~ NA_character_
  )
)

draws_grouped <- posterior_draws %>%
  left_join(group_definitions, by = "taxon") %>%
  filter(!is.na(taxon_group))  # Remove taxa not assigned to a group

draws_group_sums <- draws_grouped %>%
  group_by(.draw, dataset, Type, taxon_group, class) %>%
  summarize(epred = sum(.epred), .groups = "drop") %>%
  mutate(dataset = as.character(dataset))

pairwise_group_diffs <- draws_group_sums %>%
  rename(dataset1 = dataset, epred1 = epred) %>%
  inner_join(
    draws_group_sums %>% rename(dataset2 = dataset, epred2 = epred),
    by = c(".draw", "Type", "taxon_group", "class")
  ) %>%
  filter(dataset1 < dataset2) %>%  # Avoid redundant pairs and self-comparisons
  mutate(diff = epred1 - epred2)

summary_group_diffs <- pairwise_group_diffs %>%
  group_by(Type, taxon_group, class, dataset1, dataset2) %>%
  summarize(
    mean_diff = mean(diff),
    lower = quantile(diff, 0.025),
    upper = quantile(diff, 0.975),
    p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni"))

summary_group_diffs %>%
  filter(p_adj < 0.05) %>%
  arrange(taxon_group, class, Type) %>%
  knitr::kable()
```



## Overall taxon differences among surveys
```{r taxon_overall_survey_diffs}
# Proportion of transect area per dataset × Type × dir
survey_weights <- dfftaxon %>%
  group_by(dataset, Type, dir) %>%
  summarize(area = sum(transect_area_m2), .groups = "drop") %>%
  group_by(dataset) %>%
  mutate(weight = area / sum(area)) %>%
  select(dataset, Type, dir, weight)


posterior_weighted <- posterior_draws %>%
  mutate(dataset = as.character(dataset)) %>%
  left_join(survey_weights, by = c("dataset", "Type", "dir")) %>%
  mutate(weighted_epred = .epred * weight)


posterior_totals <- posterior_weighted %>%
  group_by(.draw, dataset, taxon) %>%
  summarize(total_epred = sum(weighted_epred), .groups = "drop")


pairwise_overall <- posterior_totals %>%
  rename(dataset1 = dataset, epred1 = total_epred) %>%
  inner_join(
    posterior_totals %>% rename(dataset2 = dataset, epred2 = total_epred),
    by = c(".draw", "taxon")
  ) %>%
  filter(dataset1 < dataset2) %>%
  mutate(diff = epred1 - epred2)

summary_overall <- pairwise_overall %>%
  group_by(taxon, dataset1, dataset2) %>%
  summarize(mean_diff = mean(diff),
            lower = quantile(diff, 0.025),
            upper = quantile(diff, 0.975),
            p_two_sided = 2 * min(mean(diff > 0), mean(diff < 0)),
            .groups = "drop")


summary_overall %>%
  mutate(p_adj = p.adjust(p_two_sided, method = "bonferroni")) %>%
  filter(p_adj < 0.01) %>%
  knitr::kable()
```

# Total corals in impact zones 

## Import Scenario 2 impact zones and NMFS consideration areas
```{r}
# Impact zone Scenario 2 and NMFS consideration areas
# Define the extraction function once
extract_placemark_info <- function(kml_file, source_label) {
  ns <- c(kml = "http://www.opengis.net/kml/2.2")
  kml <- read_xml(kml_file)
  placemarks <- xml_find_all(kml, ".//kml:Placemark", ns)

  map(placemarks, function(pm) {
    name <- xml_text(xml_find_first(pm, ".//kml:name", ns))
    coords_text <- xml_text(xml_find_first(pm, ".//kml:coordinates", ns))
    folder <- xml_text(xml_find_first(xml_parent(xml_parent(pm)), ".//kml:name", ns))

    coords <- strsplit(trimws(coords_text), "\\s+")[[1]]
    coords_mat <- do.call(rbind, lapply(coords, function(x) {
      as.numeric(strsplit(x, ",")[[1]][1:2])
    }))

    if (nrow(coords_mat) >= 3) {
      poly <- list(st_polygon(list(coords_mat)))
      tibble(folder = folder, name = name, geometry = st_sfc(poly, crs = 4326), source = source_label)
    } else {
      NULL
    }
  }) %>%
    compact() %>%
    bind_rows() %>%
    st_as_sf()
}

# File paths
impact_kml <- "data/Impact_zones_Scenario 2.kml"
nmfs_kml <- "data/Scenario2_NMFSConsiderationAreas.kml"

# Extract and label
impact_zones_sf <- extract_placemark_info(impact_kml, "Impact Zone")
nmfs_sf <- extract_placemark_info(nmfs_kml, "NMFS Consideration Area")

# Combine
all_zones_sf <- bind_rows(impact_zones_sf, nmfs_sf)

# Rename polygons based on zone folder
all_zones_sf <- all_zones_sf %>%
  mutate(
    name = if_else(
      is.na(name) | name == "",
      str_remove(str_remove(folder, "^Scenario_2_"), "\\.shp$"),
      name
    ),
    scenario = "Scenario 2 + NMFS Consideration Areas"
  )

# Create new polygon covering Rest of Monitoring Area, from 1.2 km N to 1.2 km S of the channel

# 1. Extract the 'Channel' polygon
channel_poly <- all_zones_sf %>%
  filter(str_detect(tolower(name), "channel"))

# 2. Transform all relevant layers to UTM (meters) for buffering
utm_crs <- 32617  # UTM Zone 17N — suitable for SE Florida

if (st_is_longlat(channel_poly)) {
  channel_poly <- st_transform(channel_poly, utm_crs)
  polygons_clean_m <- st_transform(polygons_clean, utm_crs)
  all_zones_sf_m <- st_transform(all_zones_sf, utm_crs)
} else {
  polygons_clean_m <- polygons_clean
  all_zones_sf_m <- all_zones_sf
}

# 3. Get midpoint of channel (north-south)
bbox_channel <- st_bbox(channel_poly)
channel_mid_y <- (bbox_channel["ymin"] + bbox_channel["ymax"]) / 2

# 4. Define rectangular box: full E-W extent of reef habitat, ±1.2 km N/S of channel
ew_range <- st_bbox(polygons_clean_m)[c("xmin", "xmax")]
ns_range <- c(channel_mid_y - 1200, channel_mid_y + 1200)
names(ns_range) <- c("ymin", "ymax")

new_box <- st_as_sfc(st_bbox(c(
  xmin = ew_range[[1]],
  xmax = ew_range[[2]],
  ymin = ns_range[[1]],
  ymax = ns_range[[2]]
), crs = st_crs(polygons_clean_m)))

# 5. Clip the box to reef habitat
intersected_box <- st_union(st_intersection(new_box, st_union(polygons_clean_m)))

# 6. Subtract all existing zones (Impact + NMFS) by taking concave hull
# 6. Create a tight, shrink-wrapped outer boundary around all existing zones
existing_zone_hull <- all_zones_sf_m %>%
  st_union() %>%
  st_buffer(40) %>%     # Buffer OUTWARD 40 meters to close gaps (adjust as needed)
  st_union() %>%
  st_buffer(-40)        # Buffer INWARD to return to approximate original size

# 7. Subtract existing zones from monitoring area
new_zone_geom <- st_difference(intersected_box, existing_zone_hull)

# 8. Build the new polygon row
new_zone <- st_sf(
  folder = "Generated",
  name = "Rest of Monitoring Area",
  scenario = "Rest of Monitoring Area",
  geometry = st_sfc(new_zone_geom),
  source = "Impact Zone",
  crs = st_crs(polygons_clean_m)
)

# 9. Transform back to match original CRS (likely WGS84)
if (st_crs(new_zone) != st_crs(all_zones_sf)) {
  new_zone <- st_transform(new_zone, st_crs(all_zones_sf))
}

# 10. Add the new zone to the combined zones object
all_zones_sf <- bind_rows(all_zones_sf, new_zone)


# Plot all zones
ggplot() +
  # Habitat polygons layer
  geom_sf(data = polygons_clean, color = "black", size = 0.2, alpha = 0.4, linewidth = 0.1) +
  
  # Impact/NMFS zones overlay
  geom_sf(data = all_zones_sf, aes(color = scenario), fill = NA, linewidth = 0.5, alpha = 0.2) +

  # Map extent
  xlim(-80.11, -80.079) +
  ylim(26.08, 26.11) +
  
  theme_minimal() +
  guides(fill = guide_legend(title = "Habitat Type"),
         color = guide_legend(title = "Zone Type"))


```


## Area of each habitat type in each impact zone
```{r}
# Set order of zone levels
zone_levels <- c("Channel", "Side Slopes", "Depth_10cm", "Depth_5cm", "Depth_1cm", "Depth_0_5cm", "Depth_0_1cm",
                 "Nearshore", "North", "East", "South", "Rest of Monitoring Area")

# Make sure 'Artificial' is changed to 'Nearshore Ridge Complex' in habitat polygons
polygons_filtered <- polygons_clean %>%
  mutate(Type = if_else(Type == "Artificial", "Nearshore Ridge Complex", Type)) %>%
  filter(Type %in% type_levels)

# Intersect with habitat polygons
habitat_in_zones <- st_intersection(polygons_filtered, all_zones_sf)

# Project for accurate area
habitat_in_zones_proj <- habitat_in_zones %>%
  st_transform(32617) %>%
  mutate(area_m2 = st_area(geometry))

# Summarize area by habitat type and zone name
area_summary <- habitat_in_zones_proj %>%
  st_drop_geometry() %>%
  mutate(name = factor(name, levels = zone_levels),
         Type = factor(Type, levels = type_levels)) %>%
  group_by(Type, name) %>%
  summarize(total_area_m2 = sum(as.numeric(area_m2)), .groups = "drop") %>%
  arrange(name)

# Print total area of each habitat in each impact zone
area_summary %>% 
  pivot_wider(names_from = name, values_from = total_area_m2, names_sort = FALSE)

# Combine for full Scenario2+NMFS, and Rest of Monitoring Area
area_summary2 <- habitat_in_zones_proj %>%
  st_drop_geometry() %>%
  mutate(Type = factor(Type, levels = type_levels)) %>%
  group_by(Type, scenario) %>%
  summarize(total_area_m2 = sum(as.numeric(area_m2)), .groups = "drop")

area_summary2 %>% 
  pivot_wider(names_from = scenario, values_from = total_area_m2)

```

## Total corals in each impact zone

Right now this does NOT differentiate N and S of the channel... This is because no significant differences based on direction were detected within any habitat in any dataset (except TT21 NRC, but this only had 2 sites N of channel, and they were right on the side of the channel...)


```{r total_impactzones}
# calc total density per dataset per habitat type, averaging across N and S
# Compute and plot total coral abundance by dataset, habitat Type, and direction from channel
# Step 1: Sum epred across taxon and class within each dataset, habitat Type, and direction
posterior_dir_totals <- posterior_draws %>%
  group_by(.draw, dataset, Type, dir) %>%
  summarize(dir_total = sum(.epred), .groups = "drop")

# Step 2: Average across directions within habitat Types
posterior_type_totals <- posterior_dir_totals %>%
  group_by(.draw, dataset, Type) %>%
  summarize(total_epred = mean(dir_total), .groups = "drop")

# Step 3: Summarize across posterior draws
total_perType <- posterior_type_totals %>%
  group_by(dataset, Type) %>%
  summarize(
    fit_mean = mean(total_epred),
    fit_sd = sd(total_epred),
    fit_lower = quantile(total_epred, 0.025),
    fit_upper = quantile(total_epred, 0.975),
    .groups = "drop"
  )

# Compute total abundances within impact zones
totals <- left_join(total_perType, area_summary, by = "Type") %>%
  mutate(tot_estimate = fit_mean * total_area_m2,
         tot_lower = fit_lower * total_area_m2,
         tot_upper = fit_upper * total_area_m2) %>%
  mutate(Type = factor(Type, levels = type_levels))

# Step 2: Add missing combinations as NA rows
totals <- totals %>%
  complete(dataset, Type, fill = list(tot_estimate = NA, tot_lower = NA, tot_upper = NA)) %>%
  mutate(name = factor(name, levels = zone_levels))

# Step 3: Plot
p1 <- ggplot(totals, aes(x = dataset, y = tot_estimate, fill = fct_rev(name))) +
  facet_grid(~Type) +
  geom_col(position = position_stack()) +
  # geom_errorbar(aes(ymin = tot_lower, ymax = tot_upper),
  #               position = position_dodge(width = 0.9), width = 0.2) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = "Habitat Type",
       y = "Total corals",
       title = "Total corals in all impact zones, by habitat") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p1





p2 <- totals %>%
  group_by(dataset) %>%
  summarize(value = sum(tot_estimate, na.rm = T),
            lower = sum(tot_lower, na.rm = T),
            upper = sum(tot_upper, na.rm = T)) %>%
  mutate(dataset = factor(dataset, levels = dataset_levels)) %>%
  ggplot(aes(x = 1, y = value, fill = dataset)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.9), width = 0.2) +
  scale_fill_discrete(labels = dataset_labels) +
  labs(x = "", y = "Total corals", title = "Total corals in all impact zones - all habitats")


p2

```


# Coral densities - across datasets, with zero-inflated negbin model
```{r}
# 🎯 Goal
# 
# Estimate: density of taxon × class within each habitat Type
# ...averaging over datasets, but not letting datasets with more transects in a given habitat dominate unfairly.
# with zero-inflation that varies per taxon*class

# mod_nb_zinf <- brm(
#   bf(n ~ taxon * class * Type +
#         (1 | Type/dataset/site) +     # Allow datasets (and sites) to have different random intercepts across habitat Types
#         #(1 | Type:dir) +  # decide later if we want to keep this
#         offset(log(transect_area_m2)),
#      zi ~ taxon * Type),
#   data = dfftaxon,   # use dataset with all structural zeros present
#   family = zero_inflated_negbinomial(),
#   prior = c(
#     prior(normal(0, 2), class = "b"),
#     prior(normal(0, 5), class = "Intercept"),
#     prior(exponential(1), class = "shape"),
#     prior(normal(0, 1), class = "b", dpar = "zi")
#   ),
#   chains = 4, cores = 4, threads = threading(5),
#   iter = 1000, warmup = 500, thin = 2,
#   backend = "cmdstanr",
#   control = list(adapt_delta = 0.95)
# )
# saveRDS(mod_nb_zinf, file = "data/processed/mod_nb_zinf.rds")
mod_nb_zinf <- readRDS("data/processed/mod_nb_zinf.rds")


# 1. Create newdata grid (1 m² for standardization)
newdata <- dfftaxon_trimmed %>%
  distinct(Type, taxon, class) %>%
  mutate(transect_area_m2 = 1)

# Posterior draws of predicted density (n / m²)
preds <- add_epred_draws(mod_nb_zinf, newdata = newdata, re_formula = NA)    # excludes random effects (gives population-level average)

summary_preds <- preds %>%
  group_by(taxon, class, Type) %>%
  mean_qi(.epred, .width = 0.95)

# Calculate total per taxclass per habitat Type and impact zone
tot1 <- left_join(summary_preds, area_summary, by = "Type") %>%
  mutate(tot = .epred * total_area_m2,
         tot_lower = .epred * total_area_m2,
         tot_upper = .epred * total_area_m2)

tot1 %>%
  select(taxon, class, Type, n_per_m2 = .epred)

# Calculate total per taxclass per impact zone (sum all habitat Types)
tot2 <- tot1 %>%
  group_by(taxon, class, name) %>%
  summarize(tot = sum(tot))

# Calculate total per impact zone (sum all taxon/class)
tot3 <- tot2 %>%
  group_by(name) %>%
  summarize(tot = sum(tot))

tot2 %>%
  mutate(ESA = if_else(taxon %in% c("ACER", "ORBI"), "ESA", "non-ESA")) %>%
  group_by(name, ESA) %>%
  summarize(tot = sum(tot), .groups = "drop") %>%
  pivot_wider(names_from = ESA, values_from = tot) %>%
  janitor::adorn_totals(where = "row") %>%
  mutate(across(where(is.numeric), ~ format(round(.), big.mark = ","))) %>%
  knitr::kable()



# Calculate total (sum all impact zones)
sum(tot3$tot)

```

```{r, fig.height = 10}

# STEP 1: Aggregate total abundance by taxon + class
tot4 <- tot2 %>%
  group_by(taxon, class) %>%
  summarize(tot = sum(tot), .groups = "drop")

# STEP 2: Set tile resolution
tile_size <- 5000

# STEP 3: Compute number of tiles per taxon + class
tile_data <- tot4 %>%
  mutate(n_tiles = round(tot / tile_size)) %>%
  filter(n_tiles > 0)

# STEP 4: Order taxa by total abundance (across size classes)
taxon_order <- tile_data %>%
  group_by(taxon) %>%
  summarize(total_tiles = sum(n_tiles), .groups = "drop") %>%
  arrange(total_tiles) %>%
  mutate(taxon_row_start = cumsum(lag(total_tiles %/% 40 + 1, default = 0)))

# STEP 5: Expand to individual tiles
tile_expanded <- tile_data %>%
  uncount(n_tiles) %>%
  mutate(class = factor(class, levels = c(">4cm", "<4cm"))) %>%
  left_join(taxon_order, by = "taxon") %>%
  group_by(taxon) %>%
  mutate(
    tile_id = row_number(),
    col = (tile_id - 1) %% 40,
    row = -(taxon_row_start + (tile_id - 1) %/% 40)
  ) %>%
  ungroup() %>%
  mutate(taxon = factor(taxon, levels = taxon_order$taxon))

# STEP 6: Assign distinct shuffled Dark2 colors per taxon
taxa <- levels(tile_expanded$taxon)
n_taxa <- length(taxa)
# Generate color palette (Dark2 can go up to 8, then extended)
raw_colors <- c("brown", "dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00", "black", "gold1", "skyblue2", "maroon", "palegreen2", "#FDBF6F", "gray70",  "orchid1", "darkturquoise", "darkorange4")
# Assign names
base_colors <- setNames(rev(raw_colors), taxa)

# STEP 7: Label positions
row_labels <- tile_expanded %>%
  group_by(taxon) %>%
  summarize(label_y = min(row), label_x_end = max(col), .groups = "drop")

# STEP 8: Taxon totals
taxon_totals <- tot4 %>%
  group_by(taxon) %>%
  summarize(total = sum(tot), .groups = "drop") %>%
  mutate(total_fmt = format(round(total), big.mark = ","))

row_labels <- row_labels %>%
  left_join(taxon_totals, by = "taxon")

# STEP 9: Plot
waffle <- ggplot(tile_expanded, aes(x = col, y = row, fill = taxon)) +
  geom_tile(width = 0.95, height = 0.95, color = "white", linewidth = 0.1) +

  # Dot for <4cm
  geom_point(
    data = filter(tile_expanded, class == "<4cm"),
    aes(x = col, y = row),
    shape = 16, color = "white", size = 0.5, inherit.aes = FALSE
  ) +

  # Taxon names (left)
  geom_text(
    data = row_labels,
    aes(x = -2, y = label_y, label = taxon),
    hjust = 1, size = 2.5, inherit.aes = FALSE
  ) +

  # Totals (right)
  geom_text(
    data = row_labels,
    aes(x = label_x_end, y = label_y, label = total_fmt),
    hjust = 0, size = 2.5, inherit.aes = FALSE
  ) +

  scale_fill_manual(values = base_colors, guide = "none") +
  coord_equal(clip = "off") +  # allow text to overflow plot area
  theme_void() +
  theme(
    plot.margin = margin(t = 10, r = 80, b = 10, l = 80),  # ensure space on both sides
    axis.text = element_blank()
  ) +
  labs(
    title = "Total abundance within 1.2-km of channel\n(1 tile = 5,000 corals)",
    caption = "White dot = smaller corals (<4cm); solid tiles = >4cm"
  )

waffle

```

```{r}
# STEP 1: Aggregate total abundance by taxon + class
tot4 <- tot2 %>%
  filter(name != "Rest of Monitoring Area") %>%
  group_by(taxon, class) %>%
  summarize(tot = sum(tot), .groups = "drop")

# STEP 2: Set tile resolution
tile_size <- 5000

# STEP 3: Compute number of tiles per taxon + class
tile_data <- tot4 %>%
  mutate(n_tiles = round(tot / tile_size)) %>%
  filter(n_tiles > 0)

# STEP 4: Order taxa by total abundance (across size classes)
taxon_order <- tile_data %>%
  group_by(taxon) %>%
  summarize(total_tiles = sum(n_tiles), .groups = "drop") %>%
  arrange(total_tiles) %>%
  mutate(taxon_row_start = cumsum(lag(total_tiles %/% 40 + 1, default = 0)))

# STEP 5: Expand to individual tiles
tile_expanded <- tile_data %>%
  uncount(n_tiles) %>%
  mutate(class = factor(class, levels = c(">4cm", "<4cm"))) %>%
  left_join(taxon_order, by = "taxon") %>%
  group_by(taxon) %>%
  mutate(
    tile_id = row_number(),
    col = (tile_id - 1) %% 40,
    row = -(taxon_row_start + (tile_id - 1) %/% 40)
  ) %>%
  ungroup() %>%
  mutate(taxon = factor(taxon, levels = taxon_order$taxon))

# STEP 6: Assign distinct shuffled Dark2 colors per taxon
taxa <- levels(tile_expanded$taxon)
n_taxa <- length(taxa)
# Generate color palette (Dark2 can go up to 8, then extended)
raw_colors <- c("brown", "dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00", "black", "gold1", "skyblue2", "maroon", "palegreen2", "#FDBF6F", "gray70",  "orchid1", "darkturquoise", "darkorange4")
# Assign names
base_colors <- setNames(rev(raw_colors), taxa)

# STEP 7: Label positions
row_labels <- tile_expanded %>%
  group_by(taxon) %>%
  summarize(label_y = min(row), label_x_end = max(col), .groups = "drop")

# STEP 8: Taxon totals
taxon_totals <- tot4 %>%
  group_by(taxon) %>%
  summarize(total = sum(tot), .groups = "drop") %>%
  mutate(total_fmt = format(round(total), big.mark = ","))

row_labels <- row_labels %>%
  left_join(taxon_totals, by = "taxon")

# STEP 9: Plot
waffle <- ggplot(tile_expanded, aes(x = col, y = row, fill = taxon)) +
  geom_tile(width = 0.95, height = 0.95, color = "white", linewidth = 0.1) +

  # Dot for <4cm
  geom_point(
    data = filter(tile_expanded, class == "<4cm"),
    aes(x = col, y = row),
    shape = 16, color = "white", size = 0.5, inherit.aes = FALSE
  ) +

  # Taxon names (left)
  geom_text(
    data = row_labels,
    aes(x = -2, y = label_y, label = taxon),
    hjust = 1, size = 2.5, inherit.aes = FALSE
  ) +

  # Totals (right)
  geom_text(
    data = row_labels,
    aes(x = label_x_end, y = label_y, label = total_fmt),
    hjust = 0, size = 2.5, inherit.aes = FALSE
  ) +

  scale_fill_manual(values = base_colors, guide = "none") +
  coord_equal(clip = "off") +  # allow text to overflow plot area
  theme_void() +
  theme(
    plot.margin = margin(t = 10, r = 80, b = 10, l = 80),  # ensure space on both sides
    axis.text = element_blank()
  ) +
  labs(
    title = "Total abundance within Scenario 2 Impact Zones\n(1 tile = 5,000 corals)",
    caption = "White dot = smaller corals (<4cm); solid tiles = >4cm"
  )

waffle

```

# Dataset-specific estimates
```{r}
newdata_ds <- dfftaxon %>%
  distinct(Type, dataset, taxon, class) %>%
  mutate(site = NA,
         transect_area_m2 = 1)


preds_ds <- add_epred_draws(
  mod_nb_zinf,
  newdata = newdata_ds,
  re_formula = NULL,
  allow_new_levels = TRUE  # Ignore unrecognized site levels
)


summary_preds_ds <- preds_ds %>%
  group_by(taxon, class, Type, dataset) %>%
  mean_qi(.epred, .width = 0.95)


ggplot(summary_preds_ds, aes(x = Type, y = .epred,
                              color = class, shape = dataset,
                              group = interaction(dataset, class))) +
  geom_point(position = position_dodge(width = 0.5), size = 2, alpha = 0.5) +
  geom_line(alpha = 0.5) +
  geom_errorbar(aes(ymin = .lower, ymax = .upper),
                position = position_dodge(width = 0.5), width = 0.2, alpha = 0.5) +
  facet_wrap(~ taxon) +
  scale_y_log10(limits = c(1e-4, 2)) +
  labs(x = "Habitat Type", y = "Estimated Density (per m²)",
       color = "Size Class", shape = "Dataset") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


dfftaxon %>%
  filter(taxon == "ACER", Type != "Nearshore Ridge Complex", n > 0)

summary_preds_ds %>%
  filter(taxon == "ACER", Type != "Nearshore Ridge Complex")
```


